### 2011-12-19 doc/Manuscript/doc.tex
Initial import.

% Add example showing that if not SSA form, then could be many iterations for
% Cousot's operator.

% Add example showing that Cousot's does not reach fix-point always.

% Implement bck/RA8 and bck/RA9 to see how closer you get to the fix-point.

% Implement the fix point operator to solve the range analysis problem.

% How many instructions you cannot handle?

\documentclass[preprint]{sigplanconf}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{skak} % for the \qside
\usepackage{mathpartir}
\usepackage{array}
\usepackage{url}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{example}[theorem]{Example}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}

\newcommand{\fun}[1]{\mbox{\textbf{#1}}}
\newcommand{\lb}[1]{#1_{\downarrow}}
\newcommand{\ub}[1]{#1_{\uparrow}}
\newcommand{\varset}[1]{\mbox{$\cal{#1}$}}

\def\qed{\unskip\kern 10pt{\unitlength1pt\linethickness{.4pt}\framebox(6,6){}}}

\begin{document}

\conferenceinfo{Crazy Horse '11}{February 30th, 2011, Tumba, CA.} 
\copyrightyear{2011} 
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}
\preprintfooter{Range Analysis of Whole Programs}

\title{Range Analysis of Whole Programs}
\subtitle{}

\authorinfo{Homer Sympson \and Luck Skywalker \and Robin Hood \and
  Frodo Beggins}
           {Computer Science Department -- UFMG -- Brazil}
           {\{hsympson,skywalker,rbhood,frodo\}@dcc.ufmg.br}

\maketitle

\begin{abstract}
This paper describes a range analysis algorithm that is linear on the
program size.
We handle comparisons between variables via a two-phases dataflow analysis, and
achieve flow sensitiveness by using the Extended Static Single Assignment
form as the intermediate representation.
We have implemented an inter-procedural version of our technique into LLVM, an
industrial strenght compiler.
\end{abstract}

\category{D - Software}{D.3 Programming Languages}{D.3.4 Processors - Optimization}

\terms
Languages, Performance, Experimentation

\keywords
Range Analyisis, Dataflow Analysis, Integer arithmetics

\section{Introduction}
\label{sec:intro}

\section{Background}
\label{sec:bck}

% Define the lattice, the constraints and the valuation I.
Following Gawlitza {\em et al.}'s notation, we shall be performing arithmetic
operations over the complete lattice
$\cal{Z} = \mathbb{Z} \cup \{-\infty, +\infty\}$, where the ordering is
naturally given by $-\infty < \ldots < -2 < -1 < 0 < 1 < 2 < \ldots +\infty$.
For any $x > -\infty$ we define:

\begin{tabular}{lcl}
$x + \infty = \infty, x \neq -\infty$ & \mbox{\hspace{0.1cm}} & $x - \infty = - \infty, x \neq +\infty$ \\
$x \times \infty = \infty$ if $x > 0$ & & $x \times \infty = -\infty$ if $x < 0$ \\
$0 \times \infty = 0$ & & $(-\infty) \times \infty = \ $ not defined $$ \\
\end{tabular}

From the lattice $\varset{Z}$ we define the product lattice
$\varset{Z}^2$, which is partially ordered by the subset relation
$\sqsubseteq$.
$\varset{Z}^2$ is defined as follows:

\begin{equation*}
\varset{Z}^2 = \emptyset \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
\ z_1 \leq z_2, \  -\infty < z_2 \}
\end{equation*}

Given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
$\ub{\iota} = u$.
We let \varset{V} be a set of variables, and
$I: \varset{V} \mapsto \varset{Z}^2$ a
mapping from variables to intervals in $\varset{Z}^2$.
Our objective is to solve a constraint system $C$, formed by instances of the
five different types of constraints seen in Figure~\ref{fig:eval_function}
(left).
Figure~\ref{fig:eval_function} (right) defines a valuation function $e$ that
computes $Y = f(\ldots)$, given I.
Armed with these concepts, we define the range analysis problem as follows:

\begin{definition}
\label{def:rcp}
\textsc{Range Analysis Problem} \\
\textbf{Input:} a set $C$ of constraints ranging over a set \varset{V} of
variables. \\
\textbf{Output:} a mapping I such that, for any variable
$V \in \varset{V}$, e(V) = I[V].
\end{definition}

\begin{figure}[t!]
\begin{small}
\begin{eqnarray*}
\begin{array}{rc}
Y = [l, u]
&
e(Y) = [l, u]
\\
\\
Y = \phi (X_1, X_2)
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [\mbox{min}(l_1, l_2), \mbox{max}(u_1, u_2)]}
\\
\\
Y = X_1 + X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [l_1 + l_2, u_1 + u_2]}
\\
\\
Y = X_1 \times X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2] \\ L = \{l_1l_2, l_1u_2, u_1l_2, u_1u_2\}}
{e(Y) = [\mbox{min}(L), \mbox{max}(L)]}
\\
\\
Y = aX + b
&
\inferrule{I[X]=[l, u] \\ k_l = al + b \\ k_u = au + b}
{e(Y) = [\mbox{min}(k_l, k_u), \mbox{max}(k_l, k_u)]}
\\
\\
Y = X \sqcap [l', u']
&
\inferrule{I[X]=[l, u]}
{e(Y) \leftarrow [\mbox{max}(l, l'), \mbox{min}(u, u')]}
\end{array}
\end{eqnarray*}
\caption{\label{fig:eval_function}
The valuation function.}
\end{small}
\end{figure}

We will use the program in Figure~\ref{fig:ex1}(a) as the running example
to illustrate our range analysis.
Figure~\ref{fig:ex1}(b) shows the same program in SSA form, and
Figure~\ref{fig:ex1}(c) outlines the constraints that we extract from this
program.
There is a clear correspondence between instructions and constraints.
A possible solution to the range analysis problem is given in
Figure~\ref{fig:ex1}(d).
This is a very conservative solution.
As we will see shortly, we can improve this solution substantially by using
a more sophisticated program representation.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/ex1}
\end{center}
\caption{\label{fig:ex1}
(a) Example program.
(b) Control Flow Graph (CFG) in SSA form.
(c) Constraints that we extract from the program.
(d) A possible solution to the range analysis problem.}
\end{figure}

\section{Polynomial Time Range Analysis}
\label{sec:algo}

% Program representation
% Extracting constraints
% Building the Constraint graph (pseudo-edges)
% Macro algorithm
% Micro algorithm

In this section we explain the algorithm that we use to solve the range
analysis problem for an entire program.
This algorithm involves a number of steps.
First, we convert the program to a suitable intermediate representation that
makes it easier to extract constraint.
From these constraints, we build a dependence graph that allows us to do
range analysis sparsely.
Finally, we solve the constraints applying different fix-point iterators on
this dependence graph.
Figure~\ref{fig:algorithm} gives a global view of this algorithm.
Some of the steps in the algorithm are optional.
They improve the precision of the range analysis, at the expenses of a longer
running time.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/algorithm}
\end{center}
\caption{\label{fig:algorithm}
Our algorithm to solve the range analysis problem.
Rounded boxes are optional steps.}
\end{figure}

\subsection{Choosing a Program Representation}
\label{sub:rep}

The solution that we found to the range analysis problem in Figure~\ref{fig:ex1}
is imprecise because we did not take conditional tests into considerations.
Branches gives us information about the ranges that some variables assume, but
only at {\em specific} program points.
For instance, given a test such as $(k_1 < 100)?$ in  Figure~\ref{fig:ex1}(b),
we know that $I[k_1] \sqsubseteq [-\infty, 99]$ whenever the condition is true.
In order to encode this information, we might split the live range of $k_1$
right after the branching point; thus, creating two new variables, one at the
path where the condition is true, and another where it is false.
There is a program representation, introduced by Bodik
{\em et al.}~\cite{Bodik00}, that performs this live range splitting:
the {\em Extended Static Single Assignment} form, or e-SSA for short.
In this paper, we have experimented with two different flavors of e-SSA form,
which we call {\em standard} and {\em non-pruned}.

We will show first how we convert a program into standard e-SSA form.
Let $(v < c)?$ be a conditional test, and let $l_t$ and $l_f$ be labels in
the program, such that $l_t$ is the target of the test if the condition is true,
and let $l_f$ is the target when the condition is false.
If $l_t$ dominates any use of $v$, then we insert at $l_f$ a copy
$v_f = v \sqcap [-\infty, c-1]$, where $v_f$ is a fresh name.
We then rename every use of $v$ that is dominated by $l_f$ to $v_f$.
Dually, if $l_t$ dominates any use of $v$, then we insert at $l_t$ a copy
$v_t = v \sqcap [c, +\infty]$, and rename variables accordingly.
If the test uses two variables, e.g., $(v_1 < v_2)?$, then we create
intersections bound to {\em futures}.
We insert, at $l_f$, $v_{1f} = v_1 \sqcap [\fun{ft}(v_2), +\infty]$,
and $v_{2f} = v_2 \sqcap [-\infty, \fun{ft}(v_1) - 1]$.
Similarly, at $l_v$ we insert
$v_{1v} = v_1 \sqcap [-\infty, \fun{ft}(v_2) - 1]$
and $v_{2f} = v_2 \sqcap [\fun{ft}(v_1), +\infty]$.
We use the notation $\fun{ft}(v)$ to denote the {\em future} bounds of a
variable.
As we will show in Section~\ref{sub:micro}, once the growth pattern of $v$ is
known, we can replace $\fun{ft}(v)$ by an actual value.
Once we are done placing copies, we insert $\phi$-functions into the
transformed program to convert it to SSA form.
This last step prevents that two different names given to the same original
variable be simultaneously alive at the program code.
Figure~\ref{fig:ex_standard_eSSA} shows our running example changed into
standard e-SSA form.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.7\columnwidth]{images/ex_standard_eSSA}
\end{center}
\caption{\label{fig:ex_standard_eSSA}
The control flow graph from Figure~\ref{fig:ex1}(b) converted to standard
e-SSA form.}
\end{figure}

The standard e-SSA form serves well analysis that need range information
at the use site of variables, such as conditional constant propagation,
redundant code elimination and detection of buffer overflows.
On the other hand, there exist compilation algorithms that require range
information at the whole live range of variables.
Examples of such algorithms include the family of
bitwidth aware register allocators~\cite{Barik06,Tallam03,Pereira08} and
hardware synthesizers~\cite{Cong05,Mahlke01,Stephenson00}.
In order to provide extra precision to these algorithms, we may work with a
non-pruned version of e-SSA form, which we produce by simply inserting copies
after every conditional test, and then reconverting the entire program to
SSA form.
Figure~\ref{fig:ex_non_pruned} illustrates the difference between these two
program representations.
Notice that the non-pruned flavor might introduce dead variables in the
program code, which can be removed by standard dead code elimination.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/ex_non_pruned}
\end{center}
\caption{\label{fig:ex_non_pruned}
(a) Example program.
(b) Standard e-SSA form.
(c) Non-pruned e-SSA form.}
\end{figure}

\subsection{Extracting Constraints}
\label{sub:constraints}

The constraints shown in Figure~\ref{fig:eval_function} let us handle
many different assembly instructions.
We use $\phi$ assignments, such as $Y = \phi(X_1, X_2)$ to represent
the $\phi$-functions typically found in SSA form programs.
As we have discussed in Section~\ref{sub:rep}, we use intersection
constraints to bound the variables created due to live range splitting
at conditionals.
The widening analysis that we will introduce in Section~\ref{sub:micro}
require monotonic transfer functions.
Many assembly operations, such as modulus or division, do not afford us
this property.
However, several of these non-monotonic instructions have a conservative
translation to our constraint system, as we show in
Figure~\ref{fig:constraints}.

\begin{figure}[t!]\begin{center}
\begin{tabular}{|l@{\hspace{0.2cm}}l@{\hspace{0.3cm}}l|} \hline
\textbf{Description} & \textbf{Operation} & \textbf{Constraint} \\ \hline
Constant & $v = c$ & $V = [c, c]$ \\
Assignment & $v_1 = v_2$ & $V_1 = V_2$ \\
Multiplication & $v_1 = v_2 * c$ & $V_1 = c V_2$ \\
Int. division & $v_1 = v_2 \ / \ v_3$ & $V_1 = V_2$ \\
 & & $V_1 = - V_2$ \\
Modulus & $v_1 = v_2 \ \% \ c$ & $V_1 = [0, c - 1]$ \\
Bitwise and & $v_1 = v_2 \ \& \ c$ & $V_1 = v_2 \sqcap [0, c]$ \\
Bitwise or & $v_1 = v_2 \ | \ v_3$ & $V_1 = V_2 + V_3$ \\
Left shift & $v_1 = v_2 \ \qside \ c$ & $V_1 = 2^c V_2$ \\ \hline
\end{tabular}
\end{center}
\caption{\label{fig:constraints}Example of constraint derivation rules.
We let $c$ denote a positive constant, $\{v, v_1, v_2\}$ are program
variables and $\{V, V_1, V_2\} \subset \cal{V}$.}
\end{figure}

\subsection{The Dependence Graph}
\label{sub:graph}

The main data structure that we use to solve the range analysis problem is
a variation of Ferrante {\em et al.}'s {\em program dependence
graph}~\cite{Ferrante87}.
For each constraint variable $V$ we create a variable node $N_v$.
For each constraint $C$ we create a constraint node $N_c$.
We add an edge from $N_v$ to $N_c$ if the name $V$ is used in $C$.
We add an edge from $N_c$ to $V$ if the constraint $C$ defines the name
$V$.
Figure~\ref{fig:ex_graph} shows the dependence graph that we build for the
e-SSA form program given in Figure~\ref{fig:ex_standard_eSSA}.
If $V$ is used by $C$ as the input of a future, then the edge from
$N_v$ to $N_c$ represents a {\em control
dependence}~\cite[p.323]{Ferrante87}.
All the other edges denote {\em data dependences}~\cite[p.322]{Ferrante87}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/ex_graph}
\end{center}
\caption{\label{fig:ex_graph}
The dependence graph that we build to the program in
Figure~\ref{fig:ex_standard_eSSA}.}
\end{figure}

\subsection{Finding Ranges: the Macro Algorithm}
\label{sub:macro}

Having built the dependence graph, we proceed to solve the range analysis
problem by finding all the strongly connected components (SCCs) of the
dependence graph, and reducing this graph to a directed acyclic graph.
We perform this last step by collapsing every SCC into a single node.
We then sort the resulting DAG topologically, and apply the analyses from
Section~\ref{sub:micro} on every SSC in topological order.
Once we solve the range analysis problem for a SCC, we propagate the
intervals that we found to the variable nodes at the {\em frontier} of this
SCC.
A variable node $N_v$ is said to be in the frontier of a strongly connected
component $S$ if:
(i) $N_v \notin S$; and
(ii) there exists a variable node $V' \in S$, and a constraint node $N_c$,
such that $V' \leftarrow N_c$, and $N_c \leftarrow V$.
This propagation gives us a very modular algorithm:
when analyzing a strongly connected component $S$ we can rest assured that
any influence that $S$ might suffer from nodes outside it has been already
taken into consideration.

\subsection{Finding Ranges: the Micro Algorithm}
\label{sub:micro}

Given a strongly connected component of the dependence graph with $N$ nodes,
we solve the range analysis problem in three-steps:
\begin{enumerate}
\item Run widening analysis: $O(N)$.

\item Fix intersections: $O(N)$.

\item Run narrowing analysis: $O(N^2)$.

\end{enumerate}
However, before we start, we remove the control dependence edges from the
strongly connected component, as they have no semantics to our transfer
functions.

\paragraph{Widening Analysis.}

The first step of our algorithm consists in determining the growth behavior of
the interval bound to each constraint variable.
We accomplish this task via Cousot and Cousot's widening
operator~\cite[p.247]{Cousot77}.
The possible behaviors of an interval are:
(i) does not change;
(ii) grows towards $+\infty$;
(iii) grows towards $-\infty$; and
(iv) grows in both directions.
The lattice of this dataflow analysis, plus its meet operator is given in
Figure~\ref{fig:growth_analysis}.
Because the lattice has height three, the intervals bound to each variable can
change at most three times.

\begin{figure}[t!]
\begin{tabular}{cc}
\begin{minipage}{2.4cm}
\includegraphics{images/growth_lattice}
\end{minipage}
&
\begin{minipage}{6cm}
\begin{small}
\begin{eqnarray*}
\begin{array}{c}
\inferrule{I[Y] = [\bot, \bot]}{I[Y] \leftarrow e(Y)}
\\
\\
\inferrule{\lb{e(Y)} < \lb{I[Y]} \\ \ub{e(Y)} > \ub{I[Y]}}
{I[Y] \leftarrow [-\infty, +\infty ]}
\\
\\
\inferrule{\lb{e(Y)} < \lb{I[Y]}}
{I[Y] \leftarrow [-\infty, \ub{I[Y]}]}
\\
\\
\inferrule{\ub{e(Y)} > \ub{I[Y]}}
{I[Y] \leftarrow [\lb{I[Y]}, +\infty]}
\end{array}
\end{eqnarray*}
\end{small}
\end{minipage}
\end{tabular}
\caption{\label{fig:growth_analysis}
(Left) The lattice of the widening analysis.
(Right) Cousot and Cousot's widening operator. We evaluate the rules from top 
to bottom, and stop upon finding a pattern matching.}
\end{figure}

\paragraph{Fixing futures.}

The ranges found by the widening analysis tells us which variables have fixed
bounds, independent on the intersections in the constraint system.
Thus, we can use actual limits to replace intersections bounded by futures.
Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
In order to correctly replace a future $\fun{ft}(V)$ that limits a variable
$V'$, we need to have already applied the widening analysis onto $V$.
Had we considered only data dependence edges, then it would be possible
that $V$ be analyzed before $V'$.
However, because of control dependence edges, this case cannot happen.
The control dependence edges ensure that any topological ordering of the
constraint graph either places $N_v$ before $N_{v'}$, or places these nodes
in the same strongly connected component.
For instance, in Figure~\ref{fig:ex_graph}, variables $J_1$ and $I_t$ are in
the same SCC only because of the control dependence edges.

\begin{figure}[t!]
\begin{center}
\begin{eqnarray*}
\begin{array}{c}
\inferrule{Y = X \sqcap [l, \fun{ft}(V) + c] \\ \ub{I[V]} = u}
{Y = X \sqcap [l, u + c]}, u, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
\\
\\
\inferrule{Y = X \sqcap [\fun{ft}(V) + c, u] \\ \lb{I[V]} = l}
{Y = X \sqcap [l + c, u]}, l, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
\end{array}
\end{eqnarray*}
\end{center}
\caption{\label{fig:fix_intersects}Rules to replace futures by actual
bounds. $S$ is the interval bound to each variable after the widening
analysis.}
\end{figure}

%\begin{figure}[t!]
%\begin{equation*}
%S[X] \leftarrow
%\begin{cases}
%``\ ? \ " \ \ \mbox{if} \ I[X] = [-\infty, +\infty]\\
%``\downarrow" \ \mbox{if} \ I[X] = [-\infty, c], c \in \mathbb{Z} \\
%``\uparrow" \ \mbox{if} \ I[X] = [c, +\infty],  c \in \mathbb{Z} \\
%`` \ 0 \ " \ \ \mbox{if} \ I[X] = [c_1, c_2], c_1, c_2 \in \mathbb{Z}
%\end{cases}
%\end{equation*}
%\caption{\label{fig:st_mem}The growth state of each variable.}
%\end{figure}

\paragraph{Narrowing Analysis.}

The widening analysis associates very conservative bounds to each variable.
Thus, the last step of our algorithm consists in narrowing these intervals.
We accomplish this step via Cousot and Cousot's narrowing
operator~\cite[248]{Cousot77}, which we show in
Figure~\ref{fig:crop_analysis}.

\begin{figure}[t!]
\begin{center}
\begin{eqnarray*}
\begin{array}{c}
\inferrule{\lb{I[Y]} = -\infty \\ \lb{e(Y)} > -\infty}
{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
\\
\\
\inferrule{\lb{I[Y]} > \lb{e(Y)}}
{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
\\
\\
\inferrule{\ub{I[Y]} = +\infty \\ \ub{e(Y)} < +\infty}
{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
\\
\\
\inferrule{\ub{I[Y]} < \ub{e(Y)}}
{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
\end{array}
\end{eqnarray*}
\end{center}
\caption{\label{fig:crop_analysis}Cousot and Cousot's narrowing operator.}
\end{figure}

\paragraph{Example}

Continuing with our example, Figure~\ref{fig:ex_partition_grow_crop} shows
the application of our algorithm on the last strongly connected component of
Figure~\ref{fig:ex_graph}.
Notice that upon meeting this SCC, we have already found the interval
$[0, 0]$ to $I_0$ and the interval $[100, 100]$ to $J_0$, as we show in
Figure~\ref{fig:ex_graph}(a).
We are not guaranteed to find the least fix point of a constraint system.
However, in this example we did it.
We emphasize that finding this tight solution was only possible because of
the topological ordering of the constraint graph in
Figure~\ref{fig:ex_graph}.
Had we applied the widening operator onto the whole graph, then we would
have found out that variable $J_0$ is bound to $[-\infty, +\infty]$,
because
(i) it receives its interval directly from variable $K_t$, which is upper
bounded by $+\infty$, and
(ii) it is part of a negative cycle.
On the other hand, because we only analyze the $J$'s SCC after we have
analyzed $K$'s, $K$ only contribute the constant range $[0, 99]$ to $J_0$.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/ex_partition_grow_crop}
\end{center}
\caption{\label{fig:ex_partition_grow_crop}
Four snapshots of the last SCC of Figure~\ref{fig:ex_partition_grow_crop}.
(a) After removing control dependence edges, immediately before applying
Cousot and Cousot's operators.
(b) After running the widening analysis.
(c) After fixing the intersections.
(d) After running the narrowing analysis.}
\end{figure}

\subsection{Alternative Narrowing Operators}
\label{sub:alt}

Cousot and Cousot's narrowing operator, given in
Figure~\ref{fig:crop_analysis} is not guaranteed to find the least fixpoint
of a system of constraints.
In order to improve the bounds, trading time for precision, we have
experimented with two other operators.
The first, is the evaluation function itself, which is given in
Figure~\ref{fig:eval_function}.
By iterating it until reaching a fixpoint, we are guaranteed to find the
optimal solution to the narrowing analysis.
However, this approach is not practical, because it is equivalent to
concretely interpreting the program.
The second alternative that we have tested consists in applying the
evaluation function on each constraint at most once per intersection.
We call this approach {\em depth-first narrowing}, because, starting from
an intersection we visit every node that might be influenced by it.
After removing control dependence edges, if we are left with a strongly
connected component containing only {\em ascencing constraints}, that
is, constraints that only increase the upper bounds of intervals, then this
approach is guaranteed to find the least fixpoint.
The same is true if we have a SCC with only {\em descending constraints}.

\bibliographystyle{plain}
\bibliography{../../../bibtex/references}

\end{document}

### 2011-12-19 prototype/PythonRangeAnalysis/Cousot.py

@@ -1,0 +1,96 @@
+from DataStructures import *
+
+def widen_meet(op):
+  """This is the meet operator of the growth analysis. The growth analysis
+     will change the bounds of each variable, if necessary. Initially, each
+     variable is bound to either the undefined interval, e.g. [., .], or to
+     a constant interval, e.g., [3, 15]. After this analysis runs, there will
+     be no undefined interval. Each variable will be either bound to a
+     constant interval, or to [-, c], or to [c, +], or to [-, +]."""
+  ij = op.sink.interval
+  kl = op.eval()
+  if isinstance(ij, BottomInterval):
+    op.sink.interval = kl
+  elif lt(kl.l, ij.l) and gt(kl.u, ij.u):
+    op.sink.interval = Interval()
+  elif lt(kl.l, ij.l):
+    op.sink.interval = Interval('-', ij.u)
+  elif gt(kl.u, ij.u):
+    op.sink.interval = Interval(ij.l, '+')
+  return ij != op.sink.interval
+
+def narrow_meet(op):
+  """This is the meet operator of the cropping analysis. Whereas the growth
+     analysis expands the bounds of each variable, regardless of intersections
+     in the constraint graph, the cropping analysis shyrinks these bounds back
+     to ranges that respect the intersections."""
+  has_changed = False
+  ij = op.sink.interval
+  kl = op.eval()
+  if ij.l == '-' and kl.l != '-':
+    ij.l = kl.l
+    has_changed = True
+  elif ij.l != min(ij.l, kl.l):
+    ij.l = min(ij.l, kl.l)
+    has_changed = True
+  if ij.u == '+' and kl.u != '+':
+    ij.u = kl.u
+    has_changed = True
+  elif ij.u != max(ij.u, kl.u):
+    ij.u = max(ij.u, kl.u)
+    has_changed = True
+  return has_changed
+
+def iterate_till_fix_point(use_map, active_vars, meet):
+  """This method finds the growth pattern of each variable. If we pass to it
+     the widen_meet, then we will have the growth analysis. If we pass the
+     narrow_meet, then we will have the cropping analysis."""
+  if len(active_vars) > 0:
+    next_variable = active_vars.pop()
+    use_list = use_map[next_variable]
+    for op in use_list:
+      if meet(op):
+        active_vars.add(op.sink.name)
+    iterate_till_fix_point(use_map, active_vars, meet)
+
+def int_op(Variables, UseMap):
+  """This method finds which operations contain non-trivial intersection.
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/CropDFS.py

@@ -1,0 +1,117 @@
+from DataStructures import *
+
+def crop_meet(op):
+  """This is the meet operator of the cropping analysis. Whereas the growth
+     analysis expands the bounds of each variable, regardless of intersections
+     in the constraint graph, the cropping analysis shyrinks these bounds back
+     to ranges that respect the intersections. Notice that we need to have
+     a reference to the original state of the variable, e.g., 0, +, - and ?.
+     We cannot store this information in the interval itself, because this
+     interval is likely to change from, say, +, to a constant."""
+  new_int = op.eval()
+  has_changed = False
+  if (op.sink.state == '-' or op.sink.state == '?') and gt(new_int.l, op.sink.interval.l):
+    op.sink.interval.l = new_int.l
+    has_changed = True
+  if (op.sink.state == '+' or op.sink.state == '?') and lt(new_int.u, op.sink.interval.u):
+    op.sink.interval.u = new_int.u
+    has_changed = True
+  return has_changed
+
+def crop_dfs(use_map, active_ops, visited):
+  """This method performs a depth-first search on the constraint graph, always
+     applying the cropping operator on the nodes that it finds."""
+  if len(active_ops) > 0:
+    op = active_ops.pop()
+    if not op.sink.name in visited:
+      crop_meet(op)
+      visited.add(op.sink.name)
+      for next_use in use_map[op.sink.name]:
+        active_ops.add(next_use)
+    crop_dfs(use_map, active_ops, visited)
+
+def growth_meet(op):
+  """This is the meet operator of the growth analysis. The growth analysis
+     will change the bounds of each variable, if necessary. Initially, each
+     variable is bound to either the undefined interval, e.g. [., .], or to
+     a constant interval, e.g., [3, 15]. After this analysis runs, there will
+     be no undefined interval. Each variable will be either bound to a
+     constant interval, or to [-, c], or to [c, +], or to [-, +]."""
+  old_int = op.sink.interval
+  new_int = op.eval()
+  if isinstance(old_int, BottomInterval):
+    op.sink.interval = new_int
+  elif lt(new_int.l, old_int.l) and gt(new_int.u, old_int.u):
+    op.sink.interval = Interval()
+  elif lt(new_int.l, old_int.l):
+    op.sink.interval = Interval('-', old_int.u)
+  elif gt(new_int.u, old_int.u):
+    op.sink.interval = Interval(old_int.l, '+')
+  return old_int != op.sink.interval
+
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/DataStructures.py

@@ -1,0 +1,460 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
+        raise InvalidIntervalException(l, u)
+    self.l = l;
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/BellmanFord.py

@@ -1,0 +1,22 @@
+V = [0, 1, 2, 3, 4, 5, 5, 6, 7]
+E = {(0, 1):10, (2, 1):1, (2, 3):1, (3, 4):3, (4, 5):-1, (5, 2):-2, (6, 5):-1,
+  (6, 1):-4, (7, 6):1, (0, 7):8, (1, 5):2}
+
+dist = range(len(V))
+prev = range(len(V))
+
+def update((u, v)):
+  dist[v] = min(dist[v], dist[u] + E[(u, v)])
+
+def shortest_path(s):
+  for u in V:
+    dist[u] = 1000000
+    prev[u] = None
+  dist[s] = 0
+  for i in range(len(V)):
+    for e in E.keys():
+      update(e)
+  for u in V:
+    print dist[u]
+
+shortest_path(0)

### 2011-12-19 prototype/PythonRangeAnalysis/bck/Constraints.py

@@ -1,0 +1,160 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
+        raise InvalidIntervalException(l, u)
+    self.l = l;
+    self.u = u;
+  def intersect(self, i):
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/Cousot1.py

@@ -1,0 +1,83 @@
+from DataStructures import *
+
+def widen_meet(op):
+  """This is the meet operator of the growth analysis. The growth analysis
+     will change the bounds of each variable, if necessary. Initially, each
+     variable is bound to either the undefined interval, e.g. [., .], or to
+     a constant interval, e.g., [3, 15]. After this analysis runs, there will
+     be no undefined interval. Each variable will be either bound to a
+     constant interval, or to [-, c], or to [c, +], or to [-, +]."""
+  ij = op.sink.interval
+  kl = op.eval()
+  if isinstance(ij, BottomInterval):
+    op.sink.interval = kl
+  elif lt(kl.l, ij.l) and gt(kl.u, ij.u):
+    op.sink.interval = Interval()
+  elif lt(kl.l, ij.l):
+    op.sink.interval = Interval('-', ij.u)
+  elif gt(kl.u, ij.u):
+    op.sink.interval = Interval(ij.l, '+')
+  return ij != op.sink.interval
+
+def narrow_meet(op):
+  """This is the meet operator of the cropping analysis. Whereas the growth
+     analysis expands the bounds of each variable, regardless of intersections
+     in the constraint graph, the cropping analysis shyrinks these bounds back
+     to ranges that respect the intersections."""
+  has_changed = False
+  ij = op.sink.interval
+  kl = op.eval()
+  if ij.l == '-' and kl.l != '-':
+    ij.l = kl.l
+    has_changed = True
+  elif ij.l != min(ij.l, kl.l):
+    ij.l = min(ij.l, kl.l)
+    has_changed = True
+  if ij.u == '+' and kl.u != '+':
+    ij.u = kl.u
+    has_changed = True
+  elif ij.u != max(ij.u, kl.u):
+    ij.u = max(ij.u, kl.u)
+    has_changed = True
+  return has_changed
+
+def iterate_till_fix_point(use_map, active_vars, meet):
+  """This method finds the growth pattern of each variable. If we pass to it
+     the widen_meet, then we will have the growth analysis. If we pass the
+     narrow_meet, then we will have the cropping analysis."""
+  if len(active_vars) > 0:
+    next_variable = active_vars.pop()
+    use_list = use_map[next_variable]
+    for op in use_list:
+      if meet(op):
+        active_vars.add(op.sink.name)
+    iterate_till_fix_point(use_map, active_vars, meet)
+
+def int_op(Variables, UseMap):
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/DS0.py

@@ -1,0 +1,346 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
+        raise InvalidIntervalException(l, u)
+    self.l = l;
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/DS1.py

@@ -1,0 +1,349 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
+        raise InvalidIntervalException(l, u)
+    self.l = l;
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/RA1.py

@@ -1,0 +1,311 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
+        raise InvalidIntervalException(l, u)
+    self.l = l;
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/RA2.py

@@ -1,0 +1,350 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
+        raise InvalidIntervalException(l, u)
+    self.l = l;
+    self.u = u;
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/RA3.py

@@ -1,0 +1,350 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
+        raise InvalidIntervalException(l, u)
+    self.l = l;
+    self.u = u;
+  def intersection(self, i):
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/RA4.py

@@ -1,0 +1,399 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+def addItv(i1, i2):
+  return Interval(add(i1.l, i2.l), add(i1.u, i2.u))
+
+class Interval:
+  def __init__(self, l='-', u='+'):
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/RA5.py

@@ -1,0 +1,346 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+def addItv(i1, i2):
+  return Interval(add(i1.l, i2.l), add(i1.u, i2.u))
+
+class Interval:
+  def __init__(self, l='-', u='+'):
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/RA6.py

@@ -1,0 +1,383 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+def addItv(i1, i2):
+  return Interval(add(i1.l, i2.l), add(i1.u, i2.u))
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/RA7.py

@@ -1,0 +1,567 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/RA8.py

@@ -1,0 +1,542 @@
+#
+# Iterates the cropping meet operator until reaching fix point.
+#
+
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/RA9.py

@@ -1,0 +1,757 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/b1/Cousot.py

@@ -1,0 +1,83 @@
+from DataStructures import *
+
+def widen_meet(op):
+  """This is the meet operator of the growth analysis. The growth analysis
+     will change the bounds of each variable, if necessary. Initially, each
+     variable is bound to either the undefined interval, e.g. [., .], or to
+     a constant interval, e.g., [3, 15]. After this analysis runs, there will
+     be no undefined interval. Each variable will be either bound to a
+     constant interval, or to [-, c], or to [c, +], or to [-, +]."""
+  ij = op.sink.interval
+  kl = op.eval()
+  if isinstance(ij, BottomInterval):
+    op.sink.interval = kl
+  elif lt(kl.l, ij.l) and gt(kl.u, ij.u):
+    op.sink.interval = Interval()
+  elif lt(kl.l, ij.l):
+    op.sink.interval = Interval('-', ij.u)
+  elif gt(kl.u, ij.u):
+    op.sink.interval = Interval(ij.l, '+')
+  return ij != op.sink.interval
+
+def narrow_meet(op):
+  """This is the meet operator of the cropping analysis. Whereas the growth
+     analysis expands the bounds of each variable, regardless of intersections
+     in the constraint graph, the cropping analysis shyrinks these bounds back
+     to ranges that respect the intersections."""
+  has_changed = False
+  ij = op.sink.interval
+  kl = op.eval()
+  if ij.l == '-' and kl.l != '-':
+    ij.l = kl.l
+    has_changed = True
+  elif ij.l != min(ij.l, kl.l):
+    ij.l = min(ij.l, kl.l)
+    has_changed = True
+  if ij.u == '+' and kl.u != '+':
+    ij.u = kl.u
+    has_changed = True
+  elif ij.u != max(ij.u, kl.u):
+    ij.u = max(ij.u, kl.u)
+    has_changed = True
+  return has_changed
+
+def iterate_till_fix_point(use_map, active_vars, meet):
+  """This method finds the growth pattern of each variable. If we pass to it
+     the widen_meet, then we will have the growth analysis. If we pass the
+     narrow_meet, then we will have the cropping analysis."""
+  if len(active_vars) > 0:
+    next_variable = active_vars.pop()
+    use_list = use_map[next_variable]
+    for op in use_list:
+      if meet(op):
+        active_vars.add(op.sink.name)
+    iterate_till_fix_point(use_map, active_vars, meet)
+
+def int_op(Variables, UseMap):
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/b1/CropDFS.py

@@ -1,0 +1,100 @@
+from DataStructures import *
+
+def crop_meet(op):
+  """This is the meet operator of the cropping analysis. Whereas the growth
+     analysis expands the bounds of each variable, regardless of intersections
+     in the constraint graph, the cropping analysis shyrinks these bounds back
+     to ranges that respect the intersections. Notice that we need to have
+     a reference to the original state of the variable, e.g., 0, +, - and ?.
+     We cannot store this information in the interval itself, because this
+     interval is likely to change from, say, +, to a constant."""
+  new_int = op.eval()
+  has_changed = False
+  if (op.sink.state == '-' or op.sink.state == '?') and gt(new_int.l, op.sink.interval.l):
+    op.sink.interval.l = new_int.l
+    has_changed = True
+  if (op.sink.state == '+' or op.sink.state == '?') and lt(new_int.u, op.sink.interval.u):
+    op.sink.interval.u = new_int.u
+    has_changed = True
+  return has_changed
+
+def crop_dfs(use_map, active_ops, visited):
+  """This method performs a depth-first search on the constraint graph, always
+     applying the cropping operator on the nodes that it finds."""
+  if len(active_ops) > 0:
+    op = active_ops.pop()
+    if not op.sink.name in visited:
+      crop_meet(op)
+      visited.add(op.sink.name)
+      for next_use in use_map[op.sink.name]:
+        active_ops.add(next_use)
+    crop_dfs(use_map, active_ops, visited)
+
+def growth_meet(op):
+  """This is the meet operator of the growth analysis. The growth analysis
+     will change the bounds of each variable, if necessary. Initially, each
+     variable is bound to either the undefined interval, e.g. [., .], or to
+     a constant interval, e.g., [3, 15]. After this analysis runs, there will
+     be no undefined interval. Each variable will be either bound to a
+     constant interval, or to [-, c], or to [c, +], or to [-, +]."""
+  old_int = op.sink.interval
+  new_int = op.eval()
+  if isinstance(old_int, BottomInterval):
+    op.sink.interval = new_int
+  elif lt(new_int.l, old_int.l) and gt(new_int.u, old_int.u):
+    op.sink.interval = Interval()
+  elif lt(new_int.l, old_int.l):
+    op.sink.interval = Interval('-', old_int.u)
+  elif gt(new_int.u, old_int.u):
+    op.sink.interval = Interval(old_int.l, '+')
+  return old_int != op.sink.interval
+
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/b1/DataStructures.py

@@ -1,0 +1,366 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
+        raise InvalidIntervalException(l, u)
+    self.l = l;
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/b2/Cousot.py

@@ -1,0 +1,100 @@
+from DataStructures import *
+
+def widen_meet(op):
+  """This is the meet operator of the growth analysis. The growth analysis
+     will change the bounds of each variable, if necessary. Initially, each
+     variable is bound to either the undefined interval, e.g. [., .], or to
+     a constant interval, e.g., [3, 15]. After this analysis runs, there will
+     be no undefined interval. Each variable will be either bound to a
+     constant interval, or to [-, c], or to [c, +], or to [-, +]."""
+  ij = op.sink.interval
+  kl = op.eval()
+  if isinstance(ij, BottomInterval):
+    op.sink.interval = kl
+  elif lt(kl.l, ij.l) and gt(kl.u, ij.u):
+    op.sink.interval = Interval()
+  elif lt(kl.l, ij.l):
+    op.sink.interval = Interval('-', ij.u)
+  elif gt(kl.u, ij.u):
+    op.sink.interval = Interval(ij.l, '+')
+  return ij != op.sink.interval
+
+def narrow_meet(op):
+  """This is the meet operator of the cropping analysis. Whereas the growth
+     analysis expands the bounds of each variable, regardless of intersections
+     in the constraint graph, the cropping analysis shyrinks these bounds back
+     to ranges that respect the intersections."""
+  has_changed = False
+  ij = op.sink.interval
+  kl = op.eval()
+  if ij.l == '-' and kl.l != '-':
+    ij.l = kl.l
+    has_changed = True
+  elif ij.l != min(ij.l, kl.l):
+    ij.l = min(ij.l, kl.l)
+    has_changed = True
+  if ij.u == '+' and kl.u != '+':
+    ij.u = kl.u
+    has_changed = True
+  elif ij.u != max(ij.u, kl.u):
+    ij.u = max(ij.u, kl.u)
+    has_changed = True
+  return has_changed
+
+def iterate_till_fix_point(use_map, active_vars, meet):
+  """This method finds the growth pattern of each variable. If we pass to it
+     the widen_meet, then we will have the growth analysis. If we pass the
+     narrow_meet, then we will have the cropping analysis."""
+  if len(active_vars) > 0:
+    next_variable = active_vars.pop()
+    use_list = use_map[next_variable]
+    for op in use_list:
+      if meet(op):
+        active_vars.add(op.sink.name)
+    iterate_till_fix_point(use_map, active_vars, meet)
+
+def int_op(Variables, UseMap):
+  """This method finds which operations contain non-trivial intersection.
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/b3/Cousot.py

@@ -1,0 +1,99 @@
+from DataStructures import *
+
+def widen_meet(op):
+  """This is the meet operator of the growth analysis. The growth analysis
+     will change the bounds of each variable, if necessary. Initially, each
+     variable is bound to either the undefined interval, e.g. [., .], or to
+     a constant interval, e.g., [3, 15]. After this analysis runs, there will
+     be no undefined interval. Each variable will be either bound to a
+     constant interval, or to [-, c], or to [c, +], or to [-, +]."""
+  ij = op.sink.interval
+  kl = op.eval()
+  if isinstance(ij, BottomInterval):
+    op.sink.interval = kl
+  elif lt(kl.l, ij.l) and gt(kl.u, ij.u):
+    op.sink.interval = Interval()
+  elif lt(kl.l, ij.l):
+    op.sink.interval = Interval('-', ij.u)
+  elif gt(kl.u, ij.u):
+    op.sink.interval = Interval(ij.l, '+')
+  return ij != op.sink.interval
+
+def narrow_meet(op):
+  """This is the meet operator of the cropping analysis. Whereas the growth
+     analysis expands the bounds of each variable, regardless of intersections
+     in the constraint graph, the cropping analysis shyrinks these bounds back
+     to ranges that respect the intersections."""
+  has_changed = False
+  ij = op.sink.interval
+  kl = op.eval()
+  if ij.l == '-' and kl.l != '-':
+    ij.l = kl.l
+    has_changed = True
+  elif ij.l != min(ij.l, kl.l):
+    ij.l = min(ij.l, kl.l)
+    has_changed = True
+  if ij.u == '+' and kl.u != '+':
+    ij.u = kl.u
+    has_changed = True
+  elif ij.u != max(ij.u, kl.u):
+    ij.u = max(ij.u, kl.u)
+    has_changed = True
+  return has_changed
+
+def iterate_till_fix_point(use_map, active_vars, meet):
+  """This method finds the growth pattern of each variable. If we pass to it
+     the widen_meet, then we will have the growth analysis. If we pass the
+     narrow_meet, then we will have the cropping analysis."""
+  if len(active_vars) > 0:
+    next_variable = active_vars.pop()
+    use_list = use_map[next_variable]
+    for op in use_list:
+      if meet(op):
+        active_vars.add(op.sink.name)
+    iterate_till_fix_point(use_map, active_vars, meet)
+
+def int_op(Variables, UseMap):
+  """This method finds which operations contain non-trivial intersection.
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/b3/CropDFS.py

@@ -1,0 +1,120 @@
+from DataStructures import *
+
+def crop_meet(op):
+  """This is the meet operator of the cropping analysis. Whereas the growth
+     analysis expands the bounds of each variable, regardless of intersections
+     in the constraint graph, the cropping analysis shyrinks these bounds back
+     to ranges that respect the intersections. Notice that we need to have
+     a reference to the original state of the variable, e.g., 0, +, - and ?.
+     We cannot store this information in the interval itself, because this
+     interval is likely to change from, say, +, to a constant."""
+  new_int = op.eval()
+  has_changed = False
+  if (op.sink.state == '-' or op.sink.state == '?') and gt(new_int.l, op.sink.interval.l):
+    op.sink.interval.l = new_int.l
+    has_changed = True
+  if (op.sink.state == '+' or op.sink.state == '?') and lt(new_int.u, op.sink.interval.u):
+    op.sink.interval.u = new_int.u
+    has_changed = True
+  return has_changed
+
+def crop_dfs(use_map, active_ops, visited):
+  """This method performs a depth-first search on the constraint graph, always
+     applying the cropping operator on the nodes that it finds."""
+  if len(active_ops) > 0:
+    op = active_ops.pop()
+    if not op.sink.name in visited:
+      crop_meet(op)
+      visited.add(op.sink.name)
+      for next_use in use_map[op.sink.name]:
+        active_ops.add(next_use)
+    crop_dfs(use_map, active_ops, visited)
+
+def growth_meet(op):
+  """This is the meet operator of the growth analysis. The growth analysis
+     will change the bounds of each variable, if necessary. Initially, each
+     variable is bound to either the undefined interval, e.g. [., .], or to
+     a constant interval, e.g., [3, 15]. After this analysis runs, there will
+     be no undefined interval. Each variable will be either bound to a
+     constant interval, or to [-, c], or to [c, +], or to [-, +]."""
+  old_int = op.sink.interval
+  new_int = op.eval()
+  if isinstance(old_int, BottomInterval):
+    op.sink.interval = new_int
+  elif lt(new_int.l, old_int.l) and gt(new_int.u, old_int.u):
+    op.sink.interval = Interval()
+  elif lt(new_int.l, old_int.l):
+    op.sink.interval = Interval('-', old_int.u)
+  elif gt(new_int.u, old_int.u):
+    op.sink.interval = Interval(old_int.l, '+')
+  return old_int != op.sink.interval
+
[…]

### 2011-12-19 prototype/PythonRangeAnalysis/bck/b3/DataStructures.py

@@ -1,0 +1,431 @@
+class InvalidIntervalException(Exception):
+  def __init__(self, l, u):
+    self.l = l
+    self.u = u
+  def __str__(self):
+    return "Invalid interval [" + str(l) + ", " + str(u) + "]"
+
+class ArithmeticException(Exception):
+  def __init__(self, msg):
+    self.msg = msg
+  def __str__(self):
+    return self.msg
+
+def lt(x, y):
+  if x == y:
+    return False
+  elif x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def leq(x, y):
+  if x == y or x == '-' or y == '+':
+    return True
+  elif y == '-' or x == '+':
+    return False
+  else:
+    return x < y
+
+def gt(x, y):
+  if x == y:
+    return False
+  elif x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def geq(x, y):
+  if x == y or x == '+' or y == '-':
+    return True
+  elif y == '+' or x == '-':
+    return False
+  else:
+    return x > y
+
+def min(x, y):
+  if lt(x, y):
+    return x
+  else:
+    return y
+
+def max(x, y):
+  if gt(x, y):
+    return x
+  else:
+    return y
+
+def add(x, y):
+  if (x == '-' and y == '+') or (y == '-' and x == '+'):
+    raise ArithmeticException("Adding minus and plus infinity.")
+  elif x == '-' or y == '-':
+    return '-'
+  elif x == '+' or y == '+':
+    return '+'
+  else:
+    return x + y
+
+def mul(x, y):
+  if (x == 0 or y == '0'):
+    return 0
+  elif (x == '-' and lt(y, 0)) or (y == '-' and lt(x, 0)):
+    return '+'
+  elif (x == '+' and gt(y, 0)) or (y == '+' and gt(x, 0)):
+    return '+'
+  elif (x == '-' and gt(y, 0)) or (y == '-' and gt(x, 0)):
+    return '-'
+  elif (x == '+' and lt(y, 0)) or (y == '+' and lt(x, 0)):
+    return '-'
+  else:
+    return x * y
+
+class Interval:
+  def __init__(self, l='-', u='+'):
+    if (l != '-' and u != '+'):
+      if (l > u):
+        raise InvalidIntervalException(l, u)
+    self.l = l;
[…]

### 2011-12-19 src/RangeAnalysis.cpp

@@ -1,0 +1,1964 @@
+//===-------------------------- RangeAnalysis.cpp -------------------------===//
+//===----- Performs a range analysis of the variables of the function -----===//
+//
+//                     The LLVM Compiler Infrastructure
+//
+// This file is distributed under the University of Illinois Open Source
+// License. See LICENSE.TXT for details.
+//
+//===----------------------------------------------------------------------===//
+
+
+#define DEBUG_TYPE "range-analysis"
+
+#include "RangeAnalysis.h"
+#include "llvm/Support/InstIterator.h"
+
+using namespace llvm;
+
+// These macros are used to get stats regarging the precision of our analysis.
+STATISTIC(usedBits, "Initial number of bits.");
+STATISTIC(needBits, "Needed bits.");
+STATISTIC(totalInstructions, "Number of instructions of the function.");
+STATISTIC(evaluatedInstructions, "Number of evaluated instructions.");
+STATISTIC(percentReduction, "Percentage of reduction of the number of bits.");
+
+// The number of bits needed to store the largest variable of the function.
+unsigned MAX_BIT_INT = 1;
+
+// This map is used to store the number of times that the widen_meet 
+// operator is called on a variable. It was a Fernando's suggestion.
+DenseMap<const Value*, unsigned> FerMap;
+
+// ========================================================================== //
+// Static global functions and definitions
+// ========================================================================== //
+
+// The min and max integer values for a given bit width.
+static APInt Min = APInt::getSignedMinValue(MAX_BIT_INT);
+static APInt Max = APInt::getSignedMaxValue(MAX_BIT_INT);
+
+/// Selects the instructions that we are going to evaluate.
+static bool isValidInstruction(const Instruction* I) {
+	// FIXME: How can I reference a Phi function by its opcode?
+	if (dyn_cast<PHINode>(I)) {
+		return true;
+	}
+
+	switch (I->getOpcode()) {
+	case Instruction::Add:   return true;
[…]

### 2011-12-19 src/RangeAnalysis.h

@@ -1,0 +1,700 @@
+//===-------------------------- RangeAnalysis.cpp -------------------------===//
+//===-----Performs the Range analysis of the variables of the function-----===//
+//
+//                     The LLVM Compiler Infrastructure
+//
+// This file is distributed under the University of Illinois Open Source
+// License. See LICENSE.TXT for details.
+//
+//===----------------------------------------------------------------------===//
+// This file contains a pass that performs range analysis. The objective of
+// the range analysis is to map each integer variable in the program to the
+// range of possible values that it might assume through out the program
+// execution. Ideally this range should be as constrained as possible, so that
+// an optimizing compiler could learn more information about each variable.
+// However, the range analysis must be conservative, that it, it will only
+// constraint the range of a variable if it can prove that it is safe to do so.
+// As an example, consider the program:
+//
+// i = read();
+// if (i < 10) {
+//   print (i + 1);
+// else {
+//   print(i - 1);
+// }
+//
+// In this program we know, from the conditional test, that the value of i in
+// the true side of the branch is in the range [-INF, 9], and in the false side
+// is in the range [10, +INF].
+//
+//
+//===----------------------------------------------------------------------===//
+
+#ifndef LLVM_TRANSFORMS_RANGEANALYSIS_RANGEANALYSIS_H_
+#define LLVM_TRANSFORMS_RANGEANALYSIS_RANGEANALYSIS_H_
+
+#include "llvm/Function.h"
+#include "llvm/Pass.h"
+#include "llvm/Constants.h"
+#include "llvm/Instructions.h"
+#include "llvm/ADT/DenseMap.h"
+#include "llvm/ADT/SmallPtrSet.h"
+#include "llvm/ADT/Statistic.h"
+#include "llvm/Support/Debug.h"
+#include "llvm/Support/raw_ostream.h"
+#include "llvm/Support/ConstantRange.h"
+#include "llvm/Transforms/Utils/SSI.h"
+#include <deque>
+#include <stack>
+#include <set>
+
+
+namespace llvm {
+
+class RangeAnalysis: public FunctionPass {
+
+public:
+	static char ID; // Pass identification, replacement for typeid
+	RangeAnalysis() : FunctionPass(&ID) {}
+	void getAnalysisUsage(AnalysisUsage &AU) const;
+	bool runOnFunction(Function &F);
+
+private:
+	/// Gets the maximum bit width of the operands in
+	/// the instructions of the function.
[…]

### 2012-02-23 src/RangeAnalysis/RangeAnalysis.cpp
Implemented an update function that iterate N times before widening.

@@ -1956,6 +1956,14 @@ void CropDFS::storeAbstractStates(const SmallPtrSet<VarNode*, 32> *component) {
 	}
 }
 
+bool Meet::fixed(BasicOp* op){
+	Range oldInterval = op->getSink()->getRange();
+	Range newInterval = op->eval();
+
+	op->getSink()->setRange(newInterval);
+	return oldInterval != newInterval;
+}
+
 /// This is the meet operator of the growth analysis. The growth analysis
 /// will change the bounds of each variable, if necessary. Initially, each
 /// variable is bound to either the undefined interval, e.g. [., .], or to
@@ -1963,7 +1971,6 @@ void CropDFS::storeAbstractStates(const SmallPtrSet<VarNode*, 32> *component) {
 /// be no undefined interval. Each variable will be either bound to a
 /// constant interval, or to [-, c], or to [c, +], or to [-, +].
 bool Meet::widen(BasicOp* op) {
-
 	Range oldInterval = op->getSink()->getRange();
 	Range newInterval = op->eval();
 
@@ -2141,44 +2148,11 @@ void CropDFS::crop(const UseMap &compUseMap, BasicOp *op) {
 	}
 }
 
-void ConstraintGraph::update(SmallPtrSet<const Value*, 6>& actv
-		, bool(*meet)(BasicOp* op)) {
-	update(*useMap, actv, meet);
-}
-
 void ConstraintGraph::update(const UseMap &compUseMap,
 		SmallPtrSet<const Value*, 6>& actv, bool(*meet)(BasicOp* op)) {
-	/*
-	 if (actv.empty()) {
-	 return;
-	 }
-
-	 const Value* V = *actv.begin();
-	 actv.erase(V);
-
-	 // The use list.
-	 SmallPtrSet<BasicOp*, 8> L = compUseMap.find(V)->second;
-	 SmallPtrSet<BasicOp*, 8>::iterator bgn = L.begin(),	end = L.end();
-	 for (; bgn != end; ++bgn) {
-	 if (meet(*bgn)) {
-	 // I want to use it as a set, but I also want
-	 // keep an order or insertions and removals.
-	 if (!actv.count((*bgn)->getSink()->getValue())) {
-	 actv.insert((*bgn)->getSink()->getValue());
-	 }
-
-	 }
-	 }
-
-	 update(compUseMap, actv, meet);
-	 */
-
-//	int i = 20;
[…]

### 2012-02-23 src/RangeAnalysis/RangeAnalysis.h

@@ -85,6 +85,8 @@
 #define ASSERT(cond,msg)
 #endif
 
+#define NUMBER_FIXED_ITERATIONS 20
+
 using namespace llvm;
 
 namespace {
@@ -668,9 +670,10 @@ private:
 	// Perform the widening and narrowing operations
 
 protected:
-	void update(SmallPtrSet<const Value*, 6>& actv, bool (*meet)(BasicOp* op));
 	void update(const UseMap &compUseMap,
 		SmallPtrSet<const Value*, 6>& actv, bool (*meet)(BasicOp* op));
+	void update(unsigned nIterations, const UseMap &compUseMap,
+			SmallPtrSet<const Value*, 6>& actv);
 
 	virtual void preUpdate(const UseMap &compUseMap,
 		SmallPtrSet<const Value*, 6>& entryPoints) = 0;
@@ -777,6 +780,7 @@ public:
 	static bool narrow(BasicOp* op);
 	static bool crop(BasicOp* op);
 	static bool growth(BasicOp* op);
+	static bool fixed(BasicOp* op);
 };
 
 class RangeAnalysis{

### 2012-02-23 tests/compile.py
Included a new test that need some iteration before widening.

@@ -3,12 +3,12 @@
 import sys
 import os
 
-passPath = "/work/victorsc/llvm-3.0/Debug/lib/"
+passPath = "~/workspace/ra/llvm-3.0/Debug/lib/"
 
 def checkArgs(args):
     if len(args) != 3:
         return False
-    if args[1] in ["-ra-inter-crop","-ra-inter-cousot","-ra-intra-crop","-ra-intra-cousot"]:
+    if args[1] in ["-ra-inter-crop","-ra-inter-cousot","-ra-intra-crop","-ra-intra-cousot","-ra-test-range"]:
         return True
     else:
         return False
@@ -25,9 +25,7 @@ if checkArgs(sys.argv):
     os.system("opt -load "+passPath+"vSSA.so -vssa "+fileName+".bc -o "+fileName+".essa.bc")
     print("Running the Range Analysis Pass")
     print("opt -load "+passPath+"RangeAnalysis.so "+sys.argv[1]+" "+fileName+".essa.bc")
-    os.system("opt -load "+passPath+"RangeAnalysis.so "+sys.argv[1]+" "+fileName+".essa.bc")
-
-#    os.system("opt -load "+passPath+"RangeAnalysis.so -load "+passPath+"Matching.so -matching "+fileName+".essa.bc")    
+    os.system("opt -load "+passPath+"RangeAnalysis.so "+sys.argv[1]+" "+fileName+".essa.bc -stats")
 else:
     print "./compile.py <OPTION> <FILE.c>"
     print "\tOPTION:"

### 2012-02-23 tests/t10.c

@@ -1,0 +1,20 @@
+/*
+   This program illustrates a situation in which an one-iteration widening is imprecise. 
+   The benchmark has a SCC with two entry-points that are phis, and single iteration followed by widening leads to very conservative bounds.
+ */
+int main(int argc, char** argv) {
+    int i = 0;
+        while (1) {
+            int tooLong = 0;
+                while (i <= argc) {
+                        if (i == argc) {
+                                tooLong = 1;
+                        }
+                }
+                if (tooLong) {
+                        break;
+                }
+        }
+        return i;
+}
+

### 2012-02-23 doc/Manuscript/paper.tex
Added new version of the manuscript, in LNCS format.

--- 2011-12-19 doc/Manuscript/doc.tex
@@
-% Add example showing that if not SSA form, then could be many iterations for
-% Cousot's operator.
+\documentclass{llncs}
 
-% Add example showing that Cousot's does not reach fix-point always.
-
-% Implement bck/RA8 and bck/RA9 to see how closer you get to the fix-point.
-
-% Implement the fix point operator to solve the range analysis problem.
-
-% How many instructions you cannot handle?
-
-\documentclass[preprint]{sigplanconf}
-
+\usepackage{proof}
 \usepackage{amssymb}
@@
 
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{example}[theorem]{Example}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{definition}[theorem]{Definition}
-
 \newcommand{\fun}[1]{\mbox{\textbf{#1}}}
@@
 
-\def\qed{\unskip\kern 10pt{\unitlength1pt\linethickness{.4pt}\framebox(6,6){}}}
+\sloppy
 
@@
 
-\conferenceinfo{Crazy Horse '11}{February 30th, 2011, Tumba, CA.} 
-\copyrightyear{2011} 
-\copyrightdata{[to be supplied]}
+\title {Range Analysis of Whole Programs}
 
-\titlebanner{banner above paper title}
-\preprintfooter{Range Analysis of Whole Programs}
+\author{Victor Sperle Campos, Igor Rafael, Douglas do Couto Teixeira,
+Fernando Magno Quint\~{a}o Pereira}
 
-\title{Range Analysis of Whole Programs}
-\subtitle{}
-
-\authorinfo{Homer Sympson \and Luck Skywalker \and Robin Hood \and
-  Frodo Beggins}
-           {Computer Science Department -- UFMG -- Brazil}
-           {\{hsympson,skywalker,rbhood,frodo\}@dcc.ufmg.br}
+\institute{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
+\email{\{victorsc,igor.rafael,douglas,fernando\}@dcc.ufmg.br}
+}
 
@@
 
+%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
 \begin{abstract}
-This paper describes a range analysis algorithm that is linear on the
-program size.
-We handle comparisons between variables via a two-phases dataflow analysis, and
-achieve flow sensitiveness by using the Extended Static Single Assignment
-form as the intermediate representation.
-We have implemented an inter-procedural version of our technique into LLVM, an
-industrial strenght compiler.
+This paper describes an inter-procedural range analysis algorithm that scales
+up to programs with millions of assembly instructions.
+Contrary to many previous techniques, we handle comparisons between variables
+without resorting to expensive algorithms.
+We achieve path sensitiveness by using Bodik's Extended Static Single
+Assignment form as the intermediate representation.
+We show that by processing the strongly connected components that constitute
+the graph of dependences between variables in topological order we not only
+gain in time, but also in precision.
+We have implemented our technique into LLVM, an industrial strenght compiler,
+and have been able to process over 4 million assembly instructions in a
+few seconds.
 \end{abstract}
-
-\category{D - Software}{D.3 Programming Languages}{D.3.4 Processors - Optimization}
-
-\terms
-Languages, Performance, Experimentation
-
-\keywords
-Range Analyisis, Dataflow Analysis, Integer arithmetics
 
@@
 \label{sec:intro}
+
+% The context: why range analysis is necessary.
+The analysis of integer variables on the interval lattice has been the
+canonical example of abstract interpretation since its introduction by
+Cousot and Cousot~\cite{Cousot77}.
+Compilers use range analysis to infer the possible values that discrete
+variables may assume during the execution of programs.
+This analysis has many uses.
+For instance, it allows the optimizing compiler to remove from the program text
+redundant overflow tests~\cite{Sol11} and unnecessary array bound
+checks~\cite{Bodik00}.
+Additionally, range analysis is essential not only to the bitwidth aware
+register allocator~\cite{Barik06,Tallam03}, but also to more traditional
+allocators that handle registers of different
+sizes~\cite{Kong98,Pereira08,Scholz02}.
+Finally, range analysis has also seen use in the static prediction of
+branches~\cite{Patterson95}, to detect buffer overflow
+vulnerabilities~\cite{Simon08,Wagner00}, to find the trip count of
+loops~\cite{Lokuciejewski09}
+and even in the synthesis of hardware~\cite{Cong05,Mahlke01}.
+
+% The problem: few implementations, not extensive experimental reports.
+Given this great importance, it comes as no surprise that the compiler
+literature is rich in works describing in details algorithmic variations of
+range analyses~\cite{Gawlitza09,Mahlke01,Stephenson00,Su05}.
+On the other hand, none of these authors provide experimental evidence that
+their approaches are able to deal with very large programs.
+There are researchers who have implemented range analyses that scale up to
+large programs~\cite{Blanchet03,Venet04}; nevertheless, because the
+algorithm itself is not the main focus of their works, they neither give
+details about their design choices nor provide experimental data about.
+This scenario was recently changed by Oh {\em et al.}~\cite{Oh12}, who
+introduced an abstract interpretation framework which handles programs with
+hundreds of thousands of lines of code.
+Nevertheless, Oh {\em et al.} have designed a very simple range analysis,
+which does not handle comparisons between variables, for instance.
+They also do not discuss the precision of their implementation, but only its
+runtime.
+
+% Our contribution.
+In this paper we provide a complete description of a range analysis algorithm,
+and show extensive experimental data that justifies our engineering choices.
+Our main algorithmic contribution on top of previous works is a three-phases
+approach to handle comparisons between variables without resorting to any
+exponential time technique.
+The few publicly available implementations of range analyses that we are
+aware of, such as those in Java Hotspot~\footnote{} and gcc~\footnote{} only
+deal with comparisons between variables and constants.
+Even the theoretical works, such as Su and Wagner's~\cite{Su05} or Gawlitza
+{\em et al.}'s~\cite{Gawlitza09} suffer from this limitation.
+This deficiency is one of the reasons explaining why none of these works has
+made their way into industrial-strength compilers.
+Two other insights allow our implementation to scale up to very large
+programs.
+Firstly, we use Bodik's Extended Static Single Assignment (e-SSA) form to
+perform range analysis sparsely.
+This program representation ensures that the interval associated with
+a variable is constant along its entire live range, and allows us to compute
+path-sensitive results.
+Secondly, we process the strongly connected components that underlines our
+constraint system in topological ordering.
+It is well-know that this technique is essential to speedup constraint solving
+algorithms~\cite[Sec 6.3]{Nielson99}; however, we show that a careful
+propagation of information along strong components not only gives us speed, but
+also improves the precision of our results.
+
+% The history of our contribution.
+This work concludes a two years long effort to produce a solid and scalable
+implementation of range analysis.
+Our first endeavor to implement such an algorithm was based on Su and
+Wagner's constraint system, which can be solved exactly in polynomial
+time~\cite{Su04,Su05}.
+However, although we could use their formula to handle a subset of C-like
+constructs, their description of how to deal with loops was
+``not very explicit"~\cite[p.422]{Gawlitza09}.
+Thus, in order to solve loops we adopted Gawlitza
+{\em et al.}'s~\cite{Gawlitza09} approach.
+This technique uses the Bellman-Ford algorithm to detect increasing or
+decreasing cycles in the constraint system, and then saturates these cycles
+via a simple widening operator.
+A detailed description of our implementation has been published by
+Couto and Pereira~\cite{Couto11}.
+Nevertheless, the inability to handle comparisons between variables, and the
+cubic complexity of the Bellman-Ford method eventually leads us to seek
+alternative solutions to range analysis.
+This quest reached a pinnacle in the present work.
+
+% Our experimental numbers.
+We have implemented our algorithm in the LLVM compiler~\cite{Lattner04}, and
+have used it to analyze a test suite with 1.4 million lines of code.
+In Section~\ref{sec:exp} we provide empirical evidence showing that our
+implementation is fast: it analyzes the gcc source code in XX seconds, for
+instance.
+It is also precise: our results are similar to Stephenson {\em et al}'s,
+even though, contrary to them, we do not assume that the program is correct.
+Furthermore, we have been able to find tight bounds to many of the examples
+used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
+{\em et al.}~\cite{Lakhdar11}, even though they use much more costly methods.
+
+% The rest of this paper.
 
@@
 Our objective is to solve a constraint system $C$, formed by instances of the
-five different types of constraints seen in Figure~\ref{fig:eval_function}
+six different types of constraints seen in Figure~\ref{fig:eval_function}
 (left).
@@
 \begin{eqnarray*}
-\begin{array}{rc}
+\begin{array}{r@{\hspace{0.5cm}}c}
 Y = [l, u]
@@
 \caption{\label{fig:eval_function}
-The valuation function.}
+The constraint system that produces instances of the range analysis problem.}
 \end{small}
@@
 \begin{center}
-\includegraphics[width=\columnwidth]{images/ex1}
+\includegraphics[width=0.7\textwidth]{images/ex1}
 \end{center}
@@
 
-\section{Polynomial Time Range Analysis}
+\section{Range Analysis in Three Steps}
 \label{sec:algo}
@@
 \begin{center}
-\includegraphics[width=\columnwidth]{images/algorithm}
+\includegraphics[width=0.8\textwidth]{images/algorithm}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\columnwidth]{images/ex_standard_eSSA}
+\includegraphics[width=0.5\textwidth]{images/ex_standard_eSSA}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=\columnwidth]{images/ex_non_pruned}
+\includegraphics[width=0.9\textwidth]{images/ex_non_pruned}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=\columnwidth]{images/ex_graph}
+\includegraphics[width=0.7\textwidth]{images/ex_graph}
 \end{center}
@@
 \begin{figure}[t!]
-\begin{tabular}{cc}
+\begin{center}
+\begin{tabular}{c@{\hspace{1.5cm}}c}
 \begin{minipage}{2.4cm}
@@
 \end{tabular}
+\end{center}
 \caption{\label{fig:growth_analysis}
@@
 \inferrule{Y = X \sqcap [l, \fun{ft}(V) + c] \\ \ub{I[V]} = u}
-{Y = X \sqcap [l, u + c]}, u, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
+{Y = X \sqcap [l, u + c]} \mbox{\hspace{0.3cm}} u, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
 \\
@@
 \inferrule{Y = X \sqcap [\fun{ft}(V) + c, u] \\ \lb{I[V]} = l}
-{Y = X \sqcap [l + c, u]}, l, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
+{Y = X \sqcap [l + c, u]} \mbox{\hspace{0.3cm}} l, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
 \end{array}
@@
 \begin{center}
-\includegraphics[width=\columnwidth]{images/ex_partition_grow_crop}
+\includegraphics[width=0.8\textwidth]{images/ex_partition_grow_crop}
 \end{center}
@@
 
-\subsection{Alternative Narrowing Operators}
-\label{sub:alt}
-
-Cousot and Cousot's narrowing operator, given in
-Figure~\ref{fig:crop_analysis} is not guaranteed to find the least fixpoint
-of a system of constraints.
-In order to improve the bounds, trading time for precision, we have
-experimented with two other operators.
-The first, is the evaluation function itself, which is given in
-Figure~\ref{fig:eval_function}.
-By iterating it until reaching a fixpoint, we are guaranteed to find the
-optimal solution to the narrowing analysis.
-However, this approach is not practical, because it is equivalent to
-concretely interpreting the program.
-The second alternative that we have tested consists in applying the
-evaluation function on each constraint at most once per intersection.
-We call this approach {\em depth-first narrowing}, because, starting from
-an intersection we visit every node that might be influenced by it.
-After removing control dependence edges, if we are left with a strongly
-connected component containing only {\em ascencing constraints}, that
-is, constraints that only increase the upper bounds of intervals, then this
-approach is guaranteed to find the least fixpoint.
-The same is true if we have a SCC with only {\em descending constraints}.
+%\subsection{Alternative Narrowing Operators}
+%\label{sub:alt}
+%
+%Cousot and Cousot's narrowing operator, given in
+%Figure~\ref{fig:crop_analysis} is not guaranteed to find the least fixpoint
+%of a system of constraints.
+%In order to improve the bounds, trading time for precision, we have
+%experimented with two other operators.
+%The first, is the evaluation function itself, which is given in
+%Figure~\ref{fig:eval_function}.
+%By iterating it until reaching a fixpoint, we are guaranteed to find the
+%optimal solution to the narrowing analysis.
+%However, this approach is not practical, because it is equivalent to
+%concretely interpreting the program.
+%The second alternative that we have tested consists in applying the
+%evaluation function on each constraint at most once per intersection.
+%We call this approach {\em depth-first narrowing}, because, starting from
+%an intersection we visit every node that might be influenced by it.
+%After removing control dependence edges, if we are left with a strongly
+%connected component containing only {\em ascencing constraints}, that
+%is, constraints that only increase the upper bounds of intervals, then this
+%approach is guaranteed to find the least fixpoint.
+%The same is true if we have a SCC with only {\em descending constraints}.
 
 \bibliographystyle{plain}
-\bibliography{../../../bibtex/references}
+\bibliography{bibtex/references}
 

### 2012-02-23 doc/Manuscript/paper.tex
Added new references to the introduction.

@@
 and show extensive experimental data that justifies our engineering choices.
-Our main algorithmic contribution on top of previous works is a three-phases
+Our first algorithmic contribution on top of previous works is a three-phases
 approach to handle comparisons between variables without resorting to any
@@
 The few publicly available implementations of range analyses that we are
-aware of, such as those in Java Hotspot~\footnote{} and gcc~\footnote{} only
-deal with comparisons between variables and constants.
-Even the theoretical works, such as Su and Wagner's~\cite{Su05} or Gawlitza
+aware of, such as those in FLEX~\footnote{The MIT's FLEX/Harpoon compiler
+provides an implementation of Stephenson's algorithm~\cite{Stephenson00}, and
+is available at \texttt{http://flex.cscott.net/Harpoon/}.} and
+gcc~\footnote{Gcc's VRP pass implements a variant of Patterson's
+algorithm~\cite{Patterson95}} only deal with comparisons between variables and
+constants.
+Even theoretical works, such as Su and Wagner's~\cite{Su05} or Gawlitza
 {\em et al.}'s~\cite{Gawlitza09} suffer from this limitation.
@@
 programs.
-Firstly, we use Bodik's Extended Static Single Assignment (e-SSA) form to
+We use Bodik's Extended Static Single Assignment (e-SSA) form~\cite{Bodik00} to
 perform range analysis sparsely.
@@
 path-sensitive results.
-Secondly, we process the strongly connected components that underlines our
+Finally, we process the strongly connected components that underline our
 constraint system in topological ordering.
@@
 instance.
-It is also precise: our results are similar to Stephenson {\em et al}'s,
-even though, contrary to them, we do not assume that the program is correct.
+It is also precise: our results are similar to Stephenson
+{\em et al}'s~\cite{Stephenson00}, even though, contrary to them, we do not
+assume that the program is correct.
 Furthermore, we have been able to find tight bounds to many of the examples
 used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
-{\em et al.}~\cite{Lakhdar11}, even though they use much more costly methods.
+{\em et al.}~\cite{Lakhdar11}, who rely on much more costly methods.
 
@@
 
-\begin{figure}[t!]
+\begin{figure}[hbt]
 \begin{small}
@@
 
+\section{Experiments}
+\label{sec:exp}
+
 \bibliographystyle{plain}
-\bibliography{bibtex/references}
+\bibliography{references}
 

### 2012-02-23 src/RangeAnalysis/RangeAnalysis.cpp
Fixed the update with fixed number of iterations. The problem was related with the generation of entry points.

@@ -2179,8 +2179,8 @@ void ConstraintGraph::update(unsigned nIterations, const UseMap &compUseMap,
 			if(nIterations == 0) break;
 			else --nIterations;
 
-			Meet::fixed(*bgn);
-			actv.insert((*bgn)->getSink()->getValue());
+			if(Meet::fixed(*bgn))
+				actv.insert((*bgn)->getSink()->getValue());
 		}
 	}
 }
@@ -2214,41 +2214,43 @@ void ConstraintGraph::findIntervals() {
 		// STATS
 		if (component.size() == 1) {
 			++numAloneSCCs;
-		}
-		if (component.size() > sizeMaxSCC) {
-			sizeMaxSCC = component.size();
-		}
+		}/*else{*/
+			if (component.size() > sizeMaxSCC) {
+				sizeMaxSCC = component.size();
+			}
 
-		//PRINTCOMPONENT(component)
+			//PRINTCOMPONENT(component)
 
-		UseMap compUseMap = buildUseMap(component);
+			UseMap compUseMap = buildUseMap(component);
 
-		// Get the entry points of the SCC
-		SmallPtrSet<const Value*, 6> entryPoints;
-		generateEntryPoints(component, entryPoints);
+			// Get the entry points of the SCC
+			SmallPtrSet<const Value*, 6> entryPoints;
 
-		//iterate a fixed number of time before widening
-		update(NUMBER_FIXED_ITERATIONS, compUseMap, entryPoints);
+			generateEntryPoints(component, entryPoints);
+			//iterate a fixed number of time before widening
+			update(component.size()*2 | NUMBER_FIXED_ITERATIONS, compUseMap, entryPoints);
 
 #ifdef PRINT_DEBUG
-		if (func)
-			printToFile(*func, "/tmp/" + func->getName() + "cgfixed.dot");
+			if (func)
+				printToFile(*func, "/tmp/" + func->getName() + "cgfixed.dot");
 #endif
-		// Primeiro iterate till fix point
-		preUpdate(compUseMap, entryPoints);
-		fixIntersects(component);
 
-		//printResultIntervals();
+			generateEntryPoints(component, entryPoints);
+			// Primeiro iterate till fix point
+			preUpdate(compUseMap, entryPoints);
+			fixIntersects(component);
+
+			//printResultIntervals();
 #ifdef PRINT_DEBUG
-		if (func)
-			printToFile(*func, "/tmp/" + func->getName() + "cgint.dot");
[…]

### 2012-02-23 src/RangeAnalysis/RangeAnalysis.h

@@ -57,7 +57,7 @@
 //#define SCC_DEBUG
 
 //TODO: comment the line below to disable the dot printing of Constraint Graphs
-#define PRINT_DEBUG
+//#define PRINT_DEBUG
 
 #define PRINTCOMPONENT(component) errs() << "\n--------------\n"; \
 		for (SmallPtrSetIterator<VarNode*> cit = component.begin(), cend = component.end(); cit != cend; ++cit) { \

### 2012-02-23 src/RangeAnalysis/RangeAnalysis.cpp
vSSA:  - Fixed bug in GEP instructions. Uses inside those instruction cannot be renamed.
RangeAnalysis:
 - Changed addVarNode function to be a little more efficient.
 - Fixed bug in init function.
 - Add time measurement about SCC resolution.

@@ -212,18 +212,20 @@ bool InterProceduralRA<CGT>::runOnModule(Module &M) {
 	updateMinMax(MAX_BIT_INT);
 
 	// Build the Constraint Graph by running on each function
+	Profile::TimeValue before = prof.timenow();
+	
 	for (Module::iterator I = M.begin(), E = M.end(); I != E; ++I) {
 		// If the function is only a declaration, or if it has variable number of arguments, do not match
 		if (I->isDeclaration() || I->isVarArg())
 			continue;
-
-		Profile::TimeValue before = prof.timenow();
+			
 		G->buildGraph(*I);
-		Profile::TimeValue elapsed = prof.timenow() - before;
-		prof.updateTime("BuildGraph", elapsed);
 		MatchParametersAndReturnValues(*I, *G);
 	}
 	G->buildVarNodes();
+	
+	Profile::TimeValue elapsed = prof.timenow() - before;
+	prof.updateTime("BuildGraph", elapsed);
 	prof.printTime("BuildGraph");
 
 #ifdef PRINT_DEBUG
@@ -1018,8 +1020,10 @@ void VarNode::init(bool outside) {
 	const Value* V = this->getValue();
 	if (const ConstantInt *CI = dyn_cast<ConstantInt>(V)) {
 		APInt tmp = CI->getValue();
-		APInt value = tmp.sextOrTrunc(MAX_BIT_INT);
-		this->setRange(Range(value, value));
+		if (tmp.getBitWidth() < MAX_BIT_INT) {
+			tmp = tmp.sext(MAX_BIT_INT);
+		}
+		this->setRange(Range(tmp, tmp));
 	} else {
 		if (!outside) {
 			// Initialize with a basic, unknown, interval.
@@ -1506,8 +1510,10 @@ ConstraintGraph::~ConstraintGraph() {
 
 /// Adds a VarNode to the graph.
 VarNode* ConstraintGraph::addVarNode(const Value* V) {
-	if (this->vars->count(V)) {
-		return this->vars->find(V)->second;
+	VarNodes::iterator vit = this->vars->find(V);
+	
+	if (vit != this->vars->end()) {
+		return vit->second;
 	}
 
 	VarNode* node = new VarNode(V);
@@ -1944,7 +1950,6 @@ void ConstraintGraph::buildVarNodes() {
 	VarNodes::iterator bgn = this->vars->begin(), end = this->vars->end();
 	for (; bgn != end; ++bgn) {
[…]

### 2012-02-23 src/vSSA/vSSA.cpp

@@ -86,6 +86,13 @@ void vSSA::createSigmasIfNeeded(BasicBlock *BB)
 		
 		if (isa<Instruction>(condition) || isa<Argument>(condition)) {
 			insertSigmas(ti, condition);
+			
+			// If the operand is a result of a indirect instruction (e.g. ZExt, SExt, Trunc),
+			// Create sigmas for the operands of the operands too
+			CastInst *cinst = NULL;
+			if ((cinst = dyn_cast<CastInst>(condition))) {
+				insertSigmas(ti, cinst->getOperand(0));
+			}
 		}
 	}
 	
@@ -191,11 +198,14 @@ void vSSA::renameUsesToSigma(Value *V, PHINode *sigma)
 		if (usepointers[i] ==  NULL) {
 			continue;
 		}
+		if (usepointers[i] == sigma) {
+			continue;
+		}
+		if (isa<GetElementPtrInst>(usepointers[i])) {
+			continue;
+		}
 		
 		BasicBlock *BB_user = usepointers[i]->getParent();
-		
-		if (usepointers[i] == sigma)
-			continue;
 		
 		// Check if the use is in the dominator tree of sigma(V)
 		if (DT_->dominates(BB_next, BB_user)){
@@ -581,7 +591,8 @@ bool vSSA::dominateOrHasInFrontier(BasicBlock *BB, BasicBlock *BB_next, Value *v
 	for (Value::use_iterator begin = value->use_begin(), end = value->use_end(); begin != end; ++begin) {
 		Instruction *I = dyn_cast<Instruction>(*begin);
 		
-		if (I == NULL) {
+		// ATTENTION: GEP uses are not taken into account
+		if (I == NULL || isa<GetElementPtrInst>(I)) {
 			continue;
 		}
 		

### 2012-02-26 doc/Manuscript/paper.tex
Edited a bit of the text in the paper.

@@
 Contrary to many previous techniques, we handle comparisons between variables
-without resorting to expensive algorithms.
+without resorting to relational lattices or expensive algorithms.
 We achieve path sensitiveness by using Bodik's Extended Static Single
@@
 gain in time, but also in precision.
-We have implemented our technique into LLVM, an industrial strenght compiler,
+We have implemented our technique in LLVM, an industrial strenght compiler,
 and have been able to process over 4 million assembly instructions in a
@@
 The analysis of integer variables on the interval lattice has been the
-canonical example of abstract interpretation since its introduction by
-Cousot and Cousot~\cite{Cousot77}.
+canonical example of abstract interpretation since its introduction in
+Cousot and Cousot's seminal paper~\cite{Cousot77}.
 Compilers use range analysis to infer the possible values that discrete
-variables may assume during the execution of programs.
+variables may assume during program execution.
 This analysis has many uses.
@@
 algorithm itself is not the main focus of their works, they neither give
-details about their design choices nor provide experimental data about.
+details about their design choices nor provide experimental data about it.
 This scenario was recently changed by Oh {\em et al.}~\cite{Oh12}, who
-introduced an abstract interpretation framework which handles programs with
+introduced an abstract interpretation framework which processes programs with
 hundreds of thousands of lines of code.
@@
 gcc~\footnote{Gcc's VRP pass implements a variant of Patterson's
-algorithm~\cite{Patterson95}} only deal with comparisons between variables and
+algorithm~\cite{Patterson95}.} only deal with comparisons between variables and
 constants.
@@
 Nevertheless, the inability to handle comparisons between variables, and the
-cubic complexity of the Bellman-Ford method eventually leads us to seek
+cubic complexity of the Bellman-Ford method eventually lead us to seek
 alternative solutions to range analysis.
@@
 We have implemented our algorithm in the LLVM compiler~\cite{Lattner04}, and
-have used it to analyze a test suite with 1.4 million lines of code.
+have used it to analyze a test suite with 1.4 million lines of C code.
 In Section~\ref{sec:exp} we provide empirical evidence showing that our
-implementation is fast: it analyzes the gcc source code in XX seconds, for
+implementation is fast: it analyzes the gcc source code in 4 seconds, for
 instance.
@@
 $\ub{\iota} = u$.
-We let \varset{V} be a set of variables, and
+We let \varset{V} be a set of constraint variables, and
 $I: \varset{V} \mapsto \varset{Z}^2$ a
-mapping from variables to intervals in $\varset{Z}^2$.
+mapping from these variables to intervals in $\varset{Z}^2$.
 Our objective is to solve a constraint system $C$, formed by instances of the
@@
 to illustrate our range analysis.
-Figure~\ref{fig:ex1}(b) shows the same program in SSA form, and
-Figure~\ref{fig:ex1}(c) outlines the constraints that we extract from this
+Figure~\ref{fig:ex1}(b) shows the same program in SSA form~\cite{Cytron91},
+and Figure~\ref{fig:ex1}(c) outlines the constraints that we extract from this
 program.
 There is a clear correspondence between instructions and constraints.
-A possible solution to the range analysis problem is given in
+A possible solution to the range analysis problem, as obtained via the
+techniques that we will introduce in Section~\ref{sec:algo}, is given in
 Figure~\ref{fig:ex1}(d).
-This is a very conservative solution.
+The SSA form, so conspicuous in modern compilers, leads to a very conservative
+solution.
 As we will see shortly, we can improve this solution substantially by using
-a more sophisticated program representation.
+a more sophisticated program representation -- the e-SSA form.
 
@@
 
-\section{Range Analysis in Three Steps}
+\section{Our Design of a Range Analysis Algorithm}
 \label{sec:algo}
@@
 transformed program to convert it to SSA form.
-This last step prevents that two different names given to the same original
+This last step avoids that two different names given to the same original
 variable be simultaneously alive at the program code.
-Figure~\ref{fig:ex_standard_eSSA} shows our running example changed into
+Figure~\ref{fig:ex_standard_eSSA}(a) shows our running example changed into
 standard e-SSA form.
+The part (b) of this figure shows the solution that we get to this new
+program.
 
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{images/ex_standard_eSSA}
+\includegraphics[width=0.9\textwidth]{images/ex_standard_eSSA}
 \end{center}
 \caption{\label{fig:ex_standard_eSSA}
-The control flow graph from Figure~\ref{fig:ex1}(b) converted to standard
-e-SSA form.}
+(a) The control flow graph from Figure~\ref{fig:ex1}(b) converted to standard
+e-SSA form.
+(b) A solution to the range analysis problem
+}
 \end{figure}
@@
 Examples of such algorithms include the family of
-bitwidth aware register allocators~\cite{Barik06,Tallam03,Pereira08} and
+bitwidth aware register allocators~\cite{Barik06,Pereira08,Tallam03} and
 hardware synthesizers~\cite{Cong05,Mahlke01,Stephenson00}.
@@
 
-The constraints shown in Figure~\ref{fig:eval_function} let us handle
-many different assembly instructions.
-We use $\phi$ assignments, such as $Y = \phi(X_1, X_2)$ to represent
+The constraints in Figure~\ref{fig:eval_function} handles
+many different programming instructions.
+The $\phi$ assignments, such as $Y = \phi(X_1, X_2)$, represent
 the $\phi$-functions typically found in SSA form programs.
-As we have discussed in Section~\ref{sub:rep}, we use intersection
+As discussed in Section~\ref{sub:rep}, we use intersection
 constraints to bound the variables created due to live range splitting
@@
 Many assembly operations, such as modulus or division, do not afford us
-this property.
-However, several of these non-monotonic instructions have a conservative
-translation to our constraint system, as we show in
-Figure~\ref{fig:constraints}.
+this monotonicity.
+However, these non-monotonic instructions have a conservative
+translation to our constraint system.
+Figure~\ref{fig:constraints} shows some examples of constraints that we
+derive from typical assembly instructions.
 
@@
 Figure~\ref{fig:ex_graph} shows the dependence graph that we build for the
-e-SSA form program given in Figure~\ref{fig:ex_standard_eSSA}.
+e-SSA form program given in Figure~\ref{fig:ex_standard_eSSA}(a).
 If $V$ is used by $C$ as the input of a future, then the edge from
-$N_v$ to $N_c$ represents a {\em control
+$N_v$ to $N_c$ represents what Ferrante {\em et al.} call a {\em control
 dependence}~\cite[p.323]{Ferrante87}.
+We use dashed lines to represent these edges.
 All the other edges denote {\em data dependences}~\cite[p.322]{Ferrante87}.
@@
 
-\subsection{Finding Ranges: the Macro Algorithm}
+\subsection{Solving Range Analysis on Strong Components}
 \label{sub:macro}
@@
 
+When finding strongly connected components, we take control dependence
+edges into consideration.
+For instance, in Figure~\ref{fig:ex_graph} the nodes that correspond to the
+variables $i_1$, $i_2$, $i_t$, $j_1$, $j_2$ and $j_t$ form a single component.
+The dashed edges, which represent control dependences, keep all these variables
+connected.
+In this way, we ensure that, upon stumbling upon an interval associated with
+future bounds, e.g., $\fun{ft}(v)$, either variable $v$ has been solved in a
+previous component, or it belongs to the current component.
+In the latter case, as we will see soon, we can still take $v$'s interval into
+consideration.
+
 \subsection{Finding Ranges: the Micro Algorithm}
@@
 \end{enumerate}
-However, before we start, we remove the control dependence edges from the
+However, before we start, we remove control dependence edges from the
 strongly connected component, as they have no semantics to our transfer
@@
 (iv) grows in both directions.
-The lattice of this dataflow analysis, plus its meet operator is given in
-Figure~\ref{fig:growth_analysis}.
+The lattice of abstract states, plus a constraint system that represents our
+widening analysis is given in Figure~\ref{fig:growth_analysis}.
 Because the lattice has height three, the intervals bound to each variable can
@@
 Figure~\ref{fig:ex_graph}.
-Notice that upon meeting this SCC, we have already found the interval
-$[0, 0]$ to $I_0$ and the interval $[100, 100]$ to $J_0$, as we show in
-Figure~\ref{fig:ex_graph}(a).
+Notice that upon meeting this SCC, we have already determined that the interval
+$[0, 0]$ is bound to $I_0$ and that the interval $[100, 100]$ is bound to
+$J_0$, as we show in Figure~\ref{fig:ex_graph}(a).
 We are not guaranteed to find the least fix point of a constraint system.
@@
 Had we applied the widening operator onto the whole graph, then we would
-have found out that variable $J_0$ is bound to $[-\infty, +\infty]$,
+have found out that variable $j_0$ is bound to $[-\infty, +\infty]$,
 because
-(i) it receives its interval directly from variable $K_t$, which is upper
+(i) it receives its interval directly from variable $k_t$, which is upper
 bounded by $+\infty$, and
 (ii) it is part of a negative cycle.
-On the other hand, because we only analyze the $J$'s SCC after we have
-analyzed $K$'s, $K$ only contribute the constant range $[0, 99]$ to $J_0$.
+On the other hand, because we only analyze $j$'s SCC after we have
+analyzed $k$'s, $k$ only contribute the constant range $[0, 99]$ to $j_0$.
+
+Figure~\ref{fig:ex_standard_eSSA}(b) shows the final solution that we find
+for this example.
+This solution is very precise, in the sense that it is the maximum fixed
+point of the constraint system given in Figure~\ref{fig:eval_function}.
+However, the solution is still an over approximation of the dynamic behavior of
+the program in Figure~\ref{fig:ex_standard_eSSA}.
+For instance, we have found that variable $i$ could reach the upper value of
+99.
+In any actual run of the program, $i$ could be at most 50.
+There exist static methods that are able to infer this tighter bound, such as
+Lakhdar-Chaouch's analysis on the relational domain~\cite{Lakhdar11}.
+However, analyses like this are much more expensive than ours, and seem, so
+far, unable to scale up to very large code bodies.
 
@@
 \begin{center}
-\includegraphics[width=0.8\textwidth]{images/ex_partition_grow_crop}
+\includegraphics[width=\textwidth]{images/ex_partition_grow_crop}
 \end{center}
@@
 
+%In our experiments we have observed that the vast majority of strongly
+%connected components that constitute typical dependence graphs contain only
+%one element.
+%Furthermore, composite components tend to be small.
+%Therefore, range analysis admit very fast solution.
+
 \bibliographystyle{plain}

### 2012-03-09 doc/Manuscript/paper.tex
Added section about experiments (layout only)
Added conclusion to the paper.

@@
 gain in time, but also in precision.
-We have implemented our technique in LLVM, an industrial strenght compiler,
+We have implemented our technique in LLVM, an industrial strength compiler,
 and have been able to process over 4 million assembly instructions in a
@@
 
+% Absolute Time and complexity
+% =============================
+
+%Chart 1: shows complexity of the algorithm, plotting program size and time
+% to run the analysis.
+
+%Chart 2: time distribution for SPEC only
+
+% Intra/Inter/Inline comparison
+% =============================
+
+% Chart 3: Time comparison between intra, inter and inlining:
+% Each bar is inter/intra, or inline/intra
+% We want to show that the whole program analysis is fast.
+
+% (I) Chart 4: Precision comparison between intra/inter/inline
+% *not for spec*: take five largest of bitwise, stanford and SPEC CINT.
+% Add the three geomeans to the chart.
+
+% (I) Chart 5: Same as (4), but this time precision in terms of [c, c], [-, c],
+% [+, c], [-, +]. Perhaps take off [-, +]
+
+% Impact of e-SSA form:
+% =====================
+
+% Chart 6: time overhead e-SSA. So, for each SPEC 2006, do time e-ssa/time ssa.
+
+% (I) Chart 7: precision with e-SSA form.
+% Percentage reduction (bitwidth only) for five largest of bitwise, stanford
+% and SPEC CINT.
+
+% Size overhead of different representations
+% ==========================================
+
+% Chart 8: Size overhead e-SSA/SSA. SPEC 2006 only.
+
+% Chart 9: Size overhead inline/non-inline. SPEC 2006 only.
+
+% Precision with different number of fixed iterations
+% ===================================================
+
+% Chart 10: Run inter + e-SSA with N*|SCC| in fixed update for five largest of
+% bitwise, stanford and SPEC CINT.
+
+% Table 1: Structural Properties for SPEC CPU 2006.
+% Remember to add %SCC's with just one component.
+
+% Impact of strongly connected components
+% =======================================
+
 %In our experiments we have observed that the vast majority of strongly
@@
 
+\section{Final Considerations}
+\label{sec:con}
+
+In this presentation we chose to omit correctness proofs for our algorithms.
+For a proof that the widening and the narrowing operators give origin to
+approximating sequences, we recommend Cousot and Cousot's work~\cite{Cousot}.
+For a proof that the e-SSA form transformation is semantics preserving, we
+invite the reader to check a recent report of ours~\cite{XXXX}.
+In this work we also show that the results that we obtain via the
+sparse analysis are equivalent to the results provided by a dense analysis.
+In other words, if the dense analysis tells us that variable $v$ is
+associated with the interval range $[l, u]$ at program point $i$, and $v'$
+is the new name of $v$, alive at $i$ in the e-SSA form program, then $v'$ is
+associated with the interval $[l, u]$ along its entire live range.
+
+Our implementation of the range analysis algorithm described in this paper is
+publicly available at \texttt{http://google.code/range-analysis/}.
+This repository contains instructions about how to deploy and use our
+implementation.
+We provide a gallery of examples, including source codes,
+CFGs and constraint graphs that we produce for meaningful programs at
+\texttt{http://google.code/range-analysis/wiki/gallery}.
+
 \bibliographystyle{plain}

### 2012-03-09 src/RangeAnalysis/RangeAnalysis.cpp
- Refactored the range test to its own file - Fixed the implementation of all binary operators.

@@ -438,11 +438,11 @@ static RegisterPass<InterProceduralRA<CropDFS> > X("ra-inter-crop",
 // ========================================================================== //
 // Range
 // ========================================================================== //
-Range::Range() :
-		l(Min), u(Max), type(Regular) {
+Range::Range() {
+	Range(Min, Max);
 }
 
-Range::Range(APInt lb, APInt ub, RangeType rType = Regular) :
+Range::Range(APInt lb, APInt ub, RangeType rType) :
 		l(lb), u(ub), type(rType) {
 }
 
@@ -539,290 +539,435 @@ Range Range::mul(const Range& other) {
 	return Range(*min, *max);
 }
 
+#define DIV_HELPER(OP,x,y) 	x.eq(Max) ? \
+						(y.slt(Zero) ? Min : (y.eq(Zero) ? Zero : Max)) \
+						: (y.eq(Max) ? \
+							(x.slt(Zero) ? Min :(x.eq(Zero) ? Zero : Max)) \
+							:(x.eq(Min) ? \
+								(y.slt(Zero) ? Max : (y.eq(Zero) ? Zero : Min)) \
+								: (y.eq(Min) ? \
+									(x.slt(Zero) ? Max : (x.eq(Zero) ? Zero : Min)) \
+									:(x.OP(y)))))
+
+// [a, b] / [c, d] = [Min(a/c, a/d, b/c, b/d), Max(a/c, a/d, b/c, b/d)]
 Range Range::udiv(const Range& other) {
-	APInt ll = Min, lu = Min, ul = Max, uu = Max;
-	if (this->getLower().ne(Min) && other.getLower().ne(Min)) {
-		if (other.getLower().ne(APInt::getNullValue(MAX_BIT_INT))) {
-			ll = this->getLower().udiv(other.getLower()); // lower lower
-		}
+	const APInt &a = this->getLower();
+	const APInt &b = this->getUpper();
+	const APInt &c = other.getLower();
+	const APInt &d = other.getUpper();
+
+	// Deal with division by 0 exception
+	if ((c.eq(Zero) || d.eq(Zero))) {
+		return Range(Min, Max);
 	}
 
-	if (this->getLower().ne(Min) && other.getUpper().ne(Max)) {
-		if (other.getUpper().ne(APInt::getNullValue(MAX_BIT_INT))) {
-			lu = this->getLower().udiv(other.getUpper()); // lower upper
-		}
[…]

### 2012-03-09 src/RangeAnalysis/RangeAnalysis.h

@@ -99,8 +99,6 @@ errs() << "\n--------------\n"; \
 
 using namespace llvm;
 
-namespace {
-
 /// In our range analysis pass we have to perform operations on ranges all the
 /// time. LLVM has a class to perform operations on ranges: the class
 /// Range. However, the class Range doesn't serve very well
@@ -125,7 +123,7 @@ private:
 
 public:
 	Range();
-	Range(APInt lb, APInt ub, RangeType type);
+	Range(APInt lb, APInt ub, RangeType type = Regular);
 	~Range();
 	APInt getLower() const {return l;}
 	APInt getUpper() const {return u;}
@@ -871,18 +869,6 @@ public:
 	bool runOnFunction(Function &F);
 	virtual Range getRange(const Value *v);
 }; // end of class RangeAnalysis
-
-class RangeUnitTest: public ModulePass{
-	unsigned total;
-	unsigned failed;
-	void printStats();
-public:
-	static char ID; // Pass identification, replacement for typeid
-	RangeUnitTest() : ModulePass(ID), total(0), failed(0) {}
-	bool runOnModule(Module & M);
-};
-
-} // end namespace
 
 #endif /* LLVM_TRANSFORMS_RANGEANALYSIS_RANGEANALYSIS_H_ */
 

### 2012-03-12 src/RangeAnalysis/RangeAnalysis.cpp
Fixed the RangeAnalysisTest refatoring.

@@ -3203,89 +3203,3 @@ bool Nuutila::checkTopologicalSort(UseMap *useMap) {
 }
 #endif
 
-#define ASSERT_TRUE(print_op,op,op1,op2,res) total++; if(op1.op(op2) != res){ \
-			failed++; \
-			errs() << "\t[" << total << "] " << print_op << ": "; \
-			op1.print(errs()); \
-			errs() << " "; \
-			op2.print(errs()); \
-			errs() << " RESULT: "; \
-			(op1.op(op2)).print(errs()); \
-			errs() << " EXPECTED: "; \
-			res.print(errs()); \
-			errs() << "\n";}
-
-void RangeUnitTest::printStats() {
-	errs() << "\n//********************** STATS *******************************//\n";
-	errs() << "\tFailed: " << failed << " (" << failed / total;
-	if (failed > 0)
-		errs() << "." << 100 / (total / failed);
-	errs() << "%)\n";
-	errs() << "\tTotal: " << total << "\n";
-	errs()
-			<< "//************************************************************//\n";
-}
-
-bool RangeUnitTest::runOnModule(Module & M) {
-	MAX_BIT_INT = InterProceduralRA<Cousot>::getMaxBitWidth(M);
-	RangeAnalysis::updateMinMax(MAX_BIT_INT);
-	errs() << "Running unit tests for Range class!\n";
-	// --------------------------- Shared Objects -------------------------//
-	Range unknown(Min, Max, Unknown);
-	Range empty(Min, Max, Empty);
-	Range zero(Zero, Zero);
-	Range infy(Min, Max);
-	Range pos(Zero, Max);
-	Range neg(Min, Zero);
-	// -------------------------------- ADD --------------------------------//
-	// [a, b] - [c, d] = [a + c, b + d]
-	ASSERT_TRUE("ADD", add, infy, infy, infy);
-	ASSERT_TRUE("ADD", add, zero, infy, infy);
-	ASSERT_TRUE("ADD", add, zero, zero, zero);
-	ASSERT_TRUE("ADD", add, neg, zero, neg);
-	ASSERT_TRUE("ADD", add, neg, infy, infy);
[…]

### 2012-03-12 src/RangeAnalysis/RangeAnalysis.h

@@ -99,8 +99,6 @@ errs() << "\n--------------\n"; \
 
 using namespace llvm;
 
-namespace {
-
 /// In our range analysis pass we have to perform operations on ranges all the
 /// time. LLVM has a class to perform operations on ranges: the class
 /// Range. However, the class Range doesn't serve very well
@@ -871,18 +869,6 @@ public:
 	bool runOnFunction(Function &F);
 	virtual Range getRange(const Value *v);
 }; // end of class RangeAnalysis
-
-class RangeUnitTest: public ModulePass{
-	unsigned total;
-	unsigned failed;
-	void printStats();
-public:
-	static char ID; // Pass identification, replacement for typeid
-	RangeUnitTest() : ModulePass(ID), total(0), failed(0) {}
-	bool runOnModule(Module & M);
-};
-
-} // end namespace
 
 #endif /* LLVM_TRANSFORMS_RANGEANALYSIS_RANGEANALYSIS_H_ */
 

### 2012-03-12 src/RangeAnalysisTest/RangeAnalysisTest.cpp

@@ -1,0 +1,282 @@
+#include "RangeAnalysisTest.h"
+
+unsigned MAX_BIT_INT = 1;
+APInt Min = APInt::getSignedMinValue(MAX_BIT_INT);
+APInt Max = APInt::getSignedMaxValue(MAX_BIT_INT);
+APInt Zero(MAX_BIT_INT, 0, true);
+
+void RangeUnitTest::printStats() {
+	errs() << "\n//********************** STATS *******************************//\n";
+	errs() << "\tFailed: " << failed << " (" << failed / total;
+	if (failed > 0)
+		errs() << "." << int(100 * ((double)failed/total));
+	errs() << "%)\n";
+	errs() << "\tTotal: " << total << "\n";
+	errs()
+			<< "//************************************************************//\n";
+}
+
+#define NEW_APINT(n) APInt(MAX_BIT_INT,n,true)
+#define NEW_RANGE(l,u) Range(l, u, Regular)
+#define NEW_RANGE_N(l,u) NEW_RANGE(NEW_APINT(l),NEW_APINT(u))
+#define NEW_RANGE_MAX(l) NEW_RANGE(NEW_APINT(l), Max)
+#define NEW_RANGE_MIN(u) NEW_RANGE(Min, NEW_APINT(u))
+
+bool RangeUnitTest::runOnModule(Module & M) {
+	MAX_BIT_INT = InterProceduralRA<Cousot>::getMaxBitWidth(M);
+	Min = APInt::getSignedMinValue(MAX_BIT_INT);
+	Max = APInt::getSignedMaxValue(MAX_BIT_INT);
+	Zero = APInt(MAX_BIT_INT, 0, true);
+
+	errs() << "Running unit tests for Range class!\n";
+	// --------------------------- Shared Objects -------------------------//
+	Range unknown(Min, Max, Unknown);
+	Range empty(Min, Max, Empty);
+	Range zero(Zero, Zero);
+	Range infy(Min, Max);
+	Range pos(Zero, Max);
+	Range neg(Min, Zero);
+	Range smallNeg(NEW_APINT(-17), Zero);
[…]

### 2012-03-12 src/RangeAnalysisTest/RangeAnalysisTest.h

@@ -1,0 +1,45 @@
+#ifndef LLVM_TRANSFORMS_RANGEANALYSIS_RANGEANALYSISTESTS_H_
+#define LLVM_TRANSFORMS_RANGEANALYSIS_RANGEANALYSISTESTS_H_
+
+#include "llvm/Function.h"
+#include "llvm/Pass.h"
+#include "llvm/Constants.h"
+#include "llvm/Instructions.h"
+#include "llvm/Module.h"
+#include "llvm/ADT/DenseMap.h"
+#include "llvm/ADT/StringMap.h"
+#include "llvm/ADT/SmallPtrSet.h"
+#include "llvm/ADT/Statistic.h"
+#include "llvm/Support/Debug.h"
+#include "llvm/Support/raw_ostream.h"
+#include "llvm/Support/ConstantRange.h"
+#include "llvm/Support/CallSite.h"
+#include "llvm/Support/TimeValue.h"
+#include "llvm/Support/Process.h"
+#include "../RangeAnalysis/RangeAnalysis.h"
+
+#define ASSERT_TRUE(print_op,op,op1,op2,eRes) {total++; \
+			Range realRes = op1.op(op2); \
+			if(realRes != eRes){ \
+			failed++; \
+			errs() << "\t[" << total << "] " << print_op << ": "; \
+			op1.print(errs()); \
+			errs() << " "; \
+			op2.print(errs()); \
+			errs() << " RESULT: "; \
+			realRes.print(errs()); \
+			errs() << " EXPECTED: "; \
+			eRes.print(errs()); \
+			errs() << "\n";}}
+
+class RangeUnitTest: public ModulePass{
+	unsigned total;
+	unsigned failed;
+	void printStats();
+public:
+	static char ID; // Pass identification, replacement for typeid
+	RangeUnitTest() : ModulePass(ID), total(0), failed(0) {}
+	bool runOnModule(Module & M);
+};
+
+#endif /* LLVM_TRANSFORMS_RANGEANALYSIS_RANGEANALYSISTESTS_H_ */

### 2012-03-13 doc/Manuscript/paper.tex
Added some text to the section about experiments.

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+We have implemented our range analysis algorithm in LLVM 3.0, and have run
+experiments on a Intel quad core CPU with a 2.40GHz clock, and 3.6GB of RAM.
+Each core has a 4,096KB L1 cache.
+The operation system is Linux Ubuntu 10.04.4.
+Our implementation of range analysis has 3,958 lines of
+commented C++ code, and our e-SSA conversion module has 672 lines.
+We have tested it on 428 C programs that constitute the LLVM test suite, plus
+the C programs from SPEC CPU 2006.
+These programs together gave us approximately 4.76 million assembly
+instructions to analyze.
+
+\noindent
+\textbf{Structural Properties.}
+Table~\ref{tab:struct} shows structural properties of the programs in
+the SPEC 2006 benchmark suite.
+Most of the strongly connected components that constitute
+the constraint graphs are singletons.
+For instance, 99.11\% of the SSCs found in gcc (\texttt{403.gcc}) have only
+one node.
+Therefore the most complex phases of our algorithm are
+confined to a few parts of the constraint graphs that we have to handle.
+Furthermore, the strongly connected components are not large -- our largest
+component, found in \texttt{403.gcc} has 2,131 nodes, and exists due to a long
+chain of mutually recursive function calls.
+These two observations imply that our runtime is likely to be linear in
+practice, as we show in Figure~\ref{fig:TimeCorr}.
+
+\begin{table}[htdp]
+\begin{center}
+\begin{scriptsize}
+\renewcommand{\tabcolsep}{0.1cm}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|} \hline
+Name	    & 464 &	473 & 483 & 458 & 429 & 471 & 403 & 445 & 462 & 401 & 456 \\ \hline
+\#I  & 139626 & 8900 & 593765 & 26947 & 2738 & 94036 & 887191 & 141330 &  6237 & 17663 & 61736 \\ \hline
+\#$N_v$ & 97494 & 5490 & 352423 & 14407 & 1853	 & 50095 & 449442 & 84846 & 3293 & 12517 & 38409 \\ \hline
+\#S  & 91731 & 5108 & 346451 & 13309 & 1751 & 48784 & 435372 & 79909 & 3003 & 11539 & 35619 \\ \hline
+L    & 38 & 49 & 120 & 116 & 16 & 258 & 2131 & 311 & 9 & 18 & 41  \\ \hline
+\#U  & 89461 & 4983 & 344132 & 12919 & 1726 & 48400 & 431703 & 78407 & 2885 & 11267 & 34631 \\ \hline
+\end{tabular}
+\end{scriptsize}
+\end{center}
+\caption{\label{tab:struct}Structural Properties of the programs in the SPEC CPU
+2006 benchmark suite.
+\#I: number of assembly instructions in the original program.
+\#$N_v$: number of variable nodes in the constraint graph.
+% TODO: add also number of constraint nodes.
+\#S: number of strongly connected components (SCC).
+L: size of the largest SCC.
+\#U: number of SCCs with only one node.}
+\end{table}%
+
 % Absolute Time and complexity
 % =============================
+
+\noindent
+\textbf{Runtime Behavior.}
+Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
+run our algorithm and the size of the input programs.
+We show data for the 100 largest benchmarks in our test suite, in number
+of variable nodes in the constraint graph.
+We perform function inlining before running our analysis, to increase program
+sizes.
+Each point in the X line corresponds to a benchmark.
+We analyze the smallest benchmark in this set, \texttt{Prolangs-C}, which has
+1,131 variable nodes in the constraint graph, in 20ms.
+We take 15,91secs to analyze our largest benchmark, \texttt{403.gcc}, which,
+after function inlining, has 1,266,273 assembly instructions, and a
+constraint graph with 679,652 variable nodes.
+For this data set, the coefficient of determination $(R^2)$ is 0.967, which
+provides very strong evidence about the linear asymptotic complexity of our
+implementation.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/TimeCorr}
+%\includegraphics[width=0.4\textwidth]{images/Complexity}
+\end{center}
+\caption{\label{fig:TimeCorr}
+Correlation between program size (measured in number of nodes in constraint
+graphs after inlining) and analysis runtime (ms).
+Coefficient of determination = 0.967.
+}
+\end{figure}
 
@@
 For a proof that the widening and the narrowing operators give origin to
-approximating sequences, we recommend Cousot and Cousot's work~\cite{Cousot}.
+approximating sequences, we recommend Cousot and Cousot's work~\cite{Cousot77}.
 For a proof that the e-SSA form transformation is semantics preserving, we

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/adpcm.c
Created an script that include Bitwise in LLVM Test Suite (bitwise-benchmarks/configure.sh)

@@ -1,0 +1,195 @@
+/* Changes from original: 9/15/99 JWB
+ * removed struct
+ * changed all types to int
+ * replaced I/O with dummy data setup
+ * added result checksum
+ * reversed direction of for loop
+ * removed decoder
+ * replaced some control flow with logic
+ * inlined coder into test code
+ * got rid of global variables
+ * 
+ * optimized for compiling with -fmacro to muxify if's
+ *
+ * This version has broken input and initialization data but should give
+ * approximately correct hardware cycle counts and area for adpcm
+ *
+ * stepSizeTable needs to be initialized correctly for true result 
+ */
+
+/*
+** Intel/DVI ADPCM coder/decoder.
+**
+** The algorithm for this coder was taken from the IMA Compatability Project
+** proceedings, Vol 2, Number 2; May 1992.
+**
+** Version 1.0, 7-Jul-92.
+*/
+
+#define NSAMPLES 2407 
+#define MAXITER 2
+#define ANS 123608
+
+#define SIZE NSAMPLES
+#define WIDTH 8
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.h"
+#endif
+
+
+int stepsizeTable[89] = {
+  7, 8, 9, 10, 11, 12, 13, 14, 16, 17,
+  19, 21, 23, 25, 28, 31, 34, 37, 41, 45,
+  50, 55, 60, 66, 73, 80, 88, 97, 107, 118,
+  130, 143, 157, 173, 190, 209, 230, 253, 279, 307,
+  337, 371, 408, 449, 494, 544, 598, 658, 724, 796,
+  876, 963, 1060, 1166, 1282, 1411, 1552, 1707, 1878, 2066,
+  2272, 2499, 2749, 3024, 3327, 3660, 4026, 4428, 4871, 5358,
+  5894, 6484, 7132, 7845, 8630, 9493, 10442, 11487, 12635, 13899,
+  15289, 16818, 18500, 20350, 22385, 24623, 27086, 29794, 32767
+};
+
+
+main() {
+
+  int indexTable[16];
+  
+  int iter, checksum, i, j;
+  
+  int	abuf[(NSAMPLES/2)+1];
+  int	sbuf[NSAMPLES];
+  
+  int val;			/* Current input sample value */
+  int sign;			/* Current adpcm sign bit */
+  int delta;			/* Current adpcm output value */
[…]

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/bilint.c

@@ -1,0 +1,113 @@
+/****************************************************************************
+ *
+ *  This benchmark was taken from Intel's MMX website.
+ *
+ ***************************************************************************/
+
+#define RED   0
+#define GREEN 1
+#define BLUE  2
+
+#define BYTE       unsigned char
+#define RGBQWORD   unsigned
+#define WORD       unsigned
+#define DWORD      unsigned
+#define LPRGBQWORD unsigned
+
+void main ()
+{
+  BYTE       TextureMap[4][128];
+  RGBQWORD   ColorLookupTable[4][4];
+  DWORD      dwUVal;
+  DWORD      dwVVal;
+  LPRGBQWORD lpRGBOut[4];
+
+  DWORD dwUDelta;
+  DWORD dwVDelta;
+  DWORD dwUInt;
+  DWORD dwVInt;  
+
+  RGBQWORD *RGB1;
+  RGBQWORD *RGB2;
+  RGBQWORD *RGB3;
+  RGBQWORD *RGB4;
+
+  LPRGBQWORD lpRGBOutR1;
+  LPRGBQWORD lpRGBOutR2;
+  LPRGBQWORD lpRGBOutR3;
+  LPRGBQWORD lpRGBOutR4;
+  LPRGBQWORD lpRGBOutG1;
+  LPRGBQWORD lpRGBOutG2;
+  LPRGBQWORD lpRGBOutG3;
+  LPRGBQWORD lpRGBOutG4;
+  LPRGBQWORD lpRGBOutB1;
+  LPRGBQWORD lpRGBOutB2;
+  LPRGBQWORD lpRGBOutB3;
+  LPRGBQWORD lpRGBOutB4;
+
+  WORD w1, w2, w3, w4;
+  DWORD dwTemp1, dwTemp2, dwTemp3, dwTemp4;
+  unsigned tmp;
+
+  /* provide some fake initialization. */
+  dwUVal = rand ();
+  dwVVal = rand ();
+
+  tmp = (dwUVal & 0x003fffffL);
+  dwUDelta = tmp >> 6;
+  dwVDelta = (dwVVal & 0x003fffffL) >> 6;
[…]

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/bubblesort.c

@@ -1,0 +1,68 @@
+/****************************************************************************
+ *
+ *  The bubblesort benchmark.
+ *
+ ***************************************************************************/
+
+#define SIZE  512
+#define WIDTH 16
+#define ANS   10880
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.h"
+#endif
+
+int main()
+{
+  int i,j,top,checksum,data,sortlist_odd[SIZE/2],sortlist_even[SIZE/2];
+  int x, s1, s2, ie, io;
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.begin"
+#endif
+
+  for (i=0; i<SIZE/2; i++) {
+    sortlist_even[i] = (SIZE-(i << 1)       ) | (1 << (WIDTH-1));
+    sortlist_odd[i]  = (SIZE-((i << 1) | 1) ) | (1 << (WIDTH-1));
+  }
+    
+  for(top=SIZE-1;top>0;top--) {
+    for(i=0;i<top;i++) {
+      io = i >> 1;
+      ie = io + (i & 1);
+      s1=sortlist_even[ie];
+      s2=sortlist_odd[io];
+      if(s1 > s2 ^ (i & 1)) {
+	sortlist_even[ie] = s2;
+	sortlist_odd[io]  = s1;
+      }
+    }
+  }
+  
+  checksum = 0;
+  for (i=0; i<SIZE/2; i++)
+    checksum = (checksum + sortlist_even[i]*i) & 0xffff;
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.end"
+#endif
+
+#ifdef PRINTOK
+  printf("RESULT: Checksum %d\n", checksum);
+#endif
+  
+  if(checksum==ANS) {
+#ifdef PRINTOK
+    printf("RESULT: Success!\n");
+    printf("RESULT: Return-value %d (%x) \n", 0xbadbeef, 0xbadbeef);
+#endif
+    return 0xbadbeef;
+  }
+  else {
+#ifdef PRINTOK
+    printf("RESULT: Failure!\n");
+    printf("RESULT: Return-value %d (%x) \n", 0xbad, 0x0bad);
+#endif
+    return 0xbad;
+  }
+}

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/convolve.c

@@ -1,0 +1,80 @@
+/****************************************************************************
+ *
+ *  A convolution algorithm.
+ *
+ ***************************************************************************/
+
+#define NTAPS   16
+#define SIZE   256
+#define WIDTH   16
+#define ANS  16128
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.h"
+#endif
+
+int main()
+{
+  int A[SIZE];
+  int B[SIZE];
+  int taps[NTAPS];
+  int checksum;
+  int i;
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.begin"
+#endif
+
+  for(i = 0; i < NTAPS;i++)
+    taps[i] = (1 << WIDTH) - 1;
+
+  for(i = 0; i < SIZE; i++)
+    A[i] = (1 << WIDTH) - 1;
+
+  for(i = 0; i < (SIZE - NTAPS); i++) {
+    B[i] =
+      A[i] * taps[0] +
+      A[i+1] * taps[1] +
+      A[i+2] * taps[2] +
+      A[i+3] * taps[3] +
+      A[i+4] * taps[4] +
+      A[i+5] * taps[5] +
+      A[i+6] * taps[6] +
+      A[i+7] * taps[7] +
+      A[i+8] * taps[8] +
+      A[i+9] * taps[9] +
+      A[i+10] * taps[10] +
+      A[i+11] * taps[11] +
+      A[i+12] * taps[12] +
+      A[i+13] * taps[13] +
+      A[i+14] * taps[14] +
+      A[i+15] * taps[15];
+  }
+
+  checksum = 0;
+  for (i=0;i<SIZE;i++)
+    checksum += A[i] & 0x3f;
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.end"
+#endif
+
+#ifdef PRINTOK
+  printf("RESULT: Checksum %d\n", checksum);
+#endif
+  
+  if(checksum==ANS) {
+#ifdef PRINTOK
+    printf("RESULT: Success!\n");
+    printf("RESULT: Return-value %d (%x) \n", 0xbadbeef, 0xbadbeef);
+#endif
+    return 0xbadbeef;
+  }
+  else {
+#ifdef PRINTOK
+    printf("RESULT: Failure!\n");
+    printf("RESULT: Return-value %d (%x) \n", 0xbad, 0x0bad);
+#endif
+    return 0xbad;
+  }
+}

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/edge_detect.c

@@ -1,0 +1,181 @@
+/****************************************************************************
+ *
+ *  This benchmark was taken from Intel's MMX website.
+ *
+ ***************************************************************************/
+
+/* ISIZE was 128, reduced to 4 to speed simulation time */
+#define ISIZE      4
+#define JSIZE    128
+
+#define SIZE    ISIZE * JSIZE
+#define WIDTH   8
+#define ANS 33150
+/* note: WIDTH shouldn't be changed */
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.h"
+#endif
+
+int main() {
+
+  int temp1, temp2, temp3, temp4;
+  int i;
+  int j;
+  int checksum, iter;
+
+  int r;
+  int c;
+  int dead_rows;
+  int dead_cols;
+  int sum;
+  int normal_factor;
+  
+  int image_buffer1[ISIZE][JSIZE];
+  int image_buffer2[ISIZE][JSIZE];
+  int image_buffer3[ISIZE][JSIZE];
+  int filter[3][3];
+  
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.begin"
+#endif
+
+  for (i = 0; i < ISIZE; i++)
+    for (j = 0; j < JSIZE; j++) {
+      image_buffer1[i][j] = (i<ISIZE/2) ? 255 : 0; /* should create an edge */
+      image_buffer2[i][j] = 0;
+      image_buffer3[i][j] = 0;
+    }
+
+  /* performs three convolutions without procedure calls */
+  for(iter=0;iter<3; iter++) {
+
+  /* if we got rid of this, then macros would work */
+    if(iter==0) {
+      filter[0][0] = 1;
+      filter[0][1] = 2;
+      filter[0][2] = 1;
+      filter[1][0] = 2;
+      filter[1][1] = 4;
+      filter[1][2] = 2;
+      filter[2][0] = 1;
+      filter[2][1] = 2;
+      filter[2][2] = 1;
+    }
+    if(iter==1) {
+      filter[0][0] = 1;
+      filter[0][1] = 0;
+      filter[0][2] = -1;
+      filter[1][0] = 2;
+      filter[1][1] = 0;
+      filter[1][2] = -2;
[…]

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/histogram.c

@@ -1,0 +1,109 @@
+/* 9/15/99 JWB
+ * no floating point, fixed point instead
+ *
+ */
+
+/* This program enhances a 256-gray-level, 128x128 pixel image by applying
+   global histogram equalization.
+
+   This program is based on the routines and algorithms found in the book
+   "C Language Algorithms for Disgital Signal Processing" by P. M. Embree
+   and B. Kimble.
+
+   Copyright (c) 1992 -- Mazen A.R. Saghir -- University of Toronto */
+/* Modified to use arrays - SMP */
+
+/*#include "traps.h" */
+
+#define         N       128
+#define         L       256
+#define SIZE N*N
+#define WIDTH  8
+#define ANS 128
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.h"
+#endif
+
+main()
+{
+  int checksum;
+  int cdf,b2;
+  int pixels;
+  int i,j,ii,b3,foo;
+
+  int image[N][N];
+  int histogram[L];
+  int gray_level_mapping[L];
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.begin"
+#endif
+
+
+  /* Get the original image */
+
+  /*  input_dsp(image, N*N, 1);*/
+  for (i = 0; i < N; i++)
+    for (j = 0; j < N; j++)
+      image[i][j]=j;
+
+  /* Initialize the histogram array. */
+
+  for (i = 0; i < L; i++)
+    histogram[i] = 0;
+
+  /* Compute the image's histogram */
+
+
+   for (i = 0; i < N; i++) {
+     for (j = 0; j < N; ++j) {
+       histogram[image[i][j]] += 1;
+     }
+   }
+
+  /* Compute the mapping from the old to the new gray levels */
+   
+   /* fixed point with 20 binary decimal places */
+
+  cdf = 0;
+  pixels = N * N;
+  for (i = 0; i < L; i++) {
+    foo = (histogram[i] << 20);
+    cdf += foo / pixels;
+    gray_level_mapping[i] = 255 * (cdf >> 20) & 0xff;
+  }
+
+  /* Map the old gray levels in the original image to the new gray levels. */
+
+  checksum = 0;
+  for (i = 0; i < N; i++) {
+    for (j = 0; j < N; ++j) {
+        image[i][j] = gray_level_mapping[image[i][j]];
[…]

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/jacobi.c

@@ -1,0 +1,67 @@
+/****************************************************************************
+ *
+ *  Jacobi
+ *
+ ***************************************************************************/
+
+
+#define ITER  32
+#define SIZEX  64
+#define SIZEY  64
+#define WIDTH   8
+
+int main() {
+  int r;
+  int i,j,a2;
+  int a[SIZEX][SIZEY];
+  int b[SIZEX][SIZEY];
+  int checksum;
+/*  int ans = 131072;*/
+  int ans = 557568;
+
+  /* Initialize data array */
+
+  a2=0;
+  for (i=0;i<SIZEX;i++) {
+    a2++;
+    for (j=0;j<SIZEY;j++) {
+
+      /* an abitrary boundary condition of the appropriate WIDTH */	  
+      a[i][j] = 1 << (WIDTH-1);
+      b[i][j] = 1 << (WIDTH-1);
+
+    }
+  }
+  
+  for (r=0;r<ITER;r+=2) {
+    for (i=1;i<SIZEX-1;i++) {
+      for (j=1;j<SIZEY-1;j++)
+	{
+	  a[i][j] = (b[i-1][j] + b[i+1][j] + b[i][j-1] + b[i][j+1])>>2;
+	}
+    }
+    for (i=1;i<SIZEX-1;i++) {
+      for (j=1;j<SIZEY-1;j++)
+	{
+	  b[i][j] = (a[i-1][j] + a[i+1][j] + a[i][j-1] + a[i][j+1])>>2;
+	}
+    }
+  }
+  
+  /* check sum */
+  checksum = 0;
+  for (i=0;i<SIZEX;i++)
+    for (j=0;j<SIZEY;j++)
+      checksum += b[i][j];
+  
+#ifdef PRINTOK
+  printf("Checksum = %d\n", checksum);
+#endif
+  
+  if(checksum==ans)
+    return 195935983; 
+  else
+    return 0;
+
+return checksum; /*195935983*/
+}

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/levdurb.c

@@ -1,0 +1,49 @@
+/****************************************************************************
+ *
+ *  This levdurb filter was taken from Intel's MMX website.
+ *
+ ***************************************************************************/
+
+#include <limits.h>
+
+void main ()
+{
+  long i;
+  long m;
+  long b[11];
+  long Rn;
+  long Rd;
+  long temp;
+  long r[11];
+  long a[11];
+  long k[11];
+  
+
+  a[0] = 8192;
+  for (i = 1; i < 10; i++) {
+    a[i] = rand() & 0xff;
+    r[i] = a[i];
+  }
+
+  for (m = 1; m < 11; m++)
+    {
+      Rn = Rd = 0;
+      
+      for (i = 0; i < m; i++)
+        {
+          Rn = Rn + (r[m-1] * (long)a[i]);
+          Rd = Rd + (r[i] * (long)a[i]);
+        }
+      
+      k[m] = -Rn / ((Rd + 0x4000) >> 15);
+      k[m] = (((long)k[m] * 0x7ff8) + 0x4000) >> 15;
+      
+      b[m] = (k[m] + 0x2) >> 2;
+      
+      for (i = 1; i < m; i++)
+        b[m] = (((long)a[i] << 15) + (k[m] * (long)a[m-i]) + 0x4000) >> 15;
+  
+      for ( i = 1; i < m+1; i++)
+        a[i] = b[i];
+    }
+}

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/life.c

@@ -1,0 +1,147 @@
+/*
+ *
+ * Life benchmark
+ *
+ * Authors: Rajeev Kumar Barua      (barua@lcs.mit.edu)
+ *          Jonathan Babb           (jbabb@lcs.mit.edu)
+ *
+ * Copyright @ 1997 MIT Laboratory for Computer Science, Cambridge, MA 02129
+ */
+
+/* Caveats for Life: Does not take advantage of CPU
+   bit twiddling to get improve software performance */
+
+#define SIZEX 1
+#define SIZEY 30
+#define IMAX  33 /* SIZEX*32+1 */  /* imax=last column (of fake boundary)*/
+#define JMAX  31 /* SIZEY+1    */  /* jmax=last row (of fake boundary)*/
+#define JMAX1 32 /* JMAX+1     */
+#define ITER  20
+#define MAXSIZE (SIZEX*32+2)*(SIZEY+2)
+
+/* for software array, indices into unpacked data, and fake boundary */
+
+#define soft_u_ref(i, j) ((j)+(i)*(SIZEY+2))
+
+/* for software array, indices into packed data, k refers to bit
+position beginning at msb at leftmost = 0, and fake boundary */
+
+#define soft_p_ref(i, j, k) ((j+1)+((i*32)+k+1)*(SIZEY+2))
+
+#define ANS 191
+
+#define SIZE SIZEX*SIZEY*32
+#define WIDTH 1
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.h"
+#endif
+
+int main()
+{
+  int it,i,j,k,data,sum;
+  int seed, checksum;
+  int data1[MAXSIZE],data2[MAXSIZE];
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.begin"
+#endif
+
+  /* Initialize data array */
+  
+  seed = 74755;
+
+  for(i=0;i<SIZEX;i++) {
+    for(j=0;j<SIZEY;j++) {
+
+      seed = (seed * 1309 + 13849);;
+      data = seed;
+	
+      /* Initialize software data after unpacking */
+
+      for (k=0;k<32;k++) {
[…]

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/median.c

@@ -1,0 +1,98 @@
+/****************************************************************************
+ *
+ *  Median filter benchmark.
+ *
+ ***************************************************************************/
+
+#define ANS 1536
+
+#define SIZE 1024
+#define WIDTH 31
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.h"
+#endif
+
+int main(void)
+{
+  int colA[SIZE+2];
+  int colB[SIZE+2];
+  int colC[SIZE+2];
+  int output[SIZE];
+  int newArray[5];
+  int sortMe[9];
+  int i,j,k, temp, checksum;
+  
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.begin"
+#endif
+  
+  /* fool bitwidth analyzer since data is initialized here */
+  colA[0]   =(1<<WIDTH)-1;
+  colB[0]   =(1<<WIDTH)-1;
+  colC[0]   =(1<<WIDTH)-1;
+  sortMe[0] =(1<<WIDTH)-1;
+
+  for (k = 0; k< SIZE+2; k++) {
+    colA[k]=k;
+    colB[k]=k+1;
+    colC[k]=k+2;
+  }
+
+  for (k = 0; k < SIZE; k++)
+  {
+    sortMe[0] = colA[k];
+    sortMe[1] = colA[k+1];
+    sortMe[2] = colA[k+2];
+    sortMe[3] = colB[k];
+    sortMe[4] = colB[k+1];
+    sortMe[5] = colB[k+2];
+    sortMe[6] = colC[k];
+    sortMe[7] = colC[k+1];
+    sortMe[8] = colC[k+2];    
+    
+    for (i = 0; i < 5; i++)
+      newArray[i] = 0;
+    
+    for (i = 0; i < 9; i++)
+    {
+      temp = sortMe[i];
+      for (j = 0; j < 5; j++)
+	{
+	int temp1 = temp;
+	int temp2 = newArray[j];
+	if (temp1 > temp2) {
+	  temp = temp2;
+	  newArray[j] = temp1;
+	}
+      }
+    }
+    output[k] = newArray[4];
+  }
+
+  checksum=0;
+  for (k = 0; k < SIZE; k++)
+    checksum = (checksum + output[k]) & 32767;
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.end"
+#endif
+
+#ifdef PRINTOK
[…]

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/milieu.h

@@ -1,0 +1,48 @@
+
+/*
+===============================================================================
+
+This C header file is part of the SoftFloat IEC/IEEE Floating-point
+Arithmetic Package, Release 2a.
+
+Written by John R. Hauser.  This work was made possible in part by the
+International Computer Science Institute, located at Suite 600, 1947 Center
+Street, Berkeley, California 94704.  Funding was partially provided by the
+National Science Foundation under grant MIP-9311980.  The original version
+of this code was written as part of a project to build a fixed-point vector
+processor in collaboration with the University of California at Berkeley,
+overseen by Profs. Nelson Morgan and John Wawrzynek.  More information
+is available through the Web page `http://HTTP.CS.Berkeley.EDU/~jhauser/
+arithmetic/SoftFloat.html'.
+
+THIS SOFTWARE IS DISTRIBUTED AS IS, FOR FREE.  Although reasonable effort
+has been made to avoid it, THIS SOFTWARE MAY CONTAIN FAULTS THAT WILL AT
+TIMES RESULT IN INCORRECT BEHAVIOR.  USE OF THIS SOFTWARE IS RESTRICTED TO
+PERSONS AND ORGANIZATIONS WHO CAN AND WILL TAKE FULL RESPONSIBILITY FOR ANY
+AND ALL LOSSES, COSTS, OR OTHER PROBLEMS ARISING FROM ITS USE.
+
+Derivative works are acceptable, even for commercial purposes, so long as
+(1) they include prominent notice that the work is derivative, and (2) they
+include prominent notice akin to these four paragraphs for those parts of
+this code that are retained.
+
+===============================================================================
+*/
+
+/*
+-------------------------------------------------------------------------------
+Include common integer types and flags.
+-------------------------------------------------------------------------------
+*/
+#include "SPARC-gcc.h"
+
+/*
+-------------------------------------------------------------------------------
+Symbolic Boolean literals.
+-------------------------------------------------------------------------------
+*/
+enum {
+    FALSE = 0,
+    TRUE  = 1
+};
+

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/motiontest.c

@@ -1,0 +1,43 @@
+/****************************************************************************
+ *
+ *  Motiontest benchmark.
+ *
+ ***************************************************************************/
+
+int  gWidth = 32;
+char gComp[16384];
+char gRef[16384];
+
+void main()
+{
+  int data;
+  int i;
+  int j;
+  char *bptr; /* pointer to start row of 16x16 pixel block being compressed.*/
+  char *cptr; /* pointer to start row of 16x16 pixel reference block.*/
+  int val=0;
+
+  for (i = 0; i < 16384; i++) {
+    gComp[i] = rand() & 0xff;
+    gRef[i] = rand() & 0xff;
+  }
+  bptr = gComp;
+  cptr = gRef;
+
+  for(i=0;i<16;i++)
+    {
+      for(j=0;j<16;j++)
+        {
+          data=(*(bptr++)-*(cptr++));
+          if (data<0) {val-=data;}
+          else    {val+=data;}
+        }
+
+      /* Fast out after this row if we've exceeded best match*/
+      if (val > 0xffff) break;
+      /* update pointer to next row*/
+      bptr += (gWidth - 16);
+      /* update pointer to next row */
+      cptr += (gWidth - 16);
+    }
+}

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/mpegcorr.c

@@ -1,0 +1,130 @@
+/********************************************************************
+ *
+ * Note: This is only a small portion of mpeg's kernel.  It is
+ *       one of the benchmarks used in the RAW benchmark suite.
+ *
+ ********************************************************************/
+
+
+#define ANS 6016
+
+#define SIZE 16 * 16
+#define WIDTH 16
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.h"
+#endif
+
+
+#define DCTSIZE     8 
+
+#define MOTION_TO_FRAME_COORD(bx1, bx2, mx1, mx2, x1, x2) { \
+	x1 = (bx1)*DCTSIZE+(mx1);			    \
+	x2 = (bx2)*DCTSIZE+(mx2);			    \
+    }
+
+#define ABS(x) (((x)<0)?-(x):(x))
+
+int main () {  
+  int currentBlock[16][16];
+  int blockSoFar[16][16];
+  int prev[32][32];
+  int mx = 4;
+  int my = 4;
+  int bx = 4;
+  int by = 4;
+  int diff = 0;
+
+  int localDiff;
+  int y;
+  int x;
+  int yHalf;
+  int xHalf;
+  int fy;
+  int fx;
+  int lumMotionBlock[16][16];
+  int checksum;
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.begin"
+#endif
+
+  for(y=0;y<16;y++) {
+    for(x=0;x<16;x++) {
+      currentBlock[x][y] = x + (1<<(WIDTH-2));
+      blockSoFar[x][y] = x+y + (1<<(WIDTH-2));
+    }
+  }
+  
+  for(y=0;y<32;y++) {
+    for(x=0;x<32;x++) {
+      prev[x][y] = x+y  + (1<<(WIDTH-2));
+    }
+  }
+
+  xHalf = (ABS(mx) % 2 == 1);
+  yHalf = (ABS(my) % 2 == 1);
+
+  MOTION_TO_FRAME_COORD(by, bx, my/2, mx/2, fy, fx);
+  
+  if ( xHalf ) {
+    if ( mx < 0 ) {
+      fx--;
+    }
+
+    if ( yHalf ) {
+      if ( my < 0 ) {
+	fy--;
+      }
+    }
+    else if ( yHalf ) {
+      if ( my < 0 ) {
+	fy--;
+      }
+      
+    }
+  }
+    
+  diff = 0;
+  
[…]

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/newlife.c

@@ -1,0 +1,120 @@
+/*
+ *
+ * Life benchmark
+ *
+ * Author: Jonathan Babb           (jbabb@lcs.mit.edu)
+ *
+ * Written specifically for compilation to hardware.
+ *
+ * Copyright @ 1999 MIT Laboratory for Computer Science, Cambridge, MA 02129
+ */
+
+#define IMAX 32
+#define JMAX 32
+#define ITER 10
+#define ANS 196
+
+#define SIZE IMAX*JMAX
+#define WIDTH 1
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.h"
+#endif
+
+int main()
+{
+  int it,i,j,sum;
+  int seed, checksum;
+  int data1[IMAX][JMAX],data2[IMAX][JMAX];
+
+
+#ifdef TIME_SOFTWARE
+#include "../../../include/time.begin"
+#endif
+
+  /* Fill with zeros */
+
+  for (i=0;i < IMAX; i++) {
+    for (j=0;j < JMAX; j++) {
+      data1[i][j]=0;
+      data2[i][j]=0;
+    }
+  }
+
+  
+  /* Initialize data array */
+  
+  seed = 74755;
+
+  for(i=1;i<IMAX-1;i++) {
+    for(j=1;j<JMAX-1;j++) {
+      seed = (seed * 1309 + 13849);
+      data1[i][j]=seed & 1;
+      data2[i][j]=seed & 1;
+    }
+  }
+
+  it = 0;
+  while(it<ITER) {
+
+    for(i=1;i<IMAX-1;i++) {
+      for(j=1;j<JMAX-1;j++) {
+	sum=
+	  ((data1[i-1][j-1]+
+	  data1[i-1][j])+
+	  (data1[i-1][j+1]+
+	  data1[i][j-1]))+
+	  ((data1[i][j+1]+
+	  data1[i+1][j-1])+
+	  (data1[i+1][j]+
+	  data1[i+1][j+1]));
+	
+	data2[i][j] = (sum==3) | (data1[i][j] & (sum==2));
+      }
+    }
+
+    for(i=1;i<IMAX-1;i++) {
+      for(j=1;j<JMAX-1;j++) {
+	sum=
+	  ((data2[i-1][j-1]+
+	  data2[i-1][j])+
+	  (data2[i-1][j+1]+
+	  data2[i][j-1]))+
+	  ((data2[i][j+1]+
+	  data2[i+1][j-1])+
[…]

### 2012-03-13 tests/bitwise-benchmarks/Bitwise/sha.c

@@ -1,0 +1,607 @@
+/* NIST Secure Hash Algorithm */
+/* heavily modified by Uwe Hollerbach uh@alumni.caltech edu */
+/* from Peter C. Gutmann's implementation as found in */
+/* Applied Cryptography by Bruce Schneier */
+
+/*****************************************************************************
+ *
+ * And further modified by Matt Frank for RAW
+ * mfrank@lcs.mit.edu Wed Jun  3 19:16:01 1998
+ *
+ ****************************************************************************/
+
+/* IF YOU CHANGE NUM_BLOCKS YOU'LL ALSO NEED TO RERUN THE PROGRAM
+ * AND CALCULATE A NEW VALUE FOR RESULT
+ */
+
+#define NUM_BLOCKS 512
+#define RESULT 0x289a69fd
+
+/* SHA f()-functions */
+
+#define f0(x,y,z)	((x & y) | (~x & z))
+#define f1(x,y,z)	(x ^ y ^ z)
+#define f2(x,y,z)	((x & y) | (x & z) | (y & z))
+#define f3(x,y,z)	(x ^ y ^ z)
+
+/* SHA constants */
+
+#define CONST0		0x5a827999L
+#define CONST1		0x6ed9eba1L
+#define CONST2		0x8f1bbcdcL
+#define CONST3		0xca62c1d6L
+
+/* 32-bit rotate */
+
+#define ROT32(x,n)	((x << n) | (x >> (32 - n)))
+
+#define FUNC(n,i)						\
+    temp = ROT32(A,5) + f##n(B,C,D) + E + W##i + CONST##n;	\
+    E = D; D = C; C = ROT32(B,30); B = A; A = temp
+
+int
+main()
+{
+  int b;
+  unsigned long temp;
+  unsigned long Aout, Bout, Cout, Dout, Eout;
+  unsigned long data[NUM_BLOCKS][16];
+
+  Aout = 0x67452301L;
+  Bout = 0xefcdab89L;
+  Cout = 0x98badcfeL;
[…]

### 2012-03-13 tests/bitwise-benchmarks/configure.sh

@@ -1,0 +1,6 @@
+llvmpath=~/workspace/ra/llvm-3.0
+testsuite=${llvmpath}/projects/test-suite/SingleSource/Benchmarks
+bitwise=$testsuite/Bitwise
+rm -rf $bitwise
+cp -R Bitwise $bitwise
+mkdir $bitwise/Output

### 2012-03-13 tests/compile.py

@@ -4,36 +4,58 @@ import sys
 import os
 
 passPath = "~/workspace/ra/llvm-3.0/Debug/lib/"
+analysisTypes = ["-ra-inter-crop","-ra-inter-cousot","-ra-intra-crop","-ra-intra-cousot"]
+testTypes = ["-ra-test-range"]
 
 def checkArgs(args):
-    if len(args) != 3:
+    if len(args) != 3: 
         return False
-    if args[1] in ["-ra-inter-crop","-ra-inter-cousot","-ra-intra-crop","-ra-intra-cousot","-ra-test-range"]:
+    elif checkIfIsAnalysis(args) or checkIfIsTest(args):
         return True
-    else:
-        return False
+    return False
 
-if checkArgs(sys.argv):
-    fileName, fileExtension = os.path.splitext(sys.argv[2])
-    print("Compiling with Clang")
-    print("clang -c -emit-llvm "+fileName+fileExtension+" -o "+fileName+".bc")
-    os.system("clang -c -emit-llvm "+fileName+fileExtension+" -o "+fileName+".bc")
-    print("Generating eSSA")
-    os.system("opt -mem2reg "+fileName+".bc -o "+fileName+".bc")
-    os.system("opt -internalize -inline "+fileName+".bc -o "+fileName+".bc")
-    print("opt -instnamer -mem2reg -break-crit-edges "+fileName+".bc -o "+fileName+".bc")
-    os.system("opt -instnamer -break-crit-edges "+fileName+".bc -o "+fileName+".bc")
-    print("opt -load "+passPath+"vSSA.so -vssa "+fileName+".bc -o "+fileName+".essa.bc")
-    os.system("opt -load "+passPath+"vSSA.so -vssa "+fileName+".bc -o "+fileName+".essa.bc")
-    print("Running the Range Analysis Pass")
-    print("opt -load "+passPath+"RangeAnalysis.so "+sys.argv[1]+" "+fileName+".essa.bc")
-    os.system("opt -load "+passPath+"RangeAnalysis.so "+sys.argv[1]+" "+fileName+".essa.bc -stats")
-else:
+def checkIfIsAnalysis(args):
+    if args[1] in analysisTypes:
+        return True
+    return False
+
+def checkIfIsTest(args):
+    if args[1] in testTypes:
+        return True
+    return False
+
+def printHelp():
     print "./compile.py <OPTION> <FILE.c>"
     print "\tOPTION:"
[…]

### 2012-03-13 tests/experiment.sh

@@ -16,9 +16,8 @@ if [ $# -eq 2 ]; then
 	# Do not forget to put the ending / in every path below
 	llvmpath=/home/igor/workspace/ra/llvm-3.0/
 	testpath=${llvmpath}projects/test-suite/$1 #External/SPEC/CINT2006/
-	outputpath=$2#/home/igor/workspace/ra/experiments/
+	outputpath=$2 #/home/igor/workspace/ra/experiments/
 	build=Debug/
-
 	mkdir -p $outputpath
 
 	# Experiment 1

### 2012-03-13 src/RangeAnalysis/RangeAnalysis.cpp
Fixed num visits map.

@@ -2254,10 +2254,6 @@ bool Meet::growth(BasicOp* op) {
 /// in the constraint graph, the cropping analysis shyrinks these bounds back
 /// to ranges that respect the intersections.
 bool Meet::narrow(BasicOp* op) {
-	// Updates Fermap
-	const Value *V = op->getSink()->getValue();
-	FerMap[V]++;
-	
 	
 	APInt oLower = op->getSink()->getRange().getLower();
 	APInt oUpper = op->getSink()->getRange().getUpper();
@@ -2392,6 +2388,13 @@ void ConstraintGraph::update(const UseMap &compUseMap,
 		const Value* V = *actv.begin();
 		actv.erase(V);
 		
+#ifdef STATS
+		// Updates Fermap
+		if (meet == Meet::narrow) {
+			FerMap[V]++;
+		}
+#endif
+
 		//errs() << "Update: " << V->getName() << "\n";
 		// The use list.
 		const SmallPtrSet<BasicOp*, 8> &L = compUseMap.find(V)->second;
@@ -2453,11 +2456,12 @@ void ConstraintGraph::findIntervals() {
 	
 	for (Nuutila::iterator nit = sccList.begin(), nend = sccList.end();
 			nit != nend; ++nit) {
-		SmallPtrSet<VarNode*, 32> &component = sccList.components[*nit];
+		SmallPtrSet<VarNode*, 32> &component = *sccList.components[*nit];
 #ifdef SCC_DEBUG
 		--numberOfSCCs;
 #endif
-		// STATS
+		//PRINTCOMPONENT(component)
+
 		if (component.size() == 1) {
 			++numAloneSCCs;
 			fixIntersects(component);
@@ -2470,8 +2474,6 @@ void ConstraintGraph::findIntervals() {
 			if (component.size() > sizeMaxSCC) {
 				sizeMaxSCC = component.size();
 			}
-
-			//PRINTCOMPONENT(component)
 
 			UseMap compUseMap = buildUseMap(component);
 
@@ -3019,13 +3021,13 @@ void Nuutila::visit(Value *V, std::stack<Value*> &stack, UseMap *useMap) {
 		// list once more.
 		worklist.push_back(V);
 
-		SmallPtrSet<VarNode*, 32> SCC;
+		SmallPtrSet<VarNode*, 32> *SCC = new SmallPtrSet<VarNode*, 32>;
 
[…]

### 2012-03-13 src/RangeAnalysis/RangeAnalysis.h

@@ -95,8 +95,6 @@ errs() << "\n--------------\n"; \
 #define ASSERT(cond,msg)
 #endif
 
-#define NUMBER_FIXED_ITERATIONS 20
-
 using namespace llvm;
 
 /// In our range analysis pass we have to perform operations on ranges all the
@@ -802,7 +800,7 @@ public:
 	DenseMap<Value*, int> dfs;
 	DenseMap<Value*, Value*> root;
 	SmallPtrSet<Value*, 32> inComponent;
-	DenseMap<Value*, SmallPtrSet<VarNode*, 32> > components;
+	DenseMap<Value*, SmallPtrSet<VarNode*, 32>* > components;
 	std::deque<Value*> worklist;
 #ifdef SCC_DEBUG
     bool checkWorklist();

### 2012-03-13 doc/Manuscript/paper.tex
Wrote the section about experiments.

@@
 \begin{center}
-\includegraphics[width=0.8\textwidth]{images/algorithm}
+\includegraphics[width=0.7\textwidth]{images/algorithm}
 \end{center}
@@
 \begin{eqnarray*}
-\begin{array}{c}
+\begin{array}{c@{\hspace{0.9cm}}c}
 \inferrule{\lb{I[Y]} = -\infty \\ \lb{e(Y)} > -\infty}
 {I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
-\\
-\\
+&
 \inferrule{\lb{I[Y]} > \lb{e(Y)}}
@@
 {I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
-\\
-\\
+&
 \inferrule{\ub{I[Y]} < \ub{e(Y)}}
@@
 chain of mutually recursive function calls.
-These two observations imply that our runtime is likely to be linear in
+We do not count control dependency edges when measuring the size of SCCs, as
+they have no impact on the algorithm explained in Section~\ref{sub:micro}.
+Finally, as we see in the table, any node was visited at most three times by
+the narrowing operator.
+These three observations imply that our runtime is likely to be linear in
 practice, as we show in Figure~\ref{fig:TimeCorr}.
@@
 \#$N_v$ & 97494 & 5490 & 352423 & 14407 & 1853	 & 50095 & 449442 & 84846 & 3293 & 12517 & 38409 \\ \hline
+\#$N_c$ & 69530 & 3799 & 213535 & 11164 & 1156 & 28943 & 309310 & 63506 & 2336 & 8731 & 24530 \\ \hline
 \#S  & 91731 & 5108 & 346451 & 13309 & 1751 & 48784 & 435372 & 79909 & 3003 & 11539 & 35619 \\ \hline
@@
 \#U  & 89461 & 4983 & 344132 & 12919 & 1726 & 48400 & 431703 & 78407 & 2885 & 11267 & 34631 \\ \hline
+\#V  & 3	2 & 2 & 2 & 2 & 2 & 2 & 3 & 2 & 2 & 2 \\ \hline
 \end{tabular}
@@
 \#$N_v$: number of variable nodes in the constraint graph.
+\#$N_c$: number of constraint nodes in the constraint graph.
 % TODO: add also number of constraint nodes.
@@
 L: size of the largest SCC.
-\#U: number of SCCs with only one node.}
+\#U: number of SCCs with only one node.
+\#V: maximum number of times any variable node was visited by narrowing.}
 \end{table}%
@@
 \noindent
-\textbf{Runtime Behavior.}
+\textbf{Asymptotic complexity.}
 Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
@@
 
-%Chart 1: shows complexity of the algorithm, plotting program size and time
-% to run the analysis.
+\noindent
+\textbf{Time distribution.}
+Figure~\ref{fig:timeComp}(Left) shows how the processing time is distributed
+among the main phases of the inter-procedural version of our algorithm.
+In this case, we have inlined functions to increase program sizes.
+We analyze the entire SPEC 2006 integer benchmarks in 41.42secs.
+The time to build the e-SSA representation ss 7.03secs; the time to extract
+constraints from the program and build the constraint graphs is 7.87secs;
+the time to run Nuutila's algorithm to separate the constraint graph into
+strong components is 10.55secs, and the time to run the analysis itself is
+15.96secs.
+Notice that the Nuutila's entry includes also the time to sort the SCCs
+topologically.
 
-%Chart 2: time distribution for SPEC only
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=0.48\textwidth]{images/timeDistr}
+\includegraphics[width=0.48\textwidth]{images/timeIntraInterInline}
+\end{center}
+\caption{\label{fig:timeComp}
+(Left) Time distribution across different phases of our inter+inline
+algorithm.
+(Right) Runtime comparison between the intra, inter e inter+inline versions of
+our algorithm.
+The bars are normalized to the time to run the intra-procedural version.
+}
+\end{figure}
 
-% Intra/Inter/Inline comparison
-% =============================
+\noindent
+\textbf{Intra vs Inter-procedural runtimes.}
+As we saw in Figure~\ref{fig:algorithm}, our implementation supports three
+different execution modes: intra-procedural, inter-procedural, and
+inter-procedural with inlining.
+The latter approach gives us some form of context-sensitiveness, because our
+baseline compiler, LLVM, is able to inline non-recursive function calls.
+Figure~\ref{fig:timeComp}(Right) compares these three different setups.
+All the times are normalized to the time to run the intra-procedural version.
+On the average, the intra-procedural mode is XX\% faster than the
+inter-procedural mode.
+If we perform function inlining, then this difference is larger: XX\%.
 
-% Chart 3: Time comparison between intra, inter and inlining:
-% Each bar is inter/intra, or inline/intra
-% We want to show that the whole program analysis is fast.
+\noindent
+\textbf{The runtime impact of the e-SSA intermediate representation.}
+Figure~\ref{fig:eSSA_data} shows how the e-SSA transformation impacts the
+runtime of our algorithm.
+As we see in Figure~\ref{fig:eSSA_data}(Left), this alternative representation
+increases the size of the SPEC 2006 integer benchmarks by a factor of XX\%,
+when compared to the original program, in SSA form.
+Such growth is due to the $\sigma$-functions used to split live ranges
+after conditionals, and to the extra $\phi$-functions necessary to preserve the
+single static assignment property.
+The e-SSA form increases the runtime of our analysis by XX\%, as we see in
+Figure~\ref{fig:eSSA_data}(Right).
+In that figure we are comparing only the time to perform range analysis.
+The bars do not include the time to perform the e-SSA conversion, which we
+show in Figure~\ref{fig:timeComp}(Left).
 
-% (I) Chart 4: Precision comparison between intra/inter/inline
-% *not for spec*: take five largest of bitwise, stanford and SPEC CINT.
-% Add the three geomeans to the chart.
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=0.48\textwidth]{images/size_e_SSA}
+\includegraphics[width=0.48\textwidth]{images/time_e_SSA}
+\end{center}
+\caption{\label{fig:eSSA_data}
+(Left) Program growth due to the conversion from the SSA to the e-SSA form.
+(Right) Runtime slowdown of the range analysis when running it on the e-SSA
+form programs relative to the time to run it on the SSA-form programs.
+}
+\end{figure}
 
-% (I) Chart 5: Same as (4), but this time precision in terms of [c, c], [-, c],
-% [+, c], [-, +]. Perhaps take off [-, +]
+\noindent
+\textbf{The impact of whole program analysis on precision.}
+Figure~\ref{fig:precision} shows that the whole program analysis increases
+moderately the precision of our analysis.
+We show results for the five largest programs in three different categories of
+benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
+Bitwise~\cite{Stephenson00}.
+Our initial goal when developing this analysis was to support a bitwidth-aware
+register allocator.
+Thus, we measure precision in terms of the average number of bits that our
+analysis allows us to save to store the program variables.
+Our results for the SPEC programs were disappointingly: on the average, the
+intra-procedural version of our analysis saves 5.23\% of the total program
+bitwidth.
+By using the inter-procedural version, with function inlining to simulate
+context-sensitiveness, we can increase this number to 8.89\%.
+A manual inspection of the SPEC programs reveal that this result is expected:
+such benchmarks manipulate files, and their code do no provide enough
+explicit constants to power our analysis.
+However, with numerical benchmarks we fare much better.
+On the average our inter procedural algorithm reduces the bitwidth of the
+Stanford benchmarks by 36.24\%.
+Notice that this number is better than the one we get with inlining (31.97\%),
+because functions in those benchmarks are called only at one location.
+Finally, for Bitwise we obtain a bitwidth reduction of 12.27\%.
+However, this average is brought down by two outliers: \texttt{edge\_detect} and
+\texttt{sha}, which cannot be reduced.
+The bitwise benchmarks were implemented by Stephenson
+{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
+Our results are on par with those found by the original authors.
+Notice that all these benchmarks contain only the \texttt{main} function; thus,
+there is no different between the different versions of our algorithm.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/precision}
+\end{center}
+\caption{\label{fig:precision}
+The impact of whole program analysis on precision.
+Each bar measures the average bitwidth reduction after we run our analysis.}
+\end{figure}
+
+\noindent
+\textbf{The impact of e-SSA on precision.}
+Figure~\ref{fig:essaDivSSA} shows that the e-SSA transformation improves the
+precision of our analysis dramatically.
+If we convert the input programs from the original SSA representation used
+by LLVM to the e-SSA format, then, in some cases, such as in
+\texttt{Stanford.queens}, in the intra-procedural case with function
+inlining, we can increase the bitwidth reduction by 154x.
+This impressive difference happens because the standard SSA representation does
+not give our analysis any subsidy to acquire information from the outcome of
+conditionals.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/essaDivSSA}
+\end{center}
+\caption{\label{fig:essaDivSSA}
+The impact of the e-SSA transformation on precision.
+Each bar gives the ratio (bitwidth reduction on e-SSA form)/(bitwidth reduction
+on SSA form).}
+\end{figure}
 

### 2012-03-14 src/RangeAnalysis/RangeAnalysis.cpp
Implemented a log of operations. Fixed overflow in Range::mul and Range::add.

@@ -42,6 +42,7 @@ unsigned MAX_BIT_INT = 1;
 // operator is called on a variable. It was a Fernando's suggestion.
 DenseMap<const Value*, unsigned> FerMap;
 
+
 // ========================================================================== //
 // Static global functions and definitions
 // ========================================================================== //
@@ -264,11 +265,13 @@ bool InterProceduralRA<CGT>::runOnModule(Module &M) {
 	prof.updateTime("BuildGraph", elapsed);
 #endif
 #ifdef PRINT_DEBUG
-	CG->printToFile(*(M.begin()), "/tmp/" + M.getModuleIdentifier() + ".cgpre.dot");
+	std::string moduleIdentifier = M.getModuleIdentifier();
+	std::string mIdentifier = moduleIdentifier.substr(moduleIdentifier.rfind("/"));
+	CG->printToFile(*(M.begin()), "/tmp/" + mIdentifier + ".cgpre.dot");
 #endif
 	CG->findIntervals();
 #ifdef PRINT_DEBUG
-	CG->printToFile(*(M.begin()), "/tmp/" + M.getModuleIdentifier() + ".cgpos.dot");
+	CG->printToFile(*(M.begin()), "/tmp/" + mIdentifier + ".cgpos.dot");
 #endif
 	// FIXME: Não sei se tem que retornar true ou false
 	return true;
@@ -477,13 +480,21 @@ bool Range::isMaxRange() const {
 /// Add and Mul are commutatives. So, they are a little different 
 /// of the other operations.
 Range Range::add(const Range& other) {
+	const APInt &a = this->getLower();
+	const APInt &b = this->getUpper();
+	const APInt &c = other.getLower();
+	const APInt &d = other.getUpper();
 	APInt l = Min, u = Max;
-	if (this->getLower().ne(Min) && other.getLower().ne(Min)) {
-		l = getLower() + other.getLower();
+	if (a.ne(Min) && c.ne(Min)) {
+		l = a + c;
+		if(a.isNegative() == c.isNegative() && a.isNegative() != l.isNegative())
+			l = Min;
 	}
 
-	if (this->getUpper().ne(Max) && other.getUpper().ne(Max)) {
-		u = getUpper() + other.getUpper();
+	if (b.ne(Max) && d.ne(Max)) {
+		u = b + d;
[…]

### 2012-03-14 src/RangeAnalysis/RangeAnalysis.h

@@ -58,7 +58,7 @@
 //#define SCC_DEBUG
 
 //TODO: comment the line below to disable the dot printing of Constraint Graphs
-//#define PRINT_DEBUG
+#define PRINT_DEBUG
 
 //Used to enable the stats computing. Comment the below line to disable it
 #define STATS
@@ -94,7 +94,20 @@ errs() << "\n--------------\n"; \
 #else
 #define ASSERT(cond,msg)
 #endif
+//************************** Log Transactions ********************************//
+//#define LOG_TRANSACTIONS
 
+#ifdef LOG_TRANSACTIONS
+std::string _log_ErrorInfo;
+llvm::Twine _log_fileName = "/tmp/ratransactions";
+llvm::raw_fd_ostream _log_file(_log_fileName.str().c_str(), _log_ErrorInfo);
+#define LOG_TRANSACTION(str) _log_file << str << "\n";
+#define FINISH_LOG _log_file.close();
+#else
+#define LOG_TRANSACTION(str)
+#define FINISH_LOG
+#endif
+//****************************************************************************//
 using namespace llvm;
 
 /// In our range analysis pass we have to perform operations on ranges all the

### 2012-03-15 src/RangeAnalysis/RangeAnalysis.cpp
Started to fix logical operations "And", "Or", and "Xor". Jorge Navas pointed out that they are currently wrong.

@@ -876,6 +876,54 @@ Range Range::ashr(const Range& other) {
 	return Range(*min, *max);
 }
 
+int64_t minAND(int64_t a, int64_t b, int64_t c, int64_t d)
+{
+	int64_t m, temp;
+	
+	m = 0x80000000;
+	while (m != 0) {
+		if (~a & ~c & m) {
+			temp = (a | m) & -m;
+			if (temp <= b) {
+				a = temp;
+				break;
+			}
+			temp = (c | m) & -m;
+			if (temp <= d) {
+				c = temp;
+				break;
+			}
+		}
+		m = m >> 1;
+	}
+	return a & c;
+}
+
+int64_t maxAND(int64_t a, int64_t b, int64_t c, int64_t d)
+{
+	int64_t m, temp;
+	
+	m = 0x80000000;
+	while (m != 0) {
+		if (b & ~d & m) {
+			temp = (b & ~m) | (m - 1);
+			if (temp >= a) {
+				b = temp;
+				break;
+			}
+		}
+		else if (~b & d & m) {
+			temp = (d & ~m) | (m - 1);
+			if (temp >= c) {
+				d = temp;
+				break;
+			}
+		}
+		m = m >> 1;
+	}
+	return b & d;
+}
+
 Range Range::And(const Range& other) {
 	const APInt &a = this->getLower();
 	const APInt &b = this->getUpper();
@@ -918,6 +966,55 @@ Range Range::And(const Range& other) {
 	return Range(*min, *max);
 }
 
+int64_t minOR(int64_t a, int64_t b, int64_t c, int64_t d)
+{
+	int64_t m, temp;
+	
+	m = 0x80000000;
+	while (m != 0) {
+		if (~a & c & m) {
+			temp = (a | m) & -m;
+			if (temp <= b) {
+				a = temp;
+				break;
+			}
+		}
+		else if (a & ~c & m) {
+			temp = (c | m) & -m;
+			if (temp <= d) {
+				c = temp;
+				break;
+			}
+		}
+		m = m >> 1;
+	}
+	return a | c;
+}
+
+int64_t maxOR(int64_t a, int64_t b, int64_t c, int64_t d)
+{
+	int64_t m, temp;
+	
+	m = 0x80000000;
+	while (m != 0) {
+		if (b & d & m) {
+			temp = (b - m) | (m - 1);
+			if (temp >= a) {
+				b = temp;
+				break;
+			}
+			temp = (d - m) | (m - 1);
[…]

### 2012-03-15 doc/Manuscript/paper.tex
Added related works section to the paper.

@@
 
-The standard e-SSA form serves well analysis that need range information
-at the use site of variables, such as conditional constant propagation,
-redundant code elimination and detection of buffer overflows.
-On the other hand, there exist compilation algorithms that require range
-information at the whole live range of variables.
-Examples of such algorithms include the family of
-bitwidth aware register allocators~\cite{Barik06,Pereira08,Tallam03} and
-hardware synthesizers~\cite{Cong05,Mahlke01,Stephenson00}.
-In order to provide extra precision to these algorithms, we may work with a
-non-pruned version of e-SSA form, which we produce by simply inserting copies
-after every conditional test, and then reconverting the entire program to
-SSA form.
-Figure~\ref{fig:ex_non_pruned} illustrates the difference between these two
-program representations.
-Notice that the non-pruned flavor might introduce dead variables in the
-program code, which can be removed by standard dead code elimination.
+%The standard e-SSA form serves well analysis that need range information
+%at the use site of variables, such as conditional constant propagation,
+%redundant code elimination and detection of buffer overflows.
+%On the other hand, there exist compilation algorithms that require range
+%information at the whole live range of variables.
+%Examples of such algorithms include the family of
+%bitwidth aware register allocators~\cite{Barik06,Pereira08,Tallam03} and
+%hardware synthesizers~\cite{Cong05,Mahlke01,Stephenson00}.
+%In order to provide extra precision to these algorithms, we may work with a
+%non-pruned version of e-SSA form, which we produce by simply inserting copies
+%after every conditional test, and then reconverting the entire program to
+%SSA form.
+%Figure~\ref{fig:ex_non_pruned} illustrates the difference between these two
+%program representations.
+%Notice that the non-pruned flavor might introduce dead variables in the
+%program code, which can be removed by standard dead code elimination.
 
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{images/ex_non_pruned}
-\end{center}
-\caption{\label{fig:ex_non_pruned}
-(a) Example program.
-(b) Standard e-SSA form.
-(c) Non-pruned e-SSA form.}
-\end{figure}
+%\begin{figure}[t!]
+%\begin{center}
+%\includegraphics[width=0.9\textwidth]{images/ex_non_pruned}
+%\end{center}
+%\caption{\label{fig:ex_non_pruned}
+%(a) Example program.
+%(b) Standard e-SSA form.
+%(c) Non-pruned e-SSA form.}
+%\end{figure}
 
-\subsection{Extracting Constraints}
-\label{sub:constraints}
-
+\paragraph{Extracting Constraints.}
 The constraints in Figure~\ref{fig:eval_function} handles
@@
 Multiplication & $v_1 = v_2 * c$ & $V_1 = c V_2$ \\
-Int. division & $v_1 = v_2 \ / \ v_3$ & $V_1 = V_2$ \\
- & & $V_1 = - V_2$ \\
+Int. division & $v_1 = v_2 \ / \ v_3$ & $V_1 = V_2; V_1 = - V_2$ \\
 Modulus & $v_1 = v_2 \ \% \ c$ & $V_1 = [0, c - 1]$ \\
@@
 
-\subsection{The Dependence Graph}
-\label{sub:graph}
-
+\paragraph{The Constraint Graph.}
 The main data structure that we use to solve the range analysis problem is
@@
 \begin{enumerate}
-\item Run widening analysis: $O(N)$.
+\item Run growth analysis: $O(N)$.
 
@@
 
-\item Run narrowing analysis: $O(N^2)$.
+\item Apply the narrowing operator: $O(N^2)$.
 
@@
 
-\paragraph{Widening Analysis.}
-
+\noindent
+\textbf{Growth Analysis.}
 The first step of our algorithm consists in determining the growth behavior of
 the interval bound to each constraint variable.
-We accomplish this task via Cousot and Cousot's widening
+We ensure termination of this phase via Cousot and Cousot's widening
 operator~\cite[p.247]{Cousot77}.
@@
 change at most three times.
+However, as we show in Figure~\ref{fig:essaDivSSA}(Bottom), we can be more
+precise if we evaluate the constraints a few times before applying widening.
 
@@
 \begin{tabular}{c@{\hspace{1.5cm}}c}
-\begin{minipage}{2.4cm}
+\begin{minipage}{2cm}
 \includegraphics{images/growth_lattice}
@@
 &
-\begin{minipage}{6cm}
+\begin{minipage}{8cm}
 \begin{small}
 \begin{eqnarray*}
-\begin{array}{c}
+\begin{array}{c@{\hspace{0.5cm}}c}
 \inferrule{I[Y] = [\bot, \bot]}{I[Y] \leftarrow e(Y)}
-\\
-\\
+&
 \inferrule{\lb{e(Y)} < \lb{I[Y]} \\ \ub{e(Y)} > \ub{I[Y]}}
@@
 \\
-\\
 \inferrule{\lb{e(Y)} < \lb{I[Y]}}
 {I[Y] \leftarrow [-\infty, \ub{I[Y]}]}
-\\
-\\
+&
 \inferrule{\ub{e(Y)} > \ub{I[Y]}}
@@
 
-\paragraph{Fixing futures.}
-
+\noindent
+\textbf{Fixing futures.}
 The ranges found by the widening analysis tells us which variables have fixed
@@
 
-\paragraph{Narrowing Analysis.}
-
+\noindent
+\textbf{Narrowing Analysis.}
 The widening analysis associates very conservative bounds to each variable.
@@
 
-\paragraph{Example}
-
+\noindent
+\textbf{Example:}
 Continuing with our example, Figure~\ref{fig:ex_partition_grow_crop} shows
-the application of our algorithm on the last strongly connected component of
+the application of our algorithm on the last strong component of
 Figure~\ref{fig:ex_graph}.
-Notice that upon meeting this SCC, we have already determined that the interval
+Upon meeting this SCC, we have already determined that the interval
 $[0, 0]$ is bound to $I_0$ and that the interval $[100, 100]$ is bound to
@@
 
-\begin{table}[htdp]
+\begin{table}[t!]
 \begin{center}
@@
 \textbf{The impact of whole program analysis on precision.}
-Figure~\ref{fig:precision} shows that the whole program analysis increases
+Figure~\ref{fig:essaDivSSA}(Top) shows that the whole program analysis increases
 moderately the precision of our analysis.
@@
 
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=1\textwidth]{images/precision}
-\end{center}
-\caption{\label{fig:precision}
-The impact of whole program analysis on precision.
-Each bar measures the average bitwidth reduction after we run our analysis.}
-\end{figure}
-
 \noindent
 \textbf{The impact of e-SSA on precision.}
-Figure~\ref{fig:essaDivSSA} shows that the e-SSA transformation improves the
-precision of our analysis dramatically.
+Figure~\ref{fig:essaDivSSA}(Middle) shows that the e-SSA transformation improves
+the precision of our analysis dramatically.
 If we convert the input programs from the original SSA representation used
@@
 \caption{\label{fig:essaDivSSA}
-The impact of the e-SSA transformation on precision.
-Each bar gives the ratio (bitwidth reduction on e-SSA form)/(bitwidth reduction
-on SSA form).}
+The impact of different design trade-offs on precision of range analysis.}
 \end{figure}
 
-% Impact of e-SSA form:
-% =====================
+\noindent
+\textbf{The impact of the number of widening iterations on precision.}
+It is well-known that premature widening might cause loss of
+information~\cite[Sec5.2]{Cousot09}.
+Figure~\ref{fig:essaDivSSA} shows how the precision of our analysis varies with
+the number of iterations used before resorting to widening.
+We have applied the widening operator in Figure~\ref{fig:growth_analysis}
+after each node was visited 1, 2 and 16 times.
 
-% Chart 6: time overhead e-SSA. So, for each SPEC 2006, do time e-ssa/time ssa.
+\section{Related works}
+\label{sec:rel}
 
-% (I) Chart 7: precision with e-SSA form.
-% Percentage reduction (bitwidth only) for five largest of bitwise, stanford
-% and SPEC CINT.
+% Early implementations of range analysis.
+Range analysis is an old ally of compiler writers.
+The notions of widening and narrowing were introduced by Cousot and Cousot in
+one of the most cited papers in computer science~\cite{Cousot77}.
+Different algorithms for range analysis have been subsequently proposed
+by Stephenson {\em et al.}~\cite{Stephenson00}, Patterson~\cite{Patterson95}
+and many other researchers.
+Recently there have been many independent efforts to find exact, polynomial
+time algorithms to solve constraints on the interval
+lattice~\cite{Gawlitza09,Costan05,Lakhdar11,Su04,Su05}.
+However, these works are still very theoretical, and have not yet been used to
+analyze large programs.
+Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
+about 25 minutes to go over a program with almost 900 basic blocks, which we
+analyze in less than one second.
+We do not claim our approach is as precise as such algorithms, even though we
+are able to exactly analyze 4/5 of the examples presented in~\cite{Lakhdar11}.
+On the contrary, this paper presents a compromise between precision and speed
+that scales to very large programs.
 
-% Size overhead of different representations
-% ==========================================
+% Scalability: Astree and others.
+There have been many practical approaches to abstract interpretation,
+with special emphasis on range analysis~\cite{Blanchet03,Jung05,Cousot09}.
+Cousot's group, for instance, has been able to globally analyze programs with
+thousands of lines of code, albeit they use domain specific tools.
+Astr\'{e}, for example, analyzes programs that do not contain recursive calls.
+The work that is the closest to us is Oh {\em et al.}'s very recent abstract
+interpretation framework~\cite{Oh12}.
+Oh {\em et al.} discuss an implementation of range analysis on the interval
+lattice that scales up to a program with 1,363K LoC (ghostscript-9.00), but
+they do not provide results about its precision.
+We could not find the benchmarks used in those experiments for a
+direct comparison -- the distribution of ghostscript-9.00 available in the
+LLVM test suite has 27K LoC.
+On the other hand, we globally analyzed our largest benchmark, SPEC CPU 2006's
+\texttt{gcc}, enabling function inlining, in less than 15secs.
+Oh {\em et al.}'s implementation took orders of magnitude more time to go over
+programs of similar size.
+However, whereas they provide a framework to develop general sparse analyses, we
+only solve range analysis on the interval lattice.
 
-% Chart 8: Size overhead e-SSA/SSA. SPEC 2006 only.
-
-% Chart 9: Size overhead inline/non-inline. SPEC 2006 only.
-
-% Precision with different number of fixed iterations
-% ===================================================
-
-% Chart 10: Run inter + e-SSA with N*|SCC| in fixed update for five largest of
-% bitwise, stanford and SPEC CINT.
-
-% Table 1: Structural Properties for SPEC CPU 2006.
-% Remember to add %SCC's with just one component.
-
-% Impact of strongly connected components
-% =======================================
-
-%In our experiments we have observed that the vast majority of strongly
-%connected components that constitute typical dependence graphs contain only
-%one element.
-%Furthermore, composite components tend to be small.
-%Therefore, range analysis admit very fast solution.
+% PLDI 2012 paper
 

### 2012-03-18 doc/Manuscript/paper.tex
Fernando has edited the section about experiments, and done minor editting in the overal text.

@@
+% Explain that we get the topological ordering for free via nuutila's algorithm.
 \documentclass{llncs}
@@
 Assignment form as the intermediate representation.
-We show that by processing the strongly connected components that constitute
-the graph of dependences between variables in topological order we not only
-gain in time, but also in precision.
-We have implemented our technique in LLVM, an industrial strength compiler,
-and have been able to process over 4 million assembly instructions in a
-few seconds.
+%We show that by processing the strongly connected components that constitute
+%the graph of dependences between variables in topological order we not only
+%gain in time, but also in precision.
+We claim to present in this paper the most extensive empirical evaluation of
+range analysis in an industrial strength compiler.
+We have implemented our technique in LLVM and have been able to process
+programs totaling almost 2 million lines of C in a few seconds.
+We do global program analysis, and achieve a simple form of context
+sensitiveness via function inlining.
 \end{abstract}
@@
 constructs, their description of how to deal with loops was
-``not very explicit"~\cite[p.422]{Gawlitza09}.
+not very explicit.
 Thus, in order to solve loops we adopted Gawlitza
@@
 We have implemented our algorithm in the LLVM compiler~\cite{Lattner04}, and
-have used it to analyze a test suite with 1.4 million lines of C code.
+have used it to analyze a test suite with almost 2 million lines of C code.
 In Section~\ref{sec:exp} we provide empirical evidence showing that our
-implementation is fast: it analyzes the gcc source code in 4 seconds, for
-instance.
+implementation is fast: it globally analyzes the gcc source code in less than
+15 seconds, for instance.
 It is also precise: our results are similar to Stephenson
@@
 assume that the program is correct.
-Furthermore, we have been able to find tight bounds to many of the examples
-used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
+Furthermore, we have been able to find tight bounds to the majority of the
+examples used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
 {\em et al.}~\cite{Lakhdar11}, who rely on much more costly methods.
@@
 mapping from these variables to intervals in $\varset{Z}^2$.
-Our objective is to solve a constraint system $C$, formed by instances of the
-six different types of constraints seen in Figure~\ref{fig:eval_function}
-(left).
-Figure~\ref{fig:eval_function} (right) defines a valuation function $e$ that
+Our objective is to solve a constraint system $C$, formed by constraints such
+as those seen in Figure~\ref{fig:eval_function}(left).
+Figure~\ref{fig:eval_function}(right) defines a valuation function $e$ that
 computes $Y = f(\ldots)$, given I.
@@
 As we will see shortly, we can improve this solution substantially by using
-a more sophisticated program representation -- the e-SSA form.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=0.7\textwidth]{images/ex1}
+a more sophisticated program representation -- the e-SSA form -- which
+gives us path-sensitiveness.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/ex1}
 \end{center}
@@
 (a) Example program.
-(b) Control Flow Graph (CFG) in SSA form.
+(b) Control Flow Graph in SSA form.
 (c) Constraints that we extract from the program.
-(d) A possible solution to the range analysis problem.}
+(d) Possible solution to the range analysis problem.}
 \end{figure}
@@
 In this section we explain the algorithm that we use to solve the range
-analysis problem for an entire program.
+analysis problem.
 This algorithm involves a number of steps.
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth]{images/algorithm}
+\includegraphics[width=1\textwidth]{images/algorithm}
 \end{center}
 \caption{\label{fig:algorithm}
-Our algorithm to solve the range analysis problem.
+Our implementation of range analysis.
 Rounded boxes are optional steps.}
@@
 
-\subsection{Choosing a Program Representation}
-\label{sub:rep}
-
+\paragraph{Choosing a Program Representation.}
 The solution that we found to the range analysis problem in Figure~\ref{fig:ex1}
 is imprecise because we did not take conditional tests into considerations.
-Branches gives us information about the ranges that some variables assume, but
+Branches give us information about the ranges that some variables assume, but
 only at {\em specific} program points.
@@
 the {\em Extended Static Single Assignment} form, or e-SSA for short.
-In this paper, we have experimented with two different flavors of e-SSA form,
-which we call {\em standard} and {\em non-pruned}.
-
-We will show first how we convert a program into standard e-SSA form.
+
+We use a non-iterative algorithm to convert a program to e-SSA form.
 Let $(v < c)?$ be a conditional test, and let $l_t$ and $l_f$ be labels in
 the program, such that $l_t$ is the target of the test if the condition is true,
-and let $l_f$ is the target when the condition is false.
-If $l_t$ dominates any use of $v$, then we insert at $l_f$ a copy
+and $l_f$ is the target when the condition is false.
+If $l_f$ dominates any use of $v$, then we insert at $l_f$ a copy
 $v_f = v \sqcap [-\infty, c-1]$, where $v_f$ is a fresh name.
@@
 program.
+The e-SSA form allows us to bind interval information directly to the live
+ranges of variables; thus, allowing us to solve range analysis sparsely.
+More traditional approaches, which we call {\em dense analyses}, bind
+interval information to pairs formed by variables and program points.
 
@@
 \paragraph{Extracting Constraints.}
-The constraints in Figure~\ref{fig:eval_function} handles
-many different programming instructions.
-The $\phi$ assignments, such as $Y = \phi(X_1, X_2)$, represent
-the $\phi$-functions typically found in SSA form programs.
-As discussed in Section~\ref{sub:rep}, we use intersection
-constraints to bound the variables created due to live range splitting
-at conditionals.
+Our implementation handles 18 different assembly instructions.
+The constraints in Figure~\ref{fig:eval_function} show some meaningful
+examples.
+These instructions include, for instance, the multiplicative operators
+\texttt{mul}, \texttt{div} and \texttt{modulus},
+the bitwise operators \texttt{and}, \texttt{or}, \texttt{xor} and \texttt{neg},
+the different types of shifts, the additive operators \texttt{add} and
+\texttt{sub}, and the logical operators \texttt{andalso}, \texttt{orelse} and
+\texttt{not}.
+We also handle different types of type conversion, e.g., converting 8-bits
+characters to 32-bits integers and vice-versa.
+Our constraint system deals with the $\phi$-functions from the SSA-form
+representation.
+Finally, we handle conditionals via intersection constraints created
+during the live range splitting of variables.
 The widening analysis that we will introduce in Section~\ref{sub:micro}
@@
 this monotonicity.
-However, these non-monotonic instructions have a conservative
-translation to our constraint system.
-Figure~\ref{fig:constraints} shows some examples of constraints that we
-derive from typical assembly instructions.
-
-\begin{figure}[t!]\begin{center}
-\begin{tabular}{|l@{\hspace{0.2cm}}l@{\hspace{0.3cm}}l|} \hline
-\textbf{Description} & \textbf{Operation} & \textbf{Constraint} \\ \hline
-Constant & $v = c$ & $V = [c, c]$ \\
-Assignment & $v_1 = v_2$ & $V_1 = V_2$ \\
-Multiplication & $v_1 = v_2 * c$ & $V_1 = c V_2$ \\
-Int. division & $v_1 = v_2 \ / \ v_3$ & $V_1 = V_2; V_1 = - V_2$ \\
-Modulus & $v_1 = v_2 \ \% \ c$ & $V_1 = [0, c - 1]$ \\
-Bitwise and & $v_1 = v_2 \ \& \ c$ & $V_1 = v_2 \sqcap [0, c]$ \\
-Bitwise or & $v_1 = v_2 \ | \ v_3$ & $V_1 = V_2 + V_3$ \\
-Left shift & $v_1 = v_2 \ \qside \ c$ & $V_1 = 2^c V_2$ \\ \hline
-\end{tabular}
-\end{center}
-\caption{\label{fig:constraints}Example of constraint derivation rules.
-We let $c$ denote a positive constant, $\{v, v_1, v_2\}$ are program
-variables and $\{V, V_1, V_2\} \subset \cal{V}$.}
-\end{figure}
+However, these non-monotonic instructions have conservative approximations.
+%TODO: fix this discussion.
+
+%Figure~\ref{fig:constraints} shows some examples of constraints that we
+%derive from typical assembly instructions.
+%
+%\begin{figure}[t!]\begin{center}
+%\begin{tabular}{|l@{\hspace{0.2cm}}l@{\hspace{0.3cm}}l|} \hline
+%\textbf{Description} & \textbf{Operation} & \textbf{Constraint} \\ \hline
+%Constant & $v = c$ & $V = [c, c]$ \\
+%Assignment & $v_1 = v_2$ & $V_1 = V_2$ \\
+%Multiplication & $v_1 = v_2 * c$ & $V_1 = c V_2$ \\
+%Int. division & $v_1 = v_2 \ / \ v_3$ & $V_1 = V_2; V_1 = - V_2$ \\
+%Modulus & $v_1 = v_2 \ \% \ c$ & $V_1 = [0, c - 1]$ \\
+%Bitwise and & $v_1 = v_2 \ \& \ c$ & $V_1 = v_2 \sqcap [0, c]$ \\
+%Bitwise or & $v_1 = v_2 \ | \ v_3$ & $V_1 = V_2 + V_3$ \\
+%Left shift & $v_1 = v_2 \ \qside \ c$ & $V_1 = 2^c V_2$ \\ \hline
+%\end{tabular}
+%\end{center}
+%\caption{\label{fig:constraints}Example of constraint derivation rules.
+%We let $c$ denote a positive constant, $\{v, v_1, v_2\}$ are program
+%variables and $\{V, V_1, V_2\} \subset \cal{V}$.}
+%\end{figure}
 
@@
 We add an edge from $N_v$ to $N_c$ if the name $V$ is used in $C$.
-We add an edge from $N_c$ to $V$ if the constraint $C$ defines the name
+We add an edge from $N_c$ to $N_v$ if the constraint $C$ defines the name
 $V$.
@@
 
-\subsection{Solving Range Analysis on Strong Components}
-\label{sub:macro}
-
+\paragraph{Solving Range Analysis on Strong Components}
 Having built the dependence graph, we proceed to solve the range analysis
@@
 (i) $N_v \notin S$; and
-(ii) there exists a variable node $V' \in S$, and a constraint node $N_c$,
-such that $V' \leftarrow N_c$, and $N_c \leftarrow V$.
-This propagation gives us a very modular algorithm:
-when analyzing a strongly connected component $S$ we can rest assured that
-any influence that $S$ might suffer from nodes outside it has been already
+(ii) there exists a variable node $N_v' \in S$, and a constraint node $N_c$,
+such that $N_v' \leftarrow N_c$, and $N_c \leftarrow N_v$.
+This propagation gives us a very modular algorithm.
+It ensures that when analyzing a strongly connected component $S$ any influence
+that $S$ might suffer from nodes outside it has been already
 taken into consideration.
@@
 consideration.
-
-\subsection{Finding Ranges: the Micro Algorithm}
+As we show in Section~\ref{sec:exp}, most of the strong components in actual
+programs are singletons.
+Furthermore, the composite components tend to be small.
+These two facts ensure that the more costly parts of our algorithm only have to
+handle small inputs.
+
+\subsection{Finding Ranges in Strongly Connected Components}
 \label{sub:micro}
@@
 \\
+\\
 \inferrule{\lb{e(Y)} < \lb{I[Y]}}
@@
 \caption{\label{fig:growth_analysis}
-(Left) The lattice of the widening analysis.
+(Left) The lattice of the growth analysis.
 (Right) Cousot and Cousot's widening operator. We evaluate the rules from top 
@@
 Each core has a 4,096KB L1 cache.
-The operation system is Linux Ubuntu 10.04.4.
+We use Linux Ubuntu 10.04.4.
 Our implementation of range analysis has 3,958 lines of
 commented C++ code, and our e-SSA conversion module has 672 lines.
-We have tested it on 428 C programs that constitute the LLVM test suite, plus
-the C programs from SPEC CPU 2006.
-These programs together gave us approximately 4.76 million assembly
-instructions to analyze.
+We have analyzed the 428 C programs in the LLVM test suite, plus the integer
+benchmarks in SPEC CPU 2006.
+Together, these programs contain 4.76 million assembly instructions.
 
@@
 \textbf{Structural Properties.}
-Table~\ref{tab:struct} shows structural properties of the programs in
-the SPEC 2006 benchmark suite.
-Most of the strongly connected components that constitute
-the constraint graphs are singletons.
-For instance, 99.11\% of the SSCs found in gcc (\texttt{403.gcc}) have only
+Table~\ref{tab:struct} shows data taken from the SPEC 2006 integer benchmark
+suite for the inter-procedural version of our analysis.
+Three observations provide evidence that our algorithm is linear in practice,
+as we will show in Figure~\ref{fig:TimeCorr}.
+First, most of the strongly connected components found in constraint graphs are
+singletons.
+For instance, 99.11\% of the SSCs in gcc (\texttt{403.gcc}) have only
 one node.
-Therefore the most complex phases of our algorithm are
-confined to a few parts of the constraint graphs that we have to handle.
-Furthermore, the strongly connected components are not large -- our largest
-component, found in \texttt{403.gcc} has 2,131 nodes, and exists due to a long
-chain of mutually recursive function calls.
+Hence, the most complex phases of our algorithm are confined to a small fraction
+of the constraint graphs.
+Second, the strongly connected components are not large -- our largest
+component, found in \texttt{403.gcc}, has 2,131 nodes.
+It exists due to a long chain of mutually recursive function calls.
 We do not count control dependency edges when measuring the size of SCCs, as
 they have no impact on the algorithm explained in Section~\ref{sub:micro}.
-Finally, as we see in the table, any node was visited at most three times by
+Third, as we see in the table, any node was visited at most three times by
 the narrowing operator.
-These three observations imply that our runtime is likely to be linear in
-practice, as we show in Figure~\ref{fig:TimeCorr}.
 
@@
 \#U  & 89461 & 4983 & 344132 & 12919 & 1726 & 48400 & 431703 & 78407 & 2885 & 11267 & 34631 \\ \hline
-\#V  & 3	2 & 2 & 2 & 2 & 2 & 2 & 3 & 2 & 2 & 2 \\ \hline
+\#V  & 3	& 2 & 2 & 2 & 2 & 2 & 2 & 3 & 2 & 2 & 2 \\ \hline
 \end{tabular}
@@
 \caption{\label{tab:struct}Structural Properties of the programs in the SPEC CPU
-2006 benchmark suite.
+2006 integer benchmark suite.
 \#I: number of assembly instructions in the original program.
@@
 \begin{center}
-\includegraphics[width=1\textwidth]{images/TimeCorr}
+\includegraphics[width=0.9\textwidth]{images/TimeCorr}
 %\includegraphics[width=0.4\textwidth]{images/Complexity}
@@
 \textbf{Time distribution.}
-Figure~\ref{fig:timeComp}(Left) shows how the processing time is distributed
+Figure~\ref{fig:timeComp}(Left) shows the distribution of processing time
 among the main phases of the inter-procedural version of our algorithm.
-In this case, we have inlined functions to increase program sizes.
-We analyze the entire SPEC 2006 integer benchmarks in 41.42secs.
-The time to build the e-SSA representation ss 7.03secs; the time to extract
+Functions are inlined to increase program sizes.
+We analyze the SPEC 2006 integer benchmarks in 41.42secs.
+The time to build the e-SSA representation is 7.03secs; the time to extract
 constraints from the program and build the constraint graphs is 7.87secs;
-the time to run Nuutila's algorithm to separate the constraint graph into
-strong components is 10.55secs, and the time to run the analysis itself is
-15.96secs.
-Notice that the Nuutila's entry includes also the time to sort the SCCs
-topologically.
+the time to find SCCs via Nuutila's algorithm and sort them topologically is
+10.55secs, and the time to run the analysis itself is 15.96secs.
 
@@
 \textbf{Intra vs Inter-procedural runtimes.}
-As we saw in Figure~\ref{fig:algorithm}, our implementation supports three
-different execution modes: intra-procedural, inter-procedural, and
-inter-procedural with inlining.
-The latter approach gives us some form of context-sensitiveness, because our
-baseline compiler, LLVM, is able to inline non-recursive function calls.
-Figure~\ref{fig:timeComp}(Right) compares these three different setups.
-All the times are normalized to the time to run the intra-procedural version.
-On the average, the intra-procedural mode is XX\% faster than the
+As we saw in Figure~\ref{fig:algorithm}, our implementation supports
+intra- and inter-procedural execution modes.
+On top of these modes, we can optionally perform function inline to obtain
+limited context-sensitiveness.
+Our baseline compiler, LLVM, is able to inline non-recursive function calls.
+Figure~\ref{fig:timeComp}(Right) compares three different execution modes.
+Times are normalized to the time to run the intra-procedural analysis
+without inlining.
+On the average, the intra-procedural mode is 28.92\% faster than the
 inter-procedural mode.
-If we perform function inlining, then this difference is larger: XX\%.
-
-\noindent
-\textbf{The runtime impact of the e-SSA intermediate representation.}
-Figure~\ref{fig:eSSA_data} shows how the e-SSA transformation impacts the
-runtime of our algorithm.
-As we see in Figure~\ref{fig:eSSA_data}(Left), this alternative representation
-increases the size of the SPEC 2006 integer benchmarks by a factor of XX\%,
-when compared to the original program, in SSA form.
-Such growth is due to the $\sigma$-functions used to split live ranges
+If we perform function inlining, then this difference is larger: 45.87\%.
+These numbers are close because we run our analysis on strong components.
+
+\noindent
+\textbf{The impact of strong components in time and precision.}
+Figure~\ref{fig:impactEssaSCC}(a) shows that strong components dramatically
+decrease the runtime of our analysis.
+For instance, this design improves the analysis time of \texttt{483.xalancbmk}
+452 times!
+Moreover, as the example in Section~\ref{sub:micro} shows, our algorithm benefits
+from the partitioning of constraint graphs into SCCs to be more precise.
+Figure~\ref{fig:impactEssaSCC}(b) measures this precision.
+Range propagation along strong components allows us, for example, to reduce
+41.5\% more bits in \texttt{401.bzip}.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{images/impactEssaSCC}
+\end{center}
+\caption{\label{fig:impactEssaSCC}
+How strong components and the e-SSA representation improve our algorithm.
+This data refer to the inter-procedural analysis with function inlining.
+}
+\end{figure}
+
+\noindent
+\textbf{Impact of the e-SSA on runtime.}
+Figure~\ref{fig:impactEssaSCC}(c) shows that the e-SSA conversion increases
+the size of the SPEC 2006 integer benchmarks by 6.52\%.
+This growth is due to the $\sigma$-functions used to split live ranges
 after conditionals, and to the extra $\phi$-functions necessary to preserve the
 single static assignment property.
-The e-SSA form increases the runtime of our analysis by XX\%, as we see in
-Figure~\ref{fig:eSSA_data}(Right).
-In that figure we are comparing only the time to perform range analysis.
+As a consequence of the program growth, the e-SSA form increases the runtime of
+our analysis by 7.11\%, as seen in Figure~\ref{fig:impactEssaSCC}(d).
+We are comparing only the time to perform range analysis.
 The bars do not include the time to perform the e-SSA conversion, which we
-show in Figure~\ref{fig:timeComp}(Left).
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=0.48\textwidth]{images/size_e_SSA}
-\includegraphics[width=0.48\textwidth]{images/time_e_SSA}
-\end{center}
-\caption{\label{fig:eSSA_data}
-(Left) Program growth due to the conversion from the SSA to the e-SSA form.
-(Right) Runtime slowdown of the range analysis when running it on the e-SSA
-form programs relative to the time to run it on the SSA-form programs.
-}
-\end{figure}
+showed in Figure~\ref{fig:timeComp}(Left).
 
@@
 register allocator.
-Thus, we measure precision in terms of the average number of bits that our
-analysis allows us to save to store the program variables.
+Thus, we measure precision by the average number of bits that our
+analysis allows us to save per program variables.
 Our results for the SPEC programs were disappointingly: on the average, the
@@
 A manual inspection of the SPEC programs reveal that this result is expected:
-such benchmarks manipulate files, and their code do no provide enough
-explicit constants to power our analysis.
+they manipulate files, and their code do no provide enough
+explicit constants to power our analysis up.
 However, with numerical benchmarks we fare much better.
@@
 Our results are on par with those found by the original authors.
-Notice that all these benchmarks contain only the \texttt{main} function; thus,
-there is no different between the different versions of our algorithm.
+The bitwise programs contain only the \texttt{main} function; thus, different
+versions of our algorithm find the same results.
 
@@
 If we convert the input programs from the original SSA representation used
-by LLVM to the e-SSA format, then, in some cases, such as in
-\texttt{Stanford.queens}, in the intra-procedural case with function
-inlining, we can increase the bitwidth reduction by 154x.
+by LLVM to the e-SSA format, then, in some cases, e.g.,
+\texttt{Stanford.queens}, we can increase bitwidth reduction by 154x.
 This impressive difference happens because the standard SSA representation does
@@
 \noindent
-\textbf{The impact of the number of widening iterations on precision.}
+\textbf{The impact of the number of iterations before widening on precision.}
 It is well-known that premature widening might cause loss of
@@
 For a proof that the e-SSA form transformation is semantics preserving, we
-invite the reader to check a recent report of ours~\cite{XXXX}.
+invite the reader to check a recent report of ours~\cite{Tavares11b}.
 In this work we also show that the results that we obtain via the
@@
 associated with the interval $[l, u]$ along its entire live range.
+We have used a dynamic profiler, available in our webpage, to empirically
+validate our results.
 

### 2012-03-21 src/RangeAnalysis/RangeAnalysis.cpp
Fixed a bug in getting the correct module name identifier.

@@ -266,7 +266,8 @@ bool InterProceduralRA<CGT>::runOnModule(Module &M) {
 #endif
 #ifdef PRINT_DEBUG
 	std::string moduleIdentifier = M.getModuleIdentifier();
-	std::string mIdentifier = moduleIdentifier.substr(moduleIdentifier.rfind("/"));
+    int pos = moduleIdentifier.rfind("/");
+	std::string mIdentifier = pos > 0 ? moduleIdentifier.substr(pos) : moduleIdentifier;
 	CG->printToFile(*(M.begin()), "/tmp/" + mIdentifier + ".cgpre.dot");
 #endif
 	CG->findIntervals();
@@ -2326,7 +2327,7 @@ bool Meet::fixed(BasicOp* op){
 	Range newInterval = op->eval();
 	
 	op->getSink()->setRange(newInterval);
-	LOG_TRANSACTION("WIDEN::" << op->getSink()->getValue()->getName() << ": " << oldInterval << " -> " << newInterval)
+	LOG_TRANSACTION("FIXED::" << op->getSink()->getValue()->getName() << ": " << oldInterval << " -> " << newInterval)
 	return oldInterval != newInterval;
 }
 
@@ -2624,7 +2625,7 @@ void ConstraintGraph::findIntervals() {
 
 			generateEntryPoints(component, entryPoints);
 			//iterate a fixed number of time before widening
-			update(component.size()*16 /*| NUMBER_FIXED_ITERATIONS*/, compUseMap, entryPoints);
+			update(component.size()*2 /*| NUMBER_FIXED_ITERATIONS*/, compUseMap, entryPoints);
 
 #ifdef PRINT_DEBUG
 			if (func)

### 2012-03-21 doc/Manuscript/paper.tex
Added Victor's fixes to the paper.

@@
 constraint system in topological ordering.
-It is well-know that this technique is essential to speedup constraint solving
+It is well-known that this technique is essential to speedup constraint solving
 algorithms~\cite[Sec 6.3]{Nielson99}; however, we show that a careful
@@
 First, we convert the program to a suitable intermediate representation that
-makes it easier to extract constraint.
+makes it easier to extract constraints.
 From these constraints, we build a dependence graph that allows us to do
@@
 Some of the steps in the algorithm are optional.
-They improve the precision of the range analysis, at the expenses of a longer
+They improve the precision of the range analysis, at the expense of a longer
 running time.
@@
 
-\paragraph{Choosing a Program Representation.}
-The solution that we found to the range analysis problem in Figure~\ref{fig:ex1}
+\noindent
+\textbf{Choosing a Program Representation.}
+The solution to the range analysis problem in Figure~\ref{fig:ex1}
 is imprecise because we did not take conditional tests into considerations.
@@
 
-We use a non-iterative algorithm to convert a program to e-SSA form.
+Given that the exact rules to convert a program to e-SSA form have never been
+explicitly stated in the literature, we describe our rules as follows.
 Let $(v < c)?$ be a conditional test, and let $l_t$ and $l_f$ be labels in
@@
 and $l_f$ is the target when the condition is false.
-If $l_f$ dominates any use of $v$, then we insert at $l_f$ a copy
-$v_f = v \sqcap [-\infty, c-1]$, where $v_f$ is a fresh name.
+We split the live range of $v$ at any of these points if at least one of two
+conditions is true:
+(i) $l_f$ or $l_t$ dominate any use of $v$;
+(ii) there exist a use of $v$ at the dominance frontier of $l_f$ or $l_t$.
+For the notions of dominance and dominance-frontier, see Aho
+{\em et al.}~\cite{Aho06}.
+To split the live range of $v$ at $l_f$ we insert at this
+program point a copy $v_f = v \sqcap [c, +\infty]$, where $v_f$ is a fresh name.
 We then rename every use of $v$ that is dominated by $l_f$ to $v_f$.
-Dually, if $l_t$ dominates any use of $v$, then we insert at $l_t$ a copy
-$v_t = v \sqcap [c, +\infty]$, and rename variables accordingly.
-If the test uses two variables, e.g., $(v_1 < v_2)?$, then we create
+Dually, if we must split at $l_t$, then we create at this point a copy
+$v_t = v \sqcap [-\infty, c-1]$, and rename variables accordingly.
+If the conditional uses two variables, e.g., $(v_1 < v_2)?$, then we create
 intersections bound to {\em futures}.
 We insert, at $l_f$, $v_{1f} = v_1 \sqcap [\fun{ft}(v_2), +\infty]$,
-and $v_{2f} = v_2 \sqcap [-\infty, \fun{ft}(v_1) - 1]$.
-Similarly, at $l_v$ we insert
+and $v_{2f} = v_2 \sqcap [-\infty, \fun{ft}(v_1)]$.
+Similarly, at $l_t$ we insert
 $v_{1v} = v_1 \sqcap [-\infty, \fun{ft}(v_2) - 1]$
-and $v_{2f} = v_2 \sqcap [\fun{ft}(v_1), +\infty]$.
+and $v_{2f} = v_2 \sqcap [\fun{ft}(v_1) + 1, +\infty]$.
+
 We use the notation $\fun{ft}(v)$ to denote the {\em future} bounds of a
@@
 The e-SSA form allows us to bind interval information directly to the live
-ranges of variables; thus, allowing us to solve range analysis sparsely.
+ranges of variables; thus, giving us the opportunity to solve range analysis
+sparsely.
 More traditional approaches, which we call {\em dense analyses}, bind
 interval information to pairs formed by variables and program points.
+
+% TODO: check with Victor ** eSSA description could also say that we create sigmas even if lt/lf does not dominate any use of v (under some conditions), i.e.
+when there is a use in dominance frontier inside a phi function **
 
@@
 
-\paragraph{Extracting Constraints.}
+\noindent
+\textbf{Extracting Constraints.}
 Our implementation handles 18 different assembly instructions.
-The constraints in Figure~\ref{fig:eval_function} show some meaningful
-examples.
-These instructions include, for instance, the multiplicative operators
-\texttt{mul}, \texttt{div} and \texttt{modulus},
+The constraints in Figure~\ref{fig:eval_function} show only a few examples.
+Instructions that we did not show include, for instance, the multiplicative
+operators \texttt{div} and \texttt{modulus},
 the bitwise operators \texttt{and}, \texttt{or}, \texttt{xor} and \texttt{neg},
-the different types of shifts, the additive operators \texttt{add} and
-\texttt{sub}, and the logical operators \texttt{andalso}, \texttt{orelse} and
-\texttt{not}.
-We also handle different types of type conversion, e.g., converting 8-bits
-characters to 32-bits integers and vice-versa.
-Our constraint system deals with the $\phi$-functions from the SSA-form
-representation.
-Finally, we handle conditionals via intersection constraints created
-during the live range splitting of variables.
+the different types of shifts, and the
+logical operators \texttt{andalso}, \texttt{orelse} and \texttt{not}.
+Most of these instructions are sign-agnostic; that is, given that numbers are
+internally represented in 2's complement, the same implementation of a
+constraint handles positive and negative numbers.
+However, there are instructions that require different constraints, depending on
+the input being signed or not.
+Examples include \texttt{modulus} and \texttt{div}.
+We also handle different kinds of type conversion, e.g., converting 8-bit
+characters to 32-bit integers and vice-versa.
+In addition to constraints that represent actual assembly instructions, we have
+constraints to represent $\phi$-functions, and intersections, as seen in
+Figure~\ref{fig:eval_function}.
 The widening analysis that we will introduce in Section~\ref{sub:micro}
@@
 this monotonicity.
-However, these non-monotonic instructions have conservative approximations.
-%TODO: fix this discussion.
+However, these non-monotonic instructions have conservative
+approximations~\cite{Warren02}.
 
@@
 
-\paragraph{The Constraint Graph.}
+\noindent
+\textbf{The Constraint Graph.}
 The main data structure that we use to solve the range analysis problem is
@@
 
-\paragraph{Solving Range Analysis on Strong Components}
-Having built the dependence graph, we proceed to solve the range analysis
+\noindent
+\textbf{Solving Range Analysis on Strong Components.}
+We solve the range analysis
 problem by finding all the strongly connected components (SCCs) of the
@@
 We then sort the resulting DAG topologically, and apply the analyses from
-Section~\ref{sub:micro} on every SSC in topological order.
+Section~\ref{sub:micro} on every SCC in topological order.
 Once we solve the range analysis problem for a SCC, we propagate the
@@
 (Left) The lattice of the growth analysis.
-(Right) Cousot and Cousot's widening operator. We evaluate the rules from top 
-to bottom, and stop upon finding a pattern matching.}
+(Right) Cousot and Cousot's widening operator. We evaluate the rules from
+left-to-right, top-to-bottom, and stop upon finding a pattern matching.}
 \end{figure}
@@
 Had we considered only data dependence edges, then it would be possible
-that $V$ be analyzed before $V'$.
+that $V'$ be analyzed before $V"$.
 However, because of control dependence edges, this case cannot happen.
@@
 commented C++ code, and our e-SSA conversion module has 672 lines.
-We have analyzed the 428 C programs in the LLVM test suite, plus the integer
-benchmarks in SPEC CPU 2006.
+We have analyzed 428 C programs that constitute the LLVM test suite plus the
+integer benchmarks in SPEC CPU 2006.
 Together, these programs contain 4.76 million assembly instructions.
@@
 singletons.
-For instance, 99.11\% of the SSCs in gcc (\texttt{403.gcc}) have only
+For instance, 99.11\% of the SCCs in gcc (\texttt{403.gcc}) have only
 one node.
-Hence, the most complex phases of our algorithm are confined to a small fraction
-of the constraint graphs.
+Hence, the most complex phases of our algorithm are confined to a small part
+of the constraint graph.
 Second, the strongly connected components are not large -- our largest
@@
 analysis allows us to save per program variables.
+It is very important to notice that we do not consider the constants in the
+program text in our statics of precision.
+In other words, we only measure bitwidth reduction in variables that a constant
+propagation step could not remove.
 Our results for the SPEC programs were disappointingly: on the average, the
@@
 However, with numerical benchmarks we fare much better.
-On the average our inter procedural algorithm reduces the bitwidth of the
+On the average our inter-procedural algorithm reduces the bitwidth of the
 Stanford benchmarks by 36.24\%.

### 2012-03-22 doc/Manuscript/paper.tex
Added small changes to the paper, to keep it under 15 pages.

@@
 
-\author{Victor Sperle Campos, Igor Rafael, Douglas do Couto Teixeira,
-Fernando Magno Quint\~{a}o Pereira}
+\author{Victor Sperle Campos, Igor Rafael de Assis Costa, Douglas do Couto
+Teixeira, Fernando Magno Quint\~{a}o Pereira}
 
@@
 We have implemented our technique in LLVM and have been able to process
-programs totaling almost 2 million lines of C in a few seconds.
+programs totaling 2.72 million lines of C in a few seconds.
 We do global program analysis, and achieve a simple form of context
@@
 time~\cite{Su04,Su05}.
-However, although we could use their formula to handle a subset of C-like
+However, although we could use their formulation to handle a subset of C-like
 constructs, their description of how to deal with loops was
@@
 We have implemented our algorithm in the LLVM compiler~\cite{Lattner04}, and
-have used it to analyze a test suite with almost 2 million lines of C code.
+have used it to analyze a test suite with 2.72 million lines of C code.
 In Section~\ref{sec:exp} we provide empirical evidence showing that our
@@
 It is also precise: our results are similar to Stephenson
-{\em et al}'s~\cite{Stephenson00}, even though, contrary to them, we do not
-assume that the program is correct.
+{\em et al}'s~\cite{Stephenson00}, even though out analysis does not require
+a backward propagation phase.
 Furthermore, we have been able to find tight bounds to the majority of the
@@
 \caption{\label{fig:eval_function}
-The constraint system that produces instances of the range analysis problem.}
+A suite of constraints that produce an instance of the range analysis problem.}
 \end{small}
@@
 For the notions of dominance and dominance-frontier, see Aho
-{\em et al.}~\cite{Aho06}.
+{\em et al.}~\cite[p.656]{Aho06}.
 To split the live range of $v$ at $l_f$ we insert at this
@@
 Figure~\ref{fig:eval_function}.
-The widening analysis that we will introduce in Section~\ref{sub:micro}
+The growth analysis that we will introduce in Section~\ref{sub:micro}
 require monotonic transfer functions.
@@
 \noindent
-\textbf{Solving Range Analysis on Strong Components.}
-We solve the range analysis
-problem by finding all the strongly connected components (SCCs) of the
-dependence graph, and reducing this graph to a directed acyclic graph.
-We perform this last step by collapsing every SCC into a single node.
+\textbf{Strongly Connected Components.}
+To solve range analysis we find all the strongly connected components (SCCs) of
+the dependence graph and collapse them into single nodes; thus, obtaining a
+directed acyclic graph.
 We then sort the resulting DAG topologically, and apply the analyses from
@@
 (ii) there exists a variable node $N_v' \in S$, and a constraint node $N_c$,
-such that $N_v' \leftarrow N_c$, and $N_c \leftarrow N_v$.
-This propagation gives us a very modular algorithm.
-It ensures that when analyzing a strongly connected component $S$ any influence
-that $S$ might suffer from nodes outside it has been already
-taken into consideration.
+such that $N_v \leftarrow N_c$, and $N_c \leftarrow N_v'$.
+This propagation ensures that when analyzing a strongly connected component $S$
+any influence that $S$ might suffer from nodes outside it has already been
+considered.
 
@@
 \textbf{Growth Analysis.}
-The first step of our algorithm consists in determining the growth behavior of
-the interval bound to each constraint variable.
+The first step of our algorithm consists in determining how the interval bound
+to each variable grows.
 We ensure termination of this phase via Cousot and Cousot's widening
@@
 (iv) grows in both directions.
-The lattice of abstract states, plus a constraint system that represents our
-widening analysis is given in Figure~\ref{fig:growth_analysis}.
+The lattice of abstract states, plus a constraint system representing our
+growth analysis is given in Figure~\ref{fig:growth_analysis}.
 Because the lattice has height three, the intervals bound to each variable can
@@
 (Right) Cousot and Cousot's widening operator. We evaluate the rules from
-left-to-right, top-to-bottom, and stop upon finding a pattern matching.}
+left-to-right, top-to-bottom, and stop upon finding a pattern matching.
+Again: given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
+$\ub{\iota} = u$}
 \end{figure}
@@
 \textbf{Fixing futures.}
-The ranges found by the widening analysis tells us which variables have fixed
+The ranges found by the growth analysis tells us which variables have fixed
 bounds, independent on the intersections in the constraint system.
@@
 In order to correctly replace a future $\fun{ft}(V)$ that limits a variable
-$V'$, we need to have already applied the widening analysis onto $V$.
+$V'$, we need to have already applied the growth analysis onto $V$.
 Had we considered only data dependence edges, then it would be possible
-that $V'$ be analyzed before $V"$.
+that $V'$ be analyzed before $V$.
 However, because of control dependence edges, this case cannot happen.
@@
 \textbf{Narrowing Analysis.}
-The widening analysis associates very conservative bounds to each variable.
+The growth analysis associates very conservative bounds to each variable.
 Thus, the last step of our algorithm consists in narrowing these intervals.
-We accomplish this step via Cousot and Cousot's narrowing
+We accomplish this step via Cousot and Cousot's classic narrowing
 operator~\cite[248]{Cousot77}, which we show in
@@
 
-Figure~\ref{fig:ex_standard_eSSA}(b) shows the final solution that we find
-for this example.
+Figure~\ref{fig:ex_standard_eSSA}(b) shows our final solution for this example.
 This solution is very precise, in the sense that it is the maximum fixed
@@
 In any actual run of the program, $i$ could be at most 50.
-There exist static methods that are able to infer this tighter bound, such as
-Lakhdar-Chaouch's analysis on the relational domain~\cite{Lakhdar11}.
-However, analyses like this are much more expensive than ours, and seem, so
-far, unable to scale up to very large code bodies.
+Analysis on relational lattices, such as the polyhedron~\cite{Cousot78} or the
+Octagon~\cite{Mine06} domains, can infer such tighter bounds, as shown by
+Lakhdar-Chaouch {\em et al.}~\cite{Lakhdar11}.
+However, analyses on these higher dimensional domains are much more
+computationally expensive than analyses on the interval domain~\cite{Oh12}.
 
@@
 \caption{\label{fig:ex_partition_grow_crop}
-Four snapshots of the last SCC of Figure~\ref{fig:ex_partition_grow_crop}.
-(a) After removing control dependence edges, immediately before applying
-Cousot and Cousot's operators.
-(b) After running the widening analysis.
-(c) After fixing the intersections.
+Four snapshots of the last SCC of Figure~\ref{fig:ex_standard_eSSA}.
+(a) After removing control dependence edges.
+(b) After running the growth analysis.
+(c) After fixing the intersections bound to futures.
 (d) After running the narrowing analysis.}
@@
 time algorithms to solve constraints on the interval
-lattice~\cite{Gawlitza09,Costan05,Lakhdar11,Su04,Su05}.
+lattice~\cite{Costan05,Gawlitza09,Lakhdar11,Su04,Su05}.
 However, these works are still very theoretical, and have not yet been used to
@@
 Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
-about 25 minutes to go over a program with almost 900 basic blocks, which we
-analyze in less than one second.
+about 25 minutes to go over a program with almost 900 basic blocks.
+We analyze programs of similar size in seconds.
 We do not claim our approach is as precise as such algorithms, even though we
@@
 There have been many practical approaches to abstract interpretation,
-with special emphasis on range analysis~\cite{Blanchet03,Jung05,Cousot09}.
+with special emphasis on range analysis~\cite{Blanchet03,Cousot09,Jung05}.
 Cousot's group, for instance, has been able to globally analyze programs with
@@
 associated with the interval $[l, u]$ along its entire live range.
-We have used a dynamic profiler, available in our webpage, to empirically
-validate our results.
+Additionally, we have used a dynamic profiler, available in our webpage, to
+empirically validate our results.
 

### 2012-03-22 src/RAInstrumentation/RAInstrumentation.cpp

@@ -1,0 +1,183 @@
+//===- RAInstrumentation.cpp - Example code from "Writing an LLVM Pass" ---===//
+//
+// This file implements the LLVM "Range Analysis Instrumentation" pass 
+//
+//===----------------------------------------------------------------------===//
+#include "RAInstrumentation.h"
+
+using namespace llvm;
+
+STATISTIC(numInstructions, "Number of instructions instrumented");
+
+static RegisterPass<RAInstrumentation> X("ra-instrumentation", "Range Analysis Instrumentation Pass");
+
+bool RAInstrumentation::isValidInst(Instruction *I)
+{
+	// Only i32 instructions are valid
+	// Exceptions: invoke instructions
+	bool conditions = !I->getType()->isIntegerTy(BIT_WIDTH) || isa<InvokeInst>(I);
+	
+	return !conditions;
+}
+
+
+void RAInstrumentation::PrintInstructionIdentifier(std::string M, std::string F, const Value *V){
+
+    
+	fprintf(file, "%s.%s.%s %d\n", M.c_str(), F.c_str(), V->getName().str().c_str(), (int)V);
+
+}
+
+
+void RAInstrumentation::InstrumentMainFunction(Function* F) {
+
+    //Put a call to printHash function before every return instruction.
+    
+    
+	// Iterate through basic blocks
+	for (Function::iterator BBit = F->begin(); BBit != F->end(); ++BBit) {
+
+    	// Iterate through instructions
+    	for (BasicBlock::iterator Iit = BBit->begin(); Iit != BBit->end(); ++Iit) {     		
+            if (isa<ReturnInst>(Iit)){
+                Function& printHash = GetPrintHashFunction();                         
+                CallInst* callPrintHash = CallInst::Create(&printHash, "", Iit);  
+                MarkAsNotOriginal(*callPrintHash);
+            }
+        }
+    };
+}
+
+
+bool RAInstrumentation::runOnModule(Module &M) {
+
+	this->module = &M;
+	this->context = &M.getContext();
+    std::string ErrorInfo;
+
+    file = fopen("RAHashNames.txt","w");
+    if (file==NULL) {
+        errs() << "Error opening file RAHashNames.txt. Range Analysis Instrumentation Pass aborted.\n";
+        return false;
+    }
+    
+    // No main, no instrumentation!
+    Function *Main = M.getFunction("main");
+    
+    // Using fortran? ... this kind of works
+    if (!Main)
[…]

### 2012-03-22 src/RAInstrumentation/RAInstrumentationHash.c

@@ -1,0 +1,192 @@
+#include <stdio.h>
+#include <limits.h>
+#include <stdlib.h>
+#include <string.h>
+
+struct List
+{
+    void* ptr;
+    int minValue;
+    int maxValue;
+    struct List* next;
+};
+
+typedef struct List List;
+
+typedef struct 
+{
+    void* ptr;
+    int minValue;
+    int maxValue;
+    List* list;
+} Hashitem;
+
+#define RA_HASH_SIZE 65536
+
+Hashitem hash[RA_HASH_SIZE];
+
+void initHash()
+{
+    memset(hash, 0, sizeof(hash));
+}
+
+void init()
+{
+    initHash();
+}
+
+List* getFromList(List* list, void* ptr)
+{
+    while (list != 0){
+        if (list->ptr == ptr) return list;
+        list = list->next;
+    }
+
+    return 0;
+}
+
+List* setCurrentMinInList(List* list, void* ptr, int minValue)
+{
+    List *tmp;
+    tmp = getFromList(list, ptr);
+    if (tmp != 0){
+        tmp->minValue = minValue;
+        return list;
+    } else {
+        tmp = malloc(sizeof(List));
+        tmp->minValue = minValue;
+        tmp->maxValue = minValue;  
+        tmp->next = list;
+        return tmp;
+    }
+}
+
+
+List* setCurrentMaxInList(List* list, void* ptr, int maxValue)
+{
+    List *tmp;
+    tmp = getFromList(list, ptr);
+    if (tmp != 0){
+        tmp->maxValue = maxValue;
+        return list;
+    } else {
+        tmp = malloc(sizeof(List));
+        tmp->minValue = maxValue;
+        tmp->maxValue = maxValue;  
+        tmp->next = list;
+        return tmp;
+    }
+}
+
+
+int getCurrentMin(void* ptr)
+{
+    List *tmp;
+    int key = (int) ptr & 0xffff;
+    if (hash[key].ptr == ptr)
+    {
+        return hash[key].minValue;
+    }
+    else
+    {
+        tmp = getFromList(hash[key].list, ptr);
+        if (tmp != 0){
+            return tmp->minValue;
+        } else {
+            return INT_MAX;
+        }
+    }
+}
+
+int getCurrentMax(void* ptr)
+{
+    List *tmp;
+    int key = (int) ptr & 0xffff;
+    if (hash[key].ptr == ptr)
+    {
+        return hash[key].maxValue;
+    }
+    else
+    {
[…]

### 2012-03-22 src/RAInstrumentation/RAInstrumentation.h

@@ -1,0 +1,66 @@
+/*
+ * RAInstrumentation.h
+ *
+ *  Created on: 22/03/2012
+ *      Author: raphael
+ */
+
+#ifndef RAINSTRUMENTATION_H_
+#define RAINSTRUMENTATION_H_
+
+#define DEBUG_TYPE "range-analysis-printer"
+
+#include "llvm/Constants.h"
+#include "llvm/DerivedTypes.h"
+#include "llvm/Function.h"
+#include "llvm/InstrTypes.h"
+#include "llvm/Instructions.h"
+#include "llvm/LLVMContext.h"
+#include "llvm/Module.h"
+#include "llvm/Pass.h"
+#include "llvm/ADT/Statistic.h"
+#include "llvm/Support/raw_ostream.h"
+#include <vector>
+#include <list>
+#include <stdio.h>
+#include "RAInstrumentation.h"
+
+#define BIT_WIDTH 32
+
+using namespace llvm;
+
+
+namespace {
+
+	struct RAInstrumentation : public ModulePass {
+		static char ID;
+		RAInstrumentation() : ModulePass(ID) {};
+
+        void printValueInfo(const Value *V);
+		void MarkAsNotOriginal(Instruction& inst);
+        void PrintInstructionIdentifier(std::string M, std::string F, const Value *V);
+
+        bool IsNotOriginal(Instruction& inst);
+        bool isValidInst(Instruction *I);
+        virtual bool runOnModule(Module &M);
+
+        Function& GetSetCurrentMinMaxFunction();
+        Function& GetPrintHashFunction();
+        Instruction* GetNextInstruction(Instruction& i);
+
+        void InstrumentMainFunction(Function* F);
+
+        Module* module;
+        LLVMContext* context;
+		FILE* file;
+
+	};
+}
+
+char RAInstrumentation::ID = 0;
+
+
+
+
+
+#endif /* RAINSTRUMENTATION_H_ */

### 2012-03-22 src/RAInstrumentation/RAPrinter.cpp

@@ -1,0 +1,73 @@
+#include "llvm/Function.h"
+#include "llvm/Pass.h"
+#include "llvm/User.h"
+#include "llvm/PassAnalysisSupport.h"
+#include "llvm/Support/raw_ostream.h"
+#include "../RangeAnalysis/RangeAnalysis.h"
+#include "RAInstrumentation.h"
+
+using namespace llvm;
+
+class RAPrinter: public llvm::ModulePass {
+public:
+        static char ID;
+        RAPrinter() : ModulePass(ID){ }
+        virtual ~RAPrinter() { }
+        virtual bool runOnModule(Module &M){
+                IntraProceduralRA<Cousot> &ra = getAnalysis<IntraProceduralRA<Cousot> >();
+
+                std::string Filename = "RAEstimatedValues.txt";
+
+                std::string ErrorInfo;
+                raw_fd_ostream File(Filename.c_str(), ErrorInfo);
+
+                if (!ErrorInfo.empty()){
+                  errs() << "Error opening file RAEstimatedValues.txt for writing! \n";
+                  return false;
+                }
+
+
+            	// Iterate through functions
+            	for (Module::iterator Fit = M.begin(), Fend = M.end(); Fit != Fend; ++Fit) {
+
+            		// If the function is empty, there is nothing to do...
+            		if (Fit->begin() == Fit->end())
+            			continue;
+
+            		// Iterate through basic blocks
+            		for (Function::iterator BBit = Fit->begin(), BBend = Fit->end(); BBit != BBend; ++BBit) {
+
+            			// Iterate through instructions
+            			for (BasicBlock::iterator Iit = BBit->begin(); Iit != BBit->end(); ++Iit) {
+
+            				llvm::Instruction* I = cast<Instruction>(Iit);
+            				if ( RAInstrumentation::isValidInst(I) ){
+
+                                const Value *v = &(*Iit);
+                                Range r = ra.getRange(v);
+
+                                //Prints the variable and the range to the output file
+                                File << M.getModuleIdentifier()
+                                     << "." << Fit->getName().str()
+                                     << "." << cast<Value>(Iit)->getName()
+                                     << " " r.getLower()
+                                     << " " r.getUpper() << "\n";
+
+            				}
+            			}
+            		}
+            	}
+
+
+
+                return false;
+        }
+
+        virtual void getAnalysisUsage(AnalysisUsage &AU) const {
[…]

### 2012-03-22 src/Instrumentation/RAInstrumentation.cpp

@@ -1,229 +1,0 @@
-//===- RAInstrumentation.cpp - Example code from "Writing an LLVM Pass" ---===//
-//
-// This file implements the LLVM "Range Analysis Instrumentation" pass 
-//
-//===----------------------------------------------------------------------===//
-
-#define DEBUG_TYPE "hello"
-
-#include "llvm/Constants.h"
-#include "llvm/DerivedTypes.h"
-#include "llvm/Function.h"
-#include "llvm/InstrTypes.h"
-#include "llvm/Instructions.h"
-#include "llvm/LLVMContext.h"
-#include "llvm/Module.h"
-#include "llvm/Pass.h"
-#include "llvm/ADT/Statistic.h"
-#include "llvm/Support/raw_ostream.h"
-#include <vector>
-#include <list>
-#include <stdio.h>
-
-STATISTIC(numInstructions, "Number of instructions instrumented");
-
-using namespace llvm;
-
-const unsigned BIT_WIDTH = 32;
-
-namespace {
-
-	struct RAInstrumentation : public ModulePass {
-		static char ID; 
-		RAInstrumentation() : ModulePass(ID) {};
-		
-        void printValueInfo(const Value *V);
-		void MarkAsNotOriginal(Instruction& inst);
-        void PrintInstructionIdentifier(std::string M, std::string F, const Value *V);
-        
-        bool IsNotOriginal(Instruction& inst);       
-        bool isValidInst(Instruction *I);
-        virtual bool runOnModule(Module &M);
-        
-        Function& GetSetCurrentMinMaxFunction();
-        Function& GetPrintHashFunction();
-        Instruction* GetNextInstruction(Instruction& i);
-        
-        void InstrumentMainFunction(Function* F);
-        
-        Module* module;
-        LLVMContext* context;
-		FILE* file;
-
-	};
-}
-
-char RAInstrumentation::ID = 0;
-static RegisterPass<RAInstrumentation> X("ra-instrumentation", "Range Analysis Instrumentation Pass");
-
-
-bool RAInstrumentation::isValidInst(Instruction *I)
-{
-	// Only i32 instructions are valid
-	// Exceptions: invoke instructions
-	bool conditions = !I->getType()->isIntegerTy(BIT_WIDTH) || isa<InvokeInst>(I);
-	
-	return !conditions;
-}
-
-
[…]

### 2012-03-22 src/Instrumentation/RAInstrumentationHash.c

@@ -1,192 +1,0 @@
-#include <stdio.h>
-#include <limits.h>
-#include <stdlib.h>
-#include <string.h>
-
-struct List
-{
-    void* ptr;
-    int minValue;
-    int maxValue;
-    struct List* next;
-};
-
-typedef struct List List;
-
-typedef struct 
-{
-    void* ptr;
-    int minValue;
-    int maxValue;
-    List* list;
-} Hashitem;
-
-#define RA_HASH_SIZE 65536
-
-Hashitem hash[RA_HASH_SIZE];
-
-void initHash()
-{
-    memset(hash, 0, sizeof(hash));
-}
-
-void init()
-{
-    initHash();
-}
-
-List* getFromList(List* list, void* ptr)
-{
-    while (list != 0){
-        if (list->ptr == ptr) return list;
-        list = list->next;
-    }
-
-    return 0;
-}
-
-List* setCurrentMinInList(List* list, void* ptr, int minValue)
-{
-    List *tmp;
-    tmp = getFromList(list, ptr);
-    if (tmp != 0){
-        tmp->minValue = minValue;
-        return list;
-    } else {
-        tmp = malloc(sizeof(List));
-        tmp->minValue = minValue;
-        tmp->maxValue = minValue;  
-        tmp->next = list;
-        return tmp;
-    }
-}
-
-
-List* setCurrentMaxInList(List* list, void* ptr, int maxValue)
-{
-    List *tmp;
-    tmp = getFromList(list, ptr);
-    if (tmp != 0){
-        tmp->maxValue = maxValue;
-        return list;
-    } else {
-        tmp = malloc(sizeof(List));
-        tmp->minValue = maxValue;
-        tmp->maxValue = maxValue;  
-        tmp->next = list;
-        return tmp;
-    }
-}
-
-
-int getCurrentMin(void* ptr)
-{
-    List *tmp;
-    int key = (int) ptr & 0xffff;
-    if (hash[key].ptr == ptr)
-    {
-        return hash[key].minValue;
-    }
-    else
-    {
-        tmp = getFromList(hash[key].list, ptr);
-        if (tmp != 0){
-            return tmp->minValue;
-        } else {
-            return INT_MAX;
-        }
-    }
-}
-
-int getCurrentMax(void* ptr)
-{
-    List *tmp;
-    int key = (int) ptr & 0xffff;
-    if (hash[key].ptr == ptr)
-    {
-        return hash[key].maxValue;
-    }
-    else
-    {
[…]

### 2012-03-22 src/RangeAnalysis/RangeAnalysis.cpp

@@ -2769,7 +2769,7 @@ void ConstraintGraph::print(const Function& F, raw_ostream& OS) const {
 	// Print the header of the .dot file.
 	OS << "digraph dotgraph {\n";
 	OS << "label=\"Constraint Graph for \'";
-	OS << F.getNameStr() << "\' function\";\n";
+	OS << F.getName() << "\' function\";\n";
 	OS << "node [shape=record,fontname=\"Times-Roman\",fontsize=14];\n";
 
 	// Print the body of the .dot file.

### 2012-03-22 src/RAInstrumentation/RAInstrumentation.cpp

@@ -7,6 +7,8 @@
 
 using namespace llvm;
 
+#define DEBUG_TYPE "range-analysis-printer"
+
 STATISTIC(numInstructions, "Number of instructions instrumented");
 
 static RegisterPass<RAInstrumentation> X("ra-instrumentation", "Range Analysis Instrumentation Pass");

### 2012-03-22 src/RAInstrumentation/RAInstrumentation.h

@@ -8,8 +8,6 @@
 #ifndef RAINSTRUMENTATION_H_
 #define RAINSTRUMENTATION_H_
 
-#define DEBUG_TYPE "range-analysis-printer"
-
 #include "llvm/Constants.h"
 #include "llvm/DerivedTypes.h"
 #include "llvm/Function.h"
@@ -41,7 +39,7 @@ namespace {
         void PrintInstructionIdentifier(std::string M, std::string F, const Value *V);
 
         bool IsNotOriginal(Instruction& inst);
-        bool isValidInst(Instruction *I);
+        static bool isValidInst(Instruction *I);
         virtual bool runOnModule(Module &M);
 
         Function& GetSetCurrentMinMaxFunction();

### 2012-03-22 src/RAInstrumentation/RAPrinter.cpp

@@ -1,73 +1,0 @@
-#include "llvm/Function.h"
-#include "llvm/Pass.h"
-#include "llvm/User.h"
-#include "llvm/PassAnalysisSupport.h"
-#include "llvm/Support/raw_ostream.h"
-#include "../RangeAnalysis/RangeAnalysis.h"
-#include "RAInstrumentation.h"
-
-using namespace llvm;
-
-class RAPrinter: public llvm::ModulePass {
-public:
-        static char ID;
-        RAPrinter() : ModulePass(ID){ }
-        virtual ~RAPrinter() { }
-        virtual bool runOnModule(Module &M){
-                IntraProceduralRA<Cousot> &ra = getAnalysis<IntraProceduralRA<Cousot> >();
-
-                std::string Filename = "RAEstimatedValues.txt";
-
-                std::string ErrorInfo;
-                raw_fd_ostream File(Filename.c_str(), ErrorInfo);
-
-                if (!ErrorInfo.empty()){
-                  errs() << "Error opening file RAEstimatedValues.txt for writing! \n";
-                  return false;
-                }
-
-
-            	// Iterate through functions
-            	for (Module::iterator Fit = M.begin(), Fend = M.end(); Fit != Fend; ++Fit) {
-
-            		// If the function is empty, there is nothing to do...
-            		if (Fit->begin() == Fit->end())
-            			continue;
-
-            		// Iterate through basic blocks
-            		for (Function::iterator BBit = Fit->begin(), BBend = Fit->end(); BBit != BBend; ++BBit) {
-
-            			// Iterate through instructions
-            			for (BasicBlock::iterator Iit = BBit->begin(); Iit != BBit->end(); ++Iit) {
-
-            				llvm::Instruction* I = cast<Instruction>(Iit);
-            				if ( RAInstrumentation::isValidInst(I) ){
-
-                                const Value *v = &(*Iit);
-                                Range r = ra.getRange(v);
-
-                                //Prints the variable and the range to the output file
-                                File << M.getModuleIdentifier()
-                                     << "." << Fit->getName().str()
-                                     << "." << cast<Value>(Iit)->getName()
-                                     << " " r.getLower()
-                                     << " " r.getUpper() << "\n";
-
-            				}
-            			}
-            		}
-            	}
-
-
-
-                return false;
-        }
-
-        virtual void getAnalysisUsage(AnalysisUsage &AU) const {
[…]

### 2012-03-23 src/RAPrinter/RAPrinter.cpp

@@ -1,0 +1,82 @@
+#include "llvm/Function.h"
+#include "llvm/Pass.h"
+#include "llvm/User.h"
+#include "llvm/PassAnalysisSupport.h"
+#include "llvm/Support/raw_ostream.h"
+#include "../RangeAnalysis/RangeAnalysis.h"
+#include "../RAInstrumentation/RAInstrumentation.h"
+
+using namespace llvm;
+
+class RAPrinter: public llvm::ModulePass {
+public:
+        static char ID;
+        RAPrinter() : ModulePass(ID){ }
+        virtual ~RAPrinter() { }
+        virtual bool runOnModule(Module &M){
+
+        		InterProceduralRA<Cousot> &ra = getAnalysis<InterProceduralRA<Cousot> >();
+
+                std::string Filename = "RAEstimatedValues.txt";
+
+                std::string ErrorInfo;
+                raw_fd_ostream File(Filename.c_str(), ErrorInfo);
+
+                if (!ErrorInfo.empty()){
+                  errs() << "Error opening file RAEstimatedValues.txt for writing! \n";
+                  return false;
+                }
+
+            	// Iterate through functions
+            	for (Module::iterator Fit = M.begin(), Fend = M.end(); Fit != Fend; ++Fit) {
+
+            		// If the function is empty, there is nothing to do...
+            		if (Fit->begin() == Fit->end())
+            			continue;
+
+            		// Iterate through basic blocks
+            		for (Function::iterator BBit = Fit->begin(), BBend = Fit->end(); BBit != BBend; ++BBit) {
+
+            			// Iterate through instructions
+            			for (BasicBlock::iterator Iit = BBit->begin(); Iit != BBit->end(); ++Iit) {
+
+            				llvm::Instruction* I = cast<Instruction>(Iit);
+            				if ( isValidInst(I) ){
+
+                                const Value *v = &(*Iit);
+                                Range r = ra.getRange(v);
+
+                                //Prints the variable and the range to the output file
+                                File << M.getModuleIdentifier()
+                                     << "." << Fit->getName().str()
+                                     << "." << cast<Value>(Iit)->getName()
+                                     << " " << r.getLower()
+                                     << " " << r.getUpper() << "\n";
+
+            				}
+            			}
+            		}
+            	}
+
+
+
+                return false;
+        }
+
[…]

### 2012-03-23 doc/Manuscript/paper.tex
Added last changes to the paper.

@@
 
-\author{Victor Sperle Campos, Igor Rafael de Assis Costa, Douglas do Couto
+\author{Victor Hugo Sperle Campos, Igor Rafael de Assis Costa, Douglas do Couto
 Teixeira, Fernando Magno Quint\~{a}o Pereira}
@@
 Assignment form as the intermediate representation.
-%We show that by processing the strongly connected components that constitute
-%the graph of dependences between variables in topological order we not only
-%gain in time, but also in precision.
 We claim to present in this paper the most extensive empirical evaluation of
@@
 loops~\cite{Lokuciejewski09}
-and even in the synthesis of hardware~\cite{Cong05,Mahlke01}.
+and even in the synthesis of hardware~\cite{Cong05,Lhairech10,Mahlke01}.
 
@@
 one of the most cited papers in computer science~\cite{Cousot77}.
-Different algorithms for range analysis have been subsequently proposed
-by Stephenson {\em et al.}~\cite{Stephenson00}, Patterson~\cite{Patterson95}
-and many other researchers.
+Different algorithms for range analysis have been later proposed
+by Patterson~\cite{Patterson95}, Stephenson {\em et al.}~\cite{Stephenson00},
+Mahlke {\em et al.}~\cite{Mahlke01} and many other researchers.
 Recently there have been many independent efforts to find exact, polynomial
@@
 There have been many practical approaches to abstract interpretation,
-with special emphasis on range analysis~\cite{Blanchet03,Cousot09,Jung05}.
+with special emphasis on range analysis~\cite{Bertrane10,Blanchet03,Cousot09,Jung05}.
 Cousot's group, for instance, has been able to globally analyze programs with

### 2012-03-24 src/RAPrinter/RAPrinter.cpp
Correcting a little bug.
Preparing file to receive other LLVM passes

@@ -8,75 +8,77 @@
 
 using namespace llvm;
 
-class RAPrinter: public llvm::ModulePass {
-public:
-        static char ID;
-        RAPrinter() : ModulePass(ID){ }
-        virtual ~RAPrinter() { }
-        virtual bool runOnModule(Module &M){
+bool isValidInst(Instruction *I)
+{
+	// Only i32 instructions are valid
+	// Exceptions: invoke instructions
+	bool conditions = !I->getType()->isIntegerTy(BIT_WIDTH) || isa<InvokeInst>(I);
 
-        		InterProceduralRA<Cousot> &ra = getAnalysis<InterProceduralRA<Cousot> >();
+	return !conditions;
+}
+namespace {
+	class RAPrinterInterProceduralCousot: public llvm::ModulePass {
+	public:
+			static char ID;
+			RAPrinterInterProceduralCousot() : ModulePass(ID){ }
+			virtual ~RAPrinterInterProceduralCousot() { }
+			virtual bool runOnModule(Module &M){
 
-                std::string Filename = "RAEstimatedValues.txt";
+					InterProceduralRA<Cousot> &ra = getAnalysis<InterProceduralRA<Cousot> >();
 
-                std::string ErrorInfo;
-                raw_fd_ostream File(Filename.c_str(), ErrorInfo);
+					std::string Filename = "RAEstimatedValues.txt";
 
-                if (!ErrorInfo.empty()){
-                  errs() << "Error opening file RAEstimatedValues.txt for writing! \n";
-                  return false;
-                }
+					std::string ErrorInfo;
+					raw_fd_ostream File(Filename.c_str(), ErrorInfo);
 
-            	// Iterate through functions
-            	for (Module::iterator Fit = M.begin(), Fend = M.end(); Fit != Fend; ++Fit) {
+					if (!ErrorInfo.empty()){
+					  errs() << "Error opening file RAEstimatedValues.txt for writing! \n";
+					  return false;
+					}
 
-            		// If the function is empty, there is nothing to do...
-            		if (Fit->begin() == Fit->end())
-            			continue;
+					// Iterate through functions
+					for (Module::iterator Fit = M.begin(), Fend = M.end(); Fit != Fend; ++Fit) {
 
-            		// Iterate through basic blocks
-            		for (Function::iterator BBit = Fit->begin(), BBend = Fit->end(); BBit != BBend; ++BBit) {
+						// If the function is empty, there is nothing to do...
[…]

### 2012-03-29 src/RangeAnalysis/RangeAnalysis.cpp
Continuing implementation of jump-set

@@ -15,7 +15,7 @@
 
 using namespace llvm;
 
-// These macros are used to get stats regarging the precision of our analysis.
+// These macros are used to get stats regarding the precision of our analysis.
 STATISTIC(usedBits, "Initial number of bits.");
 STATISTIC(needBits, "Needed bits.");
 STATISTIC(percentReduction, "Percentage of reduction of the number of bits.");
@@ -2276,33 +2276,120 @@ void ConstraintGraph::buildValueMaps(const Function& F) {
 	}
 }
 
-void ConstraintGraph::buildConstantVector(SmallPtrSet<VarNode*, 32> &component)
+/*
+ * Used to insert constant in the right position
+ */
+void ConstraintGraph::insertConstantIntoVector(APInt constantval)
 {
+	if (constantval.getBitWidth() < MAX_BIT_INT) {
+		constantval = constantval.sext(MAX_BIT_INT);
+	}
+
+	for (SmallVectorImpl<APInt>::iterator vit = constantvector.begin(), vend = constantvector.end(); vit != vend; ++vit) {
+		const APInt &vapint = *vit;
+
+		if (vapint.eq(constantval)) {
+			break;
+		}
+
+		if (constantval.slt(vapint)) {
+			constantvector.insert(vit, constantval);
+			break;
+		}
+
+		assert(constantval.ne(vapint) && "Erro no vetor de constantes para jump-set");
+	}
+}
+
+/*
+ * Get the first constant from vector greater than val
+ */
+APInt ConstraintGraph::getFirstGreaterFromVector(const APInt &val)
+{
+	for (SmallVectorImpl<APInt>::iterator vit = constantvector.begin(), vend = constantvector.end(); vit != vend; ++vit) {
+		const APInt &vapint = *vit;
+
+		if (vapint.sgt(val)) {
+			return vapint;
+		}
+	}
+
+	return val;
+}
+
+/*
+ * Create a vector containing all constants related to the component
+ * They include:
+ *   - Constants inside component
+ *   - Constants that are source of an edge to an entry point
+ *   - Constants from intersections generated by sigmas
+ */
+void ConstraintGraph::buildConstantVector(const SmallPtrSet<VarNode*, 32> &component, const UseMap &compusemap)
+{
+	// Remove all elements from the vector
 	constantvector.clear();
 	
[…]

### 2012-03-29 src/RangeAnalysis/RangeAnalysis.h

@@ -718,7 +718,7 @@ private:
 	ValuesBranchMap valuesBranchMap;
 	ValuesSwitchMap valuesSwitchMap;
 	
-	SmallVector<std::pair<const ConstantInt*, APInt>, 2 > constantvector;
+	SmallVector<APInt, 2> constantvector;
 	
 	/// Adds a BinaryOp in the graph.
 	void addBinaryOp(const Instruction* I);
@@ -732,7 +732,9 @@ private:
 	void buildValueSwitchMap(const SwitchInst *sw);
 	void buildValueMaps(const Function& F);
 	
-	void buildConstantVector(SmallPtrSet<VarNode*, 32> &component);
+	void insertConstantIntoVector(APInt constantval);
+	APInt getFirstGreaterFromVector(const APInt &val);
+	void buildConstantVector(const SmallPtrSet<VarNode*, 32> &component, const UseMap &compusemap);
 	// Perform the widening and narrowing operations
 
 protected:

### 2012-03-29 src/RangeAnalysis/RangeAnalysis.cpp
Finished jump-set optimization

@@ -2288,33 +2288,54 @@ void ConstraintGraph::insertConstantIntoVector(APInt constantval)
 	for (SmallVectorImpl<APInt>::iterator vit = constantvector.begin(), vend = constantvector.end(); vit != vend; ++vit) {
 		const APInt &vapint = *vit;
 
+		// If already in vector, get out of here
 		if (vapint.eq(constantval)) {
-			break;
+			return;
 		}
 
+		// If have found point of insertion, do it and return
 		if (constantval.slt(vapint)) {
 			constantvector.insert(vit, constantval);
-			break;
+			return;
 		}
 
 		assert(constantval.ne(vapint) && "Erro no vetor de constantes para jump-set");
 	}
+
+	// Insert at the end of the vector
+	constantvector.push_back(constantval);
 }
 
 /*
  * Get the first constant from vector greater than val
  */
-APInt ConstraintGraph::getFirstGreaterFromVector(const APInt &val)
+APInt getFirstGreaterFromVector(const SmallVector<APInt, 2> &constantvector, const APInt &val)
 {
-	for (SmallVectorImpl<APInt>::iterator vit = constantvector.begin(), vend = constantvector.end(); vit != vend; ++vit) {
+	for (SmallVectorImpl<APInt>::const_iterator vit = constantvector.begin(), vend = constantvector.end(); vit != vend; ++vit) {
 		const APInt &vapint = *vit;
 
-		if (vapint.sgt(val)) {
+		if (vapint.sge(val)) {
 			return vapint;
 		}
 	}
 
-	return val;
+	return Max;
+}
+
+/*
+ * Get the first constant from vector less than val
+ */
+APInt getFirstLessFromVector(const SmallVector<APInt, 2> &constantvector, const APInt &val)
+{
+	for (SmallVectorImpl<APInt>::const_reverse_iterator vit = constantvector.rbegin(), vend = constantvector.rend(); vit != vend; ++vit) {
+		const APInt &vapint = *vit;
+
+		if (vapint.sle(val)) {
+			return vapint;
+		}
+	}
+
+	return Min;
 }
 
 /*
@@ -2349,8 +2370,9 @@ void ConstraintGraph::buildConstantVector(const SmallPtrSet<VarNode*, 32> &compo
 			continue;
 		}
 
[…]

### 2012-03-29 src/RangeAnalysis/RangeAnalysis.h

@@ -443,6 +443,7 @@ public:
 	void addSource(const VarNode* newsrc);
 	// Return source identified by index
 	const VarNode *getSource(unsigned index) const {return sources[index];}
+	unsigned getNumSources() const {return sources.size();}
 	// Methods for RTTI
 	virtual OperationId getValueId() const {return PhiOpId;}
 	static bool classof(PhiOp const *) {
@@ -718,6 +719,8 @@ private:
 	ValuesBranchMap valuesBranchMap;
 	ValuesSwitchMap valuesSwitchMap;
 	
+	// Vector containing the constants from a SCC
+	// It is cleared at the beginning of every SCC resolution
 	SmallVector<APInt, 2> constantvector;
 	
 	/// Adds a BinaryOp in the graph.
@@ -733,13 +736,14 @@ private:
 	void buildValueMaps(const Function& F);
 	
 	void insertConstantIntoVector(APInt constantval);
-	APInt getFirstGreaterFromVector(const APInt &val);
+	APInt getFirstGreaterFromVector(const SmallVector<APInt, 2> &constantvector, const APInt &val);
+	APInt getFirstLessFromVector(const SmallVector<APInt, 2> &constantvector, const APInt &val);
 	void buildConstantVector(const SmallPtrSet<VarNode*, 32> &component, const UseMap &compusemap);
 	// Perform the widening and narrowing operations
 
 protected:
 	void update(const UseMap &compUseMap,
-		SmallPtrSet<const Value*, 6>& actv, bool (*meet)(BasicOp* op));
+		SmallPtrSet<const Value*, 6>& actv, bool (*meet)(BasicOp* op, const SmallVector<APInt, 2> *constantvector));
 	void update(unsigned nIterations, const UseMap &compUseMap,
 			SmallPtrSet<const Value*, 6>& actv);
 
@@ -839,12 +843,13 @@ public:
 };
 
 class Meet{
+
 public:
-	static bool widen(BasicOp* op);
-	static bool narrow(BasicOp* op);
-	static bool crop(BasicOp* op);
-	static bool growth(BasicOp* op);
-	static bool fixed(BasicOp* op);
+	static bool widen(BasicOp* op, const SmallVector<APInt, 2> *constantvector);
+	static bool narrow(BasicOp* op, const SmallVector<APInt, 2> *constantvector);
+	static bool crop(BasicOp* op, const SmallVector<APInt, 2> *constantvector);
[…]

### 2012-03-29 src/RangeAnalysis/RangeAnalysis.cpp
Improved algorithm to create constants vector

@@ -2277,6 +2277,14 @@ void ConstraintGraph::buildValueMaps(const Function& F) {
 }
 
 /*
+ * Comparison function used to sort the constant vector
+ */
+bool compareAPInt(const APInt &v1, const APInt &v2)
+{
+	return v1.slt(v2);
+}
+
+/*
  * Used to insert constant in the right position
  */
 void ConstraintGraph::insertConstantIntoVector(APInt constantval)
@@ -2285,24 +2293,6 @@ void ConstraintGraph::insertConstantIntoVector(APInt constantval)
 		constantval = constantval.sext(MAX_BIT_INT);
 	}
 
-	for (SmallVectorImpl<APInt>::iterator vit = constantvector.begin(), vend = constantvector.end(); vit != vend; ++vit) {
-		const APInt &vapint = *vit;
-
-		// If already in vector, get out of here
-		if (vapint.eq(constantval)) {
-			return;
-		}
-
-		// If have found point of insertion, do it and return
-		if (constantval.slt(vapint)) {
-			constantvector.insert(vit, constantval);
-			return;
-		}
-
-		assert(constantval.ne(vapint) && "Erro no vetor de constantes para jump-set");
-	}
-
-	// Insert at the end of the vector
 	constantvector.push_back(constantval);
 }
 
@@ -2428,6 +2418,10 @@ void ConstraintGraph::buildConstantVector(const SmallPtrSet<VarNode*, 32> &compo
 			}
 		}
 	}
+
+	// Sort vector in ascending order and remove duplicates
+	std::sort(constantvector.begin(), constantvector.end(), compareAPInt);
+	std::unique(constantvector.begin(), constantvector.end());
 }
 
 /// Iterates through all instructions in the function and builds the graph.
@@ -2505,10 +2499,6 @@ bool Meet::widen(BasicOp* op, const SmallVector<APInt, 2> *constantvector) {
 	// Jump-set
 	APInt nlconstant = getFirstLessFromVector(*constantvector, newLower);
 	APInt nuconstant = getFirstGreaterFromVector(*constantvector, newUpper);
-
-	// To disable jump-set, uncomment lines below
-	//nlconstant = Min;
-	//nuconstant = Max;
 
 	if (oldInterval.isUnknown()) {
 		op->getSink()->setRange(newInterval);
@@ -2764,7 +2754,7 @@ void ConstraintGraph::findIntervals() {
 #ifdef SCC_DEBUG
[…]

### 2012-03-29 src/RangeAnalysis/RangeAnalysis.h

@@ -52,6 +52,7 @@
 #include <stack>
 #include <set>
 #include <sstream>
+#include <algorithm>
 
 //TODO: comment the line below to disable the debug of SCCs and optimize the code
 // generated.

### 2012-03-29 doc/Manuscript/paper.tex
Changed the paper, and added the pdf to the svn.

@@
 Range propagation along strong components allows us, for example, to reduce
-41.5\% more bits in \texttt{401.bzip}.
+41.5\% more bits in \texttt{401.bzip2}.
 
@@
 analysis allows us to save per program variables.
-It is very important to notice that we do not consider the constants in the
-program text in our statics of precision.
+It is very important to notice that we do not consider constants in our statistics of precision.
 In other words, we only measure bitwidth reduction in variables that a constant
@@
 Our implementation of the range analysis algorithm described in this paper is
-publicly available at \texttt{http://google.code/range-analysis/}.
+publicly available at \texttt{http://code.google.com/p/range-analysis/}.
 This repository contains instructions about how to deploy and use our
@@
 CFGs and constraint graphs that we produce for meaningful programs at
-\texttt{http://google.code/range-analysis/wiki/gallery}.
+\texttt{http://code.google.com/p/range-analysis/wiki/gallery}.
 

### 2012-03-30 src/RangeAnalysis/RangeAnalysis.cpp
Created methods to measure maximum memory usage at any point of program. Created define JUMPSET so one can choose whether he wants to use jump-set or not.

@@ -172,6 +172,8 @@ bool IntraProceduralRA<CGT>::runOnFunction(Function &F) {
 #ifdef STATS
 	Profile::TimeValue elapsed = prof.timenow() - before;
 	prof.updateTime("BuildGraph", elapsed);
+
+	prof.setMemoryUsage();
 #endif
 #ifdef PRINT_DEBUG
 	CG->printToFile(F, "/tmp/" + F.getName() + "cgpre.dot");
@@ -197,6 +199,7 @@ IntraProceduralRA<CGT>::~IntraProceduralRA(){
 	prof.printTime("Nuutila");
 	prof.printTime("SCCs resolution");
 	prof.printTime("ComputeStats");
+	prof.printMemoryUsage();
 	
 	std::ostringstream formated;
 	formated << 100 * (1.0 - ((double)(needBits) / usedBits));
@@ -263,6 +266,8 @@ bool InterProceduralRA<CGT>::runOnModule(Module &M) {
 #ifdef STATS
 	Profile::TimeValue elapsed = prof.timenow() - before;
 	prof.updateTime("BuildGraph", elapsed);
+
+	prof.setMemoryUsage();
 #endif
 #ifdef PRINT_DEBUG
 	std::string moduleIdentifier = M.getModuleIdentifier();
@@ -429,7 +434,7 @@ InterProceduralRA<CGT>::~InterProceduralRA(){
 	prof.printTime("BuildGraph");
 	prof.printTime("Nuutila");
 	prof.printTime("SCCs resolution");
-	prof.printTime("ComputeStats");
+	prof.printMemoryUsage();
 	
 	std::ostringstream formated;
 	formated << 100 * (1.0 - ((double)(needBits) / usedBits));
@@ -2774,14 +2779,15 @@ void ConstraintGraph::findIntervals() {
 			// Get the entry points of the SCC
 			SmallPtrSet<const Value*, 6> entryPoints;
 			
+#ifdef JUMPSET
 			// Create vector of constants inside component
 			// Comment this line below to deactivate jump-set
 			buildConstantVector(component, compUseMap);
-			
+#endif
 
-			generateEntryPoints(component, entryPoints);
+			//generateEntryPoints(component, entryPoints);
 			//iterate a fixed number of time before widening
-			update(component.size()*2 /*| NUMBER_FIXED_ITERATIONS*/, compUseMap, entryPoints);
[…]

### 2012-03-30 src/RangeAnalysis/RangeAnalysis.h

@@ -59,10 +59,13 @@
 //#define SCC_DEBUG
 
 //TODO: comment the line below to disable the dot printing of Constraint Graphs
-#define PRINT_DEBUG
+//#define PRINT_DEBUG
 
 //Used to enable the stats computing. Comment the below line to disable it
 #define STATS
+
+// Uncomment the line below to activate jump-set
+#define JUMPSET
 
 //Used to limit the number of iterations of fixed meet operator.
 // This update runs before widening and is necessary to improve the result of
@@ -647,8 +650,12 @@ class Profile {
 	
 	private:
 		AccTimesMap accumulatedtimes;
+		size_t memory;
 	
 	public:
+		Profile():
+			memory(0) {}
+
 		TimeValue timenow() {
 			TimeValue garbage, usertime;
 			sys::Process::GetTimeUsage(garbage, usertime, garbage);
@@ -673,6 +680,25 @@ class Profile {
 			std::ostringstream formatted;
 			formatted << time;
 			errs() << formatted.str() << "\t - " << key << " elapsed time\n";
+		}
+
+		void setMemoryUsage() {
+			size_t newmemory = sys::Process::GetTotalMemoryUsage();
+			if (newmemory > memory) {
+				memory = newmemory;
+			}
+		}
+
+		size_t getMemoryUsage() {
+			return memory;
+		}
+
+		void printMemoryUsage() {
+			std::ostringstream formatted;
+			// Convert bytes to kilobytes
+			double mem = memory;
+			formatted << (mem / 1024);
+			errs() << formatted.str() << "\t - " << "Memory used in KB\n";
 		}
 };
 

### 2012-04-08 tests/Instrumentation/CompileCharts.sh

@@ -1,0 +1,146 @@
+#!/bin/bash
+
+rm -f RAEstimatedValues.Benchmark.txt
+
+#Appends every RAEstimatedValues*.txt into one single file, named RAEstimatedValues.txt 
+for file in RAEstimatedValues*.txt ; do
+
+	#echo "$file"
+	cat $file >> RAEstimatedValues.Benchmark.txt
+
+done
+
+#Joins the RAHashNames and RAHashValues of each module, into its respective RARealValues.
+for file in RAHashNames*.txt ; do
+
+	RAHashValues=`echo $file | sed -e "s/RAHashNames/RAHashValues/g"`
+	RARealValues=`echo $file | sed -e "s/RAHashNames/RARealValues/g"`
+	
+	awk 'ARGIND == 1 {vars[$2] = $1} ARGIND == 2 { if(vars[$1]!="") { print vars[$1], $2, $3 } }' $file $RAHashValues > $RARealValues
+
+done
+
+
+rm -f RARealValues.Benchmark.txt
+
+#Appends every RARealValues*.txt into one single file, named RARealValues.txt 
+for file in RARealValues*.txt ; do
+
+	cat $file >> RARealValues.Benchmark.txt
+
+done
+
+
+for file in RARealValues*.txt ; do
+
+	RARealValues=$file
+	RAEstimatedValues=`echo $file | sed -e "s/RARealValues/RAEstimatedValues/g"`
+	RAAnalysisSource=`echo $file | sed -e "s/RARealValues/RAAnalysisSource/g"`	
+
+	#Joins the files RAEstimatedValues and RARealValues in order to produce the base file of the evaluation of the quality of Range Analysis.
+	awk 'ARGIND == 1 {Emin[$1] = $2; Emax[$1] = $3} ARGIND == 2 { print $1, $2, $3, Emin[$1], Emax[$1] }' $RAEstimatedValues $RARealValues > $RAAnalysisSource
+
+done
+
+rm -f RAChartSource*.txt
+
+#For each analysis file, generate one file to be read by GNUPLOT
+for file in RAAnalysisSource*.txt ; do
+
+	RAAnalysisSource=$file
+	RAChartSource=`echo $file | sed -e "s/RAAnalysisSource/RAChartSource/g"`	
+
+	awk '	
+	     	function abs(x){
+	     		if(x<0)
[…]

### 2012-04-09 doc/Manuscript/SAS/paper.tex
Updated the files layout.

--- 2012-03-29 doc/Manuscript/paper.tex
@@
 Nevertheless, the inability to handle comparisons between variables, and the
-cubic complexity of the Bellman-Ford method eventually lead us to seek
+cubic complexity of the Bellman-Ford method eventually led us to seek
 alternative solutions to range analysis.
@@
 \noindent
-\textbf{Asymptotic complexity.}
+\textbf{Time complexity.}
 Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
@@
 Each point in the X line corresponds to a benchmark.
-We analyze the smallest benchmark in this set, \texttt{Prolangs-C}, which has
-1,131 variable nodes in the constraint graph, in 20ms.
+We analyze the smallest benchmark in this set, \texttt{Prolangs-C/deriv2}, which
+has 1,131 variable nodes in the constraint graph, in 20ms.
 We take 15,91secs to analyze our largest benchmark, \texttt{403.gcc}, which,
@@
 \includegraphics[width=0.9\textwidth]{images/TimeCorr}
-%\includegraphics[width=0.4\textwidth]{images/Complexity}
 \end{center}
 \caption{\label{fig:TimeCorr}
-Correlation between program size (measured in number of nodes in constraint
+Correlation between program size (number of var nodes in constraint
 graphs after inlining) and analysis runtime (ms).
 Coefficient of determination = 0.967.
+}
+\end{figure}
+
+\noindent
+\textbf{Memory complexity.}
+The experiments also reveal that the memory consumption of our implementation
+is linear with the program size.
+Figure~\ref{fig:MemCorr} plots these two quantities together.
+The linear correlation, in this case, is even stronger than that found in
+Figure~\ref{fig:TimeCorr}, which compares runtime and program size: the
+coefficient of determination is 0.9947.
+The figure only shows our 100 largest benchmarks.
+Again, SPEC \texttt{403.gcc} is the heaviest benchmarks, requiring
+265,588KB to run.
+Memory includes stack, heap and the executable program code.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{images/MemCorr}
+\end{center}
+\caption{\label{fig:MemCorr}
+Comparison between program size (number of var nodes in constraint
+graphs after inlining) and memory consumption (KB).
+Coefficient of determination = 0.9947.
 }
@@
 \bibliographystyle{plain}
-\bibliography{references}
+\bibliography{../references}
 

### 2012-04-09 doc/Manuscript/SBLP/paper.tex
Updated the files layout.

\documentclass{paper}
\usepackage{sbc-template}

\usepackage{proof}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}

\def\qed{\unskip\kern 10pt{\unitlength1pt\linethickness{.4pt}\framebox(6,6){}}}

\newcommand{\fun}[1]{\mbox{\textbf{#1}}}
\newcommand{\lb}[1]{#1_{\downarrow}}
\newcommand{\ub}[1]{#1_{\uparrow}}
\newcommand{\varset}[1]{\mbox{$\cal{#1}$}}

\sloppy

\begin{document}

\title{Speed and Precision in Range Analysis}

\author{Victor Hugo Sperle Campos, Igor Rafael de Assis Costa,\\
Raphael Hernani Rodrigues and Fernando Magno Quint\~{a}o Pereira}

\address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
\email{\{victorsc,igor.rafael,raphael.hernani,fernando\}@dcc.ufmg.br}
}

\maketitle

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\begin{abstract}
Range analysis is a compiler technique that determines statically the lower and
upper values that each integer variable from a target program may assume
during this program's execution.
This type of inference is very important, because it enables several compiler
optimizations, such as dead and redundant code elimination, bitwidth aware
register allocation, and detection of program vulnerabilities.
In this paper we describe an inter-procedural, context-sensitive range analysis
algorithm that we have implemented in the LLVM compiler.
During the effort to produce an industrial-quality implementation of our
algorithm, we had to face a constant tension between precision and speed.
The foremost goal of this paper is to discuss the many engineering choices
that, due to this tension, have shaped our implementation.
Given the breath of our evaluation, we believe that this paper
contains the most comprehensive empirical study of a range analysis
algorithm ever presented in the compiler related literature.
\end{abstract}

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\section{Introduction}
\label{sec:int}

% What range analysis is good for
Range analysis is a compiler technique whose objective is to determine
statically, for each program variable, limits for the minimum and maximum
values that this variable might assume during the program execution.
Range analysis is important because it enables many compiler optimizations.
Among these optimizations, the most well-known are dead and redundant
code elimination.
Examples of redundant code elimination include the removal of array bounds
checks~\cite{Bodik00,Logozzo08,Venet04} and overflow checks~\cite{Sol11}.
Additionally, range analysis is also used in bitwidth aware register
allocation~\cite{Barik06,Pereira08,Tallam03}, branch
prediction~\cite{Patterson95} and synthesis of hardware for specific
applications~\cite{Cong05,Lhairech10,Mahlke01,Stephenson00}.
Because of this importance, the programming language community has put much
effort in the design and implementation of efficient and precise range
analysis algorithms.

% The lack of a comprehensive empirical evaluation
However, the compiler related literature does not contain a comprehensive
evaluation of range analysis algorithms that scale up to entire programs.
Many works on this subject are limited to very small
programs~\cite{Mahlke01,Simon08,Stephenson00}, or, given their theoretic
perspective, have never been implemented in production
compilers~\cite{Costan05,Gawlitza09,Lakhdar11,Su04,Su05}.
There are implementations of range analysis that deal with very large
programs~\cite{Bertrane10,Cousot09,Logozzo08,Oh11}; nevertheless, because
these papers focus on applications of range analysis,
and not on its implementation, they do not provide a thorough discussion
about their engineering decisions.
A noticeable exception is the recent work of Oh {\em et al.}~\cite{Oh12},
which discusses a range analysis algorithm developed for C programs that can
handle very large benchmarks.
Oh {\em et al.} present an evaluation of the speed and memory consumption of
their implementation.
In this paper we claim to push this discussion considerably further.

% Summary of our experiments
We have implemented an industrial-quality range analysis algorithm in the
LLVM compiler~\cite{Lattner04}.
As we show in Section~\ref{sub:showdown}, our implementation, which is
publicly available, is able to analyze programs with over one million assembly
instructions in ten seconds.
And our implementation is not a straw-man: it produces very precise results.
We have compared the results produced by our implementation with the results
obtained via a dynamic profiler, which we have also implemented.
As we show in Section~\ref{sub:showdown}, when analyzing well-known numeric
benchmarks we are able to estimate tight ranges for almost half of all the
integer variables present in these programs.

% Design space exploration:
While designing and implementing our algorithm we had to face several important
engineering choices.
Many approaches that we have used in an attempt to increase the precision of
our implementation would result in runtime slowdowns.
Although we cannot determine the optimum spot in this design space, given the
vast number of possibilities, we discuss our most important implementation
decisions in Section~\ref{sec:design}.
Section~\ref{sub:sccs} shows how we could improve runtime and precision
substantially by processing data-flow information in the strongly connected
components that underly our constraint system.
Section~\ref{sub:program_rep} discuss the importance of choosing a suitable
intermediate representation when implementing a sparse data-flow framework.
Section~\ref{sub:whole} compares the intra-procedural and the inter-procedural
versions of our algorithm.
The role of context sensitiveness is discussed in Section~\ref{sub:context}.
Finally, Section~\ref{sub:widen} discusses the different widening strategies
that we have experimented with.

% The history of our implementation
This work concludes a two years long effort to produce a solid and scalable
implementation of range analysis.
Our first endeavor to implement such an algorithm was based on Su and
Wagner's constraint system, which can be solved exactly in polynomial
time~\cite{Su04,Su05}.
However, although we could use their formulation to handle a subset of C-like
constructs, their description of how to deal with loops was
not very explicit.
Thus, in order to solve loops we adopted Gawlitza
{\em et al.}'s~\cite{Gawlitza09} approach.
This technique uses the Bellman-Ford algorithm to detect increasing or
decreasing cycles in the constraint system, and then saturates these cycles
via a simple widening operator.
A detailed description of our implementation has been published by
Couto and Pereira~\cite{Couto11}.
Nevertheless, the inability to handle comparisons between variables, and the
cubic complexity of the Bellman-Ford method eventually led us to seek
alternative solutions to range analysis.
This quest reached a pinnacle in the present work, which we summarize in this
paper.

\section{Brief Description of our Range Analysis Algorithm}
\label{sec:desc}

% Define the lattice, the constraints and the valuation I.
\paragraph{The Interval Lattice.}
Following Gawlitza {\em et al.}'s notation, we shall be performing arithmetic
operations over the complete lattice
$\cal{Z} = \mathbb{Z} \cup \{-\infty, +\infty\}$, where the ordering is
naturally given by $-\infty < \ldots < -2 < -1 < 0 < 1 < 2 < \ldots +\infty$.
For any $x > -\infty$ we define:

\begin{tabular}{lcl}
$x + \infty = \infty, x \neq -\infty$ & \mbox{\hspace{0.1cm}} & $x - \infty = - \infty, x \neq +\infty$ \\
$x \times \infty = \infty$ if $x > 0$ & & $x \times \infty = -\infty$ if $x < 0$ \\
$0 \times \infty = 0$ & & $(-\infty) \times \infty = \ $ not defined $$ \\
\end{tabular}

From the lattice $\varset{Z}$ we define the product lattice
$\varset{Z}^2$, which is partially ordered by the subset relation
$\sqsubseteq$.
$\varset{Z}^2$ is defined as follows:
%
\begin{equation*}
\varset{Z}^2 = \emptyset \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
\ z_1 \leq z_2, \  -\infty < z_2 \}
\end{equation*}
%
The objective of range analysis is to determine a mapping
$I: \varset{V} \mapsto \varset{Z}^2$ from the set of integer program variables
$V$ to intervals, such that, for any variable $v \in V$, if
$I(v) = [l, u]$, the, during the execution of the target program, any
valued $i$ assigned to $v$ is such that $l \leq i \leq u$.

\paragraph{A Holistic View of our Range Analysis Algorithm.}
We perform range analysis in a number of steps.
First, we convert the program to a suitable intermediate representation that
makes it easier to extract constraints.
From these constraints, we build a dependence graph that allows us to do
range analysis sparsely.
Finally, we solve the constraints applying different fix-point iterators on
this dependence graph.
Figure~\ref{fig:algorithm} gives a global view of this algorithm.
Some of the steps in the algorithm are optional.
They improve the precision of the range analysis, at the expense of a longer
running time.
In Section~\ref{sec:design} we discuss in more details these tradeoffs.
The last phase, which we call the {\em micro algorithm}, happens per
strong component; however, if we opted for not building these components,
then it happens once for the entire constraint graph.
Nevertheless, the use of strongly connected components
is so essential for performance and precision, as we show in
Section~\ref{sub:sccs}, that it is considered optional only because we
can easily build our implementation without this module.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/algorithm}
\end{center}
\caption{\label{fig:algorithm}
Our implementation of range analysis. Rounded boxes are optional steps.}
\end{figure}

We will illustrate the mandatory parts of the
algorithm via the example program in Figure~\ref{fig:overall_view}.
More details about each phase of the algorithm will be introduced in
Section~\ref{sec:design}, when we discuss our engineering decisions.
Figure~\ref{fig:overall_view}(a) shows an example program taken from the
partition function of the quicksort algorithm used by Bodik
{\em et al.}~\cite{Bodik00}.
We have removed the code that performs array manipulation from this program,
as it plays no role in our explanation.
Figure~\ref{fig:overall_view}(b) shows one possible way to represent this
program internally.
As we explain in Section~\ref{sub:program_rep}, a good program
representation allows us to find more precise results.
In this example we chose a program representation called
Extended Static Single Assignment form, which lets us to solve range
analysis via a path sensitive algorithm.
Figure~\ref{fig:overall_view}(c) shows the constraints that we extract from
the intermediate representation seen in part (b) of this figure.
From these constraints we build the {\em constraint graph} in
Figure~\ref{fig:overall_view}(d).
This graph is the main data-structure that we use to solve range analysis.
For each variable $v$ in the constraint system, the constraint graph has a node
$n_v$.
Similarly, for each constraint $v = f(\ldots, u, \ldots)$ in the constraint
system, the graph has an {\em operation node} $n_f$.
For each constraint $v = f(\ldots, u, \ldots)$ we add two edges to the
graph: $\overrightarrow{n_un_f}$ and $\overrightarrow{n_fn_v}$.
Some edges in the constraint graph are dashed.
These are called {\em control dependence edges}.
If a constraint $v = f(\ldots, \fun{ft}(u), \ldots)$ uses a {\em future}
bound from a variable $u$, then we add to the constraint graph a control
dependence edge $\overrightarrow{n_un_f}$.
The final solution to this instance of the range analysis problem is
given in Figure~\ref{fig:overall_view}(e).

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/overall_view}
\end{center}
\caption{\label{fig:overall_view}
Range analysis by example.
(a) Input program.
(b) Internal compiler representation.
(c) Constraints of the range analysis problem.
(d) The constraint graph.
(e) The final solution.}
\end{figure}

\paragraph{The Micro Algorithm.}
We find the solution given in Figure~\ref{fig:overall_view}(e) in a
process that we call the micro algorithm.
This phase is further divided into three sub-steps:
(i) growth analysis;
(ii) future resolution and
(iii) narrowing analysis.

\noindent
\textbf{Growth analysis: }
The objective of growth analysis is to determine the growth behavior of
each program variable.
There are four possible behaviors:
(a) the variable is bound to a constant interval,
such as $k_0$ in Figure~\ref{fig:overall_view}(b).
(b) The variable is bound to a decreasing interval, i.e., an interval whose
lower bound decreases.
This is the case of $j_1$ in our example.
(c) The variable is bound to an increasing interval, i.e., its upper bound
increases.
This is the case of $i_1$ in the example.
(d) The variable is bound to an interval that expands in both directions.
The growth analysis uses an infinite lattice, i.e., $\varset{Z}^2$.
Thus, a careless implementation of an algorithm that infers growth patterns
might not terminate.
In order to ensure termination, we must reckon on a technique called
{\em widening}, first introduced by Cousot and Cousot as a key component
of abstract interpretation~\cite{Cousot77}.
There are many different widening strategies.
We discuss some of them in Section~\ref{sub:widen}.

\noindent
\textbf{Future resolution: }
In order to learn information from comparisons between variables, such as
\texttt{i < j} in Figure~\ref{fig:overall_view}(a), we bind some intervals
to {\em futures}.
Futures are symbolic limits, which will be replaced by actual numbers once
we finish the growth analysis.
The ranges found by the growth analysis tells us which variables have fixed
bounds, independent on the intersections in the constraint system.
Thus, we can use actual limits to replace intersections bounded by futures.
Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
In order to correctly replace a future $\fun{ft}(V)$ that limits a variable
$V'$, we need to have already applied the growth analysis onto $V$.
Had we considered only data dependence edges, then it would be possible
that $V'$ be analyzed before $V$.
However, because of control dependence edges, this case cannot happen.
The control dependence edges ensure that any topological ordering of the
constraint graph either places $N_v$ before $N_{v'}$, or places these nodes
in the same strongly connected component.
For instance, in Figure~\ref{fig:overall_view}(b), variables $j_1$ and $i_t$
are in the same SCC only because of the control dependence edges.

\begin{figure}[t!]
\begin{center}
\begin{eqnarray*}
\begin{array}{c}
\inferrule{Y = X \sqcap [l, \fun{ft}(V) + c] \\ \ub{I[V]} = u}
{Y = X \sqcap [l, u + c]} \mbox{\hspace{0.3cm}} u, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
\\
\\
\inferrule{Y = X \sqcap [\fun{ft}(V) + c, u] \\ \lb{I[V]} = l}
{Y = X \sqcap [l + c, u]} \mbox{\hspace{0.3cm}} l, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
\end{array}
\end{eqnarray*}
\end{center}
\caption{\label{fig:fix_intersects}Rules to replace futures by actual
bounds. $S$ is the interval bound to each variable after the widening
analysis.}
\end{figure}

\noindent
\textbf{Narrowing Analysis.}
The growth analysis associates very conservative bounds to each variable.
Thus, the last step of our algorithm consists in narrowing these intervals.
We accomplish this step via Cousot and Cousot's classic narrowing
operator~\cite[248]{Cousot77}, which we show in
Figure~\ref{fig:crop_analysis}.

\begin{figure}[t!]
\begin{center}
\begin{eqnarray*}
\begin{array}{c@{\hspace{0.9cm}}c}
\inferrule{\lb{I[Y]} = -\infty \\ \lb{e(Y)} > -\infty}
{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
&
\inferrule{\lb{I[Y]} > \lb{e(Y)}}
{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
\\
\\
\inferrule{\ub{I[Y]} = +\infty \\ \ub{e(Y)} < +\infty}
{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
&
\inferrule{\ub{I[Y]} < \ub{e(Y)}}
{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
\end{array}
\end{eqnarray*}
\end{center}
\caption{\label{fig:crop_analysis}Cousot and Cousot's narrowing operator.}
\end{figure}

\noindent
\textbf{Example:}
Continuing with our example, Figure~\ref{fig:ex_partition_grow_crop} shows
the application of our algorithm on the last strong component of
Figure~\ref{fig:overall_view}(d).
Upon meeting this SCC, we have already determined that the interval
$[0, 0]$ is bound to $i_0$ and that the interval $[100, 100]$ is bound to
$j_0$.
We are not guaranteed to find the least fix point of a constraint system.
However, in this example we did it.
We emphasize that finding this tight solution was only possible because of
the topological ordering of the constraint graph in
Figure~\ref{fig:overall_view}(d).
Had we applied the widening operator onto the whole graph, then we would
have found out that variable $j_0$ is bound to $[-\infty, +\infty]$,
because
(i) it receives its interval directly from variable $k_t$, which is upper
bounded by $+\infty$, and
(ii) it is part of a negative cycle.
On the other hand, by only analyzing $j$'s SCC after we have
analyzed $k$'s, $k$ only contribute the constant range $[0, 99]$ to $j_0$.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/ex_partition_grow_crop}
\end{center}
\caption{\label{fig:ex_partition_grow_crop}
Four snapshots of the last SCC of Figure~\ref{fig:overall_view}(d).
(a) After removing control dependence edges.
(b) After running the growth analysis.
(c) After fixing the intersections bound to futures.
(d) After running the narrowing analysis.}
\end{figure}

\subsection{Range Analysis Showdown}
\label{sub:showdown}

The objective of this section is to show, via experimental numbers, that our
implementation of range analysis is fast, economic and effective.
We have used it to analyze a test suite with 2.72 million lines of C code,
which includes, in addition to all the benchmarks distributed with LLVM,
the programs in the SPEC CPU 2006 collection.

\paragraph{Time and Memory Complexity}

Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
run our algorithm and the size of the input programs.
We show data for the 100 largest benchmarks in our test suite, in number
of variable nodes in the constraint graph.
We perform function inlining before running our analysis, to increase program
sizes.
Each point in the X line corresponds to a benchmark.
We analyze the smallest benchmark in this set, \texttt{Prolangs-C/deriv2}, which
has 1,131 variable nodes in the constraint graph, in 20ms.
We take 15,91secs to analyze our largest benchmark, \texttt{403.gcc}, which,
after function inlining, has 1,266,273 assembly instructions, and a
constraint graph with 679,652 variable nodes.
For this data set, the coefficient of determination $(R^2)$ is 0.967, which
provides very strong evidence about the linear asymptotic complexity of our
implementation.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.9\textwidth]{images/TimeCorr}
\end{center}
\caption{\label{fig:TimeCorr}
Correlation between program size (number of var nodes in constraint
graphs after inlining) and analysis runtime (ms).
Coefficient of determination = 0.967.
}
\end{figure}

\noindent
\textbf{Memory complexity.}
The experiments also reveal that the memory consumption of our implementation
is linear with the program size.
Figure~\ref{fig:MemCorr} plots these two quantities together.
The linear correlation, in this case, is even stronger than that found in
Figure~\ref{fig:TimeCorr}, which compares runtime and program size: the
coefficient of determination is 0.9947.
The figure only shows our 100 largest benchmarks.
Again, SPEC \texttt{403.gcc} is the heaviest benchmarks, requiring
265,588KB to run.
Memory includes stack, heap and the executable program code.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.9\textwidth]{images/MemCorr}
\end{center}
\caption{\label{fig:MemCorr}
Comparison between program size (number of var nodes in constraint
graphs after inlining) and memory consumption (KB).
Coefficient of determination = 0.9947.
}
\end{figure}

\paragraph{Precision}

Our implementation of range analysis is remarkably precise, considering its
runtime.
Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
about 25 minutes to go over a program with almost 900 basic blocks.
We analyze programs of similar size less than one second.
We do not claim our approach is as precise as such algorithms, even though we
are able to find exact bounds to 4/5 of the examples presented
in~\cite{Lakhdar11}.
On the contrary, this paper presents a compromise between precision and speed
that scales to very large programs.
Nevertheless, our results are far from being trivial.
We have implemented a dynamic profiler that measures, for each variable,
its upper and lower limits, given an execution of the target program.
Figure~\ref{fig:precision} compares our results with those measured
dynamically for the Stanford benchmark suite, which is publicly
available~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/precUpperBound}
\includegraphics[width=\textwidth]{images/precLowerBound}
\end{center}
\caption{\label{fig:precision}
(Upper) Comparison between static range analysis and dynamic profiler for
upper bounds.
(Lower) Comparison between static range analysis and dynamic profiler for
lower bounds. The numbers above the benchmark names give the number of
variables in each program.}
\end{figure}

We have classified the bounds estimated by the static analysis into four
categories.
The first category, which we call $1$, contains those bounds that are tight:
during the execution of the program, the variable has been assigned an upper,
or lower limit, that equals the limit inferred statically.
The second category, which we call $n$, contains the estimated bounds that are
within twice the value inferred statically.
For instance, if the range analysis estimates that a variable $v$ is in the
range $[0, 100]$, and during the execution the dynamic profiler finds that
its maximum value is $51$, then $v$ falls into this category.
The third category, $n^2$, contains variables whose actual value is within
a quadratic factor from the estimated value.
In our example, $v$'s upper bound would have to be at most $10$ for it to
be in this category.
Finally, the fourth category contains variables whose estimated value lays
outside a quadratic factor of the actual value.
We call this category {\em imprecise}, and it contains mostly the limits that
our static analysis has marked as either $+\infty$ or $-\infty$.

As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
we have estimated statically are exact.
Similarly, 51.99\% of our upper bounds are also tight.
The figure also shows that, on average, 37.39\% of our lower limits are
imprecise, and 35.40\% of our upper limits are imprecise.
This result is on pair with those obtained by more costly analysis, such as
Stephenson {\em et al.}'s~\cite{Stephenson00}.
However, whereas that approach have not been used with programs larger than
the Stanford benchmark suite, we, as shown before, have been able to
deal with remarkably large programs.

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\section{Design Space}
\label{sec:design}

As we see from a cursory glance at Figure~\ref{fig:algorithm}, our range
analysis algorithm has many optional modules.
These modules provide the user with the possibility to choose between
more precise results, or a faster analysis.
Given the number of options, the design space of a range analysis algorithm
is vast.
In this section we try to cover some of the most important tradeoffs.
Figure~\ref{fig:space} plots, for the integer programs in the SPEC CPU 2006
benchmark suite, precision versus speed for different configurations of
our implementation.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/space}
\end{center}
\caption{\label{fig:space}
Design space exploration: precision versus speed for different configurations
of our algorithm.}
\end{figure}

\subsection{Strongly Connected Components}
\label{sub:sccs}

\subsection{The choice of a program representation}
\label{sub:program_rep}

\subsection{Intra versus Inter-procedural Analysis}
\label{sub:whole}

\subsection{Context Sensitive versus Context Insensitive Analysis}
\label{sub:context}

\subsection{Choosing a Widening Strategy}
\label{sub:widen}

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\section{Conclusion}
\label{sec:con}

\bibliographystyle{plain}
\bibliography{../references}

\end{document}

### 2012-04-10 doc/Manuscript/SBLP/paper.tex
Fixed email address.
Added newest changes to the paper, including more charts.

@@
 \author{Victor Hugo Sperle Campos, Igor Rafael de Assis Costa,\\
-Raphael Hernani Rodrigues and Fernando Magno Quint\~{a}o Pereira}
+Raphael Ernani Rodrigues and Fernando Magno Quint\~{a}o Pereira}
 
@@
 
-\subsection{The choice of a program representation}
-\label{sub:program_rep}
+The greatest source of improvement in our implementation is the use of strongly
+connected components.
+In order to propagate ranges across the constraint graph, we fragment it
+into strongly connected components, collapse each of these components into
+single nodes, and sort the resulting directed acyclic graph topologically.
+We then solve the range analysis problem for each component individually.
+Once we have solved a component, we propagate its ranges to the next
+components, and repeat the process until we walk over the entire constraint
+graph.
+It is well-known that this technique is essential to speedup constraint solving
+algorithms~\cite[Sec 6.3]{Nielson99}.
+In our case, the results are dramatic, mostly in terms of speed, but also in
+terms of precision.
+Figure~\ref{fig:impactSCC} shows the speed up that we gain by using strong
+components.
+We show results for the integer programs in the SPEC CPU 2006 benchmark suite.
+In some cases, as in \texttt{xalancbmk} the analysis on strong components is
+450x faster.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{images/impactSCC}
+\end{center}
+\caption{\label{fig:impactSCC}
+(Left) Bars give time to run our analysis without building strong components
+divided by time to run the analysis on strongly connected components.
+(Right) Bars give precision, in bitwidth reduction, that we obtain with strong
+components, dividid by the precision that we obtain without them.
+}
+\end{figure}
+
+The strong components improve the precision of our growth analysis.
+As we see in Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
+it is 40\% more precise.
+The gains in precision happen because, by completely resolving a component,
+we are able to propagate constant intervals to the next components, instead
+of propagating intervals that can grow in both directions.
+An an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
+$[0, 99]$ from variable $k$ to the component that contains variable $j$.
+Had we run the analysis in the entire constraint graph, by the time we
+applied the growth analysis on $j$ we would still find $k$ bound to
+$0, +\infty$.
 
@@
 
+A naive implementation of range analysis would solve it once for each
+function.
+However, we can gain in precision by performing it inter-procedurally; that
+is, by allowing the results found for a function $f$ to flow into other
+functions that $f$ calls.
+Figure~\ref{fig:intra} illustrates the inter-procedural analysis for the
+program seen in Figure~\ref{fig:overall_view}(a).
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{images/intra}
+\end{center}
+\caption{\label{fig:intra}
+Example where an intra-procedural implementation would lead to imprecise
+results.
+}
+\end{figure}
+
+Figure~\ref{fig:wholeImpact} shows that the whole program analysis increases
+moderately the precision of our analysis.
+We show results for the five largest programs in three different categories of
+benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
+Bitwise~\cite{Stephenson00}.
+Our initial goal when developing this analysis was to support a bitwidth-aware
+register allocator.
+Thus, we measure precision by the average number of bits that our
+analysis allows us to save per program variables.
+It is very important to notice that we do not consider constants in our statistics of precision.
+In other words, we only measure bitwidth reduction in variables that a constant
+propagation step could not remove.
+Our results for the SPEC programs were disappointingly: on the average, the
+intra-procedural version of our analysis saves 5.23\% of the total program
+bitwidth.
+By using the inter-procedural version we can increase this number to 8.89\%.
+A manual inspection of the SPEC programs reveal that this result is expected:
+they manipulate files, and their texts do no provide enough
+explicit constants to power our analysis up.
+However, with numerical benchmarks we fare much better.
+On the average our inter-procedural algorithm reduces the bitwidth of the
+Stanford benchmarks by 36.24\%.
+Finally, for Bitwise we obtain a bitwidth reduction of 12.27\%.
+However, this average is brought down by two outliers: \texttt{edge\_detect} and
+\texttt{sha}, which cannot be reduced.
+The bitwise benchmarks were implemented by Stephenson
+{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
+Our results are on par with those found by the original authors.
+The bitwise programs contain only the \texttt{main} function; thus, different
+versions of our algorithm find the same results.
+
 \subsection{Context Sensitive versus Context Insensitive Analysis}
 \label{sub:context}
+
+Another way to increase the precision of range analysis is by using a
+context-sensitive implementation.
+Context-sensitiveness allows us to distinguish different calling sites of
+the same function.
+Figure~\ref{context} shows way the ability to make this distinction is
+important for precision.
+In Figure~\ref{fig:context}(a) we have two different calls of function
+\texttt{foo}.
+A usual way to perform a data-flow analysis inter-procedurally is to
+create assignments between formal and actual parameters, as we show in
+Figure~\ref{fig:context}(b).
+However, in this case the multiple assignment of values to parameters
+makes the ranges of these parameters very large, whereas in reality they
+are not.
+A way to circumvent this source of imprecision is via function inlining,
+as we show in Figure~\ref{fig:context}(c).
+The results that we can derive for the transformed program are much more
+precise, as each input parameter is assigned a single value.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{images/context}
+\end{center}
+\caption{\label{fig:context}
+Example where a context-sensitive implementation improves the results of
+range analysis.
+}
+\end{figure}
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{images/wholeImpact}
+\end{center}
+\caption{\label{fig:wholeImpact}
+The impact of whole program analysis on precision. Each bar gives precision in
+\%bitwidth reduciton.}
+\end{figure}
+
+\subsection{The choice of a program representation}
+\label{sub:program_rep}
+
+If strong components account for the largest gains in speed, the choice of
+a suitable program representation are responsible for the largest gains in
+precision.
+However, here we no longer have a win-win condition: a more expressive
+program representation decreases our runtime, because it increases the
+size of the target program.
+Nevertheless, the slowdown is inexpressive.
+We chose to run our analysis in programs in the Extended Static Single
+Assignment (e-SSA) form~\cite{Bodik00}.
+The alternative, which gives us a faster, albeit imprecise, analysis, is to run
+our algorithm in programs in Static Single Assignment (SSA)
+form~\cite{Cytron91}.
+Any program in e-SSA form has also the SSA core property: any variable name
+has at most one definition site.
+The contrary is not true: SSA form programs do not have the core e-SSA
+property: any use site of a variable that appears in a conditional test
+post-dominates its definition.
+The program in Figure~\ref{fig:overall_view}(b) is in e-SSA form.
+The live ranges of variables $i_1$ and $j_1$ have been split right after the
+conditional test via the assertions that created variables $i_t$ and $j_t$.
+The e-SSA format serves well analyses that extract information from definition
+sites and conditional tests, and propagate this information forwardly.
+Examples include, in addition to range analysis, tainted flow
+analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/impactESSA}
+\end{center}
+\caption{\label{fig:impactESSA}
+(Left) Bars give the time to run analysis on e-SSA form programs divided by
+the time to run analysis on SSA form programs.
+(Right) Bars give the size of the e-SSA form program, in number of
+assembly instructions, divided by the size of the SSA form program.}
+\end{figure}
+
+Figure~\ref{fig:impactESSA} compares these two program representations in
+terms of runtime.
+As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
+analysis.
+In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
+time by 71\%.
+Runtime increases because the e-SSA form programs are larger than the SSA
+form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
+However, this growth is small: in none of the integer programs in SPEC
+CPU 2006 we verified an increase in code size of more than 9\%.
+But, if the e-SSA form slowdowns the analysis runtime, its gains in precision
+are remarkable, as we show in Figure~\ref{fig:precESSA}.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/precESSA}
+\end{center}
+\caption{\label{fig:precESSA}
+The impact of the e-SSA transformation on precision for three different
+benchmark suites. Bars give the ratio of precision (in bitwidth reduction),
+obtained with e-SSA form conversion divided by precision without e-SSA form
+conversion.}
+\end{figure}
 

### 2012-04-11 doc/Manuscript/SBLP/paper.tex
Added new space chart.

@@
 
-\noindent
-\textbf{Memory complexity.}
 The experiments also reveal that the memory consumption of our implementation
@@
 about 25 minutes to go over a program with almost 900 basic blocks.
-We analyze programs of similar size less than one second.
+We analyze programs of similar size in less than one second.
 We do not claim our approach is as precise as such algorithms, even though we
@@
 our implementation.
+Our initial goal when developing this analysis was to support a bitwidth-aware
+register allocator.
+Thus, we measure precision by the average number of bits that our
+analysis allows us to save per program variables.
+It is very important to notice that we do not consider constants in our statistics of precision.
+In other words, we only measure bitwidth reduction in variables that a constant
+propagation step could not remove.
 
@@
 \caption{\label{fig:space}
-Design space exploration: precision versus speed for different configurations
-of our algorithm.}
+Design space exploration: precision (percentage of bitwidth reduction)
+versus speed (secs) for different configurations of our algorithm analyzing
+the SPEC CPU 2006 integer benchmarks.}
 \end{figure}
@@
 
-A naive implementation of range analysis would solve it once for each
-function.
-However, we can gain in precision by performing it inter-procedurally; that
-is, by allowing the results found for a function $f$ to flow into other
-functions that $f$ calls.
+A naive implementation of range analysis would be intra-procedural; that is,
+would solve the range analysis problem once for each function.
+However, we can gain in precision by performing it inter-procedurally.
+An inter-procedural implementation allows the results found for a function $f$
+to flow into other functions that $f$ calls.
 Figure~\ref{fig:intra} illustrates the inter-procedural analysis for the
 program seen in Figure~\ref{fig:overall_view}(a).
+The trivial way to produce an inter-procedural implementation is to insert
+into the constraint system assignments from the actual parameter names to the
+actual parameter names.
+In our example of Figure~\ref{fig:intra}, our constraint graph contains a flow
+of information from $0$, the actual parameter, to $k_0$, the formal parameter
+of function \texttt{foo}.
 
@@
 Bitwise~\cite{Stephenson00}.
-Our initial goal when developing this analysis was to support a bitwidth-aware
-register allocator.
-Thus, we measure precision by the average number of bits that our
-analysis allows us to save per program variables.
-It is very important to notice that we do not consider constants in our statistics of precision.
-In other words, we only measure bitwidth reduction in variables that a constant
-propagation step could not remove.
-Our results for the SPEC programs were disappointingly: on the average, the
-intra-procedural version of our analysis saves 5.23\% of the total program
-bitwidth.
+Our results for the SPEC programs were disappointingly: on the average for
+the five largest programs, the intra-procedural version of our analysis saves
+5.23\% of the total program bitwidth.
 By using the inter-procedural version we can increase this number to 8.89\%.
@@
 the same function.
-Figure~\ref{context} shows way the ability to make this distinction is
+Figure~\ref{fig:context} shows why the ability to make this distinction is
 important for precision.
@@
 as we show in Figure~\ref{fig:context}(c).
-The results that we can derive for the transformed program are much more
+The results that we can derive for the transformed program are more
 precise, as each input parameter is assigned a single value.
@@
 
+Figure~\ref{fig:wholeImpact} also shows how function inlining modifies the
+precision of our results.
+It is difficult to find an adequate way to compare the precision of
+our analysis with, and without inlining.
+This difficulty stems from the fact that this transformation tends to change
+the target program too much.
+In absolute numbers, we always reduce the bitwidth of more variables after
+function inlining.
+However, proportionally function inlining leads to a smaller percentage of
+bitwidth reduction for many benchmarks.
+In the Stanford Collection, for instance, where most of the functions are
+called in only one location, inlining leads to worst precision results.
+On the other hand, for the SPEC programs inlining, even in terms of
+percentage of reduction, tends to increase our measure of precision.
+
 \begin{figure}[t!]
@@
 \%bitwidth reduciton.}
+\end{figure}
+
+\noindent
+\textbf{Intra vs Inter-procedural runtimes.}
+Figure~\ref{fig:timeComp}(Right) compares three different execution modes.
+Bars are normalized to the time to run the intra-procedural analysis
+without inlining.
+On the average, the intra-procedural mode is 28.92\% faster than the
+inter-procedural mode.
+If we perform function inlining, then this difference is larger: 45.87\%.
+These numbers are close because we run our analysis on strong components.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=0.45\textwidth]{images/timeIntraInterInline}
+\end{center}
+\caption{\label{fig:timeComp}
+Runtime comparison between the intra, inter e inter+inline versions of
+our algorithm.
+The bars are normalized to the time to run the intra-procedural analysis.
+}
 \end{figure}

### 2012-04-11 doc/Manuscript/SBLP/paper.tex
Alterei meu e-mail
Alterando o e-mail do Igor
Added discussion about the widening operator to the paper.

@@
 \address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
-\email{\{victorsc,igor.rafael,raphael.hernani,fernando\}@dcc.ufmg.br}
+\email{\{victorsc,raphael.ernani,igor.rafael,fernando\}@dcc.ufmg.br}
 }
@@
 % Define the lattice, the constraints and the valuation I.
-\paragraph{The Interval Lattice.}
+\noindent
+\textbf{The Interval Lattice.}
 Following Gawlitza {\em et al.}'s notation, we shall be performing arithmetic
@@
 
-\paragraph{A Holistic View of our Range Analysis Algorithm.}
+\noindent
+\textbf{A Holistic View of our Range Analysis Algorithm.}
 We perform range analysis in a number of steps.
@@
 
-\paragraph{The Micro Algorithm.}
+\noindent
+\textbf{The Micro Algorithm.}
 We find the solution given in Figure~\ref{fig:overall_view}(e) in a
@@
 
-\paragraph{Time and Memory Complexity}
+\noindent
+\textbf{Time and Memory Complexity}
 
@@
 
-\paragraph{Precision}
+\noindent
+\textbf{Precision}
 
@@
 size of the target program.
-Nevertheless, the slowdown is inexpressive.
-We chose to run our analysis in programs in the Extended Static Single
-Assignment (e-SSA) form~\cite{Bodik00}.
-The alternative, which gives us a faster, albeit imprecise, analysis, is to run
-our algorithm in programs in Static Single Assignment (SSA)
-form~\cite{Cytron91}.
+We have tried our analysis in two different program representations: the
+Static Single Assignment (SSA) Form~\cite{Cytron91}, and the Extended Static
+Single Assignment (e-SSA) form~\cite{Bodik00}.
+The SSA form gives us a faster, albeit more imprecise, analysis.
 Any program in e-SSA form has also the SSA core property: any variable name
@@
 
+We have implemented the widening operator used in the growth analysis in
+two different ways.
+The first way, which we call {\em simple}, is based on Cousot and
+Cousot's original widening operator~\cite{Cousot77}.
+This operator is shown in Figure~\ref{fig:widening}, and it is the one
+used in Figure~\ref{fig:ex_partition_grow_crop}(b).
+The second widening strategy, which we call {\em jump-set widening} consists
+in using the constants that appear in the program text, in sorted order, as
+the next limits of each interval after widening is applied.
+This operator is common in implementations of range
+analysis~\cite[p.228]{Nielson99}.
+There are situations in which jump-set widening produces better results
+than the simple operator.
+Figure~\ref{fig:jumpSet} shows an example taken from the code of
+SPEC CPU \texttt{bzip2}.
+Part of the constraint graph of the program in Figure~\ref{fig:jumpSet}(a)
+is given in Figure~\ref{fig:jumpSet}(b).
+The result of applying the simple operator is shown in
+Figure~\ref{fig:jumpSet}(c).
+Jump-set widening would use the lattice in Figure~\ref{fig:jumpSet}(d),
+instead of the lattice in Figure~\ref{fig:widening}(Right).
+This lattice would yield the result given in Figure~\ref{fig:jumpSet}(e),
+which is more precise.
+
+\begin{figure}[t!]
+\begin{center}
+\begin{tabular}{c@{\hspace{1.5cm}}c}
+\begin{minipage}{2cm}
+\includegraphics{images/growth_lattice}
+\end{minipage}
+&
+\begin{minipage}{8cm}
+\begin{small}
+\begin{eqnarray*}
+\begin{array}{c@{\hspace{0.5cm}}c}
+\inferrule{I[Y] = [\bot, \bot]}{I[Y] \leftarrow e(Y)}
+&
+\inferrule{\lb{e(Y)} < \lb{I[Y]} \\ \ub{e(Y)} > \ub{I[Y]}}
+{I[Y] \leftarrow [-\infty, +\infty ]}
+\\
+\\
+\inferrule{\lb{e(Y)} < \lb{I[Y]}}
+{I[Y] \leftarrow [-\infty, \ub{I[Y]}]}
+&
+\inferrule{\ub{e(Y)} > \ub{I[Y]}}
+{I[Y] \leftarrow [\lb{I[Y]}, +\infty]}
+\end{array}
+\end{eqnarray*}
+\end{small}
+\end{minipage}
+\end{tabular}
+\end{center}
+\caption{\label{fig:widening}
+(Left) The lattice used in the simple widening strategy.
+(Right) Cousot and Cousot's widening operator. We evaluate the rules from
+left-to-right, top-to-bottom, and stop upon finding a pattern matching.
+Again: given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
+$\ub{\iota} = u$}
+\end{figure}
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/jumpSet}
+\end{center}
+\caption{\label{fig:jumpSet}
+An example where jump-set widening is more precise.}
+\end{figure}
+
+Another way to improve the precision of growth analysis is to perform a few
+rounds of abstract interpretation on the constraint graph, and, in case the
+process does not reach a fix point, only then applying the widening operator.
+Each round consists in evaluating all the constraints, and updating the
+intervals that change from one evaluation to the other.
+The simple widening strategy, coupled with one round of abstract interpretation
+would give us precise results in the example from Figure~\ref{fig:jumpSet}.
+We have experimented with 0 and 16 iterations before doing widening, and the
+overall result, for the programs in the SPEC CPU 2006 suite is given in
+Figure~\ref{fig:space}.
+Figure~\ref{fig:wideningPrec} shows some of these results in more details
+for a broader set of benchmarks.
+In general jump-set widening improves the precision of our results in
+non-trivial ways.
+Nevertheless, the simple widening operator preceded by 16 rounds of
+abstract interpretation in general is more precise than jump-set widening
+without any cycle of pre-evaluation, as we see in Figure~\ref{fig:space}.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/wideningPrec}
+\end{center}
+\caption{\label{fig:wideningPrec}
+The impact of the widening strategy on precision.
+Each bar gives precision in \%bitwidth reduciton.}
+\end{figure}
+
 %mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
@@
 \label{sec:con}
+
+This paper presents what we believe is the most comprehensive evaluation of
+range analysis in the literature.
+Altogether we have experimented with 32 different configurations of our
+range analysis algorithm.
+Our implementation is
+publicly available at \texttt{http://code.google.com/p/range-analysis/}.
+This repository contains instructions about how to deploy and use our
+implementation.
+We provide a gallery of examples, including source codes,
+CFGs and constraint graphs that we produce for meaningful programs at
+\texttt{http://code.google.com/p/range-analysis/wiki/gallery}.
 

### 2012-04-17 doc/Manuscript/SBLP/paper.tex
Fixed erros in the SBLP paper.

@@
 $V$ to intervals, such that, for any variable $v \in V$, if
-$I(v) = [l, u]$, the, during the execution of the target program, any
+$I(v) = [l, u]$, then, during the execution of the target program, any
 valued $i$ assigned to $v$ is such that $l \leq i \leq u$.
@@
 \noindent
-\textbf{Growth analysis: }
+\textbf{Growth analysis. }
 The objective of growth analysis is to determine the growth behavior of
@@
 \noindent
-\textbf{Future resolution: }
+\textbf{Future resolution. }
 In order to learn information from comparisons between variables, such as
@@
 \noindent
-\textbf{Narrowing Analysis.}
+\textbf{Narrowing analysis.}
 The growth analysis associates very conservative bounds to each variable.
@@
 \noindent
-\textbf{Example:}
+\textbf{Example.}
 Continuing with our example, Figure~\ref{fig:ex_partition_grow_crop} shows
@@
 The figure only shows our 100 largest benchmarks.
-Again, SPEC \texttt{403.gcc} is the heaviest benchmarks, requiring
+Again, SPEC \texttt{403.gcc} is the heaviest benchmark, requiring
 265,588KB to run.
@@
 Comparison between program size (number of var nodes in constraint
-graphs after inlining) and memory consumption (KB).
+graphs) and memory consumption (KB).
 Coefficient of determination = 0.9947.
@@
 the Stanford benchmark suite, we, as shown before, have been able to
-deal with remarkably large programs.
+deal with remarkably larger programs.
 
@@
 (Right) Bars give precision, in bitwidth reduction, that we obtain with strong
-components, dividid by the precision that we obtain without them.
+components, divided by the precision that we obtain without them.
 }
@@
 into the constraint system assignments from the actual parameter names to the
-actual parameter names.
+formal parameter names.
 In our example of Figure~\ref{fig:intra}, our constraint graph contains a flow
@@
 \texttt{foo}.
-A usual way to perform a data-flow analysis inter-procedurally is to
+An usual way to perform a data-flow analysis inter-procedurally is to
 create assignments between formal and actual parameters, as we show in
@@
 called in only one location, inlining leads to worst precision results.
-On the other hand, for the SPEC programs inlining, even in terms of
+On the other hand, for the SPEC programs, inlining, even in terms of
 percentage of reduction, tends to increase our measure of precision.
@@
 The impact of whole program analysis on precision. Each bar gives precision in
-\%bitwidth reduciton.}
+\%bitwidth reduction.}
 \end{figure}
@@
 \caption{\label{fig:timeComp}
-Runtime comparison between the intra, inter e inter+inline versions of
+Runtime comparison between intra, inter and inter+inline versions of
 our algorithm.
@@
 If strong components account for the largest gains in speed, the choice of
-a suitable program representation are responsible for the largest gains in
+a suitable program representation is responsible for the largest gains in
 precision.
 However, here we no longer have a win-win condition: a more expressive
-program representation decreases our runtime, because it increases the
+program representation decreases our speed, because it increases the
 size of the target program.
@@
 The impact of the widening strategy on precision.
-Each bar gives precision in \%bitwidth reduciton.}
+Each bar gives precision in \%bitwidth reduction.}
 \end{figure}

### 2012-04-18 doc/Manuscript/SBLP/paper.tex
Alterando meu e-mail
Pequenas correções.

@@
 \address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
-\email{\{victorsc,raphael.ernani,igor.rafael,fernando\}@dcc.ufmg.br}
+\email{\{victorsc,raphael,igor.rafael,fernando\}@dcc.ufmg.br}
 }
@@
 We accomplish this step via Cousot and Cousot's classic narrowing
-operator~\cite[248]{Cousot77}, which we show in
+operator~\cite[p.248]{Cousot77}, which we show in
 Figure~\ref{fig:crop_analysis}.
@@
 Again, SPEC \texttt{403.gcc} is the heaviest benchmark, requiring
-265,588KB to run.
+265.588KB to run.
 Memory includes stack, heap and the executable program code.

### 2012-04-18 doc/Manuscript/SBLP/paper.tex
Fixed my email again.
Changed the tex, and a the figure about context sensitiveness.

@@
 \address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
-\email{\{victorsc,raphael,igor.rafael,fernando\}@dcc.ufmg.br}
+\email{\{victorsc,raphael,igor,fernando\}@dcc.ufmg.br}
 }
@@
 publicly available, is able to analyze programs with over one million assembly
-instructions in ten seconds.
+instructions in sixteen seconds.
 And our implementation is not a straw-man: it produces very precise results.
@@
 analysis algorithm has many optional modules.
-These modules provide the user with the possibility to choose between
-more precise results, or a faster analysis.
+These modules give the user the chance to choose between more precise results,
+or a faster analysis.
 Given the number of options, the design space of a range analysis algorithm
@@
 Thus, we measure precision by the average number of bits that our
-analysis allows us to save per program variables.
+analysis allows us to save per program variable.
 It is very important to notice that we do not consider constants in our statistics of precision.
@@
 terms of precision.
-Figure~\ref{fig:impactSCC} shows the speed up that we gain by using strong
+Figure~\ref{fig:impactSCC} shows the speedup that we gain by using strong
 components.
@@
 The strong components improve the precision of our growth analysis.
-As we see in Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
-it is 40\% more precise.
+According to Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
+strong components increase our precision by 40\%.
 The gains in precision happen because, by completely resolving a component,
@@
 
+\subsection{The choice of a program representation}
+\label{sub:program_rep}
+
+If strong components account for the largest gains in speed, the choice of
+a suitable program representation is responsible for the largest gains in
+precision.
+However, here we no longer have a win-win condition: a more expressive
+program representation decreases our speed, because it increases the
+size of the target program.
+We have tried our analysis in two different program representations: the
+Static Single Assignment (SSA) Form~\cite{Cytron91}, and the Extended Static
+Single Assignment (e-SSA) form~\cite{Bodik00}.
+The SSA form gives us a faster, albeit more imprecise, analysis.
+Any program in e-SSA form has also the SSA core property: any variable name
+has at most one definition site.
+The contrary is not true: SSA form programs do not have the core e-SSA
+property: any use site of a variable that appears in a conditional test
+post-dominates its definition.
+The program in Figure~\ref{fig:overall_view}(b) is in e-SSA form.
+The live ranges of variables $i_1$ and $j_1$ have been split right after the
+conditional test via the assertions that creates variables $i_t$ and $j_t$.
+The e-SSA format serves well analyses that extract information from definition
+sites and conditional tests, and propagate this information forwardly.
+Examples include, in addition to range analysis, tainted flow
+analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/impactESSA}
+\end{center}
+\caption{\label{fig:impactESSA}
+(Left) Bars give the time to run analysis on e-SSA form programs divided by
+the time to run analysis on SSA form programs.
+(Right) Bars give the size of the e-SSA form program, in number of
+assembly instructions, divided by the size of the SSA form program.}
+\end{figure}
+
+Figure~\ref{fig:impactESSA} compares these two program representations in
+terms of runtime.
+As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
+analysis.
+In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
+time by 71\%.
+Runtime increases because the e-SSA form programs are larger than the SSA
+form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
+However, this growth is small: in none of the integer programs in SPEC
+CPU 2006 we verified an increase in code size of more than 9\%.
+But, if the e-SSA form slowdowns the analysis runtime, its gains in precision
+are remarkable, as seen in Figure~\ref{fig:precESSA}.
+These gains happen because the e-SSA format lets the analysis to use the
+results of comparisons to narrow the ranges of variables.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/precESSA}
+\end{center}
+\caption{\label{fig:precESSA}
+The impact of the e-SSA transformation on precision for three different
+benchmark suites. Bars give the ratio of precision (in bitwidth reduction),
+obtained with e-SSA form conversion divided by precision without e-SSA form
+conversion.}
+\end{figure}
+
 \subsection{Intra versus Inter-procedural Analysis}
@@
 A naive implementation of range analysis would be intra-procedural; that is,
-would solve the range analysis problem once for each function.
+would solve the range analysis problem once per each function.
 However, we can gain in precision by performing it inter-procedurally.
@@
 
-Figure~\ref{fig:wholeImpact} shows that the whole program analysis increases
-moderately the precision of our analysis.
-We show results for the five largest programs in three different categories of
-benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
+Figure~\ref{fig:wholeImpact} compares the precision of the intra and
+inter-procedural analyses for the five largest programs in three different
+categories of benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
 Bitwise~\cite{Stephenson00}.
@@
 the five largest programs, the intra-procedural version of our analysis saves
-5.23\% of the total program bitwidth.
-By using the inter-procedural version we can increase this number to 8.89\%.
-A manual inspection of the SPEC programs reveal that this result is expected:
-they manipulate files, and their texts do no provide enough
+5.23\% of bits per variable.
+The inter-procedural version increases this number to 8.89\%.
+A manual inspection of the SPEC programs reveals that this result is expected:
+these programs manipulate files, and their source codes do no provide enough
 explicit constants to power our analysis up.
@@
 Stanford benchmarks by 36.24\%.
-Finally, for Bitwise we obtain a bitwidth reduction of 12.27\%.
-However, this average is brought down by two outliers: \texttt{edge\_detect} and
+For Bitwise we obtain a bitwidth reduction of 12.27\%.
+However, this average is lowered by two outliers: \texttt{edge\_detect} and
 \texttt{sha}, which cannot be reduced.
@@
 The bitwise programs contain only the \texttt{main} function; thus, different
-versions of our algorithm find the same results.
+versions of our algorithm find the same results when applied onto these
+programs.
 
@@
 
-Another way to increase the precision of range analysis is by using a
+Another way to increase the precision of range analysis is via a
 context-sensitive implementation.
@@
 Figure~\ref{fig:context}(b).
+If a function is called more than once, then its formal parameters will receive
+information from many actual parameters.
+We use $\phi$-functions to bind this information together into a single
+flow.
 However, in this case the multiple assignment of values to parameters
@@
 inter-procedural mode.
-If we perform function inlining, then this difference is larger: 45.87\%.
-These numbers are close because we run our analysis on strong components.
+If we perform function inlining, then this difference is 45.87\%.
+These numbers are close because our runtime is bound to the
+size of the strong components.
+We have observed that function inlining does not increase too much these
+components.
 
@@
 }
-\end{figure}
-
-\subsection{The choice of a program representation}
-\label{sub:program_rep}
-
-If strong components account for the largest gains in speed, the choice of
-a suitable program representation is responsible for the largest gains in
-precision.
-However, here we no longer have a win-win condition: a more expressive
-program representation decreases our speed, because it increases the
-size of the target program.
-We have tried our analysis in two different program representations: the
-Static Single Assignment (SSA) Form~\cite{Cytron91}, and the Extended Static
-Single Assignment (e-SSA) form~\cite{Bodik00}.
-The SSA form gives us a faster, albeit more imprecise, analysis.
-Any program in e-SSA form has also the SSA core property: any variable name
-has at most one definition site.
-The contrary is not true: SSA form programs do not have the core e-SSA
-property: any use site of a variable that appears in a conditional test
-post-dominates its definition.
-The program in Figure~\ref{fig:overall_view}(b) is in e-SSA form.
-The live ranges of variables $i_1$ and $j_1$ have been split right after the
-conditional test via the assertions that created variables $i_t$ and $j_t$.
-The e-SSA format serves well analyses that extract information from definition
-sites and conditional tests, and propagate this information forwardly.
-Examples include, in addition to range analysis, tainted flow
-analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=1\textwidth]{images/impactESSA}
-\end{center}
-\caption{\label{fig:impactESSA}
-(Left) Bars give the time to run analysis on e-SSA form programs divided by
-the time to run analysis on SSA form programs.
-(Right) Bars give the size of the e-SSA form program, in number of
-assembly instructions, divided by the size of the SSA form program.}
-\end{figure}
-
-Figure~\ref{fig:impactESSA} compares these two program representations in
-terms of runtime.
-As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
-analysis.
-In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
-time by 71\%.
-Runtime increases because the e-SSA form programs are larger than the SSA
-form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
-However, this growth is small: in none of the integer programs in SPEC
-CPU 2006 we verified an increase in code size of more than 9\%.
-But, if the e-SSA form slowdowns the analysis runtime, its gains in precision
-are remarkable, as we show in Figure~\ref{fig:precESSA}.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=1\textwidth]{images/precESSA}
-\end{center}
-\caption{\label{fig:precESSA}
-The impact of the e-SSA transformation on precision for three different
-benchmark suites. Bars give the ratio of precision (in bitwidth reduction),
-obtained with e-SSA form conversion divided by precision without e-SSA form
-conversion.}
 \end{figure}
@@
 instead of the lattice in Figure~\ref{fig:widening}(Right).
-This lattice would yield the result given in Figure~\ref{fig:jumpSet}(e),
+This lattice yields the result given in Figure~\ref{fig:jumpSet}(e),
 which is more precise.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Another way to improve the precision of growth analysis is to perform a few
 rounds of abstract interpretation on the constraint graph, and, in case the
-process does not reach a fix point, only then applying the widening operator.
-Each round consists in evaluating all the constraints, and updating the
-intervals that change from one evaluation to the other.
-The simple widening strategy, coupled with one round of abstract interpretation
-would give us precise results in the example from Figure~\ref{fig:jumpSet}.
+process does not reach a fix point, only then to apply the widening operator.
+Each round of abstract interpretation consists in evaluating all the
+constraints, and then updating the intervals that change from one evaluation to
+the other.
+For instance, in Figure~\ref{fig:jumpSet} one round of abstract interpretation,
+coupled with the simple widening operator, would be enough to reach the
+fix point of that constraint system.
 We have experimented with 0 and 16 iterations before doing widening, and the
@@
 Figure~\ref{fig:wideningPrec} shows some of these results in more details
-for a broader set of benchmarks.
+for the five largest benchmarks in this collection.
 In general jump-set widening improves the precision of our results in
@@
 abstract interpretation in general is more precise than jump-set widening
-without any cycle of pre-evaluation, as we see in Figure~\ref{fig:space}.
+without any cycle of pre-evaluation, as we see in Figure~\ref{fig:wideningPrec}.
 
 \begin{figure}[t!]
+\begin{small}
 \begin{center}
-\includegraphics[width=1\textwidth]{images/wideningPrec}
+\renewcommand{\tabcolsep}{0.2cm}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|} \hline
+Benchmark &    Size & 0 + Simple &   16 + Simple &     0 + Jump & 16 + Jump \\ \hline
+hmmer     &  38,409 &       9.98 & 11.40 (12.45) & 10.98 (9.11) & 11.40 (12.45) \\ \hline
+gobmk     &  84,846 &       8.15 &  9.93 (17.92) &  9.02 (9.64) & 10.13 (19.54) \\ \hline
+h264ref   &  97,494 &      12.58 &  13.11 (4.04) & 13.00 (3.23) & 13.11 (4.04) \\ \hline
+xalancbmk & 352,423 &       7.71 &   7.98 (3.38) &  7.95 (3.02) & 7.98 (3.38) \\ \hline
+gcc       & 449,442 &      16.09 &  16.63 (3.25) & 16.41 (1.95) & 16.64 (3.31) \\ \hline
+\end{tabular}
 \end{center}
+\end{small}
 \caption{\label{fig:wideningPrec}
-The impact of the widening strategy on precision.
-Each bar gives precision in \%bitwidth reduction.}
+Variation in the precision of our analysis given the widening strategy.
+The size of each benchmark is given in number of variable nodes in the
+constraint graph.
+Precision is given in percentage of bitwidth reduction.
+Numbers in parenthesis are percentage of gain over 0 + Simple.}
 \end{figure}

### 2012-05-02 doc/Manuscript/SBLP/paper.tex
Changed the author list.
Fixed English errors in SBLP paper.

@@
 
-\author{Victor Hugo Sperle Campos, Igor Rafael de Assis Costa,\\
-Raphael Ernani Rodrigues and Fernando Magno Quint\~{a}o Pereira}
+\author{Victor Hugo Sperle Campos, Raphael Ernani Rodrigues, \\
+Igor Rafael de Assis Costa and Fernando Magno Quint\~{a}o Pereira}
 
@@
 On the other hand, by only analyzing $j$'s SCC after we have
-analyzed $k$'s, $k$ only contribute the constant range $[0, 99]$ to $j_0$.
+analyzed $k$'s, $k$ only contributes the constant range $[0, 99]$ to $j_0$.
 
@@
 Again, SPEC \texttt{403.gcc} is the heaviest benchmark, requiring
-265.588KB to run.
+265,588KB to run.
 Memory includes stack, heap and the executable program code.

### 2012-05-06 doc/Manuscript/DTool/paper.tex

--- 2012-04-18 doc/Manuscript/SBLP/paper.tex
@@
 \usepackage{graphicx}
+\usepackage{listings}
+\usepackage{float}
 
@@
 \begin{document}
+\lstset{language=, basicstyle=\small}
 
-\title{Speed and Precision in Range Analysis}
+\title{Tool for Static Range Analysis of Whole Programs}
 
-\author{Victor Hugo Sperle Campos, Igor Rafael de Assis Costa,\\
-Raphael Ernani Rodrigues and Fernando Magno Quint\~{a}o Pereira}
+\author{Victor Hugo Sperle Campos, Raphael Ernani Rodrigues, \\
+Igor Rafael de Assis Costa and Fernando Magno Quint\~{a}o Pereira}
 
 \address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
-\email{\{victorsc,raphael,igor.rafael,fernando\}@dcc.ufmg.br}
+\email{\{victorsc,raphael,igor,fernando\}@dcc.ufmg.br}
 }
@@
 
-%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
 \section{Introduction}
-\label{sec:int}
+\label{sec:intro}
 
-% What range analysis is good for
-Range analysis is a compiler technique whose objective is to determine
-statically, for each program variable, limits for the minimum and maximum
-values that this variable might assume during the program execution.
-Range analysis is important because it enables many compiler optimizations.
-Among these optimizations, the most well-known are dead and redundant
-code elimination.
-Examples of redundant code elimination include the removal of array bounds
-checks~\cite{Bodik00,Logozzo08,Venet04} and overflow checks~\cite{Sol11}.
-Additionally, range analysis is also used in bitwidth aware register
-allocation~\cite{Barik06,Pereira08,Tallam03}, branch
-prediction~\cite{Patterson95} and synthesis of hardware for specific
-applications~\cite{Cong05,Lhairech10,Mahlke01,Stephenson00}.
-Because of this importance, the programming language community has put much
-effort in the design and implementation of efficient and precise range
-analysis algorithms.
+% The context: why range analysis is necessary.
+The analysis of integer variables on the interval lattice has been the
+canonical example of abstract interpretation since its introduction in
+Cousot and Cousot's seminal paper~\cite{Cousot77}.
+Compilers use range analysis to infer the possible values that discrete
+variables may assume during program execution.
+This analysis has many uses.
+For instance, it allows the optimizing compiler to remove from the program text
+redundant overflow tests~\cite{Sol11} and unnecessary array bound
+checks~\cite{Bodik00}.
+Additionally, range analysis is essential not only to the bitwidth aware
+register allocator~\cite{Barik06,Tallam03}, but also to more traditional
+allocators that handle registers of different
+sizes~\cite{Kong98,Pereira08,Scholz02}.
+Finally, range analysis has also seen use in the static prediction of
+branches~\cite{Patterson95}, to detect buffer overflow
+vulnerabilities~\cite{Simon08,Wagner00}, to find the trip count of
+loops~\cite{Lokuciejewski09}
+and even in the synthesis of hardware~\cite{Cong05,Lhairech10,Mahlke01}.
 
-% The lack of a comprehensive empirical evaluation
-However, the compiler related literature does not contain a comprehensive
-evaluation of range analysis algorithms that scale up to entire programs.
-Many works on this subject are limited to very small
-programs~\cite{Mahlke01,Simon08,Stephenson00}, or, given their theoretic
-perspective, have never been implemented in production
-compilers~\cite{Costan05,Gawlitza09,Lakhdar11,Su04,Su05}.
-There are implementations of range analysis that deal with very large
-programs~\cite{Bertrane10,Cousot09,Logozzo08,Oh11}; nevertheless, because
-these papers focus on applications of range analysis,
-and not on its implementation, they do not provide a thorough discussion
-about their engineering decisions.
-A noticeable exception is the recent work of Oh {\em et al.}~\cite{Oh12},
-which discusses a range analysis algorithm developed for C programs that can
-handle very large benchmarks.
-Oh {\em et al.} present an evaluation of the speed and memory consumption of
-their implementation.
-In this paper we claim to push this discussion considerably further.
+% The problem: few implementations, not extensive experimental reports.
+Given this great importance, it comes as no surprise that the compiler
+literature is rich in works describing in details algorithmic variations of
+range analyses~\cite{Gawlitza09,Mahlke01,Stephenson00,Su05}.
+On the other hand, none of these authors provide experimental evidence that
+their approaches are able to deal with very large programs.
+There are researchers who have implemented range analyses that scale up to
+large programs~\cite{Blanchet03,Venet04}; nevertheless, because the
+algorithm itself is not the main focus of their works, they neither give
+details about their design choices nor provide experimental data about it.
+This scenario was recently changed by Oh {\em et al.}~\cite{Oh12}, who
+introduced an abstract interpretation framework which processes programs with
+hundreds of thousands of lines of code.
+Nevertheless, Oh {\em et al.} have designed a very simple range analysis,
+which does not handle comparisons between variables, for instance.
+They also do not discuss the precision of their implementation, but only its
+runtime.
 
-% Summary of our experiments
-We have implemented an industrial-quality range analysis algorithm in the
-LLVM compiler~\cite{Lattner04}.
-As we show in Section~\ref{sub:showdown}, our implementation, which is
-publicly available, is able to analyze programs with over one million assembly
-instructions in ten seconds.
-And our implementation is not a straw-man: it produces very precise results.
-We have compared the results produced by our implementation with the results
-obtained via a dynamic profiler, which we have also implemented.
-As we show in Section~\ref{sub:showdown}, when analyzing well-known numeric
-benchmarks we are able to estimate tight ranges for almost half of all the
-integer variables present in these programs.
+% Tool capabilities.
+We have implemented our algorithm in the LLVM compiler~\cite{Lattner04} as a pass, 
+that can be used either as a stand alone analysis pass or called by a client pass. 
+We have used it to analyze a test suite with 2.72 million lines of C code.
+Our implementation is fast: it globally analyzes the gcc source code in less than
+15 seconds, for instance. 
+It is also precise: our results are similar to Stephenson
+{\em et al}'s~\cite{Stephenson00}, even though our analysis does not require
+a backward propagation phase.
+Furthermore, we have been able to find tight bounds to the majority of the
+examples used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
+{\em et al.}~\cite{Lakhdar11}, who rely on much more costly methods.
+In Section~\ref{sec:dsc} we provide a description of our implementation.
 
-% Design space exploration:
-While designing and implementing our algorithm we had to face several important
-engineering choices.
-Many approaches that we have used in an attempt to increase the precision of
-our implementation would result in runtime slowdowns.
-Although we cannot determine the optimum spot in this design space, given the
-vast number of possibilities, we discuss our most important implementation
-decisions in Section~\ref{sec:design}.
-Section~\ref{sub:sccs} shows how we could improve runtime and precision
-substantially by processing data-flow information in the strongly connected
-components that underly our constraint system.
-Section~\ref{sub:program_rep} discuss the importance of choosing a suitable
-intermediate representation when implementing a sparse data-flow framework.
-Section~\ref{sub:whole} compares the intra-procedural and the inter-procedural
-versions of our algorithm.
-The role of context sensitiveness is discussed in Section~\ref{sub:context}.
-Finally, Section~\ref{sub:widen} discusses the different widening strategies
-that we have experimented with.
+% The demo tool.
+Our demo tool provides a visual interface that allows the user to try our
+range analysis. The tool runs our analysis in the source
+code of user's programs. The user can produce the CFG, the Constraint Graph
+and evaluate the analysis precision with the execution of an instrumented
+version of his program.
 
-% The history of our implementation
-This work concludes a two years long effort to produce a solid and scalable
-implementation of range analysis.
-Our first endeavor to implement such an algorithm was based on Su and
-Wagner's constraint system, which can be solved exactly in polynomial
-time~\cite{Su04,Su05}.
-However, although we could use their formulation to handle a subset of C-like
-constructs, their description of how to deal with loops was
-not very explicit.
-Thus, in order to solve loops we adopted Gawlitza
-{\em et al.}'s~\cite{Gawlitza09} approach.
-This technique uses the Bellman-Ford algorithm to detect increasing or
-decreasing cycles in the constraint system, and then saturates these cycles
-via a simple widening operator.
-A detailed description of our implementation has been published by
-Couto and Pereira~\cite{Couto11}.
-Nevertheless, the inability to handle comparisons between variables, and the
-cubic complexity of the Bellman-Ford method eventually led us to seek
-alternative solutions to range analysis.
-This quest reached a pinnacle in the present work, which we summarize in this
-paper.
+% The rest of this paper.
 
-\section{Brief Description of our Range Analysis Algorithm}
-\label{sec:desc}
+\section{Algorithm Description}
+\label{sec:dsc}
 
@@
 $x \times \infty = \infty$ if $x > 0$ & & $x \times \infty = -\infty$ if $x < 0$ \\
-$0 \times \infty = 0$ & & $(-\infty) \times \infty = \ $ not defined $$ \\
+$0 \times \infty = 0$ & & $(-\infty) \times \infty = \ $ not defined  \\
 \end{tabular}
@@
 running time.
-In Section~\ref{sec:design} we discuss in more details these tradeoffs.
-The last phase, which we call the {\em micro algorithm}, happens per
-strong component; however, if we opted for not building these components,
-then it happens once for the entire constraint graph.
+The last phase happens per strong component; however, if we opted for not building 
+these components, then it happens once for the entire constraint graph.
 Nevertheless, the use of strongly connected components
-is so essential for performance and precision, as we show in
-Section~\ref{sub:sccs}, that it is considered optional only because we
+is so essential for performance and precision that it is considered optional only because we
 can easily build our implementation without this module.
 
-\begin{figure}[t!]
+\begin{figure}[H]
 \begin{center}
@@
 algorithm via the example program in Figure~\ref{fig:overall_view}.
-More details about each phase of the algorithm will be introduced in
-Section~\ref{sec:design}, when we discuss our engineering decisions.
 Figure~\ref{fig:overall_view}(a) shows an example program taken from the
@@
 program internally.
-As we explain in Section~\ref{sub:program_rep}, a good program
-representation allows us to find more precise results.
+A good program representation allows us to find more precise results.
 In this example we chose a program representation called
@@
 
-\begin{figure}[t!]
+\begin{figure}[h!]
 \begin{center}
@@
 
-\noindent
-\textbf{The Micro Algorithm.}
-We find the solution given in Figure~\ref{fig:overall_view}(e) in a
-process that we call the micro algorithm.
-This phase is further divided into three sub-steps:
-(i) growth analysis;
-(ii) future resolution and
-(iii) narrowing analysis.
+\section{How to Create a LLVM pass using our Analysis}
+\label{sec:howto}
 
-\noindent
-\textbf{Growth analysis. }
-The objective of growth analysis is to determine the growth behavior of
-each program variable.
-There are four possible behaviors:
-(a) the variable is bound to a constant interval,
-such as $k_0$ in Figure~\ref{fig:overall_view}(b).
-(b) The variable is bound to a decreasing interval, i.e., an interval whose
-lower bound decreases.
-This is the case of $j_1$ in our example.
-(c) The variable is bound to an increasing interval, i.e., its upper bound
-increases.
-This is the case of $i_1$ in the example.
-(d) The variable is bound to an interval that expands in both directions.
-The growth analysis uses an infinite lattice, i.e., $\varset{Z}^2$.
-Thus, a careless implementation of an algorithm that infers growth patterns
-might not terminate.
-In order to ensure termination, we must reckon on a technique called
-{\em widening}, first introduced by Cousot and Cousot as a key component
-of abstract interpretation~\cite{Cousot77}.
-There are many different widening strategies.
-We discuss some of them in Section~\ref{sub:widen}.
+Our analysis can be used stand-alone, to identify logical problems in the analyzed code.
+However, the Range Analysis can be used as a tool to identify dead code, memory overflow, redundant checks and many other analyses.
+Then, our analysis tool was designed to allow being called from client passes.
+In order to use our range analysis, one can write a LLVM pass that calls it.
+There is vast documentation \footnote{http://llvm.org/docs/WritingAnLLVMPass.html} about how to write LLVM passes in the web.
+The program below, which is self-contained, is an example of such a pass.
 
-\noindent
-\textbf{Future resolution. }
-In order to learn information from comparisons between variables, such as
-\texttt{i < j} in Figure~\ref{fig:overall_view}(a), we bind some intervals
-to {\em futures}.
-Futures are symbolic limits, which will be replaced by actual numbers once
-we finish the growth analysis.
-The ranges found by the growth analysis tells us which variables have fixed
-bounds, independent on the intersections in the constraint system.
-Thus, we can use actual limits to replace intersections bounded by futures.
-Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
-In order to correctly replace a future $\fun{ft}(V)$ that limits a variable
-$V'$, we need to have already applied the growth analysis onto $V$.
-Had we considered only data dependence edges, then it would be possible
-that $V'$ be analyzed before $V$.
-However, because of control dependence edges, this case cannot happen.
-The control dependence edges ensure that any topological ordering of the
-constraint graph either places $N_v$ before $N_{v'}$, or places these nodes
-in the same strongly connected component.
-For instance, in Figure~\ref{fig:overall_view}(b), variables $j_1$ and $i_t$
-are in the same SCC only because of the control dependence edges.
+%Example pass: Source code
+\begin{lstlisting}[frame=single]
+//We are ommiting the LLVM includes
+#include "../RangeAnalysis/RangeAnalysis.h"
 
-\begin{figure}[t!]
-\begin{center}
-\begin{eqnarray*}
-\begin{array}{c}
-\inferrule{Y = X \sqcap [l, \fun{ft}(V) + c] \\ \ub{I[V]} = u}
-{Y = X \sqcap [l, u + c]} \mbox{\hspace{0.3cm}} u, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
-\\
-\\
-\inferrule{Y = X \sqcap [\fun{ft}(V) + c, u] \\ \lb{I[V]} = l}
-{Y = X \sqcap [l + c, u]} \mbox{\hspace{0.3cm}} l, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
-\end{array}
-\end{eqnarray*}
-\end{center}
-\caption{\label{fig:fix_intersects}Rules to replace futures by actual
-bounds. $S$ is the interval bound to each variable after the widening
-analysis.}
-\end{figure}
+using namespace llvm;
 
-\noindent
-\textbf{Narrowing analysis.}
-The growth analysis associates very conservative bounds to each variable.
-Thus, the last step of our algorithm consists in narrowing these intervals.
-We accomplish this step via Cousot and Cousot's classic narrowing
-operator~\cite[p.248]{Cousot77}, which we show in
-Figure~\ref{fig:crop_analysis}.
+class ClientRA: public llvm::FunctionPass {
+public:
+  static char ID;
+  ClientRA() : FunctionPass(ID){ }
+  virtual ~ClientRA() { }
+  virtual bool runOnFunction(Function &F){
+    IntraProceduralRA<Cousot> &ra = 
+      getAnalysis<IntraProceduralRA<Cousot> >();
 
-\begin{figure}[t!]
-\begin{center}
-\begin{eqnarray*}
-\begin{array}{c@{\hspace{0.9cm}}c}
-\inferrule{\lb{I[Y]} = -\infty \\ \lb{e(Y)} > -\infty}
-{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
-&
-\inferrule{\lb{I[Y]} > \lb{e(Y)}}
-{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
-\\
-\\
-\inferrule{\ub{I[Y]} = +\infty \\ \ub{e(Y)} < +\infty}
-{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
-&
-\inferrule{\ub{I[Y]} < \ub{e(Y)}}
-{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
-\end{array}
-\end{eqnarray*}
-\end{center}
-\caption{\label{fig:crop_analysis}Cousot and Cousot's narrowing operator.}
-\end{figure}
+    errs() << "\nCousot Intra Procedural analysis 
+               (Values -> Ranges) of " << F.getName() << ":\n";
+    for(Function::iterator bb=F.begin(), bbEnd=F.end(); bb!=bbEnd; ++bb){
+      for(BasicBlock::iterator I=bb->begin(), IEnd=bb->end(); I!=IEnd; ++I){
+        if(I->getOpcode() == Instruction::Store){
+          const Value *v = &(*I);
+          Range r = ra.getRange(v);
+          r.print(errs());
+          I->dump();
+        }
+      }
+    }
+    return false;
+  }
 
-\noindent
-\textbf{Example.}
-Continuing with our example, Figure~\ref{fig:ex_partition_grow_crop} shows
-the application of our algorithm on the last strong component of
-Figure~\ref{fig:overall_view}(d).
-Upon meeting this SCC, we have already determined that the interval
-$[0, 0]$ is bound to $i_0$ and that the interval $[100, 100]$ is bound to
-$j_0$.
-We are not guaranteed to find the least fix point of a constraint system.
-However, in this example we did it.
-We emphasize that finding this tight solution was only possible because of
-the topological ordering of the constraint graph in
-Figure~\ref{fig:overall_view}(d).
-Had we applied the widening operator onto the whole graph, then we would
-have found out that variable $j_0$ is bound to $[-\infty, +\infty]$,
-because
-(i) it receives its interval directly from variable $k_t$, which is upper
-bounded by $+\infty$, and
-(ii) it is part of a negative cycle.
-On the other hand, by only analyzing $j$'s SCC after we have
-analyzed $k$'s, $k$ only contribute the constant range $[0, 99]$ to $j_0$.
+  virtual void getAnalysisUsage(AnalysisUsage &AU) const {
+    AU.setPreservesAll();
+    AU.addRequired<IntraProceduralRA<Cousot> >();
+  }
+};
 
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/ex_partition_grow_crop}
-\end{center}
-\caption{\label{fig:ex_partition_grow_crop}
-Four snapshots of the last SCC of Figure~\ref{fig:overall_view}(d).
-(a) After removing control dependence edges.
-(b) After running the growth analysis.
-(c) After fixing the intersections bound to futures.
-(d) After running the narrowing analysis.}
-\end{figure}
+char ClientRA::ID = 0;
+static RegisterPass<ClientRA> 
+	X("client-ra", "A client that uses RangeAnalysis", false, false);
+\end{lstlisting}
 
-\subsection{Range Analysis Showdown}
-\label{sub:showdown}
+Our Range Analysis interface provides a method, \textit{getRange}, that returns a \textit{Range} object for any variable in the original code.
+This object of type \textit{Range} contains the range information related to the variable.
+There are many versions of our range analysis pass, e.g., intra/inter procedural, with different narrowing operators, etc.
+In this example we are using the intra-procedural version using Cousot & Cousot's original narrowing operator.
 
-The objective of this section is to show, via experimental numbers, that our
-implementation of range analysis is fast, economic and effective.
-We have used it to analyze a test suite with 2.72 million lines of C code,
-which includes, in addition to all the benchmarks distributed with LLVM,
-the programs in the SPEC CPU 2006 collection.
+%Steps to compile and run
+In order to use the example client, you need to give it a bitcode input file. Below we show how to do this.
+First, we can translate a c file into a bitcode file using clang:
+\begin{lstlisting}[frame=single]
+clang -c -emit-llvm test.c -o test.bc
+\end{lstlisting}
 
-\noindent
-\textbf{Time and Memory Complexity}
+Next thing: we must convert the bitcode file to e-SSA form. We do it using the `vssa` pass.
+\begin{lstlisting}[frame=single]
+opt -instnamer -mem2reg -break-crit-edges test.bc -o test.bc
+opt -load LLVM_SRC_PATH/Debug/lib/vSSA.so -vssa test.bc -o test.essa.bc
+\end{lstlisting}
 
-Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
-run our algorithm and the size of the input programs.
-We show data for the 100 largest benchmarks in our test suite, in number
-of variable nodes in the constraint graph.
-We perform function inlining before running our analysis, to increase program
-sizes.
-Each point in the X line corresponds to a benchmark.
-We analyze the smallest benchmark in this set, \texttt{Prolangs-C/deriv2}, which
-has 1,131 variable nodes in the constraint graph, in 20ms.
-We take 15,91secs to analyze our largest benchmark, \texttt{403.gcc}, which,
-after function inlining, has 1,266,273 assembly instructions, and a
-constraint graph with 679,652 variable nodes.
-For this data set, the coefficient of determination $(R^2)$ is 0.967, which
-provides very strong evidence about the linear asymptotic complexity of our
-implementation.
+Notice that we use a number of other passes too, to improve the quality of the code that we are producing:
+\textit{instnamer} just assigns strings to each variable. This will make the dot files that we produce to look nicer. We only use this pass for aesthetic reasons.
+\textit{mem2reg} maps variables allocated in the stack to virtual registers.
+Without this pass everything is mapped into memory, and then our range analysis will not be able to find any meaningful ranges to the variables.
+\textit{break-crit-edges} removes the critical edges from the control flow graph of the input program.
+This will increase the precision of our range analysis (just a tiny bit though), because the e-SSA transformation will be able to insert more sigma-functions into the code.
 
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{images/TimeCorr}
-\end{center}
-\caption{\label{fig:TimeCorr}
-Correlation between program size (number of var nodes in constraint
-graphs after inlining) and analysis runtime (ms).
-Coefficient of determination = 0.967.
-}
-\end{figure}
-
-The experiments also reveal that the memory consumption of our implementation
-is linear with the program size.
-Figure~\ref{fig:MemCorr} plots these two quantities together.
-The linear correlation, in this case, is even stronger than that found in
-Figure~\ref{fig:TimeCorr}, which compares runtime and program size: the
-coefficient of determination is 0.9947.
-The figure only shows our 100 largest benchmarks.
-Again, SPEC \texttt{403.gcc} is the heaviest benchmark, requiring
-265.588KB to run.
-Memory includes stack, heap and the executable program code.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{images/MemCorr}
-\end{center}
-\caption{\label{fig:MemCorr}
-Comparison between program size (number of var nodes in constraint
-graphs) and memory consumption (KB).
-Coefficient of determination = 0.9947.
-}
-\end{figure}
-
-\noindent
-\textbf{Precision}
-
-Our implementation of range analysis is remarkably precise, considering its
-runtime.
-Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
-about 25 minutes to go over a program with almost 900 basic blocks.
-We analyze programs of similar size in less than one second.
-We do not claim our approach is as precise as such algorithms, even though we
-are able to find exact bounds to 4/5 of the examples presented
-in~\cite{Lakhdar11}.
-On the contrary, this paper presents a compromise between precision and speed
-that scales to very large programs.
-Nevertheless, our results are far from being trivial.
-We have implemented a dynamic profiler that measures, for each variable,
-its upper and lower limits, given an execution of the target program.
-Figure~\ref{fig:precision} compares our results with those measured
-dynamically for the Stanford benchmark suite, which is publicly
-available~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/precUpperBound}
-\includegraphics[width=\textwidth]{images/precLowerBound}
-\end{center}
-\caption{\label{fig:precision}
-(Upper) Comparison between static range analysis and dynamic profiler for
-upper bounds.
-(Lower) Comparison between static range analysis and dynamic profiler for
-lower bounds. The numbers above the benchmark names give the number of
-variables in each program.}
-\end{figure}
-
-We have classified the bounds estimated by the static analysis into four
-categories.
-The first category, which we call $1$, contains those bounds that are tight:
-during the execution of the program, the variable has been assigned an upper,
-or lower limit, that equals the limit inferred statically.
-The second category, which we call $n$, contains the estimated bounds that are
-within twice the value inferred statically.
-For instance, if the range analysis estimates that a variable $v$ is in the
-range $[0, 100]$, and during the execution the dynamic profiler finds that
-its maximum value is $51$, then $v$ falls into this category.
-The third category, $n^2$, contains variables whose actual value is within
-a quadratic factor from the estimated value.
-In our example, $v$'s upper bound would have to be at most $10$ for it to
-be in this category.
-Finally, the fourth category contains variables whose estimated value lays
-outside a quadratic factor of the actual value.
-We call this category {\em imprecise}, and it contains mostly the limits that
-our static analysis has marked as either $+\infty$ or $-\infty$.
-
-As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
-we have estimated statically are exact.
-Similarly, 51.99\% of our upper bounds are also tight.
-The figure also shows that, on average, 37.39\% of our lower limits are
-imprecise, and 35.40\% of our upper limits are imprecise.
-This result is on pair with those obtained by more costly analysis, such as
-Stephenson {\em et al.}'s~\cite{Stephenson00}.
-However, whereas that approach have not been used with programs larger than
-the Stanford benchmark suite, we, as shown before, have been able to
-deal with remarkably larger programs.
-
-%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
-\section{Design Space}
-\label{sec:design}
-
-As we see from a cursory glance at Figure~\ref{fig:algorithm}, our range
-analysis algorithm has many optional modules.
-These modules provide the user with the possibility to choose between
-more precise results, or a faster analysis.
-Given the number of options, the design space of a range analysis algorithm
-is vast.
-In this section we try to cover some of the most important tradeoffs.
-Figure~\ref{fig:space} plots, for the integer programs in the SPEC CPU 2006
-benchmark suite, precision versus speed for different configurations of
-our implementation.
-Our initial goal when developing this analysis was to support a bitwidth-aware
-register allocator.
-Thus, we measure precision by the average number of bits that our
-analysis allows us to save per program variables.
-It is very important to notice that we do not consider constants in our statistics of precision.
-In other words, we only measure bitwidth reduction in variables that a constant
-propagation step could not remove.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/space}
-\end{center}
-\caption{\label{fig:space}
-Design space exploration: precision (percentage of bitwidth reduction)
-versus speed (secs) for different configurations of our algorithm analyzing
-the SPEC CPU 2006 integer benchmarks.}
-\end{figure}
-
-\subsection{Strongly Connected Components}
-\label{sub:sccs}
-
-The greatest source of improvement in our implementation is the use of strongly
-connected components.
-In order to propagate ranges across the constraint graph, we fragment it
-into strongly connected components, collapse each of these components into
-single nodes, and sort the resulting directed acyclic graph topologically.
-We then solve the range analysis problem for each component individually.
-Once we have solved a component, we propagate its ranges to the next
-components, and repeat the process until we walk over the entire constraint
-graph.
-It is well-known that this technique is essential to speedup constraint solving
-algorithms~\cite[Sec 6.3]{Nielson99}.
-In our case, the results are dramatic, mostly in terms of speed, but also in
-terms of precision.
-Figure~\ref{fig:impactSCC} shows the speed up that we gain by using strong
-components.
-We show results for the integer programs in the SPEC CPU 2006 benchmark suite.
-In some cases, as in \texttt{xalancbmk} the analysis on strong components is
-450x faster.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/impactSCC}
-\end{center}
-\caption{\label{fig:impactSCC}
-(Left) Bars give time to run our analysis without building strong components
-divided by time to run the analysis on strongly connected components.
-(Right) Bars give precision, in bitwidth reduction, that we obtain with strong
-components, divided by the precision that we obtain without them.
-}
-\end{figure}
-
-The strong components improve the precision of our growth analysis.
-As we see in Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
-it is 40\% more precise.
-The gains in precision happen because, by completely resolving a component,
-we are able to propagate constant intervals to the next components, instead
-of propagating intervals that can grow in both directions.
-An an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
-$[0, 99]$ from variable $k$ to the component that contains variable $j$.
-Had we run the analysis in the entire constraint graph, by the time we
-applied the growth analysis on $j$ we would still find $k$ bound to
-$0, +\infty$.
-
-\subsection{Intra versus Inter-procedural Analysis}
-\label{sub:whole}
-
-A naive implementation of range analysis would be intra-procedural; that is,
-would solve the range analysis problem once for each function.
-However, we can gain in precision by performing it inter-procedurally.
-An inter-procedural implementation allows the results found for a function $f$
-to flow into other functions that $f$ calls.
-Figure~\ref{fig:intra} illustrates the inter-procedural analysis for the
-program seen in Figure~\ref{fig:overall_view}(a).
-The trivial way to produce an inter-procedural implementation is to insert
-into the constraint system assignments from the actual parameter names to the
-formal parameter names.
-In our example of Figure~\ref{fig:intra}, our constraint graph contains a flow
-of information from $0$, the actual parameter, to $k_0$, the formal parameter
-of function \texttt{foo}.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/intra}
-\end{center}
-\caption{\label{fig:intra}
-Example where an intra-procedural implementation would lead to imprecise
-results.
-}
-\end{figure}
-
-Figure~\ref{fig:wholeImpact} shows that the whole program analysis increases
-moderately the precision of our analysis.
-We show results for the five largest programs in three different categories of
-benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
-Bitwise~\cite{Stephenson00}.
-Our results for the SPEC programs were disappointingly: on the average for
-the five largest programs, the intra-procedural version of our analysis saves
-5.23\% of the total program bitwidth.
-By using the inter-procedural version we can increase this number to 8.89\%.
-A manual inspection of the SPEC programs reveal that this result is expected:
-they manipulate files, and their texts do no provide enough
-explicit constants to power our analysis up.
-However, with numerical benchmarks we fare much better.
-On the average our inter-procedural algorithm reduces the bitwidth of the
-Stanford benchmarks by 36.24\%.
-Finally, for Bitwise we obtain a bitwidth reduction of 12.27\%.
-However, this average is brought down by two outliers: \texttt{edge\_detect} and
-\texttt{sha}, which cannot be reduced.
-The bitwise benchmarks were implemented by Stephenson
-{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
-Our results are on par with those found by the original authors.
-The bitwise programs contain only the \texttt{main} function; thus, different
-versions of our algorithm find the same results.
-
-\subsection{Context Sensitive versus Context Insensitive Analysis}
-\label{sub:context}
-
-Another way to increase the precision of range analysis is by using a
-context-sensitive implementation.
-Context-sensitiveness allows us to distinguish different calling sites of
-the same function.
-Figure~\ref{fig:context} shows why the ability to make this distinction is
-important for precision.
-In Figure~\ref{fig:context}(a) we have two different calls of function
-\texttt{foo}.
-An usual way to perform a data-flow analysis inter-procedurally is to
-create assignments between formal and actual parameters, as we show in
-Figure~\ref{fig:context}(b).
-However, in this case the multiple assignment of values to parameters
-makes the ranges of these parameters very large, whereas in reality they
-are not.
-A way to circumvent this source of imprecision is via function inlining,
-as we show in Figure~\ref{fig:context}(c).
-The results that we can derive for the transformed program are more
-precise, as each input parameter is assigned a single value.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/context}
-\end{center}
-\caption{\label{fig:context}
-Example where a context-sensitive implementation improves the results of
-range analysis.
-}
-\end{figure}
-
-Figure~\ref{fig:wholeImpact} also shows how function inlining modifies the
-precision of our results.
-It is difficult to find an adequate way to compare the precision of
-our analysis with, and without inlining.
-This difficulty stems from the fact that this transformation tends to change
-the target program too much.
-In absolute numbers, we always reduce the bitwidth of more variables after
-function inlining.
-However, proportionally function inlining leads to a smaller percentage of
-bitwidth reduction for many benchmarks.
-In the Stanford Collection, for instance, where most of the functions are
-called in only one location, inlining leads to worst precision results.
-On the other hand, for the SPEC programs, inlining, even in terms of
-percentage of reduction, tends to increase our measure of precision.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/wholeImpact}
-\end{center}
-\caption{\label{fig:wholeImpact}
-The impact of whole program analysis on precision. Each bar gives precision in
-\%bitwidth reduction.}
-\end{figure}
-
-\noindent
-\textbf{Intra vs Inter-procedural runtimes.}
-Figure~\ref{fig:timeComp}(Right) compares three different execution modes.
-Bars are normalized to the time to run the intra-procedural analysis
-without inlining.
-On the average, the intra-procedural mode is 28.92\% faster than the
-inter-procedural mode.
-If we perform function inlining, then this difference is larger: 45.87\%.
-These numbers are close because we run our analysis on strong components.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=0.45\textwidth]{images/timeIntraInterInline}
-\end{center}
-\caption{\label{fig:timeComp}
-Runtime comparison between intra, inter and inter+inline versions of
-our algorithm.
-The bars are normalized to the time to run the intra-procedural analysis.
-}
-\end{figure}
-
-\subsection{The choice of a program representation}
-\label{sub:program_rep}
-
-If strong components account for the largest gains in speed, the choice of
-a suitable program representation is responsible for the largest gains in
-precision.
-However, here we no longer have a win-win condition: a more expressive
-program representation decreases our speed, because it increases the
-size of the target program.
-We have tried our analysis in two different program representations: the
-Static Single Assignment (SSA) Form~\cite{Cytron91}, and the Extended Static
-Single Assignment (e-SSA) form~\cite{Bodik00}.
-The SSA form gives us a faster, albeit more imprecise, analysis.
-Any program in e-SSA form has also the SSA core property: any variable name
-has at most one definition site.
-The contrary is not true: SSA form programs do not have the core e-SSA
-property: any use site of a variable that appears in a conditional test
-post-dominates its definition.
-The program in Figure~\ref{fig:overall_view}(b) is in e-SSA form.
-The live ranges of variables $i_1$ and $j_1$ have been split right after the
-conditional test via the assertions that created variables $i_t$ and $j_t$.
-The e-SSA format serves well analyses that extract information from definition
-sites and conditional tests, and propagate this information forwardly.
-Examples include, in addition to range analysis, tainted flow
-analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=1\textwidth]{images/impactESSA}
-\end{center}
-\caption{\label{fig:impactESSA}
-(Left) Bars give the time to run analysis on e-SSA form programs divided by
-the time to run analysis on SSA form programs.
-(Right) Bars give the size of the e-SSA form program, in number of
-assembly instructions, divided by the size of the SSA form program.}
-\end{figure}
-
-Figure~\ref{fig:impactESSA} compares these two program representations in
-terms of runtime.
-As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
-analysis.
-In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
-time by 71\%.
-Runtime increases because the e-SSA form programs are larger than the SSA
-form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
-However, this growth is small: in none of the integer programs in SPEC
-CPU 2006 we verified an increase in code size of more than 9\%.
-But, if the e-SSA form slowdowns the analysis runtime, its gains in precision
-are remarkable, as we show in Figure~\ref{fig:precESSA}.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=1\textwidth]{images/precESSA}
-\end{center}
-\caption{\label{fig:precESSA}
-The impact of the e-SSA transformation on precision for three different
-benchmark suites. Bars give the ratio of precision (in bitwidth reduction),
-obtained with e-SSA form conversion divided by precision without e-SSA form
-conversion.}
-\end{figure}
-
-\subsection{Choosing a Widening Strategy}
-\label{sub:widen}
-
-We have implemented the widening operator used in the growth analysis in
-two different ways.
-The first way, which we call {\em simple}, is based on Cousot and
-Cousot's original widening operator~\cite{Cousot77}.
-This operator is shown in Figure~\ref{fig:widening}, and it is the one
-used in Figure~\ref{fig:ex_partition_grow_crop}(b).
-The second widening strategy, which we call {\em jump-set widening} consists
-in using the constants that appear in the program text, in sorted order, as
-the next limits of each interval after widening is applied.
-This operator is common in implementations of range
-analysis~\cite[p.228]{Nielson99}.
-There are situations in which jump-set widening produces better results
-than the simple operator.
-Figure~\ref{fig:jumpSet} shows an example taken from the code of
-SPEC CPU \texttt{bzip2}.
-Part of the constraint graph of the program in Figure~\ref{fig:jumpSet}(a)
-is given in Figure~\ref{fig:jumpSet}(b).
-The result of applying the simple operator is shown in
-Figure~\ref{fig:jumpSet}(c).
-Jump-set widening would use the lattice in Figure~\ref{fig:jumpSet}(d),
-instead of the lattice in Figure~\ref{fig:widening}(Right).
-This lattice would yield the result given in Figure~\ref{fig:jumpSet}(e),
-which is more precise.
-
-\begin{figure}[t!]
-\begin{center}
-\begin{tabular}{c@{\hspace{1.5cm}}c}
-\begin{minipage}{2cm}
-\includegraphics{images/growth_lattice}
-\end{minipage}
-&
-\begin{minipage}{8cm}
-\begin{small}
-\begin{eqnarray*}
-\begin{array}{c@{\hspace{0.5cm}}c}
-\inferrule{I[Y] = [\bot, \bot]}{I[Y] \leftarrow e(Y)}
-&
-\inferrule{\lb{e(Y)} < \lb{I[Y]} \\ \ub{e(Y)} > \ub{I[Y]}}
-{I[Y] \leftarrow [-\infty, +\infty ]}
-\\
-\\
-\inferrule{\lb{e(Y)} < \lb{I[Y]}}
-{I[Y] \leftarrow [-\infty, \ub{I[Y]}]}
-&
-\inferrule{\ub{e(Y)} > \ub{I[Y]}}
-{I[Y] \leftarrow [\lb{I[Y]}, +\infty]}
-\end{array}
-\end{eqnarray*}
-\end{small}
-\end{minipage}
-\end{tabular}
-\end{center}
-\caption{\label{fig:widening}
-(Left) The lattice used in the simple widening strategy.
-(Right) Cousot and Cousot's widening operator. We evaluate the rules from
-left-to-right, top-to-bottom, and stop upon finding a pattern matching.
-Again: given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
-$\ub{\iota} = u$}
-\end{figure}
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=1\textwidth]{images/jumpSet}
-\end{center}
-\caption{\label{fig:jumpSet}
-An example where jump-set widening is more precise.}
-\end{figure}
-
-Another way to improve the precision of growth analysis is to perform a few
-rounds of abstract interpretation on the constraint graph, and, in case the
-process does not reach a fix point, only then applying the widening operator.
-Each round consists in evaluating all the constraints, and updating the
-intervals that change from one evaluation to the other.
-The simple widening strategy, coupled with one round of abstract interpretation
-would give us precise results in the example from Figure~\ref{fig:jumpSet}.
-We have experimented with 0 and 16 iterations before doing widening, and the
-overall result, for the programs in the SPEC CPU 2006 suite is given in
-Figure~\ref{fig:space}.
-Figure~\ref{fig:wideningPrec} shows some of these results in more details
-for a broader set of benchmarks.
-In general jump-set widening improves the precision of our results in
-non-trivial ways.
-Nevertheless, the simple widening operator preceded by 16 rounds of
-abstract interpretation in general is more precise than jump-set widening
-without any cycle of pre-evaluation, as we see in Figure~\ref{fig:space}.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=1\textwidth]{images/wideningPrec}
-\end{center}
-\caption{\label{fig:wideningPrec}
-The impact of the widening strategy on precision.
-Each bar gives precision in \%bitwidth reduction.}
-\end{figure}
-
-%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
-\section{Conclusion}
-\label{sec:con}
-
-This paper presents what we believe is the most comprehensive evaluation of
-range analysis in the literature.
-Altogether we have experimented with 32 different configurations of our
-range analysis algorithm.
-Our implementation is
-publicly available at \texttt{http://code.google.com/p/range-analysis/}.
-This repository contains instructions about how to deploy and use our
-implementation.
-We provide a gallery of examples, including source codes,
-CFGs and constraint graphs that we produce for meaningful programs at
-\texttt{http://code.google.com/p/range-analysis/wiki/gallery}.
+Now, we can run our example client. We can do this with the code below:
+\begin{lstlisting}[frame=single]
+opt -load LLVM_SRC_PATH/Debug/lib/RangeAnalysis.so 
+    -load LLVM_SRC_PATH/Debug/lib/ClientRA.so -client-ra test.essa.bc
+\end{lstlisting}
 

### 2012-05-07 doc/Manuscript/DTool/paper.tex

@@
 
-\title{Tool for Static Range Analysis of Whole Programs}
+\title{Tool for Range Analysis of Whole Programs}
 
@@
 register allocation, and detection of program vulnerabilities.
-In this paper we describe an inter-procedural, context-sensitive range analysis
-algorithm that we have implemented in the LLVM compiler.
-During the effort to produce an industrial-quality implementation of our
-algorithm, we had to face a constant tension between precision and speed.
-The foremost goal of this paper is to discuss the many engineering choices
-that, due to this tension, have shaped our implementation.
-Given the breath of our evaluation, we believe that this paper
-contains the most comprehensive empirical study of a range analysis
-algorithm ever presented in the compiler related literature.
+In this paper we present a tool that provides an inter-procedural, context-sensitive 
+range analysis algorithm for the LLVM compiler. This algorithm is a result
+of a two years work, in which we had to face a constant tension between precision 
+and speed.
+The foremost goal of this paper is to show an easy way other optimization passes
+can use the information provided by our range analysis tool.
 \end{abstract}
@@
 
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=\textwidth]{images/overall_view}
+\end{center}
+\caption{\label{fig:overall_view}
+Range analysis by example.
+(a) Input program.
+(b) Internal compiler representation.
+(c) Constraints of the range analysis problem.
+(d) The constraint graph.
+(e) The final solution.}
+\end{figure}
+
 We will illustrate the mandatory parts of the
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/overall_view}
-\end{center}
-\caption{\label{fig:overall_view}
-Range analysis by example.
-(a) Input program.
-(b) Internal compiler representation.
-(c) Constraints of the range analysis problem.
-(d) The constraint graph.
-(e) The final solution.}
-\end{figure}
-
 \section{How to Create a LLVM pass using our Analysis}
@@
 #include "../RangeAnalysis/RangeAnalysis.h"
-
 using namespace llvm;
-
 class ClientRA: public llvm::FunctionPass {
@@
   virtual bool runOnFunction(Function &F){
-    IntraProceduralRA<Cousot> &ra = 
-      getAnalysis<IntraProceduralRA<Cousot> >();
-
+    IntraProceduralRA<Cousot> &ra = getAnalysis<IntraProceduralRA<Cousot>>();
     errs() << "\nCousot Intra Procedural analysis 
@@
 
-Next thing: we must convert the bitcode file to e-SSA form. We do it using the `vssa` pass.
+Next thing: we must convert the bitcode file to e-SSA form. We do it using the \textit{vssa} pass.
 \begin{lstlisting}[frame=single]

### 2012-05-07 doc/Manuscript/DTool/paper.tex
Added changes to the paper.

@@
 
-\title{Tool for Range Analysis of Whole Programs}
+\title{A Tool for Range Analysis of Whole Programs}
 
 \author{Victor Hugo Sperle Campos, Raphael Ernani Rodrigues, \\
-Igor Rafael de Assis Costa and Fernando Magno Quint\~{a}o Pereira}
+Igor Rafael de Assis Costa, Douglas do Couto Teixeira \\
+and Fernando Magno Quint\~{a}o Pereira}
 
 \address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
-\email{\{victorsc,raphael,igor,fernando\}@dcc.ufmg.br}
+\email{\{victorsc,raphael,igor,douglas,fernando\}@dcc.ufmg.br}
 }
@@
 register allocation, and detection of program vulnerabilities.
-In this paper we present a tool that provides an inter-procedural, context-sensitive 
-range analysis algorithm for the LLVM compiler. This algorithm is a result
-of a two years work, in which we had to face a constant tension between precision 
-and speed.
-The foremost goal of this paper is to show an easy way other optimization passes
-can use the information provided by our range analysis tool.
+In this paper we present a tool that implements an inter-procedural,
+context-sensitive  range analysis algorithm for the LLVM compiler.
+Our tool can be used as a standalone compiler pass that gives the user
+static information about the variables in a program, or it can be used to
+enable other compiler optimizations.
+This tool has been implemented to be able to scale to large code bodies.
+As an example, we have used it to analyze the whole gcc compiler is less than
+15 seconds.
 \end{abstract}
@@
 Additionally, range analysis is essential not only to the bitwidth aware
-register allocator~\cite{Barik06,Tallam03}, but also to more traditional
+register allocator~\cite{Barik06}, but also to more traditional
 allocators that handle registers of different
-sizes~\cite{Kong98,Pereira08,Scholz02}.
+sizes~\cite{Pereira08}.
 Finally, range analysis has also seen use in the static prediction of
 branches~\cite{Patterson95}, to detect buffer overflow
-vulnerabilities~\cite{Simon08,Wagner00}, to find the trip count of
+vulnerabilities~\cite{Simon08}, to find the trip count of
 loops~\cite{Lokuciejewski09}
-and even in the synthesis of hardware~\cite{Cong05,Lhairech10,Mahlke01}.
-
-% The problem: few implementations, not extensive experimental reports.
-Given this great importance, it comes as no surprise that the compiler
-literature is rich in works describing in details algorithmic variations of
-range analyses~\cite{Gawlitza09,Mahlke01,Stephenson00,Su05}.
-On the other hand, none of these authors provide experimental evidence that
-their approaches are able to deal with very large programs.
-There are researchers who have implemented range analyses that scale up to
-large programs~\cite{Blanchet03,Venet04}; nevertheless, because the
-algorithm itself is not the main focus of their works, they neither give
-details about their design choices nor provide experimental data about it.
-This scenario was recently changed by Oh {\em et al.}~\cite{Oh12}, who
-introduced an abstract interpretation framework which processes programs with
-hundreds of thousands of lines of code.
-Nevertheless, Oh {\em et al.} have designed a very simple range analysis,
-which does not handle comparisons between variables, for instance.
-They also do not discuss the precision of their implementation, but only its
-runtime.
+and even in the synthesis of hardware~\cite{Mahlke01}.
 
 % Tool capabilities.
-We have implemented our algorithm in the LLVM compiler~\cite{Lattner04} as a pass, 
-that can be used either as a stand alone analysis pass or called by a client pass. 
+We have implemented a range analysis tool on top of the LLVM
+compiler~\cite{Lattner04}.
+This tool can be used either as a standalone analysis, or it can be called by a
+client pass in order to feed this client with more information about the
+program.
+One of our main concerns when developing our range analysis was scalability,
+and we believe that our final product meets the requirements of
+industrial-quality software.
 We have used it to analyze a test suite with 2.72 million lines of C code.
@@
 % The demo tool.
-Our demo tool provides a visual interface that allows the user to try our
-range analysis. The tool runs our analysis in the source
-code of user's programs. The user can produce the CFG, the Constraint Graph
-and evaluate the analysis precision with the execution of an instrumented
-version of his program.
+\noindent
+\textbf{Software: }
+Our tool is publicly available for
+download at \texttt{http://code.google.com/p/range-analysis/}.
+In our webpage we provide, in addition to the static range analysis itself,
+a profiler that records the minimum and maximum values assigned to each
+variable during program execution.
+We also provide a visual interval to our tool, which lets the user to
+see the control flow graph of the target program.
+This visual interface also lets the user to see the integer intervals that
+the analysis estimates to each variable.
 
@@
 \textbf{The Interval Lattice.}
-Following Gawlitza {\em et al.}'s notation, we shall be performing arithmetic
-operations over the complete lattice
+We perform arithmetic operations over the complete lattice
 $\cal{Z} = \mathbb{Z} \cup \{-\infty, +\infty\}$, where the ordering is
@@
 $I(v) = [l, u]$, then, during the execution of the target program, any
-valued $i$ assigned to $v$ is such that $l \leq i \leq u$.
+value $i$ assigned to $v$ is such that $l \leq i \leq u$.
 
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 \begin{center}
@@
 algorithm via the example program in Figure~\ref{fig:overall_view}.
-Figure~\ref{fig:overall_view}(a) shows an example program taken from the
+Figure~\ref{fig:overall_view}(a) shows a program taken from the
 partition function of the quicksort algorithm used by Bodik
@@
 In this example we chose a program representation called
-Extended Static Single Assignment form, which lets us to solve range
-analysis via a path sensitive algorithm.
+Extended Static Single Assignment form~\cite{Bodik00}, which lets us to solve
+range analysis via a path sensitive algorithm.
 Figure~\ref{fig:overall_view}(c) shows the constraints that we extract from
@@
 $n_v$.
-Similarly, for each constraint $v = f(\ldots, u, \ldots)$ in the constraint
-system, the graph has an {\em operation node} $n_f$.
-For each constraint $v = f(\ldots, u, \ldots)$ we add two edges to the
+Similarly, for each operation $v = f(\ldots, u, \ldots)$ in the program, the
+graph has an {\em operation node} $n_f$.
+For each operation $v = f(\ldots, u, \ldots)$ we add two edges to the
 graph: $\overrightarrow{n_un_f}$ and $\overrightarrow{n_fn_v}$.
@@
 dependence edge $\overrightarrow{n_un_f}$.
+Futures are symbolic expressions, which denote intervals whose limits are
+not known immediately, but might be known after the analysis starts solving
+the constraints.
 The final solution to this instance of the range analysis problem is
@@
 
-Our analysis can be used stand-alone, to identify logical problems in the analyzed code.
-However, the Range Analysis can be used as a tool to identify dead code, memory overflow, redundant checks and many other analyses.
-Then, our analysis tool was designed to allow being called from client passes.
+Our analysis can be used independently from other compiler clients, to identify
+logical problems in the analyzed code.
+However, the Range Analysis is more often used as a tool to identify dead code,
+memory overflow and redundant checks; hence, enabling other compiler analysis.
+We will illustrate this last mode via an example.
 In order to use our range analysis, one can write a LLVM pass that calls it.
-There is vast documentation \footnote{http://llvm.org/docs/WritingAnLLVMPass.html} about how to write LLVM passes in the web.
+There is vast documentation~\footnote{\texttt{http://llvm.org/docs/WritingAnLLVMPass.html}} about how to write LLVM passes in the web.
 The program below, which is self-contained, is an example of such a pass.
+This program is a LLVM pass; that is, it is called as part of the LLVM
+tool chain, and receives as input a data-structure of the \texttt{Function}
+type, that describes the program.
+An iterator over such a data-structure returns a list of basic blocks.
+Iterators over basic-blocks return a list of instructions.
 
@@
 
-Our Range Analysis interface provides a method, \textit{getRange}, that returns a \textit{Range} object for any variable in the original code.
-This object of type \textit{Range} contains the range information related to the variable.
-There are many versions of our range analysis pass, e.g., intra/inter procedural, with different narrowing operators, etc.
-In this example we are using the intra-procedural version using Cousot & Cousot's original narrowing operator.
+Our Range Analysis interface provides a method, \textit{getRange}, that returns
+a \textit{Range} object for any variable in the original code.
+This object, of type \textit{Range}, contains the range information related to
+the variable.
+There are many versions of our range analysis pass, e.g., intra/inter procedural,
+with different narrowing operators, etc.
+In this example we are using the intra-procedural version using
+Cousot \& Cousot's original narrowing operator~\cite{Cousot77}.
 
@@
 \end{lstlisting}
-
-Next thing: we must convert the bitcode file to e-SSA form. We do it using the \textit{vssa} pass.
+Next thing: we must convert the bitcode file to e-SSA form.
+We do it using the \textit{vssa} pass.
+The e-SSA conversion module is distributed together with our analysis; however,
+its use is not mandatory.
+It increases the precision of the analysis, at the expenses of a small slowdown.
+In our experiments, we have observed that this slowdown is very small.
 \begin{lstlisting}[frame=single]
@@
 
-Notice that we use a number of other passes too, to improve the quality of the code that we are producing:
-\textit{instnamer} just assigns strings to each variable. This will make the dot files that we produce to look nicer. We only use this pass for aesthetic reasons.
+Notice that we use a number of other passes too, to improve the quality of the
+code that we are producing:
+\textit{instnamer} just assigns strings to each variable.
+We only use this pass for aesthetic reasons.
+This pass will help our visual interface to look nicer.
 \textit{mem2reg} maps variables allocated in the stack to virtual registers.
-Without this pass everything is mapped into memory, and then our range analysis will not be able to find any meaningful ranges to the variables.
-\textit{break-crit-edges} removes the critical edges from the control flow graph of the input program.
-This will increase the precision of our range analysis (just a tiny bit though), because the e-SSA transformation will be able to insert more sigma-functions into the code.
-
-Now, we can run our example client. We can do this with the code below:
+Without this pass everything is mapped into memory, and then our range analysis
+will not be able to find any meaningful ranges to the variables.
+\textit{break-crit-edges} removes the critical edges from the control flow graph
+of the input program.
+This pass will increase the precision of our range analysis (just a tiny bit
+though), because the e-SSA transformation will be able to convert more variables
+into a better format.
+We run the example client with the commands below:
 \begin{lstlisting}[frame=single]
@@
 \end{lstlisting}
+This sequence of calls will cause our example pass to be dynamically loaded
+by the LLVM framework.
+It will receive the bitcodes that we had produced before, and will produce,
+as output, a list of variables, followed by their intervals.
 

### 2012-05-08 doc/Manuscript/DTool/paper.tex
Fixed some English errors. Improved cohesion by adding some connectives between sentences. Fixed commands syntax in last section.

@@
 We have used it to analyze a test suite with 2.72 million lines of C code.
-Our implementation is fast: it globally analyzes the gcc source code in less than
+Our implementation is fast: it globally analyzes the GCC source code in less than
 15 seconds, for instance. 
@@
 variable during program execution.
-We also provide a visual interval to our tool, which lets the user to
+We also provide a visual interface to our tool, which lets the user to
 see the control flow graph of the target program.
@@
 running time.
-The last phase happens per strong component; however, if we opted for not building 
-these components, then it happens once for the entire constraint graph.
+The last phase happens per strong component; however, if we opted not for building 
+these components, then it would happen once for the entire constraint graph.
 Nevertheless, the use of strongly connected components
@@
 
-Our analysis can be used independently from other compiler clients, to identify
+Our analysis can be used independently from other compiler clients to identify
 logical problems in the analyzed code.
 However, the Range Analysis is more often used as a tool to identify dead code,
-memory overflow and redundant checks; hence, enabling other compiler analysis.
+memory overflow and redundant checks; hence, enabling other compiler analyses.
 We will illustrate this last mode via an example.
 In order to use our range analysis, one can write a LLVM pass that calls it.
-There is vast documentation~\footnote{\texttt{http://llvm.org/docs/WritingAnLLVMPass.html}} about how to write LLVM passes in the web.
-The program below, which is self-contained, is an example of such a pass.
-This program is a LLVM pass; that is, it is called as part of the LLVM
+For this purpose, there is vast documentation~\footnote{\texttt{http://llvm.org/docs/WritingAnLLVMPass.html}} about how to write LLVM passes.
+This program below is an example of LLVM client pass, that is, it is called as part of the LLVM
 tool chain, and receives as input a data-structure of the \texttt{Function}
@@
 An iterator over such a data-structure returns a list of basic blocks.
-Iterators over basic-blocks return a list of instructions.
+Consequently, iterators over basic-blocks return a list of instructions.
 
@@
 \begin{lstlisting}[frame=single]
-//We are ommiting the LLVM includes
+//We are omitting the LLVM includes
 #include "../RangeAnalysis/RangeAnalysis.h"
@@
 In order to use the example client, you need to give it a bitcode input file. Below we show how to do this.
-First, we can translate a c file into a bitcode file using clang:
+Firstly, we must translate a c file into a bitcode file using clang:
 \begin{lstlisting}[frame=single]
@@
 \end{lstlisting}
-Next thing: we must convert the bitcode file to e-SSA form.
+
+Next step: we must convert the bitcode file to e-SSA form.
 We do it using the \textit{vssa} pass.
@@
 its use is not mandatory.
-It increases the precision of the analysis, at the expenses of a small slowdown.
-In our experiments, we have observed that this slowdown is very small.
+It increases the precision of the analysis, at the expense of a small slowdown.
 \begin{lstlisting}[frame=single]
-opt -instnamer -mem2reg -break-crit-edges test.bc -o test.bc
-opt -load LLVM_SRC_PATH/Debug/lib/vSSA.so -vssa test.bc -o test.essa.bc
+opt -mem2reg -instnamer -break-crit-edges test.bc -o test.bc
+opt -load LLVM_SRC_PATH/BUILD_TYPE/lib/vSSA.so -vssa test.bc -o test.essa.bc
 \end{lstlisting}
@@
 \textit{instnamer} just assigns strings to each variable.
-We only use this pass for aesthetic reasons.
-This pass will help our visual interface to look nicer.
+We only use this pass for aesthetic reasons, as it 
+will help our visual interface to look nicer.
 \textit{mem2reg} maps variables allocated in the stack to virtual registers.
@@
 of the input program.
-This pass will increase the precision of our range analysis (just a tiny bit
+Therefore, this pass will increase the precision of our range analysis (just a tiny bit
 though), because the e-SSA transformation will be able to convert more variables
@@
 \begin{lstlisting}[frame=single]
-opt -load LLVM_SRC_PATH/Debug/lib/RangeAnalysis.so 
-    -load LLVM_SRC_PATH/Debug/lib/ClientRA.so -client-ra test.essa.bc
+opt -load LLVM_SRC_PATH/BUILD_TYPE/lib/RangeAnalysis.so 
+    -load LLVM_SRC_PATH/BUILD_TYPE/lib/ClientRA.so -client-ra test.essa.bc
 \end{lstlisting}
+
 This sequence of calls will cause our example pass to be dynamically loaded

### 2012-05-08 src/vSSA/vSSA.cpp
Continuing to develop vSSA. Little refactoring has been done.

@@ -33,23 +33,6 @@ bool vSSA::runOnFunction(Function &F) {
 void vSSA::createSigmasIfNeeded(BasicBlock *BB)
 {
 	TerminatorInst *ti = BB->getTerminator();
-	// If the condition used in the terminator instruction is a Comparison instruction:
-	//for each operand of the CmpInst, create sigmas, depending on some conditions
-	/*
-	if(isa<BranchInst>(ti)){
-		BranchInst * bc = cast<BranchInst>(ti);
-		if(bc->isConditional()){
-			Value * cond = bc->getCondition();
-			CmpInst *comparison = dyn_cast<CmpInst>(cond);
-			for (User::const_op_iterator it = comparison->op_begin(), e = comparison->op_end(); it != e; ++it) {
-				Value *operand = *it;
-				if (isa<Instruction>(operand) || isa<Argument>(operand)) {
-					insertSigmas(ti, operand);
-				}
-			}
-		}
-	}
-	*/
 	
 	// CASE 1: Branch Instruction
 	BranchInst *bi = NULL;
@@ -148,6 +131,7 @@ void vSSA::insertSigmas(TerminatorInst *TI, Value *V)
 			if (firstSigma) {
 				// Insert vSSA_phi, if necessary
 				vssaphi_created = insertPhisForSigma(V, sigma);
+				insertSigmaAsOperandOfPhis(vssaphi_created, sigma);
 			}
 			else {
 				// If it was not the first sigma created for V, then the phis already have been created and are stored in the deque, so we need only to insert the sigma as operand of them
@@ -163,7 +147,7 @@ void vSSA::insertSigmas(TerminatorInst *TI, Value *V)
 	// Populate the phi functions created
 	populatePhis(vssaphi_created, V);
 	
-	// Insertion of vSSA_phi functions may require the creation of vSSA_phi functions too, so we call this function recursively
+	// Insertion of vSSA_phi functions may require the creation of vSSA_phi functions too
 	for (SmallVectorImpl<PHINode*>::iterator vit = vssaphi_created.begin(), vend = vssaphi_created.end(); vit != vend; ++vit) {
 		// Rename uses of V to phi
 		renameUsesToPhi(V, *vit);
@@ -260,21 +244,8 @@ SmallVector<PHINode*, 25> vSSA::insertPhisForSigma(Value *V, PHINode *sigma)
 	
[…]

### 2012-05-08 src/vSSA/vSSA.h

@@ -1,3 +1,17 @@
+//#####################################################
+//           vSSA Transformation Pass
+//
+//	This pass is an implementation of Bodik's
+//	Extended Static Single Assignment form.
+//	It is based on a completely different
+//	algorithm from the previous version, which was
+//	present in LLVM 2.7.
+//	Thus, it is extremely faster than that one.
+//
+//	Author: Victor Campos (vhscampos at gmail dot com)
+//#####################################################
+
+
 #include "llvm/Pass.h"
 #include "llvm/ADT/Statistic.h"
 #include "llvm/Analysis/Dominators.h"

### 2012-05-09 doc/Manuscript/DTool/paper.tex
Changed the example a little bit.

@@
 
-\title{A Tool for Range Analysis of Whole Programs}
+\title{A Tool for the Range Analysis of Whole Programs}
 
@@
 Our tool is publicly available for
-download at \texttt{http://code.google.com/p/range-analysis/}.
+download at \texttt{http://code.google.} \texttt{com/p/range-analysis/}.
 In our webpage we provide, in addition to the static range analysis itself,
@@
 as output, a list of variables, followed by their intervals.
+Below we have the output that our analysis prints for the input program in Figure~\ref{fig:overall_view}(a).
+%Example pass: Source code
+\begin{lstlisting}[frame=single]
+Cousot Intra Procedural analysis (Values -> Ranges) of foo:
+[0, 100]  %k.0 = phi i32 [ 0, %bb ], [ %tmp9, %bb8 ]
+[0, 99]  %vSSA_sigma = phi i32 [ %k.0, %bb1 ]
+[-1, 99]  %j.0 = phi i32 [ %vSSA_sigma, %bb2 ], [ %tmp7, %bb5 ]
+[0, 99]  %i.0 = phi i32 [ 0, %bb2 ], [ %tmp6, %bb5 ]
+[0, 99]  %vSSA_sigma3 = phi i32 [ %j.0, %bb3 ]
+[0, 98]  %vSSA_sigma2 = phi i32 [ %i.0, %bb3 ]
+[1, 99]  %tmp6 = add nsw i32 %vSSA_sigma2, 1
+[-1, 98]  %tmp7 = sub nsw i32 %vSSA_sigma3, 1
+[1, 100]  %tmp9 = add nsw i32 %vSSA_sigma, 1
+[100, 100]  %vSSA_sigma1 = phi i32 [ %k.0, %bb1 ]
+
+Cousot Intra Procedural analysis (Values -> Ranges) of main:
+\end{lstlisting}
+Notice that we get a number of output lines for function \texttt{foo}, and zero
+for function \texttt{main}.
+Each output line is the range of a given variable, using LLVM's internal
+representation.
+It is important to use the \texttt{instnamer} pass, like we did before, to get
+some meaningful names for the variables.
+We did not get any line for \texttt{main} just because this function did not have
+any variable with a known range.
 

### 2012-05-10 doc/Manuscript/IST/paper.tex
Added first version of IST paper.

--- 2012-04-09 doc/Manuscript/SAS/paper.tex
@@
-% Explain that we get the topological ordering for free via nuutila's algorithm.
-\documentclass{llncs}
-
-\usepackage{proof}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{skak} % for the \qside
-\usepackage{mathpartir}
-\usepackage{array}
-\usepackage{url}
-
+\documentclass[11pt]{article}
+
+% Packages that I must import to compile this latex file:
+\usepackage{amssymb}    % To use \mathbb
+\usepackage{amsmath}    % To use the equation environment
+\usepackage{graphicx}   % For includegraphics
+\usepackage{mathpartir} % To use \inferrule
+
+% New environments that I use to organize the text:
+\newtheorem{definition}[section]{Definition}
+
+% Definition of new commands that I use in this text:
 \newcommand{\fun}[1]{\mbox{\textbf{#1}}}
@@
 
-\sloppy
-
 \begin{document}
 
-\title {Range Analysis of Whole Programs}
-
-\author{Victor Hugo Sperle Campos, Igor Rafael de Assis Costa, Douglas do Couto
-Teixeira, Fernando Magno Quint\~{a}o Pereira}
-
-\institute{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
-\email{\{victorsc,igor.rafael,douglas,fernando\}@dcc.ufmg.br}
-}
-
+\title{Global Range Analysis}
+\author{Victor Hugo Sperle Campos, Douglas do Couto Teixeira \\
+Igor Rafael Assis Costa, Raphael Ernani Rodrigues and\\
+Fernando Magno Quint\~{a}o Pereira}
+\date{}
 \maketitle
 
-%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
 \begin{abstract}
-This paper describes an inter-procedural range analysis algorithm that scales
-up to programs with millions of assembly instructions.
-Contrary to many previous techniques, we handle comparisons between variables
+Range analysis estimates statically the lower and upper values that
+the variables may assume during a program's execution.
+This analysis is used to detect program vulnerabilities, and to enable compiler
+optimizations such as dead and redundant code elimination.
+This paper describes an inter-procedural range analysis of integer variables that
+scales up to programs with millions of assembly instructions.
+Contrary to previous techniques, we handle comparisons between variables
 without resorting to relational lattices or expensive algorithms.
 We achieve path sensitiveness by using Bodik's Extended Static Single
-Assignment form as the intermediate representation.
-We claim to present in this paper the most extensive empirical evaluation of
-range analysis in an industrial strength compiler.
+Assignment intermediate representation, and we obtain partial
+context sensitiveness via function inlining.
 We have implemented our technique in LLVM and have been able to process
 programs totaling 2.72 million lines of C in a few seconds.
-We do global program analysis, and achieve a simple form of context
-sensitiveness via function inlining.
+During the effort to produce an industrial-quality implementation of our
+algorithm, we had to face a constant tension between precision and speed.
+In this paper we discuss the many engineering choices that, due to this tension,
+have shaped our implementation.
+Given the breath of our evaluation, we believe that this paper contains the most
+comprehensive empirical study of a range analysis algorithm ever presented in the
+compiler related literature.
 \end{abstract}
@@
 \section{Introduction}
-\label{sec:intro}
 
@@
 sizes~\cite{Kong98,Pereira08,Scholz02}.
-Finally, range analysis has also seen use in the static prediction of
-branches~\cite{Patterson95}, to detect buffer overflow
+Additionally, range analysis has also been used to statically predict the
+outcome of branches~\cite{Patterson95}, to detect buffer overflow
 vulnerabilities~\cite{Simon08,Wagner00}, to find the trip count of
 loops~\cite{Lokuciejewski09}
-and even in the synthesis of hardware~\cite{Cong05,Lhairech10,Mahlke01}.
+and even to synthesize hardware~\cite{Cong05,Lhairech10,Mahlke01}.
 
@@
 There are researchers who have implemented range analyses that scale up to
-large programs~\cite{Blanchet03,Venet04}; nevertheless, because the
+large programs~\cite{Blanchet03,Patterson95,Venet04}; nevertheless, because the
 algorithm itself is not the main focus of their works, they neither give
@@
 hundreds of thousands of lines of code.
-Nevertheless, Oh {\em et al.} have designed a very simple range analysis,
+Nevertheless, Oh {\em et al.} have designed a simple range analysis,
 which does not handle comparisons between variables, for instance.
 They also do not discuss the precision of their implementation, but only its
-runtime.
+runtime and memory consumption.
+In this paper we claim to push this discussion considerably further.
 
@@
 is available at \texttt{http://flex.cscott.net/Harpoon/}.} and
-gcc~\footnote{Gcc's VRP pass implements a variant of Patterson's
+gcc~\footnote{Gcc's VRP pass (at \texttt{http://gcc.gnu.org/svn/gcc/trunk/gcc/tree-vrp.c}) implements a variant of Patterson's
 algorithm~\cite{Patterson95}.} only deal with comparisons between variables and
@@
 We use Bodik's Extended Static Single Assignment (e-SSA) form~\cite{Bodik00} to
-perform range analysis sparsely.
+perform path-sensitive range analysis sparsely.
 This program representation ensures that the interval associated with
-a variable is constant along its entire live range, and allows us to compute
-path-sensitive results.
+a variable is constant along its entire live range.
 Finally, we process the strongly connected components that underline our
-constraint system in topological ordering.
+constraint system in topological order.
 It is well-known that this technique is essential to speedup constraint solving
-algorithms~\cite[Sec 6.3]{Nielson99}; however, we show that a careful
-propagation of information along strong components not only gives us speed, but
-also improves the precision of our results.
+algorithms~\cite[Sec 6.3]{Nielson99}; however, due to our three-phases
+approach, a careful propagation of information along strong components not only
+gives us speed, but also improves the precision of our results.
+
+% Our experimental numbers.
+We have implemented our algorithm in the LLVM compiler~\cite{Lattner04}, and
+have used it to process a test suite with 2.72 million lines of C code.
+As we show in Section~\ref{sub:showdown}, our implementation, which is
+publicly available, is able to analyze programs with over one million assembly
+instructions within fifteen seconds.
+And our implementation is not a straw-man: it produces very precise results.
+We have compared the ranges that our implementation finds with the results
+obtained via a dynamic profiler, which we have also implemented.
+As we show in Section~\ref{sub:showdown}, when analyzing well-known numeric
+benchmarks we are able to estimate tight ranges for almost half of all the
+integer variables present in these programs.
+Our results are similar to Stephenson
+{\em et al}'s~\cite{Stephenson00}, even though out analysis does not require
+a backward propagation phase.
+Furthermore, we have been able to find tight bounds to the majority of the
+examples used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
+{\em et al.}~\cite{Lakhdar11}, who rely on much more costly methods.
+
+% Design space exploration:
+While designing and implementing our algorithm we had to face several important
+engineering choices.
+Many approaches that we have used to increase the precision of
+our results would result in runtime slowdowns.
+We cannot determine the optimum spot in this design space, given the
+vast number of possibilities that it contains, but we discuss our most important
+implementation decisions in Section~\ref{sec:design}.
+Section~\ref{sub:sccs} shows how we could improve runtime and precision
+substantially by processing data-flow information in the strongly connected
+components that underly our constraint system.
+Section~\ref{sub:program_rep} discuss the importance of choosing a suitable
+intermediate representation when implementing a sparse data-flow framework.
+Section~\ref{sub:whole} compares the intra-procedural and the inter-procedural
+versions of our algorithm.
+The role of context sensitiveness is discussed in Section~\ref{sub:context}.
+Finally, Section~\ref{sub:widen} discusses the different widening strategies
+that we have experimented with.
 
@@
 However, although we could use their formulation to handle a subset of C-like
-constructs, their description of how to deal with loops was
-not very explicit.
+constructs, their description of how to deal with loops was not very explicit.
 Thus, in order to solve loops we adopted Gawlitza
@@
 This quest reached a pinnacle in the present work.
-
-% Our experimental numbers.
-We have implemented our algorithm in the LLVM compiler~\cite{Lattner04}, and
-have used it to analyze a test suite with 2.72 million lines of C code.
-In Section~\ref{sec:exp} we provide empirical evidence showing that our
-implementation is fast: it globally analyzes the gcc source code in less than
-15 seconds, for instance.
-It is also precise: our results are similar to Stephenson
-{\em et al}'s~\cite{Stephenson00}, even though out analysis does not require
-a backward propagation phase.
-Furthermore, we have been able to find tight bounds to the majority of the
-examples used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
-{\em et al.}~\cite{Lakhdar11}, who rely on much more costly methods.
-
-% The rest of this paper.
 
@@
 consideration.
-As we show in Section~\ref{sec:exp}, most of the strong components in actual
-programs are singletons.
+As we show in Section~\ref{sub:showdown}, most of the strong components in
+actual programs are singletons.
 Furthermore, the composite components tend to be small.
@@
 to each variable grows.
-We ensure termination of this phase via Cousot and Cousot's widening
-operator~\cite[p.247]{Cousot77}.
+We ensure termination of this phase via the widening operators that we
+discuss in Section~\ref{sub:widen}.
 The possible behaviors of an interval are:
@@
 (iv) grows in both directions.
-The lattice of abstract states, plus a constraint system representing our
-growth analysis is given in Figure~\ref{fig:growth_analysis}.
+The lattice of abstract states, plus a constraint system representing
+Cousot and Cousot's widening operator~\cite[p.247]{Cousot77} is given in
+Figure~\ref{fig:growth_analysis}.
 Because the lattice has height three, the intervals bound to each variable can
 change at most three times.
-However, as we show in Figure~\ref{fig:essaDivSSA}(Bottom), we can be more
-precise if we evaluate the constraints a few times before applying widening.
 
@@
 
-%\subsection{Alternative Narrowing Operators}
-%\label{sub:alt}
-%
-%Cousot and Cousot's narrowing operator, given in
-%Figure~\ref{fig:crop_analysis} is not guaranteed to find the least fixpoint
-%of a system of constraints.
-%In order to improve the bounds, trading time for precision, we have
-%experimented with two other operators.
-%The first, is the evaluation function itself, which is given in
-%Figure~\ref{fig:eval_function}.
-%By iterating it until reaching a fixpoint, we are guaranteed to find the
-%optimal solution to the narrowing analysis.
-%However, this approach is not practical, because it is equivalent to
-%concretely interpreting the program.
-%The second alternative that we have tested consists in applying the
-%evaluation function on each constraint at most once per intersection.
-%We call this approach {\em depth-first narrowing}, because, starting from
-%an intersection we visit every node that might be influenced by it.
-%After removing control dependence edges, if we are left with a strongly
-%connected component containing only {\em ascencing constraints}, that
-%is, constraints that only increase the upper bounds of intervals, then this
-%approach is guaranteed to find the least fixpoint.
-%The same is true if we have a SCC with only {\em descending constraints}.
-
-\section{Experiments}
-\label{sec:exp}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We have implemented our range analysis algorithm in LLVM 3.0, and have run
-experiments on a Intel quad core CPU with a 2.40GHz clock, and 3.6GB of RAM.
-Each core has a 4,096KB L1 cache.
-We use Linux Ubuntu 10.04.4.
-Our implementation of range analysis has 3,958 lines of
-commented C++ code, and our e-SSA conversion module has 672 lines.
-We have analyzed 428 C programs that constitute the LLVM test suite plus the
-integer benchmarks in SPEC CPU 2006.
-Together, these programs contain 4.76 million assembly instructions.
+\subsection{Range Analysis Showdown}
+\label{sub:showdown}
+
+The objective of this section is to show, via experimental numbers, that our
+implementation of range analysis is fast, economic and effective.
+We have used it to analyze a test suite with 2.72 million lines of C code,
+which includes, in addition to all the benchmarks distributed with LLVM,
+the programs in the SPEC CPU 2006 collection.
 
 \noindent
-\textbf{Structural Properties.}
-Table~\ref{tab:struct} shows data taken from the SPEC 2006 integer benchmark
-suite for the inter-procedural version of our analysis.
-Three observations provide evidence that our algorithm is linear in practice,
-as we will show in Figure~\ref{fig:TimeCorr}.
-First, most of the strongly connected components found in constraint graphs are
-singletons.
-For instance, 99.11\% of the SCCs in gcc (\texttt{403.gcc}) have only
-one node.
-Hence, the most complex phases of our algorithm are confined to a small part
-of the constraint graph.
-Second, the strongly connected components are not large -- our largest
-component, found in \texttt{403.gcc}, has 2,131 nodes.
-It exists due to a long chain of mutually recursive function calls.
-We do not count control dependency edges when measuring the size of SCCs, as
-they have no impact on the algorithm explained in Section~\ref{sub:micro}.
-Third, as we see in the table, any node was visited at most three times by
-the narrowing operator.
-
-\begin{table}[t!]
-\begin{center}
-\begin{scriptsize}
-\renewcommand{\tabcolsep}{0.1cm}
-\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|} \hline
-Name	    & 464 &	473 & 483 & 458 & 429 & 471 & 403 & 445 & 462 & 401 & 456 \\ \hline
-\#I  & 139626 & 8900 & 593765 & 26947 & 2738 & 94036 & 887191 & 141330 &  6237 & 17663 & 61736 \\ \hline
-\#$N_v$ & 97494 & 5490 & 352423 & 14407 & 1853	 & 50095 & 449442 & 84846 & 3293 & 12517 & 38409 \\ \hline
-\#$N_c$ & 69530 & 3799 & 213535 & 11164 & 1156 & 28943 & 309310 & 63506 & 2336 & 8731 & 24530 \\ \hline
-\#S  & 91731 & 5108 & 346451 & 13309 & 1751 & 48784 & 435372 & 79909 & 3003 & 11539 & 35619 \\ \hline
-L    & 38 & 49 & 120 & 116 & 16 & 258 & 2131 & 311 & 9 & 18 & 41  \\ \hline
-\#U  & 89461 & 4983 & 344132 & 12919 & 1726 & 48400 & 431703 & 78407 & 2885 & 11267 & 34631 \\ \hline
-\#V  & 3	& 2 & 2 & 2 & 2 & 2 & 2 & 3 & 2 & 2 & 2 \\ \hline
-\end{tabular}
-\end{scriptsize}
-\end{center}
-\caption{\label{tab:struct}Structural Properties of the programs in the SPEC CPU
-2006 integer benchmark suite.
-\#I: number of assembly instructions in the original program.
-\#$N_v$: number of variable nodes in the constraint graph.
-\#$N_c$: number of constraint nodes in the constraint graph.
-% TODO: add also number of constraint nodes.
-\#S: number of strongly connected components (SCC).
-L: size of the largest SCC.
-\#U: number of SCCs with only one node.
-\#V: maximum number of times any variable node was visited by narrowing.}
-\end{table}%
-
-% Absolute Time and complexity
-% =============================
-
-\noindent
-\textbf{Time complexity.}
+\textbf{Time and Memory Complexity.}
 Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
@@
 \noindent
-\textbf{Time distribution.}
-Figure~\ref{fig:timeComp}(Left) shows the distribution of processing time
-among the main phases of the inter-procedural version of our algorithm.
-Functions are inlined to increase program sizes.
-We analyze the SPEC 2006 integer benchmarks in 41.42secs.
-The time to build the e-SSA representation is 7.03secs; the time to extract
-constraints from the program and build the constraint graphs is 7.87secs;
-the time to find SCCs via Nuutila's algorithm and sort them topologically is
-10.55secs, and the time to run the analysis itself is 15.96secs.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=0.48\textwidth]{images/timeDistr}
-\includegraphics[width=0.48\textwidth]{images/timeIntraInterInline}
-\end{center}
-\caption{\label{fig:timeComp}
-(Left) Time distribution across different phases of our inter+inline
-algorithm.
-(Right) Runtime comparison between the intra, inter e inter+inline versions of
-our algorithm.
-The bars are normalized to the time to run the intra-procedural version.
-}
-\end{figure}
-
-\noindent
-\textbf{Intra vs Inter-procedural runtimes.}
-As we saw in Figure~\ref{fig:algorithm}, our implementation supports
-intra- and inter-procedural execution modes.
-On top of these modes, we can optionally perform function inline to obtain
-limited context-sensitiveness.
-Our baseline compiler, LLVM, is able to inline non-recursive function calls.
-Figure~\ref{fig:timeComp}(Right) compares three different execution modes.
-Times are normalized to the time to run the intra-procedural analysis
-without inlining.
-On the average, the intra-procedural mode is 28.92\% faster than the
-inter-procedural mode.
-If we perform function inlining, then this difference is larger: 45.87\%.
-These numbers are close because we run our analysis on strong components.
-
-\noindent
-\textbf{The impact of strong components in time and precision.}
-Figure~\ref{fig:impactEssaSCC}(a) shows that strong components dramatically
-decrease the runtime of our analysis.
-For instance, this design improves the analysis time of \texttt{483.xalancbmk}
-452 times!
-Moreover, as the example in Section~\ref{sub:micro} shows, our algorithm benefits
-from the partitioning of constraint graphs into SCCs to be more precise.
-Figure~\ref{fig:impactEssaSCC}(b) measures this precision.
-Range propagation along strong components allows us, for example, to reduce
-41.5\% more bits in \texttt{401.bzip2}.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/impactEssaSCC}
-\end{center}
-\caption{\label{fig:impactEssaSCC}
-How strong components and the e-SSA representation improve our algorithm.
-This data refer to the inter-procedural analysis with function inlining.
-}
-\end{figure}
-
-\noindent
-\textbf{Impact of the e-SSA on runtime.}
-Figure~\ref{fig:impactEssaSCC}(c) shows that the e-SSA conversion increases
-the size of the SPEC 2006 integer benchmarks by 6.52\%.
-This growth is due to the $\sigma$-functions used to split live ranges
-after conditionals, and to the extra $\phi$-functions necessary to preserve the
-single static assignment property.
-As a consequence of the program growth, the e-SSA form increases the runtime of
-our analysis by 7.11\%, as seen in Figure~\ref{fig:impactEssaSCC}(d).
-We are comparing only the time to perform range analysis.
-The bars do not include the time to perform the e-SSA conversion, which we
-showed in Figure~\ref{fig:timeComp}(Left).
-
-\noindent
-\textbf{The impact of whole program analysis on precision.}
-Figure~\ref{fig:essaDivSSA}(Top) shows that the whole program analysis increases
-moderately the precision of our analysis.
-We show results for the five largest programs in three different categories of
-benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
-Bitwise~\cite{Stephenson00}.
+\textbf{Precision.}
+Our implementation of range analysis is remarkably precise, considering its
+runtime.
+Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
+about 25 minutes to go over a program with almost 900 basic blocks.
+We analyze programs of similar size in less than one second.
+We do not claim our approach is as precise as such algorithms, even though we
+are able to find exact bounds to 4/5 of the examples presented
+in~\cite{Lakhdar11}.
+On the contrary, this paper presents a compromise between precision and speed
+that scales to very large programs.
+Nevertheless, our results are far from being trivial.
+We have implemented a dynamic profiler that measures, for each variable,
+its upper and lower limits, given an execution of the target program.
+Figure~\ref{fig:precision} compares our results with those measured
+dynamically for the Stanford benchmark suite, which is publicly
+available~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{images/precUpperBound}
+\includegraphics[width=\textwidth]{images/precLowerBound}
+\end{center}
+\caption{\label{fig:precision}
+(Upper) Comparison between static range analysis and dynamic profiler for
+upper bounds.
+(Lower) Comparison between static range analysis and dynamic profiler for
+lower bounds. The numbers above the benchmark names give the number of
+variables in each program.}
+\end{figure}
+
+We have classified the bounds estimated by the static analysis into four
+categories.
+The first category, which we call $1$, contains those bounds that are tight:
+during the execution of the program, the variable has been assigned an upper,
+or lower limit, that equals the limit inferred statically.
+The second category, which we call $n$, contains the bounds that are
+within twice the value inferred statically.
+For instance, if the range analysis estimates that a variable $v$ is in the
+range $[0, 100]$, and during the execution the dynamic profiler finds that
+its maximum value is $51$, then $v$ falls into this category.
+The third category, $n^2$, contains variables whose actual value is within
+a quadratic factor from the estimated value.
+In our example, $v$'s upper bound would have to be at most $10$ for it to
+be in this category.
+Finally, the fourth category contains variables whose estimated value lays
+outside a quadratic factor of the actual value.
+We call this category {\em imprecise}, and it contains mostly the limits that
+our static analysis has marked as either $+\infty$ or $-\infty$.
+
+As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
+we have estimated statically are exact.
+Similarly, 51.99\% of our upper bounds are also tight.
+The figure also shows that, on average, 37.39\% of our lower limits are
+imprecise, and 35.40\% of our upper limits are imprecise.
+This result is on pair with those obtained by more costly analysis, such as
+Stephenson {\em et al.}'s~\cite{Stephenson00}.
+However, whereas that approach have not been used with programs larger than
+the Stanford benchmark suite, we, as shown before, have been able to
+deal with remarkably larger programs.
+
+%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
+\section{Design Space}
+\label{sec:design}
+
+As we see from a cursory glance at Figure~\ref{fig:algorithm}, our range
+analysis algorithm has many optional modules.
+These modules give the user the chance to choose between more precise results,
+or a faster analysis.
+Given the number of options, the design space of a range analysis algorithm
+is vast.
+In this section we try to cover some of the most important tradeoffs.
+Figure~\ref{fig:space} plots, for the integer programs in the SPEC CPU 2006
+benchmark suite, precision versus speed for different configurations of
+our implementation.
 Our initial goal when developing this analysis was to support a bitwidth-aware
@@
 propagation step could not remove.
-Our results for the SPEC programs were disappointingly: on the average, the
-intra-procedural version of our analysis saves 5.23\% of the total program
-bitwidth.
-By using the inter-procedural version, with function inlining to simulate
-context-sensitiveness, we can increase this number to 8.89\%.
-A manual inspection of the SPEC programs reveal that this result is expected:
-they manipulate files, and their code do no provide enough
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{images/space}
+\end{center}
+\caption{\label{fig:space}
+Design space exploration: precision (percentage of bitwidth reduction)
+versus speed (secs) for different configurations of our algorithm analyzing
+the SPEC CPU 2006 integer benchmarks.}
+\end{figure}
+
+\subsection{Strongly Connected Components}
+\label{sub:sccs}
+
+The greatest source of improvement in our implementation is the use of strongly
+connected components.
+In order to propagate ranges across the constraint graph, we fragment it
+into strongly connected components, collapse each of these components into
+single nodes, and sort the resulting directed acyclic graph topologically.
+We then solve the range analysis problem for each component individually.
+Once we have solved a component, we propagate its ranges to the next
+components, and repeat the process until we walk over the entire constraint
+graph.
+It is well-known that this technique is essential to speedup constraint solving
+algorithms~\cite[Sec 6.3]{Nielson99}.
+In our case, the results are dramatic, mostly in terms of speed, but also in
+terms of precision.
+Figure~\ref{fig:impactSCC} shows the speedup that we gain by using strong
+components.
+We show results for the integer programs in the SPEC CPU 2006 benchmark suite.
+In some cases, as in \texttt{xalancbmk} the analysis on strong components is
+450x faster.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{images/impactSCC}
+\end{center}
+\caption{\label{fig:impactSCC}
+(Left) Bars give time to run our analysis without building strong components
+divided by time to run the analysis on strongly connected components.
+(Right) Bars give precision, in bitwidth reduction, that we obtain with strong
+components, divided by the precision that we obtain without them.
+}
+\end{figure}
+
+The strong components improve the precision of our growth analysis.
+According to Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
+strong components increase our precision by 40\%.
+The gains in precision happen because, by completely resolving a component,
+we are able to propagate constant intervals to the next components, instead
+of propagating intervals that can grow in both directions.
+An an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
+$[0, 99]$ from variable $k$ to the component that contains variable $j$.
+Had we run the analysis in the entire constraint graph, by the time we
+applied the growth analysis on $j$ we would still find $k$ bound to
+$0, +\infty$.
+
+\subsection{The choice of a program representation}
+\label{sub:program_rep}
+
+If strong components account for the largest gains in speed, the choice of
+a suitable program representation is responsible for the largest gains in
+precision.
+However, here we no longer have a win-win condition: a more expressive
+program representation decreases our speed, because it increases the
+size of the target program.
+We have tried our analysis in two different program representations: the
+Static Single Assignment (SSA) Form~\cite{Cytron91}, and the Extended Static
+Single Assignment (e-SSA) form~\cite{Bodik00}.
+The SSA form gives us a faster, albeit more imprecise, analysis.
+Any program in e-SSA form has also the SSA core property: any variable name
+has at most one definition site.
+The contrary is not true: SSA form programs do not have the core e-SSA
+property: any use site of a variable that appears in a conditional test
+post-dominates its definition.
+The program in Figure~\ref{fig:ex1}(b) is in e-SSA form.
+The live ranges of variables $i_1$ and $j_1$ have been split right after the
+conditional test via the assertions that creates variables $i_t$ and $j_t$.
+The e-SSA format serves well analyses that extract information from definition
+sites and conditional tests, and propagate this information forwardly.
+Examples include, in addition to range analysis, tainted flow
+analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/impactESSA}
+\end{center}
+\caption{\label{fig:impactESSA}
+(Left) Bars give the time to run analysis on e-SSA form programs divided by
+the time to run analysis on SSA form programs.
+(Right) Bars give the size of the e-SSA form program, in number of
+assembly instructions, divided by the size of the SSA form program.}
+\end{figure}
+
+Figure~\ref{fig:impactESSA} compares these two program representations in
+terms of runtime.
+As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
+analysis.
+In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
+time by 71\%.
+Runtime increases because the e-SSA form programs are larger than the SSA
+form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
+However, this growth is small: in none of the integer programs in SPEC
+CPU 2006 we verified an increase in code size of more than 9\%.
+But, if the e-SSA form slowdowns the analysis runtime, its gains in precision
+are remarkable, as seen in Figure~\ref{fig:precESSA}.
+These gains happen because the e-SSA format lets the analysis to use the
+results of comparisons to narrow the ranges of variables.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/precESSA}
+\end{center}
+\caption{\label{fig:precESSA}
+The impact of the e-SSA transformation on precision for three different
+benchmark suites. Bars give the ratio of precision (in bitwidth reduction),
+obtained with e-SSA form conversion divided by precision without e-SSA form
+conversion.}
+\end{figure}
+
+\subsection{Intra versus Inter-procedural Analysis}
+\label{sub:whole}
+
+A naive implementation of range analysis would be intra-procedural; that is,
+would solve the range analysis problem once per each function.
+However, we can gain in precision by performing it inter-procedurally.
+An inter-procedural implementation allows the results found for a function $f$
+to flow into other functions that $f$ calls.
+Figure~\ref{fig:intra} illustrates the inter-procedural analysis for the
+program seen in Figure~\ref{fig:ex1}(a).
+The trivial way to produce an inter-procedural implementation is to insert
+into the constraint system assignments from the actual parameter names to the
+formal parameter names.
+In our example of Figure~\ref{fig:intra}, our constraint graph contains a flow
+of information from $0$, the actual parameter, to $k_0$, the formal parameter
+of function \texttt{foo}.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{images/intra}
+\end{center}
+\caption{\label{fig:intra}
+Example where an intra-procedural implementation would lead to imprecise
+results.
+}
+\end{figure}
+
+Figure~\ref{fig:wholeImpact} compares the precision of the intra and
+inter-procedural analyses for the five largest programs in three different
+categories of benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
+Bitwise~\cite{Stephenson00}.
+Our results for the SPEC programs were disappointingly: on the average for
+the five largest programs, the intra-procedural version of our analysis saves
+5.23\% of bits per variable.
+The inter-procedural version increases this number to 8.89\%.
+A manual inspection of the SPEC programs reveals that this result is expected:
+these programs manipulate files, and their source codes do no provide enough
 explicit constants to power our analysis up.
@@
 Stanford benchmarks by 36.24\%.
-Notice that this number is better than the one we get with inlining (31.97\%),
-because functions in those benchmarks are called only at one location.
-Finally, for Bitwise we obtain a bitwidth reduction of 12.27\%.
-However, this average is brought down by two outliers: \texttt{edge\_detect} and
+For Bitwise we obtain a bitwidth reduction of 12.27\%.
+However, this average is lowered by two outliers: \texttt{edge\_detect} and
 \texttt{sha}, which cannot be reduced.
@@
 The bitwise programs contain only the \texttt{main} function; thus, different
-versions of our algorithm find the same results.
+versions of our algorithm find the same results when applied onto these
+programs.
+
+\subsection{Context Sensitive versus Context Insensitive Analysis}
+\label{sub:context}
+
+Another way to increase the precision of range analysis is via a
+context-sensitive implementation.
+Context-sensitiveness allows us to distinguish different calling sites of
+the same function.
+Figure~\ref{fig:context} shows why the ability to make this distinction is
+important for precision.
+In Figure~\ref{fig:context}(a) we have two different calls of function
+\texttt{foo}.
+An usual way to perform a data-flow analysis inter-procedurally is to
+create assignments between formal and actual parameters, as we show in
+Figure~\ref{fig:context}(b).
+If a function is called more than once, then its formal parameters will receive
+information from many actual parameters.
+We use $\phi$-functions to bind this information together into a single
+flow.
+However, in this case the multiple assignment of values to parameters
+makes the ranges of these parameters very large, whereas in reality they
+are not.
+A way to circumvent this source of imprecision is via function inlining,
+as we show in Figure~\ref{fig:context}(c).
+The results that we can derive for the transformed program are more
+precise, as each input parameter is assigned a single value.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{images/context}
+\end{center}
+\caption{\label{fig:context}
+Example where a context-sensitive implementation improves the results of
+range analysis.
+}
+\end{figure}
+
+Figure~\ref{fig:wholeImpact} also shows how function inlining modifies the
+precision of our results.
+It is difficult to find an adequate way to compare the precision of
+our analysis with, and without inlining.
+This difficulty stems from the fact that this transformation tends to change
+the target program too much.
+In absolute numbers, we always reduce the bitwidth of more variables after
+function inlining.
+However, proportionally function inlining leads to a smaller percentage of
+bitwidth reduction for many benchmarks.
+In the Stanford Collection, for instance, where most of the functions are
+called in only one location, inlining leads to worst precision results.
+On the other hand, for the SPEC programs, inlining, even in terms of
+percentage of reduction, tends to increase our measure of precision.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{images/wholeImpact}
+\end{center}
+\caption{\label{fig:wholeImpact}
+The impact of whole program analysis on precision. Each bar gives precision in
+\%bitwidth reduction.}
+\end{figure}
 
 \noindent
-\textbf{The impact of e-SSA on precision.}
-Figure~\ref{fig:essaDivSSA}(Middle) shows that the e-SSA transformation improves
-the precision of our analysis dramatically.
-If we convert the input programs from the original SSA representation used
-by LLVM to the e-SSA format, then, in some cases, e.g.,
-\texttt{Stanford.queens}, we can increase bitwidth reduction by 154x.
-This impressive difference happens because the standard SSA representation does
-not give our analysis any subsidy to acquire information from the outcome of
-conditionals.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=1\textwidth]{images/essaDivSSA}
-\end{center}
-\caption{\label{fig:essaDivSSA}
-The impact of different design trade-offs on precision of range analysis.}
-\end{figure}
-
-\noindent
-\textbf{The impact of the number of iterations before widening on precision.}
-It is well-known that premature widening might cause loss of
-information~\cite[Sec5.2]{Cousot09}.
-Figure~\ref{fig:essaDivSSA} shows how the precision of our analysis varies with
-the number of iterations used before resorting to widening.
-We have applied the widening operator in Figure~\ref{fig:growth_analysis}
-after each node was visited 1, 2 and 16 times.
+\textbf{Intra vs Inter-procedural runtimes.}
+Figure~\ref{fig:timeComp}(Right) compares three different execution modes.
+Bars are normalized to the time to run the intra-procedural analysis
+without inlining.
+On the average, the intra-procedural mode is 28.92\% faster than the
+inter-procedural mode.
+If we perform function inlining, then this difference is 45.87\%.
+These numbers are close because our runtime is bound to the
+size of the strong components.
+We have observed that function inlining does not increase too much these
+components.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=0.45\textwidth]{images/timeIntraInterInline}
+\end{center}
+\caption{\label{fig:timeComp}
+Runtime comparison between intra, inter and inter+inline versions of
+our algorithm.
+The bars are normalized to the time to run the intra-procedural analysis.
+}
+\end{figure}
+
+\subsection{Choosing a Widening Strategy}
+\label{sub:widen}
+
+We have implemented the widening operator used in the growth analysis in
+two different ways.
+The first way, which we call {\em simple}, is based on Cousot and
+Cousot's original widening operator~\cite{Cousot77}, and we have shown it
+in Figure~\ref{fig:growth_analysis}(Right).
+The second widening strategy, which we call {\em jump-set widening} consists
+in using the constants that appear in the program text, in sorted order, as
+the next limits of each interval after widening is applied.
+This operator is common in implementations of range
+analysis~\cite[p.228]{Nielson99}.
+Jump-set widening never produces worse results than the simple operator, and
+sometimes it does better.
+Figure~\ref{fig:jumpSet} shows an example taken from the code of
+SPEC CPU \texttt{bzip2}.
+Part of the constraint graph of the program in Figure~\ref{fig:jumpSet}(a)
+is given in Figure~\ref{fig:jumpSet}(b).
+The result of applying the simple operator is shown in
+Figure~\ref{fig:jumpSet}(c).
+Jump-set widening would use the lattice in Figure~\ref{fig:jumpSet}(d),
+instead of the lattice in Figure~\ref{fig:growth_analysis}(Left).
+This lattice yields the result given in Figure~\ref{fig:jumpSet}(e),
+which is more precise.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/jumpSet}
+\end{center}
+\caption{\label{fig:jumpSet}
+An example where jump-set widening is more precise.}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Another way to improve the precision of growth analysis is to perform a few
+rounds of abstract interpretation on the constraint graph, and, in case the
+process does not reach a fix point, only then to apply the widening operator.
+Each round of abstract interpretation consists in evaluating all the
+constraints, and then updating the intervals that change from one evaluation to
+the other.
+For instance, in Figure~\ref{fig:jumpSet} one round of abstract interpretation,
+coupled with the simple widening operator, would be enough to reach the
+fix point of that constraint system.
+We have experimented with 0 and 16 iterations before doing widening, and the
+overall result, for the programs in the SPEC CPU 2006 suite is given in
+Figure~\ref{fig:space}.
+Figure~\ref{fig:wideningPrec} shows some of these results in more details
+for the five largest benchmarks in this collection.
+In general jump-set widening improves the precision of our results in
+non-trivial ways.
+Nevertheless, the simple widening operator preceded by 16 rounds of
+abstract interpretation in general is more precise than jump-set widening
+without any cycle of pre-evaluation, as we see in Figure~\ref{fig:wideningPrec}.
+
+\begin{figure}[t!]
+\begin{small}
+\begin{center}
+\renewcommand{\tabcolsep}{0.2cm}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|} \hline
+Benchmark &    Size & 0 + Simple &   16 + Simple &     0 + Jump & 16 + Jump \\ \hline
+hmmer     &  38,409 &       9.98 & 11.40 (12.45) & 10.98 (9.11) & 11.40 (12.45) \\ \hline
+gobmk     &  84,846 &       8.15 &  9.93 (17.92) &  9.02 (9.64) & 10.13 (19.54) \\ \hline
+h264ref   &  97,494 &      12.58 &  13.11 (4.04) & 13.00 (3.23) & 13.11 (4.04) \\ \hline
+xalancbmk & 352,423 &       7.71 &   7.98 (3.38) &  7.95 (3.02) & 7.98 (3.38) \\ \hline
+gcc       & 449,442 &      16.09 &  16.63 (3.25) & 16.41 (1.95) & 16.64 (3.31) \\ \hline
+\end{tabular}
+\end{center}
+\end{small}
+\caption{\label{fig:wideningPrec}
+Variation in the precision of our analysis given the widening strategy.
+The size of each benchmark is given in number of variable nodes in the
+constraint graph.
+Precision is given in percentage of bitwidth reduction.
+Numbers in parenthesis are percentage of gain over 0 + Simple.}
+\end{figure}
 
@@
 analyze large programs.
-Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
-about 25 minutes to go over a program with almost 900 basic blocks.
-We analyze programs of similar size in seconds.
-We do not claim our approach is as precise as such algorithms, even though we
-are able to exactly analyze 4/5 of the examples presented in~\cite{Lakhdar11}.
-On the contrary, this paper presents a compromise between precision and speed
-that scales to very large programs.
 
@@
 On the other hand, we globally analyzed our largest benchmark, SPEC CPU 2006's
-\texttt{gcc}, enabling function inlining, in less than 15secs.
+\texttt{gcc}, enabling function inlining, in less than 15 seconds.
 Oh {\em et al.}'s implementation took orders of magnitude more time to go over
@@
 only solve range analysis on the interval lattice.
-
-% PLDI 2012 paper
 
@@
 invite the reader to check a recent report of ours~\cite{Tavares11b}.
-In this work we also show that the results that we obtain via the
+In that paper we also show that the results that we obtain via the
 sparse analysis are equivalent to the results provided by a dense analysis.

### 2012-05-16 doc/Manuscript/SPE/paper.tex
Added Vic's new chart.

--- 2012-05-10 doc/Manuscript/IST/paper.tex
@@
-\documentclass[11pt]{article}
+\documentclass[times]{speauth}
 
@@
 \usepackage{mathpartir} % To use \inferrule
+\usepackage{moreverb}   % Requirement from SPE. See if you can remove it.
+
+\usepackage[dvips,colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlco
+lor=red]{hyperref}      % Anther requirement from SPE.
+
+\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
+T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
+\def\volumeyear{2012}
 
@@
 
-\title{Global Range Analysis}
+\runningheads{V. H. S. Campos {\em et al.}}{Speed and Precision in Range Analysis}
+
+\title{Speed and Precision in Range Analysis}
+
 \author{Victor Hugo Sperle Campos, Douglas do Couto Teixeira \\
 Igor Rafael Assis Costa, Raphael Ernani Rodrigues and\\
-Fernando Magno Quint\~{a}o Pereira}
-\date{}
-\maketitle
+Fernando Magno Quint\~{a}o Pereira\corrauth}
+
+\address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil}
+
+\corraddr{fernando@dcc.ufmg.br}
 
@@
 
+\maketitle
+
 \section{Introduction}
@@
 redundant overflow tests~\cite{Sol11} and unnecessary array bound
-checks~\cite{Bodik00}.
+checks~\cite{Bodik00,Gampe11}.
 Additionally, range analysis is essential not only to the bitwidth aware
@@
 \begin{figure}[hbt]
+\begin{center}
 \begin{small}
 \begin{eqnarray*}
-\begin{array}{r@{\hspace{0.5cm}}c}
+\begin{array}{r@{\hspace{1.5cm}}c}
 Y = [l, u]
@@
 \end{small}
+\end{center}
 \end{figure}
@@
 \includegraphics[width=0.45\textwidth]{images/timeIntraInterInline}
+\includegraphics[width=0.485\textwidth]{images/timeWideningOperators}
 \end{center}
 \caption{\label{fig:timeComp}
-Runtime comparison between intra, inter and inter+inline versions of
+(Left) Runtime comparison between intra, inter and inter+inline versions of
 our algorithm.
+(Right) Runtime comparison between different widening operators.
 The bars are normalized to the time to run the intra-procedural analysis.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+By combining the different widening operators -- simple or jump-set -- with
+the different number of pre-evaluations -- in our case 0 or 16, we have four
+different widening strategies.
+Figure~\ref{fig:timeComp}(Right) compares the runtime of all these strategies
+for the integer programs in the SPEC CPU 2006 collection.
+We have observed no measurable different between the jump-set and the simple
+operator: the latter is 0.93\% faster, e.g., 1.0093x faster.
+The strategy that precedes the simple operator with 16 rounds of pre-evaluation
+is 7\% slower than the strategy that does not do any pre-evaluation.
+Finally, the combination of 16 rounds of pre-evaluation, plus jump-set widening
+is 13\% slower than the simple widening strategy.
+We observe an anomalous behavior in \texttt{astar} and \texttt{mcf}: the simple
+strategy results in a small slowdown.
+These benchmarks have the two fastest runtimes in the benchmark suite; e.g.,
+we analyze \texttt{mcf} in 0.02 seconds.
+Thus, we believe that the unexpected behavior is due to runtime noise that
+outlives a sequence of 100 executions of each benchmark.
+
 \section{Related works}
@@
 There have been many practical approaches to abstract interpretation,
-with special emphasis on range analysis~\cite{Bertrane10,Blanchet03,Cousot09,Jung05}.
+with special emphasis on range analysis~\cite{Bertrane10,Blanchet03,Cousot09,Gampe11,Jung05}.
 Cousot's group, for instance, has been able to globally analyze programs with
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{wileyj}
 \bibliography{../references}

### 2012-05-24 doc/Manuscript/SPE/paper.tex
Changed the journal paper a tiny bit.

@@
 which does not handle comparisons between variables, for instance.
-They also do not discuss the precision of their implementation, but only its
+They do not discuss the precision of their implementation, but only its
 runtime and memory consumption.
@@
 Our results are similar to Stephenson
-{\em et al}'s~\cite{Stephenson00}, even though out analysis does not require
+{\em et al}'s~\cite{Stephenson00}, even though our analysis does not require
 a backward propagation phase.
@@
 From the lattice $\varset{Z}$ we define the product lattice
-$\varset{Z}^2$, which is partially ordered by the subset relation
-$\sqsubseteq$.
-$\varset{Z}^2$ is defined as follows:
-
+$\varset{Z}^2$, which is defined as follows:
+%
 \begin{equation*}
-\varset{Z}^2 = \emptyset \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
+\varset{Z}^2 = \{ \emptyset \} \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
 \ z_1 \leq z_2, \  -\infty < z_2 \}
 \end{equation*}
+%
+This interval lattice is partially ordered by the subset relation, which we
+denote by ``$\sqsubseteq$".
+The meet operator ``$\sqcap$" is defined by:
+\[
+[a_1, a_2] \sqcap [b_1, b_2] =
+\begin{cases}
+[\mbox{max}(a_1, b_1), \mbox{min}(a_2, b_2)], \ \mbox{if} \ a_1 \leq b_1 \leq a_2 \ \mbox{or} \ b_1 \leq a_1 \leq b_2 \\
+[a_1, a_2] \sqcap [b_1, b_2] = \emptyset, \ \mbox{otherwise}
+\end{cases}
+\]
 
@@
 as those seen in Figure~\ref{fig:eval_function}(left).
-Figure~\ref{fig:eval_function}(right) defines a valuation function $e$ that
-computes $Y = f(\ldots)$, given I.
+We let the $\phi$-functions be as defined by Cytron
+{\em et al.}~\cite{Cytron91}: they join different variable names into a single
+definition.
+Figure~\ref{fig:eval_function}(right) defines a valuation function $e$ on the
+interval domain.
 Armed with these concepts, we define the range analysis problem as follows:
@@
 
-\begin{figure}[hbt]
+\begin{figure}[t!]
 \begin{center}
@@
 Figure~\ref{fig:ex1}(d).
-The SSA form, so conspicuous in modern compilers, leads to a very conservative
+The SSA form, so common in modern compilers, leads to a very conservative
 solution.
@@
 and $v_{2f} = v_2 \sqcap [\fun{ft}(v_1) + 1, +\infty]$.
+Notice that a variable $v$ can never be associated with a future bound to
+itself, e.g., $\fun{ft}(v)$.
+This invariant holds because whenever the e-SSA conversion associates a variable
+$u$ with $\fun{ft}(v)$, then $u$ is a fresh name created to split the live range
+of $v$, given that $v$ was used in a conditional.
 
@@
 known, we can replace $\fun{ft}(v)$ by an actual value.
-Once we are done placing copies, we insert $\phi$-functions into the
-transformed program to convert it to SSA form.
+After splitting the live ranges according to the rules stated above, we might
+have to insert $\phi$-functions into the transformed program to re-convert it to
+SSA form.
 This last step avoids that two different names given to the same original
@@
 standard e-SSA form.
+We have not created variable names for $i_f$ and $j_f$, because neither
+$i_1$ nor $i_j$ are dominated by the target of the conditional's else case.
+In this example, new $\phi$-functions are not necessary: new variable
+names are not alive together with the original variables.
 The part (b) of this figure shows the solution that we get to this new
@@
 To solve range analysis we find all the strongly connected components (SCCs) of
-the dependence graph and collapse them into single nodes; thus, obtaining a
+the dependence graph and collapse them into single nodes, obtaining a
 directed acyclic graph.
@@
 consideration.
+This flexibility is possible because we first discover how each variable in
+a strong component grows, before resolving future bounds.
 As we show in Section~\ref{sub:showdown}, most of the strong components in
@@
 handle small inputs.
+
+An interesting question is: can a variable $v$ be associated with a future
+bound of itself, e.g., $\fun{ft}(v)$?
+The answer is yes, but only if the program contains pathological comparisons
+between a variable and itself, e.g., $v < v$, or $v == v$.
+All these comparisons
 
@@
 Upon meeting this SCC, we have already determined that the interval
-$[0, 0]$ is bound to $I_0$ and that the interval $[100, 100]$ is bound to
+$[0, 0]$ is bound to $I_0$ and that the interval $[0, 99]$ is bound to
 $J_0$, as we show in Figure~\ref{fig:ex_graph}(a).
@@
 However, this average is lowered by two outliers: \texttt{edge\_detect} and
-\texttt{sha}, which cannot be reduced.
+\texttt{sha}, which have been purposely engineered to be resilient against
+range analyses~\cite{Stephenson00}.
 The bitwise benchmarks were implemented by Stephenson
-{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
+{\em et al.}~\cite{Stephenson00} to validate their intra-procedural bitwidth
+analysis.
 Our results are on par with those found by the original authors.
@@
 
-\subsection{Context Sensitive versus Context Insensitive Analysis}
+\subsection{Achieving Partial Context-Sensitiveness via Function Inlining}
 \label{sub:context}
@@
 \texttt{foo}.
-An usual way to perform a data-flow analysis inter-procedurally is to
-create assignments between formal and actual parameters, as we show in
-Figure~\ref{fig:context}(b).
-If a function is called more than once, then its formal parameters will receive
-information from many actual parameters.
+If we apply the trivial intra-procedural approach of Section~\ref{sub:whole},
+then we get the graph shown in Figure~\ref{fig:context}(b).
+In other words, if a function is called more than once, then its formal
+parameters will receive information from many actual parameters.
 We use $\phi$-functions to bind this information together into a single
@@
 size of the strong components.
-We have observed that function inlining does not increase too much these
-components.
+Although function inlining can increase the number of strongly connected
+components in the constraint graph, it cannot increase the size of the largest
+component, when compared to the simple inter-procedural analysis of
+Section~\ref{sub:whole}.
 
@@
 thousands of lines of code, albeit they use domain specific tools.
-Astr\'{e}, for example, analyzes programs that do not contain recursive calls.
+Astr\'{e}e, for example, analyzes programs that do not contain recursive calls.
 The work that is the closest to us is Oh {\em et al.}'s very recent abstract

### 2012-05-25 doc/Manuscript/SPE/paper.tex
Added more changes to the paper.

@@
+% TODO: discuss more related works: sparsity, and actual implementations.
+% TODO: Explain why we are using Stanford
 \documentclass[times]{speauth}
@@
 optimizations such as dead and redundant code elimination.
-This paper describes an inter-procedural range analysis of integer variables that
-scales up to programs with millions of assembly instructions.
+This paper describes an inter-procedural range analysis of integer variables
+that scales up to programs with millions of assembly instructions.
 Contrary to previous techniques, we handle comparisons between variables
 without resorting to relational lattices or expensive algorithms.
-We achieve path sensitiveness by using Bodik's Extended Static Single
-Assignment intermediate representation, and we obtain partial
+We gain precision from conditional tests by using Bodik's Extended Static
+Single Assignment intermediate representation, and we obtain partial
 context sensitiveness via function inlining.
@@
 Given the breath of our evaluation, we believe that this paper contains the most
-comprehensive empirical study of a range analysis algorithm ever presented in the
-compiler related literature.
+comprehensive empirical study of a range analysis algorithm ever presented in
+the compiler related literature.
 \end{abstract}
@@
 checks~\cite{Bodik00,Gampe11}.
-Additionally, range analysis is essential not only to the bitwidth aware
+Furthermore, range analysis is essential not only to the bitwidth aware
 register allocator~\cite{Barik06,Tallam03}, but also to more traditional
@@
 There are researchers who have implemented range analyses that scale up to
-large programs~\cite{Blanchet03,Patterson95,Venet04}; nevertheless, because the
+large programs~\cite{Patterson95,Blanchet03,Venet04}; nevertheless, because the
 algorithm itself is not the main focus of their works, they neither give
@@
 examples used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
-{\em et al.}~\cite{Lakhdar11}, who rely on much more costly methods.
+{\em et al.}~\cite{Lakhdar11}, who rely on more costly methods.
 
@@
 While designing and implementing our algorithm we had to face several important
-engineering choices.
+engineering decisions.
 Many approaches that we have used to increase the precision of
@@
 vast number of possibilities that it contains, but we discuss our most important
-implementation decisions in Section~\ref{sec:design}.
+implementation choices in Section~\ref{sec:design}.
 Section~\ref{sub:sccs} shows how we could improve runtime and precision
@@
 Wagner's constraint system, which can be solved exactly in polynomial
-time~\cite{Su04,Su05}.
+time~\cite{Su05,Su04}.
 However, although we could use their formulation to handle a subset of C-like
@@
 
-An interesting question is: can a variable $v$ be associated with a future
-bound of itself, e.g., $\fun{ft}(v)$?
-The answer is yes, but only if the program contains pathological comparisons
-between a variable and itself, e.g., $v < v$, or $v == v$.
-All these comparisons
-
 \subsection{Finding Ranges in Strongly Connected Components}
@@
 to each variable grows.
-We ensure termination of this phase via the widening operators that we
-discuss in Section~\ref{sub:widen}.
 The possible behaviors of an interval are:
@@
 (iv) grows in both directions.
-The lattice of abstract states, plus a constraint system representing
-Cousot and Cousot's widening operator~\cite[p.247]{Cousot77} is given in
-Figure~\ref{fig:growth_analysis}.
+We ensure termination of this phase via a {\em widening operator}.
+We have experimented with four different widening strategies, which we
+discuss in Section~\ref{sub:widen}.
+One of these strategies is based on Cousot and Cousot's widening
+operator~\cite[p.247]{Cousot77}.
+The lattice of abstract states, plus a constraint system representing this
+operator is given in Figure~\ref{fig:growth_analysis}.
 Because the lattice has height three, the intervals bound to each variable can
@@
 
-\noindent
-\textbf{Fixing futures.}
-The ranges found by the growth analysis tells us which variables have fixed
-bounds, independent on the intersections in the constraint system.
-Thus, we can use actual limits to replace intersections bounded by futures.
-Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
-In order to correctly replace a future $\fun{ft}(V)$ that limits a variable
-$V'$, we need to have already applied the growth analysis onto $V$.
-Had we considered only data dependence edges, then it would be possible
-that $V'$ be analyzed before $V$.
-However, because of control dependence edges, this case cannot happen.
-The control dependence edges ensure that any topological ordering of the
-constraint graph either places $N_v$ before $N_{v'}$, or places these nodes
-in the same strongly connected component.
-For instance, in Figure~\ref{fig:ex_graph}, variables $J_1$ and $I_t$ are in
-the same SCC only because of the control dependence edges.
-
 \begin{figure}[t!]
@@
 \caption{\label{fig:fix_intersects}Rules to replace futures by actual
-bounds. $S$ is the interval bound to each variable after the widening
-analysis.}
+bounds.}
 \end{figure}
@@
 
-\noindent
-\textbf{Narrowing Analysis.}
-The growth analysis associates very conservative bounds to each variable.
-Thus, the last step of our algorithm consists in narrowing these intervals.
-We accomplish this step via Cousot and Cousot's classic narrowing
-operator~\cite[248]{Cousot77}, which we show in
-Figure~\ref{fig:crop_analysis}.
-
 \begin{figure}[t!]
@@
 \noindent
-\textbf{Example:}
-Continuing with our example, Figure~\ref{fig:ex_partition_grow_crop} shows
-the application of our algorithm on the last strong component of
-Figure~\ref{fig:ex_graph}.
-Upon meeting this SCC, we have already determined that the interval
-$[0, 0]$ is bound to $I_0$ and that the interval $[0, 99]$ is bound to
-$J_0$, as we show in Figure~\ref{fig:ex_graph}(a).
-We are not guaranteed to find the least fix point of a constraint system.
-However, in this example we did it.
-We emphasize that finding this tight solution was only possible because of
-the topological ordering of the constraint graph in
-Figure~\ref{fig:ex_graph}.
-Had we applied the widening operator onto the whole graph, then we would
-have found out that variable $j_0$ is bound to $[-\infty, +\infty]$,
-because
-(i) it receives its interval directly from variable $k_t$, which is upper
-bounded by $+\infty$, and
-(ii) it is part of a negative cycle.
-On the other hand, because we only analyze $j$'s SCC after we have
-analyzed $k$'s, $k$ only contribute the constant range $[0, 99]$ to $j_0$.
-
-Figure~\ref{fig:ex_standard_eSSA}(b) shows our final solution for this example.
-This solution is very precise, in the sense that it is the maximum fixed
-point of the constraint system given in Figure~\ref{fig:eval_function}.
-However, the solution is still an over approximation of the dynamic behavior of
-the program in Figure~\ref{fig:ex_standard_eSSA}.
-For instance, we have found that variable $i$ could reach the upper value of
-99.
-In any actual run of the program, $i$ could be at most 50.
-Analysis on relational lattices, such as the polyhedron~\cite{Cousot78} or the
-Octagon~\cite{Mine06} domains, can infer such tighter bounds, as shown by
-Lakhdar-Chaouch {\em et al.}~\cite{Lakhdar11}.
-However, analyses on these higher dimensional domains are much more
-computationally expensive than analyses on the interval domain~\cite{Oh12}.
+\textbf{Fixing futures.}
+The ranges found by the growth analysis tells us which variables have fixed
+bounds, independent on the intersections in the constraint system.
+Thus, we can use actual limits to replace intersections bounded by futures.
+Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
+In order to correctly replace a future $\fun{ft}(v)$ that limits a variable
+$v'$, we need to have already applied the growth analysis onto $v$.
+Had we considered only data dependence edges, then it would be possible
+that $N_v'$ be analyzed before $N_v$.
+However, because of control dependence edges, this case cannot happen.
+The control dependence edges ensure that any topological ordering of the
+constraint graph either places $N_v$ before $N_{v'}$, or places these nodes
+in the same strongly connected component.
+For instance, in Figure~\ref{fig:ex_graph}, variables $j_1$ and $i_t$ are in
+the same SCC only because of the control dependence edges.
+
+\noindent
+\textbf{Narrowing Analysis.}
+The growth analysis associates very conservative bounds to each variable.
+Thus, the last step of our algorithm consists in narrowing these intervals.
+We accomplish this step via Cousot and Cousot's classic narrowing
+operator~\cite[p.248]{Cousot77}, which we show in
+Figure~\ref{fig:crop_analysis}.
 
@@
 
+% TODO: check if we say fix point or fixed point.
+\noindent
+\textbf{Example:}
+Continuing with our example, Figure~\ref{fig:ex_partition_grow_crop} shows
+the application of our algorithm on the last strong component of
+Figure~\ref{fig:ex_graph}.
+We are not guaranteed to find the least fixed point of a constraint system.
+However, in this example we did it.
+Figure~\ref{fig:ex_standard_eSSA}(b) shows our final solution for this example.
+This solution is very precise, in the sense that it is the least fixed
+point of the constraint system given in Figure~\ref{fig:eval_function}.
+However, the solution is still an over approximation of the dynamic behavior of
+the program in Figure~\ref{fig:ex1}(a).
+For instance, we have found that variable $i$ could reach the upper value of
+99.
+In any actual run of the program, $i$ could be at most 50.
+Analyses on relational lattices, such as the polyhedron~\cite{Cousot78} or the
+Octagon~\cite{Mine06} domains, can infer such tighter bounds, as shown by
+Lakhdar-Chaouch {\em et al.}~\cite{Lakhdar11}.
+However, analyses on these higher dimensional domains are much more
+computationally expensive than analyses on the interval domain~\cite{Oh12}.
+
+We emphasize that finding this tight solution was only possible because of
+the topological ordering of the constraint graph in Figure~\ref{fig:ex_graph}.
+Upon meeting the constraint graph's last SCC, shown in
+Figure~\ref{fig:ex_partition_grow_crop}, we had already determined that the
+interval $[0, 0]$ is bound to $i_0$ and that the interval $[0, 99]$ is bound to
+$j_0$, as we show in Figure~\ref{fig:ex_partition_grow_crop}(a).
+Had we applied the widening operator onto the whole graph, then we would
+have found out that variable $j_1$ is bound to $[-\infty, +\infty]$.
+This imprecision happens because, on one hand $j_1$'s interval is influenced
+by $k_t$'s, which is upper bounded by $+\infty$.
+On the other hand $j_1$ is part of a decreasing cycle of dependences formed by
+variables $j_t$ and $j_2$ in addition to itself.
+Therefore, if we had applied the widening phase over the entire program followed
+by a global narrowing phase, then we would not be able to recover some of
+widening's precision loss.
+However, because in this example we only analyze $j$'s SCC after we have
+analyzed $k$'s, $k$ only contribute the constant range $[0, 99]$ to $j_0$.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 of variable nodes in the constraint graph.
-We perform function inlining before running our analysis, to increase program
-sizes.
+We perform function inlining before running our analysis.
 Each point in the X line corresponds to a benchmark.
@@
 has 1,131 variable nodes in the constraint graph, in 20ms.
-We take 15,91secs to analyze our largest benchmark, \texttt{403.gcc}, which,
+We take 15.91 sec to analyze our largest benchmark, \texttt{403.gcc}, which,
 after function inlining, has 1,266,273 assembly instructions, and a
@@
 \caption{\label{fig:impactSCC}
-(Left) Bars give time to run our analysis without building strong components
+(Left) Time to run our analysis without building strong components
 divided by time to run the analysis on strongly connected components.
-(Right) Bars give precision, in bitwidth reduction, that we obtain with strong
+(Right) Precision, in bitwidth reduction, that we obtain with strong
 components, divided by the precision that we obtain without them.
@@
 form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
+The increase in size is due to the extra instructions used to partition the
+live ranges of variables after conditionals.
 However, this growth is small: in none of the integer programs in SPEC
@@
 These gains happen because the e-SSA format lets the analysis to use the
-results of comparisons to narrow the ranges of variables.
+results of conditional tests to narrow the ranges of variables.
 
@@
 \caption{\label{fig:wholeImpact}
-The impact of whole program analysis on precision. Each bar gives precision in
+The impact of whole program analysis on precision. Each bar shows precision in
 \%bitwidth reduction.}
@@
 rounds of abstract interpretation on the constraint graph, and, in case the
-process does not reach a fix point, only then to apply the widening operator.
+process does not reach a fixed point, only then to apply the widening operator.
 Each round of abstract interpretation consists in evaluating all the
@@
 coupled with the simple widening operator, would be enough to reach the
-fix point of that constraint system.
+fixed point of that constraint system.
 We have experimented with 0 and 16 iterations before doing widening, and the
@@
 We provide a gallery of examples, including source codes,
-CFGs and constraint graphs that we produce for meaningful programs at
-\texttt{http://code.google.com/p/range-analysis/wiki/gallery}.
+control flow graphs and constraint graphs that we produce for meaningful
+programs at \texttt{http://code.google.com/p/range-analysis/wiki/gallery}.
 

### 2012-05-26 doc/Manuscript/SPE/paper.tex
Last updates in the paper.

@@
 This last step avoids that two different names given to the same original
-variable be simultaneously alive at the program code.
+variable be simultaneously {\em alive} at the program code.
+A variable $v$ is alive at a program point $p$ if the program's control flow
+graph contains a path from $p$ to a site where $v$ is used, that does not
+go across any re-definition of $v$.
 Figure~\ref{fig:ex_standard_eSSA}(a) shows our running example changed into
@@
 
-\section{Related works}
+\section{Related work}
 \label{sec:rel}
 
+\noindent
+\textbf{The Scalability of Range Analysis. }
 % Early implementations of range analysis.
@@
 time algorithms to solve constraints on the interval
-lattice~\cite{Costan05,Gawlitza09,Lakhdar11,Su04,Su05}.
+lattice~\cite{Gawlitza09,Su05,Costan05,Lakhdar11,Su04}.
 However, these works are still very theoretical, and have not yet been used to
@@
 There have been many practical approaches to abstract interpretation,
-with special emphasis on range analysis~\cite{Bertrane10,Blanchet03,Cousot09,Gampe11,Jung05}.
+with special emphasis on range analysis~\cite{Gampe11,Blanchet03,Bertrane10,Cousot09,Jung05}.
 Cousot's group, for instance, has been able to globally analyze programs with
-thousands of lines of code, albeit they use domain specific tools.
+thousands of lines of code, albeit using domain specific tools.
 Astr\'{e}e, for example, analyzes programs that do not contain recursive calls.
@@
 Oh {\em et al.} discuss an implementation of range analysis on the interval
-lattice that scales up to a program with 1,363K LoC (ghostscript-9.00), but
-they do not provide results about its precision.
+lattice that scales up to a program with 1,363K LoC (ghostscript-9.00).
+Because their focus is speed, they do not provide results about precision.
 We could not find the benchmarks used in those experiments for a
@@
 
+\noindent
+\textbf{Sparse Data-Flow Analyses. }
+In this paper we provide a {\em sparse} implementation of range analysis.
+Sparsity, in our context, means that we associate points in the lattice of
+interest -- intervals in our case -- directly to variables.
+Dense analyses map such information to pairs formed by variables and program
+points.
+The compiler related literature contains many descriptions of sparse
+data-flow analyses.
+Some among these analyses obtain sparsity by using specific program
+representations, like we did.
+Others rely on data-structures.
+In terms of data-structures, the first, and best known method proposed to
+support sparse data-flow analyses is Choi {\em et al.}'s {\em Sparse Evaluation
+Graph} (SEG)~\cite{Choi91}.
+The nodes of this graph represent program regions where information produced by
+the data-flow analysis might change.
+Choi {\em et al.}'s ideas have been further expanded, for example, by Johnson
+{\em et al.}'s {\em Quick Propagation Graphs}~\cite{Johnson93}, or Ramalingan's
+{\em Compact Evaluation Graphs}~\cite{Ramalingan02}.
+Building upon Choi's pioneering work, researchers have developed many
+efficient ways to build such graphs~\cite{Pingali95,Pingali97,Johnson94}.
+These data-structures have been shown to improve many data-flow analyses in
+terms of runtime and memory consumption.
+Nevertheless, the elegance of SEGs and its successors have not, so far, been
+enough to attract the attention of mainstream compiler writers.
+Compilers such as gcc, LLVM or Java Hotspot rely, instead, on several types of
+program representations to provide support to sparse data-flow analyses.
+
+Most eminent among these representations is the Static Single Assignment
+form~\cite{Cytron91}, which suits well forward flow analyses, such as reaching
+definitions.
+Since its debut, the SSA form has been expanded in different ways.
+For instance, the Gated SSA form allows the static association of logical
+predicates with data-flow paths~\cite{Ottenstein90,Tu95}.
+Scott Ananian has introduced in the late nineties the {\em Static Single
+Information} (SSI) form, a program representation that supports both forward and
+backward analyses~\cite{Ananian99}.
+This representation was later revisited by Jeremy Singer~\cite{Singer06} and,
+a few years later, by Boissinot {\em et al.}~\cite{Benoit09}.
+Singer provided new algorithms plus examples of applications that benefit from
+the SSI form, and Boissinot {\em et al.}, in an effort to clarify some
+misconceptions about this program representation, introduced the notions of
+{\em week} and {\em strong} SSA form.
+Another important representation, which supports data-flow analyses that
+acquire information from uses, is the \emph{Static Single Use} form (SSU).
+There exists many variants of
+SSU~\cite{Plevyak96,George03,Lo98}.
+For instance, the ``strict'' SSU form enforces that each definition reaches a
+single use, whereas SSI and other variations of SSU allow two consecutive uses
+of a variable on the same path.
+The program representation that we have used in this paper -- the {\em Extended
+Static Single Assignment} (e-SSA) form -- was introduced by Bodik
+{\em et al.}~\cite{Bodik00}.
+
+There are so many different program representations because they fit specific
+data-flow problems.
+Each representation, given a domain of application, provides the following
+property: the information associated with the live range of a variable is
+invariant along every program point where this variable is alive.
+There are two key aspects that distinguish one representation from the others:
+firstly, where the information about a variable is acquired, and secondly,
+how this information is propagated.
+The e-SSA form, for instance, supports flow analyses that obtain information
+both from variable definitions and conditional tests and propagate this
+information forwardly.
+Such analyses are also supported by the SSI form; hence, we could have used this
+other representation too.
+However, as we have shown in previous work, the e-SSA form is considerably
+more economical~\cite{Tavares11b}.
+
 \section{Final Considerations}
@@
 In other words, if the dense analysis tells us that variable $v$ is
-associated with the interval range $[l, u]$ at program point $i$, and $v'$
+associated with the interval $[l, u]$ at program point $i$, and $v'$
 is the new name of $v$, alive at $i$ in the e-SSA form program, then $v'$ is
@@
 publicly available at \texttt{http://code.google.com/p/range-analysis/}.
-This repository contains instructions about how to deploy and use our
+This repository contains instructions to deploy and use our
 implementation.
@@
 control flow graphs and constraint graphs that we produce for meaningful
-programs at \texttt{http://code.google.com/p/range-analysis/wiki/gallery}.
+programs at \\ \texttt{http://code.google.com/p/range-analysis/wiki/gallery}.
 

### 2012-06-30 doc/Manuscript/SBLP/paper.tex
Applied the reviews to the paper.

@@
-\documentclass{paper}
-\usepackage{sbc-template}
+\documentclass{llncs}
 
@@
 
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{definition}[theorem]{Definition}
+%\newtheorem{theorem}{Theorem}[section]
+%\newtheorem{corollary}[theorem]{Corollary}
+%\newtheorem{lemma}[theorem]{Lemma}
+%\newtheorem{definition}[theorem]{Definition}
 
@@
 
-\address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
+\institute{Department of Computer Science -- UFMG -- Brazil
 \email{\{victorsc,raphael,igor,fernando\}@dcc.ufmg.br}
@@
 perspective, have never been implemented in production
-compilers~\cite{Costan05,Gawlitza09,Lakhdar11,Su04,Su05}.
+compilers~\cite{Gawlitza09,Lakhdar11,Su04,Su05}.
 There are implementations of range analysis that deal with very large
@@
 LLVM compiler~\cite{Lattner04}.
-As we show in Section~\ref{sub:showdown}, our implementation, which is
-publicly available, is able to analyze programs with over one million assembly
-instructions in sixteen seconds.
-And our implementation is not a straw-man: it produces very precise results.
-We have compared the results produced by our implementation with the results
-obtained via a dynamic profiler, which we have also implemented.
-As we show in Section~\ref{sub:showdown}, when analyzing well-known numeric
-benchmarks we are able to estimate tight ranges for almost half of all the
-integer variables present in these programs.
-
-% Design space exploration:
 While designing and implementing our algorithm we had to face several important
@@
 decisions in Section~\ref{sec:design}.
-Section~\ref{sub:sccs} shows how we could improve runtime and precision
+Section~\ref{sub:sccs} shows how we can improve runtime and precision
 substantially by processing data-flow information in the strongly connected
@@
 Our first endeavor to implement such an algorithm was based on Su and
-Wagner's constraint system, which can be solved exactly in polynomial
-time~\cite{Su04,Su05}.
+Wagner's constraint system~\cite{Su04,Su05}.
 However, although we could use their formulation to handle a subset of C-like
@@
 Following Gawlitza {\em et al.}'s notation, we shall be performing arithmetic
-operations over the complete lattice
+operations over the lattice
 $\cal{Z} = \mathbb{Z} \cup \{-\infty, +\infty\}$, where the ordering is
-naturally given by $-\infty < \ldots < -2 < -1 < 0 < 1 < 2 < \ldots +\infty$.
+naturally given by $-\infty < \ldots -1 < 0 < 1 \ldots +\infty$.
+We let meet and join be the min and max operators respectively.
 For any $x > -\infty$ we define:
@@
 \begin{tabular}{lcl}
-$x + \infty = \infty, x \neq -\infty$ & \mbox{\hspace{0.1cm}} & $x - \infty = - \infty, x \neq +\infty$ \\
+$x + \infty = \infty$ & \mbox{\hspace{0.1cm}} & $x - \infty = - \infty$ \\
 $x \times \infty = \infty$ if $x > 0$ & & $x \times \infty = -\infty$ if $x < 0$ \\
@@
 From the lattice $\varset{Z}$ we define the product lattice
-$\varset{Z}^2$, which is partially ordered by the subset relation
-$\sqsubseteq$.
-$\varset{Z}^2$ is defined as follows:
-%
-\begin{equation*}
-\varset{Z}^2 = \emptyset \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
-\ z_1 \leq z_2, \  -\infty < z_2 \}
-\end{equation*}
-%
-The objective of range analysis is to determine a mapping
-$I: \varset{V} \mapsto \varset{Z}^2$ from the set of integer program variables
+$\varset{Z}^2$, partially ordered by the subset relation $\sqsubseteq$, and
+defined as $\varset{Z}^2 = \emptyset \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z}, \ z_1 \leq z_2, \  -\infty < z_2 \}$. The objective of range analysis is to determine a mapping
+$I: V \mapsto \varset{Z}^2$ from the set of integer program variables
 $V$ to intervals, such that, for any variable $v \in V$, if
 $I(v) = [l, u]$, then, during the execution of the target program, any
-valued $i$ assigned to $v$ is such that $l \leq i \leq u$.
+value $i$ assigned to $v$ is such that $l \leq i \leq u$.
 
@@
 \textbf{A Holistic View of our Range Analysis Algorithm.}
-We perform range analysis in a number of steps.
-First, we convert the program to a suitable intermediate representation that
-makes it easier to extract constraints.
-From these constraints, we build a dependence graph that allows us to do
-range analysis sparsely.
-Finally, we solve the constraints applying different fix-point iterators on
-this dependence graph.
-Figure~\ref{fig:algorithm} gives a global view of this algorithm.
-Some of the steps in the algorithm are optional.
-They improve the precision of the range analysis, at the expense of a longer
-running time.
+Figure~\ref{fig:algorithm} gives a global view of our range analysis
+algorithm.
+We perform range analysis in a number of steps, some of which are optional.
+The optional parts improve the precision of the range analysis, at the expense of
+a longer running time.
 In Section~\ref{sec:design} we discuss in more details these tradeoffs.
-The last phase, which we call the {\em micro algorithm}, happens per
-strong component; however, if we opted for not building these components,
-then it happens once for the entire constraint graph.
-Nevertheless, the use of strongly connected components
-is so essential for performance and precision, as we show in
-Section~\ref{sub:sccs}, that it is considered optional only because we
-can easily build our implementation without this module.
 
@@
 \caption{\label{fig:algorithm}
-Our implementation of range analysis. Rounded boxes are optional steps.}
+Our implementation of range analysis. Rounded boxes are optional modules.
+The grey box is a module implemented in LLVM; the other parts are our
+contributions.}
 \end{figure}
@@
 {\em et al.}~\cite{Bodik00}.
-We have removed the code that performs array manipulation from this program,
-as it plays no role in our explanation.
 Figure~\ref{fig:overall_view}(b) shows one possible way to represent this
@@
 As we explain in Section~\ref{sub:program_rep}, a good program
-representation allows us to find more precise results.
+representation helps range analysis to find more precise results.
 In this example we chose a program representation called
-Extended Static Single Assignment form, which lets us to solve range
+Extended Static Single Assignment form~\cite{Bodik00}, which lets us solve range
 analysis via a path sensitive algorithm.
+This representation uses the $\phi$-functions typical in SSA form
+programs~\cite{Cytron91}, plus {\em futures} (\textbf{ft}), which we shall
+define later.
 Figure~\ref{fig:overall_view}(c) shows the constraints that we extract from
@@
 process that we call the micro algorithm.
-This phase is further divided into three sub-steps:
+This process is divided into three sub-steps:
 (i) growth analysis;
@@
 might not terminate.
-In order to ensure termination, we must reckon on a technique called
+In order to ensure termination, we must rely on a technique called
 {\em widening}, first introduced by Cousot and Cousot as a key component
@@
 Figure~\ref{fig:fix_intersects} shows the rules to perform these substitutions.
-In order to correctly replace a future $\fun{ft}(V)$ that limits a variable
-$V'$, we need to have already applied the growth analysis onto $V$.
+In order to correctly replace a future $\fun{ft}(v)$ that limits a variable
+$v'$, we need to have already applied the growth analysis onto $v$.
 Had we considered only data dependence edges, then it would be possible
-that $V'$ be analyzed before $V$.
+that $v'$ be analyzed before $v$.
 However, because of control dependence edges, this case cannot happen.
@@
 in the same strongly connected component.
-For instance, in Figure~\ref{fig:overall_view}(b), variables $j_1$ and $i_t$
+For instance, in Figure~\ref{fig:overall_view}(d), variables $j_1$ and $i_t$
 are in the same SCC only because of the control dependence edges.
@@
 \begin{array}{c}
-\inferrule{Y = X \sqcap [l, \fun{ft}(V) + c] \\ \ub{I[V]} = u}
+\inferrule{Y = X \sqcap [l, \fun{ft}(V) + c] \\ \ub{I(V)} = u}
 {Y = X \sqcap [l, u + c]} \mbox{\hspace{0.3cm}} u, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
@@
 \\
-\inferrule{Y = X \sqcap [\fun{ft}(V) + c, u] \\ \lb{I[V]} = l}
+\inferrule{Y = X \sqcap [\fun{ft}(V) + c, u] \\ \lb{I(V)} = l}
 {Y = X \sqcap [l + c, u]} \mbox{\hspace{0.3cm}} l, c \in \mathbb{Z} \cup \{-\infty, +\infty\}
@@
 \caption{\label{fig:fix_intersects}Rules to replace futures by actual
-bounds. $S$ is the interval bound to each variable after the widening
-analysis.}
+bounds. Given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
+$\ub{\iota} = u$}
 \end{figure}
@@
 \begin{array}{c@{\hspace{0.9cm}}c}
-\inferrule{\lb{I[Y]} = -\infty \\ \lb{e(Y)} > -\infty}
-{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
+\inferrule{\lb{I(V)} = -\infty \\ \lb{e(V)} > -\infty}
+{I(V) \leftarrow [\lb{e(V)}, \ub{I(V)}]}
 &
-\inferrule{\lb{I[Y]} > \lb{e(Y)}}
-{I[Y] \leftarrow [\lb{e(Y)}, \ub{I[Y]}]}
+\inferrule{\lb{I(V)} > \lb{e(V)}}
+{I(V) \leftarrow [\lb{e(V)}, \ub{I(V)}]}
 \\
 \\
-\inferrule{\ub{I[Y]} = +\infty \\ \ub{e(Y)} < +\infty}
-{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
+\inferrule{\ub{I(V)} = +\infty \\ \ub{e(V)} < +\infty}
+{I(V) \leftarrow [\lb{I(V)}, \ub{e(V)}]}
 &
-\inferrule{\ub{I[Y]} < \ub{e(Y)}}
-{I[Y] \leftarrow [\lb{I[Y]}, \ub{e(Y)}]}
+\inferrule{\ub{I(V)} < \ub{e(V)}}
+{I(V) \leftarrow [\lb{I(V)}, \ub{e(V)}]}
 \end{array}
@@
 \end{center}
-\caption{\label{fig:crop_analysis}Cousot and Cousot's narrowing operator.}
+\caption{\label{fig:crop_analysis}Cousot and Cousot's narrowing operator.
+Function $e(V)$ is an abstract evaluation, on the interval lattice, of the
+instruction that produces $V$.}
 \end{figure}
@@
 $j_0$.
-We are not guaranteed to find the least fix point of a constraint system.
+We are not guaranteed to find the least fixed point of a constraint system.
 However, in this example we did it.
@@
 
-\subsection{Range Analysis Showdown}
-\label{sub:showdown}
-
-The objective of this section is to show, via experimental numbers, that our
-implementation of range analysis is fast, economic and effective.
-We have used it to analyze a test suite with 2.72 million lines of C code,
-which includes, in addition to all the benchmarks distributed with LLVM,
-the programs in the SPEC CPU 2006 collection.
-
-\noindent
-\textbf{Time and Memory Complexity}
-
-Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
-run our algorithm and the size of the input programs.
-We show data for the 100 largest benchmarks in our test suite, in number
-of variable nodes in the constraint graph.
-We perform function inlining before running our analysis, to increase program
-sizes.
-Each point in the X line corresponds to a benchmark.
-We analyze the smallest benchmark in this set, \texttt{Prolangs-C/deriv2}, which
-has 1,131 variable nodes in the constraint graph, in 20ms.
-We take 15,91secs to analyze our largest benchmark, \texttt{403.gcc}, which,
-after function inlining, has 1,266,273 assembly instructions, and a
-constraint graph with 679,652 variable nodes.
-For this data set, the coefficient of determination $(R^2)$ is 0.967, which
-provides very strong evidence about the linear asymptotic complexity of our
-implementation.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{images/TimeCorr}
-\end{center}
-\caption{\label{fig:TimeCorr}
-Correlation between program size (number of var nodes in constraint
-graphs after inlining) and analysis runtime (ms).
-Coefficient of determination = 0.967.
-}
-\end{figure}
-
-The experiments also reveal that the memory consumption of our implementation
-is linear with the program size.
-Figure~\ref{fig:MemCorr} plots these two quantities together.
-The linear correlation, in this case, is even stronger than that found in
-Figure~\ref{fig:TimeCorr}, which compares runtime and program size: the
-coefficient of determination is 0.9947.
-The figure only shows our 100 largest benchmarks.
-Again, SPEC \texttt{403.gcc} is the heaviest benchmark, requiring
-265,588KB to run.
-Memory includes stack, heap and the executable program code.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{images/MemCorr}
-\end{center}
-\caption{\label{fig:MemCorr}
-Comparison between program size (number of var nodes in constraint
-graphs) and memory consumption (KB).
-Coefficient of determination = 0.9947.
-}
-\end{figure}
-
-\noindent
-\textbf{Precision}
-
-Our implementation of range analysis is remarkably precise, considering its
-runtime.
-Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
-about 25 minutes to go over a program with almost 900 basic blocks.
-We analyze programs of similar size in less than one second.
-We do not claim our approach is as precise as such algorithms, even though we
-are able to find exact bounds to 4/5 of the examples presented
-in~\cite{Lakhdar11}.
-On the contrary, this paper presents a compromise between precision and speed
-that scales to very large programs.
-Nevertheless, our results are far from being trivial.
-We have implemented a dynamic profiler that measures, for each variable,
-its upper and lower limits, given an execution of the target program.
-Figure~\ref{fig:precision} compares our results with those measured
-dynamically for the Stanford benchmark suite, which is publicly
-available~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/precUpperBound}
-\includegraphics[width=\textwidth]{images/precLowerBound}
-\end{center}
-\caption{\label{fig:precision}
-(Upper) Comparison between static range analysis and dynamic profiler for
-upper bounds.
-(Lower) Comparison between static range analysis and dynamic profiler for
-lower bounds. The numbers above the benchmark names give the number of
-variables in each program.}
-\end{figure}
-
-We have classified the bounds estimated by the static analysis into four
-categories.
-The first category, which we call $1$, contains those bounds that are tight:
-during the execution of the program, the variable has been assigned an upper,
-or lower limit, that equals the limit inferred statically.
-The second category, which we call $n$, contains the estimated bounds that are
-within twice the value inferred statically.
-For instance, if the range analysis estimates that a variable $v$ is in the
-range $[0, 100]$, and during the execution the dynamic profiler finds that
-its maximum value is $51$, then $v$ falls into this category.
-The third category, $n^2$, contains variables whose actual value is within
-a quadratic factor from the estimated value.
-In our example, $v$'s upper bound would have to be at most $10$ for it to
-be in this category.
-Finally, the fourth category contains variables whose estimated value lays
-outside a quadratic factor of the actual value.
-We call this category {\em imprecise}, and it contains mostly the limits that
-our static analysis has marked as either $+\infty$ or $-\infty$.
-
-As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
-we have estimated statically are exact.
-Similarly, 51.99\% of our upper bounds are also tight.
-The figure also shows that, on average, 37.39\% of our lower limits are
-imprecise, and 35.40\% of our upper limits are imprecise.
-This result is on pair with those obtained by more costly analysis, such as
-Stephenson {\em et al.}'s~\cite{Stephenson00}.
-However, whereas that approach have not been used with programs larger than
-the Stanford benchmark suite, we, as shown before, have been able to
-deal with remarkably larger programs.
-
 %mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
@@
 In this section we try to cover some of the most important tradeoffs.
+All the numbers that we show have been obtained as the average of 15 runs in
+an Intel Duo CPU with a 3.00GHz clock, and 2.04GHz main memory.
 Figure~\ref{fig:space} plots, for the integer programs in the SPEC CPU 2006
@@
 connected components.
-In order to propagate ranges across the constraint graph, we fragment it
+To propagate ranges across the constraint graph, we fragment it
 into strongly connected components, collapse each of these components into
@@
 We show results for the integer programs in the SPEC CPU 2006 benchmark suite.
-In some cases, as in \texttt{xalancbmk} the analysis on strong components is
-450x faster.
+In \texttt{xalancbmk}, the analysis on strong components is 450x faster.
 
@@
 
-The strong components improve the precision of our growth analysis.
 According to Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
@@
 We have tried our analysis in two different program representations: the
-Static Single Assignment (SSA) Form~\cite{Cytron91}, and the Extended Static
+Static Single Assignment (SSA) form~\cite{Cytron91}, and the Extended Static
 Single Assignment (e-SSA) form~\cite{Bodik00}.
@@
 time by 71\%.
-Runtime increases because the e-SSA form programs are larger than the SSA
-form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
-However, this growth is small: in none of the integer programs in SPEC
-CPU 2006 we verified an increase in code size of more than 9\%.
-But, if the e-SSA form slowdowns the analysis runtime, its gains in precision
-are remarkable, as seen in Figure~\ref{fig:precESSA}.
+Runtime increases for two reasons.
+Firstly, the e-SSA form programs are larger than the SSA form programs, as we
+show in Figure~\ref{fig:impactESSA}(Right).
+However, this growth is small: we did not verify any growth larger than 9\% in
+any integer program of SPEC CPU 2006.
+Second, the e-SSA form program has futures; hence requiring the future resolution
+phase of our algorithm, which is not necessary in SSA form programs.
+Nevertheless, if the e-SSA form slow downs the analysis runtime, its gains in
+precision are remarkable, as seen in Figure~\ref{fig:precESSA}.
 These gains happen because the e-SSA format lets the analysis to use the
@@
 \caption{\label{fig:intra}
-Example where an intra-procedural implementation would lead to imprecise
-results.
+Example where an inter-procedural analysis is more precise than an
+intra-procedural analysis.
 }
@@
 Bitwise~\cite{Stephenson00}.
-Our results for the SPEC programs were disappointingly: on the average for
+Our results for the SPEC programs were disappointing: on average, for
 the five largest programs, the intra-procedural version of our analysis saves
@@
 A manual inspection of the SPEC programs reveals that this result is expected:
-these programs manipulate files, and their source codes do no provide enough
-explicit constants to power our analysis up.
+these programs use may external library functions, which we cannot analyze, and
+their source codes do no provide enough explicit constants to power our analysis
+up.
 However, with numerical benchmarks we fare much better.
-On the average our inter-procedural algorithm reduces the bitwidth of the
+On average, our inter-procedural algorithm reduces the bitwidth of the
 Stanford benchmarks by 36.24\%.
@@
 information from many actual parameters.
-We use $\phi$-functions to bind this information together into a single
+We use the SSA's $\phi$-functions to bind this information together into a single
 flow.
@@
 are not.
+As an example, in Figure~\ref{fig:context}(b), variable \texttt{k} ends up
+associated with the range $[0, 10^5]$, but in reality this variable contains
+an interval that is only 100 units long.
 A way to circumvent this source of imprecision is via function inlining,
@@
 
-Figure~\ref{fig:wholeImpact} also shows how function inlining modifies the
+Figure~\ref{fig:wholeImpact} shows how function inlining modifies the
 precision of our results.
@@
 without inlining.
-On the average, the intra-procedural mode is 28.92\% faster than the
-inter-procedural mode.
+On average, the intra-procedural mode is 28.92\% faster than the inter-procedural
+mode.
 If we perform function inlining, then this difference is 45.87\%.
@@
 \begin{array}{c@{\hspace{0.5cm}}c}
-\inferrule{I[Y] = [\bot, \bot]}{I[Y] \leftarrow e(Y)}
+\inferrule{I(V) = [\bot, \bot]}{I(V) \leftarrow e(V)}
 &
-\inferrule{\lb{e(Y)} < \lb{I[Y]} \\ \ub{e(Y)} > \ub{I[Y]}}
-{I[Y] \leftarrow [-\infty, +\infty ]}
+\inferrule{\lb{e(V)} < \lb{I(V)} \\ \ub{e(V)} > \ub{I(V)}}
+{I(V) \leftarrow [-\infty, +\infty ]}
 \\
 \\
-\inferrule{\lb{e(Y)} < \lb{I[Y]}}
-{I[Y] \leftarrow [-\infty, \ub{I[Y]}]}
+\inferrule{\lb{e(V)} < \lb{I(V)}}
+{I(V) \leftarrow [-\infty, \ub{I(V)}]}
 &
-\inferrule{\ub{e(Y)} > \ub{I[Y]}}
-{I[Y] \leftarrow [\lb{I[Y]}, +\infty]}
+\inferrule{\ub{e(V)} > \ub{I(V)}}
+{I(V) \leftarrow [\lb{I(V)}, +\infty]}
 \end{array}
@@
 (Right) Cousot and Cousot's widening operator. We evaluate the rules from
-left-to-right, top-to-bottom, and stop upon finding a pattern matching.
-Again: given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
-$\ub{\iota} = u$}
+left-to-right, top-to-bottom, and stop upon finding a pattern matching.}
 \end{figure}
@@
 Another way to improve the precision of growth analysis is to perform a few
-rounds of abstract interpretation on the constraint graph, and, in case the
-process does not reach a fix point, only then to apply the widening operator.
+rounds of abstract interpretation on the constraint graph, and to apply widening
+only if this process does not reach a fixed point.
 Each round of abstract interpretation consists in evaluating all the
@@
 Figure~\ref{fig:space}.
-Figure~\ref{fig:wideningPrec} shows some of these results in more details
+Figure~\ref{fig:wideningPrec} shows some of these results in more detail
 for the five largest benchmarks in this collection.
@@
 %mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
-\section{Conclusion}
+\section{Final Remarks}
 \label{sec:con}

### 2012-06-30 doc/Manuscript/SPE/paper.tex
Correção
Applied the reviews to the paper.

@@
 checks~\cite{Bodik00,Gampe11}.
-Furthermore, range analysis is essential not only to the bitwidth aware
-register allocator~\cite{Barik06,Tallam03}, but also to more traditional
-allocators that handle registers of different
-sizes~\cite{Kong98,Pereira08,Scholz02}.
+Furthermore, range analysis is essential to bitwidth aware register
+allocators~\cite{Barik06,Tallam03}, register allocators that handle registers of
+different sizes~\cite{Kong98,Pereira08,Scholz02}, and scratchpad cache
+allocators~\cite{Yanqin11}.
 Additionally, range analysis has also been used to statically predict the
@@
 literature is rich in works describing in details algorithmic variations of
-range analyses~\cite{Gawlitza09,Mahlke01,Stephenson00,Su05}.
+range analyses~\cite{Mahlke01,Gawlitza09,Stephenson00,Su05}.
 On the other hand, none of these authors provide experimental evidence that
@@
 provides an implementation of Stephenson's algorithm~\cite{Stephenson00}, and
-is available at \texttt{http://flex.cscott.net/Harpoon/}.} and
+is available at \texttt{http://flex.cscott.net/Harpoon/}.},
 gcc~\footnote{Gcc's VRP pass (at \texttt{http://gcc.gnu.org/svn/gcc/trunk/gcc/tree-vrp.c}) implements a variant of Patterson's
-algorithm~\cite{Patterson95}.} only deal with comparisons between variables and
+algorithm~\cite{Patterson95}.} or Mozilla's IonMonkey~\footnote{\texttt{hg.mozilla.org/projects/ionmonkey/file/629d1e6251b9/js/src/ion/RangeAnalysis.cpp}} only deal with comparisons between variables and
 constants.
@@
 We have not created variable names for $i_f$ and $j_f$, because neither
-$i_1$ nor $i_j$ are dominated by the target of the conditional's else case.
+$i_1$ nor $j_1$, the variables that have been renamed, are dominated by the
+target of the conditional's else case.
 In this example, new $\phi$-functions are not necessary: new variable
@@
 Had we considered only data dependence edges, then it would be possible
-that $N_v'$ be analyzed before $N_v$.
+that $N_{v'}$ be analyzed before $N_v$.
 However, because of control dependence edges, this case cannot happen.
@@
 available~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/precUpperBound}
+We chose Stanford because these benchmarks do not read data from external
+files; hence, imprecisions are due to library functions that we cannot
+analyze.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=0.97\textwidth]{images/precUpperBound}
 \includegraphics[width=\textwidth]{images/precLowerBound}
@@
 Stephenson {\em et al.}'s~\cite{Stephenson00}.
-However, whereas that approach have not been used with programs larger than
-the Stanford benchmark suite, we, as shown before, have been able to
-deal with remarkably larger programs.
+However, whereas that approach has only been tested on single functions,
+we have been able to deal with remarkably larger programs.
 
@@
 of propagating intervals that can grow in both directions.
-An an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
+As an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
 $[0, 99]$ from variable $k$ to the component that contains variable $j$.
@@
 time by 71\%.
-Runtime increases because the e-SSA form programs are larger than the SSA
-form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
-The increase in size is due to the extra instructions used to partition the
-live ranges of variables after conditionals.
+Runtime increases for two reasons.
+Firstly, the e-SSA form programs are larger than the SSA form programs, as we
+show in Figure~\ref{fig:impactESSA}(Right).
 However, this growth is small: in none of the integer programs in SPEC
 CPU 2006 we verified an increase in code size of more than 9\%.
-But, if the e-SSA form slowdowns the analysis runtime, its gains in precision
+Secondly, the e-SSA form program has futures; hence requiring the future
+resolution phase of our algorithm, which is not necessary in SSA form programs.
+Nevertheless, if the e-SSA form slowdowns the analysis runtime, its gains in precision
 are remarkable, as seen in Figure~\ref{fig:precESSA}.
@@
 Bitwise~\cite{Stephenson00}.
-Our results for the SPEC programs were disappointingly: on the average for
+Our results for the SPEC programs were disappointing: on the average for
 the five largest programs, the intra-procedural version of our analysis saves
@@
 A manual inspection of the SPEC programs reveals that this result is expected:
-these programs manipulate files, and their source codes do no provide enough
+these programs manipulate files, and their source codes do not provide enough
 explicit constants to power our analysis up.
@@
 \texttt{foo}.
-If we apply the trivial intra-procedural approach of Section~\ref{sub:whole},
+If we apply the trivial inter-procedural approach of Section~\ref{sub:whole},
 then we get the graph shown in Figure~\ref{fig:context}(b).
@@
 In the Stanford Collection, for instance, where most of the functions are
-called in only one location, inlining leads to worst precision results.
+called in only one location, inlining leads to worse precision results.
 On the other hand, for the SPEC programs, inlining, even in terms of
@@
 \caption{\label{fig:wholeImpact}
-The impact of whole program analysis on precision. Each bar shows precision in
-\%bitwidth reduction.}
+The impact of inter-procedural analysis on precision. Each bar shows precision
+in \%bitwidth reduction.}
 \end{figure}
@@
 \textbf{Intra vs Inter-procedural runtimes.}
-Figure~\ref{fig:timeComp}(Right) compares three different execution modes.
+Figure~\ref{fig:timeComp}(Left) compares three different execution modes.
 Bars are normalized to the time to run the intra-procedural analysis
@@
 for the integer programs in the SPEC CPU 2006 collection.
-We have observed no measurable different between the jump-set and the simple
+We have observed no measurable difference between the jump-set and the simple
 operator: the latter is 0.93\% faster, e.g., 1.0093x faster.
@@
 misconceptions about this program representation, introduced the notions of
-{\em week} and {\em strong} SSA form.
+{\em weak} and {\em strong} SSI form.
 Another important representation, which supports data-flow analyses that

### 2012-07-01 doc/Manuscript/SBLP/paper.tex
My revision of the paper.

@@
 $\cal{Z} = \mathbb{Z} \cup \{-\infty, +\infty\}$, where the ordering is
-naturally given by $-\infty < \ldots -1 < 0 < 1 \ldots +\infty$.
+naturally given by $-\infty < \ldots -1 < 0 < 1 \ldots < +\infty$.
 We let meet and join be the min and max operators respectively.
@@
 a longer running time.
-In Section~\ref{sec:design} we discuss in more details these tradeoffs.
+In Section~\ref{sec:design} we discuss in more detail these tradeoffs.
 
@@
 All the numbers that we show have been obtained as the average of 15 runs in
-an Intel Duo CPU with a 3.00GHz clock, and 2.04GHz main memory.
+an Intel Core 2 Quad processor with 2.4 GHz, and 3.5 GB of main memory.
 Figure~\ref{fig:space} plots, for the integer programs in the SPEC CPU 2006
@@
 applied the growth analysis on $j$ we would still find $k$ bound to
-$0, +\infty$.
+$[0, +\infty]$.
 
@@
 phase of our algorithm, which is not necessary in SSA form programs.
-Nevertheless, if the e-SSA form slow downs the analysis runtime, its gains in
+Nevertheless, whereas the e-SSA form slows down the analysis runtime, its gains in
 precision are remarkable, as seen in Figure~\ref{fig:precESSA}.
@@
 A manual inspection of the SPEC programs reveals that this result is expected:
-these programs use may external library functions, which we cannot analyze, and
-their source codes do no provide enough explicit constants to power our analysis
+these programs use many external library functions, which we cannot analyze, and
+their source codes do not provide enough explicit constants to power our analysis
 up.
@@
 are not.
-As an example, in Figure~\ref{fig:context}(b), variable \texttt{k} ends up
+As an example, in Figure~\ref{fig:context}(b), variable $k$ ends up
 associated with the range $[0, 10^5]$, but in reality this variable contains

### 2012-07-01 doc/Manuscript/SBLP/paper.tex
Fixed some errors.

@@
 \texttt{sha}, which cannot be reduced.
-The bitwise benchmarks were implemented by Stephenson
+The Bitwise benchmarks were implemented by Stephenson
 {\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
 Our results are on par with those found by the original authors.
-The bitwise programs contain only the \texttt{main} function; thus, different
+The Bitwise programs contain only the \texttt{main} function; thus, different
 versions of our algorithm find the same results when applied onto these

### 2012-07-05 doc/Manuscript/CLSC/paper.tex
Added CLSC journal paper.

--- 2012-06-30 doc/Manuscript/SPE/paper.tex
@@
 % TODO: Explain why we are using Stanford
-\documentclass[times]{speauth}
+\documentclass[preprint]{elsarticle}
 
@@
 \usepackage{mathpartir} % To use \inferrule
-\usepackage{moreverb}   % Requirement from SPE. See if you can remove it.
-
-\usepackage[dvips,colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlco
-lor=red]{hyperref}      % Anther requirement from SPE.
 
@@
 
+\journal{Computer Languages, Systems and Structures}
+
 \begin{document}
 
-\runningheads{V. H. S. Campos {\em et al.}}{Speed and Precision in Range Analysis}
+\begin{frontmatter}
 
@@
 Igor Rafael Assis Costa, Raphael Ernani Rodrigues and\\
-Fernando Magno Quint\~{a}o Pereira\corrauth}
+Fernando Magno Quint\~{a}o Pereira%\corrauth
+}
 
 \address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil}
-
-\corraddr{fernando@dcc.ufmg.br}
 
@@
 
-\maketitle
+\begin{keyword}
+kw1 \sep kw2 \sep kw3
+\end{keyword}
+
+\end{frontmatter}
 
@@
 
-\bibliographystyle{wileyj}
+\bibliographystyle{model3-num-names}
 \bibliography{../references}

### 2012-07-05 doc/Manuscript/CLSC/paper.tex
CLSC: Fixed authors and affiliation

@@
 
-\author{Victor Hugo Sperle Campos, Douglas do Couto Teixeira \\
-Igor Rafael Assis Costa, Raphael Ernani Rodrigues and\\
-Fernando Magno Quint\~{a}o Pereira%\corrauth
-}
-
-\address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil}
+%\author{Victor Hugo Sperle Campos, Douglas do Couto Teixeira \\
+%Igor Rafael Assis Costa, Raphael Ernani Rodrigues and\\
+%Fernando Magno Quint\~{a}o Pereira%\corrauth
+%}
+
+\author[llp]{Victor Hugo Sperle Campos}
+\ead{victorsc@dcc.ufmg.br}
+
+\author[llp]{Douglas do Couto Teixeira}
+\ead{douglas@dcc.ufmg.br}
+
+\author[llp]{Igor Rafael de Assis Costa}
+\ead{igor@dcc.ufmg.br}
+
+\author[llp]{Raphael Ernani Rodrigues}
+\ead{raphael@dcc.ufmg.br}
+
+\author[llp]{Fernando Magno Quint\~{a}o Pereira\corref{cor1}}
+\ead{fernando@dcc.ufmg.br}
+
+\cortext[cor1]{Corresponding author}
+
+\address[llp]{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil}
 

### 2012-08-09 doc/Manuscript/CGO/paper.tex
Added the CGO paper.

\documentclass[preprint]{sigplanconf}[10pt]

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{mathpartir} % To use \inferrule
\usepackage{amssymb}    % To use \mathbb

\newtheorem{definition}[section]{Definition}

% Definition of new commands that I use in this text:
\newcommand{\fun}[1]{\mbox{\textbf{#1}}}
\newcommand{\lb}[1]{#1_{\downarrow}}
\newcommand{\ub}[1]{#1_{\uparrow}}
\newcommand{\varset}[1]{\mbox{$\cal{#1}$}}
\newcommand{\code}[1]{\textbf{#1}}

\begin{document}

\conferenceinfo{CGO '13}{February 23 - February 27. Shenzhen, China} 
\copyrightyear{2013}
\copyrightdata{[to be supplied]}

\titlebanner{Dinamica EGO}        % These are ignored unless
\preprintfooter{Dinamica EGO}   % 'preprint' option specified.

\title{An Algorithm to Prune Integer Overflow Tests}
%\subtitle{A case study of the Dinamica EGO toolkit}

\authorinfo{Removed due to blind review}
           {No affiliation given}
           {No e-mail given}

\maketitle

\begin{abstract}
The integer primitive type has upper and lower bounds in many programming
languages, including C, C++ and Java.
This limitation might lead programs that manipulate large integer numbers to
produce unexpected results due to overflows.
There exists a plethora of works that instrument programs to track the occurrence
of these overflows.
In this paper we present an algorithm that relies on static range analysis to
eliminate this instrumentation whenever possible.
We present a novel live range splitting strategy that refine the precision of
our range analysis.
We have used this algorithm to improve the quality of a dynamic instrumentation
library that we have implemented in LLVM.
This framework has been used to detect overflows in hundreds of C/C++ programs.
As a testimony of its effectiveness, our range analysis has been able to speedup
the SPEC CPU 2006 instrumented programs by XX\%.
\end{abstract}

\category{D.3.4}{Processors}{Compilers}

\terms
Languages, Performance

\keywords
Integer arithmetics, Overflow, Compiler, Range analysis

\section{Introduction}
\label{sec:int}

% What are integer overflows.
The most popular programming languages, including C, C++ and Java, limit the
size of primitive numeric types.
For instance, the \texttt{int} type, in Java, ranges from $-2^{31}-1$ to
$2^{31}$.
Consequently, there exist numbers that cannot be represented by these types.
In general, these programming languages resort to a {\em wrapping-arithmetics}
semantics~\cite{TODO} to perform integer operations.
If a number $n$ is too large to fit into a primitive data type $T$, then $N$'s
value wraps around, and $N \mbox{module} T_{max}$ wounds up represented
instead.
There are situations in which this semantics is acceptable~\cite{TODO}.
For instance, programers might rely on this behavior to implement hash functions
and random number generators.
On the other hand, there exist also situations in which this behavior might
lead a program to produce unexpected results.
As an example, in 19XX, XX happened because of an integer overflow~\cite{TODO}.
% TODO: we need to do some research to improve this paragraph:
% - We need to explain better the wrapping-arithmetics, and we need to cite it
% properly. What could be a good (official) citation?
% And we need to cite

% Why there is still the need to devise efficient methods to eliminate them.
A considerable amount of work has been done in the dynamic detection of
integer overflows~\cite{TODO}.
This approach consists in instrumenting the program in such a way that the
transformed binary is overflow aware.
Thus, if an overflow happens, then the instrumented program can take some
action, such as logging that event, or terminating the program.
There has been also some work aiming at reducing the instrumentation overhead
via static analysis~\cite{TODO}.
However, many authors, such as XX {\em et al.}~\cite[p.xx]{TODO} and
XX {\em et al.}~\cite[p.x]{TODO}, leave instrumentation pruning as future
work.
The goal of this paper is to bring this future back to the present.

% Our contribution: range analysis.
In this paper we discuss a simple and efficient range analysis algorithm that
we have developed to eliminate overflow checks in instrumented programs.
Our algorithm has three core insights.
Firstly, we rely on a three-phases approach to extract information from
comparisons between variables, e.g., $x < y$.
Previous algorithms either deal with these comparisons by resorting to
expensive relational analyses~\cite{Cousot78,Lakhdar11,Mine06}, or only
consider comparisons between variables and
constants~\cite{Mahlke01,Patterson95,Stephenson00}.
Secondly, we rely on strongly connected components to achieve better speed and
precision.
It is well-known that this technique is effective in speeding up static analysis
algorithms~\cite[Sec 6.3]{Nielson99}.
However, given our three-phases approach, we also achieve better precision
by solving strong components in a topological ordering.
Finally, we propose a program representation that allows us to perform a sparse
range analysis while taking benefit from the knowledge that the execution of
an instrumented program is overflow-free.
This extensive live range splitting strategy is only valid if the instrumented
program terminates whenever an integer overflow is detected.
If we cannot rely on this guarantee, then our live range splitting strategy
is more conservative, and produces the program representation that Bodik
{\em et al.}~\cite{Bodik00} have called the Extended Static Single Assignment
form.

% Our experimental results.
We use our range analysis to reduce the runtime overhead imposed by a dynamic
instrumentation library.
This instrumentation framework, also a contribution of this paper, has been
implemented in the LLVM compiler~\cite{Lattner04}.
We have used it to log overflows in a vast number of programs, and in this
paper we focus on SPEC CPU 2006.
Our implementation has been able to reproduce the results recently obtained by XX
{\em et al.}~\cite{TODO}.
Our range analysis algorithm has been able to remove XX\% of the overflow
checks created by the dynamic instrumentation framework.
All our experiments use an interprocedural version of range analysis.
We inline small functions to make our implementation more context sensitive.
Our experiments show that the speed and memory consumption of our range analysis
grows linearly with the program size.
By comparing the ranges that we determine statically with the results found by
a profiler, we see that more than 50\% of the bounds that we found are within a
factor of 50\% of the values that variables actually assume.

\section{Background}
\label{sec:bck}

% A brief overview about integer overflows.

% Harmless and malefic overflows.

\section{Range Analysis}
\label{sec:range}

\subsection{The Interval Lattice}
\label{sub:lattice}

% Define the lattice, the constraints and the valuation I.
Following Gawlitza {\em et al.}'s notation~\cite{Gawlitza09}, we shall be
performing arithmetic operations over the complete lattice
$\cal{Z} = \mathbb{Z} \cup \{-\infty, +\infty\}$, where the ordering is
naturally given by $-\infty < \ldots < -2 < -1 < 0 < 1 < 2 < \ldots +\infty$.
For any $x > -\infty$ we define:

\begin{tabular}{lcl}
$x + \infty = \infty, x \neq -\infty$ & \mbox{\hspace{0.1cm}} & $x - \infty = - \infty, x \neq +\infty$ \\
$x \times \infty = \infty$ if $x > 0$ & & $x \times \infty = -\infty$ if $x < 0$ \\
$0 \times \infty = 0$ & & $(-\infty) \times \infty = \ $ not defined $$ \\
\end{tabular}

From the lattice $\varset{Z}$ we define the product lattice
$\varset{Z}^2$, which is defined as follows:
%
\begin{equation*}
\varset{Z}^2 = \{ \emptyset \} \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
\ z_1 \leq z_2, \  -\infty < z_2 \}
\end{equation*}
%
This interval lattice is partially ordered by the subset relation, which we
denote by ``$\sqsubseteq$".
The meet operator ``$\sqcap$" is defined by:
\[
[a_1, a_2] \sqcap [b_1, b_2] =
\begin{cases}
[\mbox{max}(a_1, b_1), \mbox{min}(a_2, b_2)], \ \mbox{if} \ a_1 \leq b_1 \leq a_2  \\ \mbox{or} \ b_1 \leq a_1 \leq b_2 \\
[a_1, a_2] \sqcap [b_1, b_2] = \emptyset, \ \mbox{otherwise}
\end{cases}
\]

Given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
$\ub{\iota} = u$.
We let \varset{V} be a set of constraint variables, and
$I: \varset{V} \mapsto \varset{Z}^2$ a
mapping from these variables to intervals in $\varset{Z}^2$.
Our objective is to solve a constraint system $C$, formed by constraints such
as those seen in Figure~\ref{fig:eval_function}(left).
We let the $\phi$-functions be as defined by Cytron
{\em et al.}~\cite{Cytron91}: they join different variable names into a single
definition.
Figure~\ref{fig:eval_function}(right) defines a valuation function $e$ on the
interval domain.
Armed with these concepts, we define the range analysis problem as follows:

\begin{definition}
\label{def:rcp}
\textsc{Range Analysis Problem} \\
\textbf{Input:} a set $C$ of constraints ranging over a set \varset{V} of
variables. \\
\textbf{Output:} a mapping I such that, for any variable
$V \in \varset{V}$, e(V) = I[V].
\end{definition}

\begin{figure}[t!]
\begin{center}
\begin{small}
\begin{eqnarray*}
\begin{array}{r@{\hspace{.5cm}}c}
Y = [l, u]
&
e(Y) = [l, u]
\\
\\
Y = \phi (X_1, X_2)
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [l_1, u_1] \sqcap [l_2, u_2]}
\\
\\
Y = X_1 + X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [l_1 + l_2, u_1 + u_2]}
\\
\\
Y = X_1 \times X_2
&
\inferrule{L = \{l_1l_2, l_1u_2, u_1l_2, u_1u_2\} \\ I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [\mbox{min}(L), \mbox{max}(L)]}
\\
\\
Y = aX + b
&
\inferrule{I[X]=[l, u] \\ k_l = al + b \\ k_u = au + b}
{e(Y) = [\mbox{min}(k_l, k_u), \mbox{max}(k_l, k_u)]}
\\
\\
Y = X \sqcap [l', u']
&
\inferrule{I[X]=[l, u]}
{e(Y) \leftarrow [\mbox{max}(l, l'), \mbox{min}(u, u')]}
\end{array}
\end{eqnarray*}
\caption{\label{fig:eval_function}
A suite of constraints that produce an instance of the range analysis problem.}
\end{small}
\end{center}
\end{figure}

We will use the program in Figure~\ref{fig:ex1}(a) to illustrate our range
analysis.
Figure~\ref{fig:ex1}(b) shows the same program in e-SSA form~\cite{Bodik00},
and Figure~\ref{fig:ex1}(c) outlines the constraints that we extract from this
program.
There is a clear correspondence between instructions and constraints.
Our analysis is sparse~\cite{Choi91}; thus, we associate one, and only one,
constraint with each integer variable defined in the program.
A possible solution to the range analysis problem, as obtained via the
techniques that we will introduce in Section~\ref{sub:algo}, is given in
Figure~\ref{fig:ex1}(d).

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/ex_eSSA_cgo}
\end{center}
\caption{\label{fig:ex_eSSA_cgo}
(a) Example program.
(b) Control Flow Graph in SSA form.
(c) Constraints that we extract from the program.
(d) Possible solution to the range analysis problem.}
\end{figure}

\subsection{Widening, Future Resolution and Narrowing}
\label{sub:algo}

% The macro algorithm
The range analysis algorithm that we propose in this paper works in a number
of steps.
Firstly, we convert the program to an intermediate representation that gives
us subsidies to perform a sparse analysis.
We have tested our algorithm with two different representations, as we discuss
in Section~\ref{sub:splitting}.
Secondly, we extract constraints from the program representation.
Thirdly, we build a constraint graph, following the strategy pointed by
Su and Wagner~\cite{Su05}.
However, contrary to them, in a next phase we find the strongly connected
components in this graph, collapse them into super-nodes, and sort the
resulting digraph topologically.
Finally, for each strong component, we apply a three-phases approach to determine
the ranges of the variables.

\noindent
\textbf{Building the Constraint Graph. }
Given a set $\varset{C}$ of constraints, which define and/or use constraint
variables from a set $\varset{V}$, we build a constraint graph
$G = (\varset{C} \cup \varset{V}, E)$.
If variable $V \in \varset{V}$ is used in constraint $C \in \varset{C}$, then
we create an edge from $V$ to $C$.
If constraint $C \in \varset{C}$ defines variable $\varset{V} \in V$, then we
create an edge from $C$ to $V$.
Figure~\ref{fig:ex_graph} shows the constraint graph that we build for the
program in Figure~\ref{fig:ex_eSSA_cgo}(b).

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/ex_graph}
\end{center}
\caption{\label{fig:ex_graph}
The constraint graph that we build to the program in
Figure~\ref{fig:ex_eSSA_cgo}(b).}
\end{figure}

\subsection{Splitting Live Ranges after Uses of Variables}
\label{sub:splitting}

\section{The Dynamic Instrumentation Library}
\label{sec:dyn}

\section{Experimental Results}
\label{sec:exp}

\section{Related Work}
\label{sec:rel}

\section{Final Remarks}
\label{sec:rem}

\bibliographystyle{plain}
\bibliography{../references}

\end{document}

### 2012-08-10 doc/Manuscript/CGO/paper.tex
Added some more text in the paper.

@@
 % Why there is still the need to devise efficient methods to eliminate them.
-A considerable amount of work has been done in the dynamic detection of
-integer overflows~\cite{TODO}.
+There exist recent research about the dynamic detection of
+integer overflows~\cite{Brumley07,Dietz12}.
 This approach consists in instrumenting the program in such a way that the
@@
 There has been also some work aiming at reducing the instrumentation overhead
-via static analysis~\cite{TODO}.
+via static analysis~\cite{Molnar09,Zhang09}.
 However, many authors, such as XX {\em et al.}~\cite[p.xx]{TODO} and
@@
 
+\noindent
+\textbf{Propagating Ranges in Topological Order. }
+After building the constraint graph, we find its strongly connected components.
+We collapse these components in super nodes, and then proceed to propagate
+ranges along the resulting digraph.
+This approach is essential for scalability, because all the complexity of our
+algorithm lies in the resolution of strong components.
+As we have showed elsewhere~\cite{Victor12}, the vast majority of the strongly
+connected components are singletons.
+For instance, 99.11\% of the SCCs in SPEC CPU 2006 gcc (\texttt{403.gcc}) have
+only one node.
+Moreover, the composite components usually contain a small number of nodes.
+As an example, the largest component in \texttt{403.gcc}, has 2,131 nodes,
+even though gcc's constraint graph has over one million nodes.
+This large SCC exists due to a long chain of mutually recursive function calls.
+
 \subsection{Splitting Live Ranges after Uses of Variables}
@@
 
+Zhang {\em et al.}~\cite{Zhang09} have designed a static analysis to sanitize
+programs against integer overflow based vulnerabilities.
+They instrument integer operations in paths from a source to a sink.
+In Zhang {\em et al.}'s context, sources are functions that read values from
+users, and sinks are memory allocation operations.
+Thus, contrary to our work, Zhang {\em et al.}'s only need to instrument
+about 10\% of the integer operations in the program.
+However, they do not use any form of range analysis to limit the number of
+checks inserted in the transformed code.
+
 \section{Final Remarks}

### 2012-08-13 doc/Manuscript/CGO/paper.tex
Added more text to the paper.

@@
 techniques that we will introduce in Section~\ref{sub:algo}, is given in
-Figure~\ref{fig:ex1}(d).
+Figure~\ref{fig:ex_eSSA_cgo}(d).
 
@@
 
-\subsection{Widening, Future Resolution and Narrowing}
-\label{sub:algo}
+\subsection{Range Propagation}
+\label{sub:prop}
 
@@
 program in Figure~\ref{fig:ex_eSSA_cgo}(b).
+If $V$ is used by $C$ as the input of a future, then the edge from
+$V$ to $C$ represents what Ferrante {\em et al.} call a {\em control
+dependence}~\cite[p.323]{Ferrante87}.
+We use dashed lines to represent these edges.
+All the other edges denote {\em data dependences}~\cite[p.322]{Ferrante87}.
+As we will show later, control dependence edges increase the precision of our
+algorithm to solve future bounds.
 
@@
 algorithm lies in the resolution of strong components.
-As we have showed elsewhere~\cite{Victor12}, the vast majority of the strongly
-connected components are singletons.
+Our tests show that the vast majority of the strongly connected components are
+singletons.
 For instance, 99.11\% of the SCCs in SPEC CPU 2006 gcc (\texttt{403.gcc}) have
@@
 This large SCC exists due to a long chain of mutually recursive function calls.
+
+\subsection{A Three-Phases Approach to Solve Strong Components}
+\label{sub:micro}
+
+We determine the ranges for each strongly connected component in a three phases
+approach.
+For each SCC, first we determine the growth pattern of each variable via a
+widening phase.
+In the second step, once we know in which directions the values stored in each
+variable might grow, we replace futures by actual bounds.
+Finally, we have a narrowing phase that uses conditional tests to improve the
+precision of our results.
+
+\noindent
+\textbf{Widening: } we start solving constraints by determining how each program
+variable might grow.
+For instance, if a variable is only updated by sums with positive numbers, then
+we say that it only grows up.
+If, instead, a variable is only updated by sums with negative numbers, then it
+grows down.
+Some variables can also grow in both directions.
+We discover these growth patterns by abstractly interpreting the constraints that
+constitute the strongly connected component.
+We ensure termination via a widening operator.
+Our implementation uses {\em jump-set widening}, which is typically used in
+range analysis~\cite[p.228]{Nielson99}.
+This operator is a generalization of Cousot and Cousot's original widening
+operator~\cite{Cousot77}, which we describe below:
+%
+\begin{equation*}
+I[Y] =
+\begin{cases}
+\mbox{if} \ I[Y] = [\bot, \bot] \ \mbox{then} \ e(Y) \\
+\mbox{elif} \ \lb{e(Y)} < \lb{I[Y]}  \ \mbox{and} \ \ub{e(Y)} > \ub{I[Y]} \ \mbox{then} \ [-\infty, \infty] \\
+\mbox{elif} \ \lb{e(Y)} < \lb{I[Y]} \ \mbox{then} \ [-\infty, \ub{I[Y]}] \\
+\mbox{elif} \ \ub{e(Y)} > \ub{I[Y]} \ \mbox{then} \ [\lb{I[Y]}, \infty]
+\end{cases}
+\end{equation*}
+%
+We let $\lb{[l, u]} = l$ and $\ub{[l, u]} = u$.
+The map $I$ and the abstract evaluation function $e$ are determined as in
+Figure~\ref{fig:eval_function}.
+We have a different implementation of $e$ for each operation that the
+target programming language provides.
+Our current LLVM implementation has 18 different instances of $e$, including
+signed and unsigned addition, subtraction, multiplication and division, plus
+truncation, the bitwise integer operators and $\phi$-functions.
+
+If we use the widening operator above, then the abstract state of any constraint
+variable can only change three times, e.g., $[\bot, \bot] \rightarrow [c_1, c_2]
+\rightarrow [c_1, \infty] \rightarrow [-\infty, \infty]$, or
+$[\bot, \bot] \rightarrow [c_1, c_2] \rightarrow [-\infty, c_2]
+\rightarrow [-\infty, \infty]$.
+Therefore, we can determine the growth behavior of each constraint variable in
+a strong component in linear time on the number of constraints in that
+component.
+Figure~\ref{fig:ex_partition_grow_crop}(b) shows the abstract state of the
+variables in the largest SCC of the graph in Figure~\ref{fig:ex_graph}.
+As we see in the figure, this step of our algorithm has been able to determine
+that variables \texttt{i}$_1$, \texttt{i}$_2$ and \texttt{i}$_t$ can only
+increase, and that variables \texttt{j}$_1$, \texttt{j}$_2$ and \texttt{j}$_t$
+can only decrease.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=0.9\columnwidth]{images/ex_partition_grow_crop}
+\end{center}
+\caption{\label{fig:ex_partition_grow_crop}
+Four snapshots of the last SCC of Figure~\ref{fig:ex_graph}.
+(a) After removing control dependence edges.
+(b) After running the growth analysis.
+(c) After fixing the intersections bound to futures.
+(d) After running the narrowing analysis.}
+\end{figure}
+
+\noindent
+\textbf{Future resolution: }
+The next phase of the algorithm to determine intervals inside a strong
+component consists in replacing futures by actual bounds, a task that we
+accomplish by using the rules below:
+%
+\begin{eqnarray*}
+\begin{array}{c}
+\inferrule{Y = X \sqcap [l, \fun{ft}(V) + c] \\ \ub{I[V]} = u}
+{Y = X \sqcap [l, u + c]} \mbox{\hspace{0.3cm}} u, c \in \mathbb{Z} \cup \{-\infty, \infty\}
+\\
+\\
+\inferrule{Y = X \sqcap [\fun{ft}(V) + c, u] \\ \lb{I[V]} = l}
+{Y = X \sqcap [l + c, u]} \mbox{\hspace{0.3cm}} l, c \in \mathbb{Z} \cup \{-\infty, \infty\}
+\end{array}
+\end{eqnarray*}
+%
+In order to correctly replace a future $\fun{ft}(V)$ that limits a constraint
+variable $V'$, we need to have already applied the growth analysis onto $V$.
+Had we considered only data dependence edges, then it would be possible
+that $V'$ be analyzed before $V$.
+However, because of control dependence edges, this case cannot happen.
+The control dependence edges ensure that any topological ordering of the
+constraint graph either places $V$ before $V'$, or places these nodes
+in the same strongly connected component.
+For instance, in Figure~\ref{fig:ex_graph}, variables $j_1$ and $i_t$ are in
+the same SCC only because of the control dependence edges.
+Figure~\ref{fig:ex_partition_grow_crop}(c) shows the result of resolving
+futures in our running example.
+The information that we acquire from the growth analysis is essential in this
+phase.
+For instance, the growth analysis has found out that the value stored in variable
+\texttt{i}$_i$ can only increase.
+Given that this variable is assigned the initial value zero, we can replace
+$\fun{ft}(I_1)$ with this value.
+
+\noindent
+\textbf{Narrowing: } the last step that we apply on the strongly connected
+component is the narrowing phase.
+In this step we use values extracted from conditional tests to restrict the
+bounds of the constraint variables.
+We use the narrowing operator firstly proposed by Cousot and
+Cousot~\cite{Cousot77}, which we show below:
+%
+\begin{equation*}
+I[Y] =
+\begin{cases}
+\mbox{if} \ \lb{I[Y]} = -\infty  \ \mbox{and} \ \lb{e(Y)} > -\infty \ \mbox{then} \ [\lb{e(Y)}, \ub{I[Y]}] \\
+\mbox{elif} \ \ub{I[Y]} = \infty \ \mbox{and} \ \ub{e(Y)} < \infty \ \mbox{then}
+\ [\lb{I[Y]}, \ub{e(Y)}] \\
+\mbox{elif} \ \lb{I[Y]} > \lb{e(Y)} \ \mbox{then} \ [\lb{e(Y)}, \ub{I[Y]}] \\
+\mbox{elif} \ \ub{I[Y]} < \ub{e(Y)} \ \mbox{then} \ [\lb{I[Y]}, \ub{e(Y)}]
+\end{cases}
+\end{equation*}
+%
+Figure~\ref{fig:ex_partition_grow_crop}(d) show the result of our narrowing
+operator in our running example.
+Ranges initially improve due to the two conditional tests in the program.
+Firstly, we have that $I[I_t] = I[I_1] \sqcap [-\infty, 98]$, which gives us
+that $I[I_t] = [0, 98]$.
+We also have that $I[J_t] = I[J_1] \sqcap [0, \infty]$, giving
+$I[J_t] = [0, 99]$.
+From these new intervals, we can narrow the ranges bound to the other constraint
+variables.
+
+The combination of widening, futures and narrowing, plus use of strong components
+gives us, in this example, a very precise solution.
+We emphasize that finding this tight solution was only possible because of
+the topological ordering of the constraint graph in Figure~\ref{fig:ex_graph}.
+Upon meeting the constraint graph's last SCC, shown in
+Figure~\ref{fig:ex_partition_grow_crop}, we had already determined that the
+interval $[0, 0]$ is bound to $i_0$ and that the interval $[0, 99]$ is bound to
+$j_0$, as we show in Figure~\ref{fig:ex_partition_grow_crop}(a).
+Had we applied the widening operator onto the whole graph, then we would
+have found out that variable $j_1$ is bound to $[-\infty, +\infty]$.
+This imprecision happens because, on one hand $j_1$'s interval is influenced
+by $k_t$'s, which is upper bounded by $+\infty$.
+On the other hand $j_1$ is part of a decreasing cycle of dependences formed by
+variables $j_t$ and $j_2$ in addition to itself.
+Therefore, if we had applied the widening phase over the entire program followed
+by a global narrowing phase, then we would not be able to recover some of
+widening's precision loss.
+However, because in this example we only analyze $j$'s SCC after we have
+analyzed $k$'s, $k$ only contribute the constant range $[0, 99]$ to $j_0$.
 

### 2012-08-14 doc/Manuscript/CGO/paper.tex
Added new text to the paper.

@@
 
-\titlebanner{Dinamica EGO}        % These are ignored unless
+\titlebanner{Code Generation and Optimization}        % These are ignored unless
 \preprintfooter{Dinamica EGO}   % 'preprint' option specified.
@@
 denote by ``$\sqsubseteq$".
-The meet operator ``$\sqcap$" is defined by:
+Range intersection, ``$\sqcap$", is defined by:
 \[
@@
 \end{cases}
+\]
+And range union, ``$\sqcup$", is given by:
+\[
+[a_1, a_2] \sqcup [b_1, b_2] = [\mbox{min}(a_1, b_1), \mbox{max}(a_2, b_2)]
 \]
@@
 \inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
-{e(Y) = [l_1, u_1] \sqcap [l_2, u_2]}
+{e(Y) = [l_1, u_1] \sqcup [l_2, u_2]}
 \\
@@
 \inferrule{I[X]=[l, u]}
-{e(Y) \leftarrow [\mbox{max}(l, l'), \mbox{min}(u, u')]}
+{e(Y) \leftarrow [l, u] \sqcap [l', u']}
 \end{array}
@@
 
-\subsection{Splitting Live Ranges after Uses of Variables}
+\subsection{Live Range Splitting Strategies}
 \label{sub:splitting}
+
+A dense dataflow analysis associates information, i.e., a point in a lattice,
+with each pair formed by a variable plus a program point.
+If this information is invariant along every program point where the variable
+is alive, then we can associate the information with the variable itself.
+In this case, we say that the dataflow analysis is {\em sparse}~\cite{Choi91}.
+A dense dataflow analysis can be transformed into a sparse one via a suitable
+intermediate representation.
+A compiler builds this intermediate representation by splitting the live ranges
+of variables at the program points where the information associated with these
+variables might change.
+To split the live range of a variable $v$, at a program point $p$, we insert
+a copy $v' = v$ at $p$, and rename every use of $v$ that is dominated by $p$.
+In this paper we have experimented with two different live range splitting
+alternatives.
+
+The first strategy is the {\em Extended Static Single Assignment} (e-SSA) form,
+proposed by Bodik {\em et al.}~\cite{Bodik00}.
+We build the e-SSA representation by splitting live ranges at definition sites
+-- hence it subsumes the SSA form -- and at conditional tests.
+The program in Figure~\ref{fig:ex_eSSA_cgo}(b) is in e-SSA form.
+Let $(v < c)?$ be a conditional test, and let $l_t$ and $l_f$ be labels in
+the program, such that $l_t$ is the target of the test if the condition is true,
+and $l_f$ is the target when the condition is false.
+We split the live range of $v$ at any of these points if at least one of two
+conditions is true:
+(i) $l_f$ or $l_t$ dominate any use of $v$;
+(ii) there exist a use of $v$ at the dominance frontier of $l_f$ or $l_t$.
+For the notions of dominance and dominance-frontier, see Aho
+{\em et al.}~\cite[p.656]{Aho06}.
+To split the live range of $v$ at $l_f$ we insert at this
+program point a copy $v_f = v \sqcap [c, +\infty]$, where $v_f$ is a fresh name.
+We then rename every use of $v$ that is dominated by $l_f$ to $v_f$.
+Dually, if we must split at $l_t$, then we create at this point a copy
+$v_t = v \sqcap [-\infty, c-1]$, and rename variables accordingly.
+If the conditional uses two variables, e.g., $(v_1 < v_2)?$, then we create
+intersections bound to {\em futures}.
+We insert, at $l_f$, $v_{1f} = v_1 \sqcap [\fun{ft}(v_2), +\infty]$,
+and $v_{2f} = v_2 \sqcap [-\infty, \fun{ft}(v_1)]$.
+Similarly, at $l_t$ we insert
+$v_{1v} = v_1 \sqcap [-\infty, \fun{ft}(v_2) - 1]$
+and $v_{2f} = v_2 \sqcap [\fun{ft}(v_1) + 1, +\infty]$.
+Notice that a variable $v$ can never be associated with a future bound to
+itself, e.g., $\fun{ft}(v)$.
+This invariant holds because whenever the e-SSA conversion associates a variable
+$u$ with $\fun{ft}(v)$, then $u$ is a fresh name created to split the live range
+of $v$, given that $v$ was used in a conditional.
+
+The second intermediate representation consists in splitting live ranges at
+(i) definition sites -- it subsumes SSA, (ii) at conditional tests -- it
+subsumes e-SSA, and at some use sites.
+This representation, which we henceforth call u-SSA, is only valid if we
+assume that integer overflows cannot happen.
+We can provide this guarantee by using our dynamic instrumentation to terminate
+a program in face of an overflow.
+The rationale behind u-SSA is as follows: we know that past an instruction such
+as $v = u + c, c \in \mathbb{Z}$ at a program point $p$, variable $u$ must be
+less than $\mathit{MaxInt} - c$.
+If that were not the case, an overflow would have happened and the
+program would have terminated.
+Therefore, we split the live range of $u$ past its use point $p$, producing the
+sequence $v = u + c; u' = u$, and renaming every use of $u$ that is dominated
+by $p$ to $u'$.
+We then associate $u$ with the constraint $I[U'] \sqsubseteq I[U] \sqcap [-\infty, \mathit{MaxInt} - c]$.
 

### 2012-08-14 src/OverflowDetect/OverflowDetect.cpp
Renaming joao Pass to uSSA; Inserting cl::opt arguments and removing conditional defines Inserting new statistics in OverflowDetect Inserting static analysis that identifies possible overflows in compile-time

@@ -3,6 +3,7 @@
 // This file implements the LLVM "Range Analysis Instrumentation" pass 
 //
 //===----------------------------------------------------------------------===//
+#include "llvm/Type.h"
 #include "llvm/Constants.h"
 #include "llvm/Operator.h"
 #include "llvm/DerivedTypes.h"
@@ -16,19 +17,26 @@
 #include "llvm/Analysis/DebugInfo.h"
 #include "llvm/ADT/Statistic.h"
 #include "llvm/Support/raw_ostream.h"
+#include "llvm/Support/CommandLine.h"
+#include "../uSSA/uSSA.h"
 #include <vector>
 #include <list>
 #include <map>
 #include <stdio.h>
 
-// Use range analysis to avoid unnecessary overflow tests
-//#define RANGEANALYSIS
 
-#ifdef RANGEANALYSIS
 #include "../RangeAnalysis/RangeAnalysis.h"
-#else
-#define DEBUG_TYPE "OverflowDetect"
-#endif
+
+typedef enum { OvUnknown, OvCanHappen, OvWillHappen, OvWillNotHappen } OvfPrediction;
+
+static cl::opt<bool, false>
+UseRaPrunning("use-ra-prunning", cl::desc("Use range analysis to avoid inserting unnecessary checks."), cl::NotHidden);
+
+static cl::opt<bool, false>
+InsertAborts("insert-ovf-aborts", cl::desc("Abort the instrumented program when an overflow occurs."), cl::NotHidden);
+
+//Declaration of this argument is in uSSA.cpp. Public function is in uSSA.h
+static bool TruncInstrumentation = getTruncInstrumentation();
 
 using namespace llvm;
 
@@ -41,9 +49,9 @@ namespace {
         void printValueInfo(const Value *V);
 		void MarkAsNotOriginal(Instruction& inst);
 		void InsertGlobalDeclarations();
-		BasicBlock* NewOverflowOccurrenceBlock(Instruction* I, BasicBlock* NextBlock);
+		BasicBlock* NewOverflowOccurrenceBlock(Instruction* I, BasicBlock* NextBlock, Value *messagePtr);
 
-		void insertInstrumentation(Instruction* I);
+		void insertInstrumentation(Instruction* I, BasicBlock* AbortBB);
 		Constant* getSourceFile(Instruction* I);
 		Constant* getLineNumber(Instruction* I);
 
@@ -56,31 +64,29 @@ namespace {
[…]

### 2012-08-14 src/uSSA/uSSA.cpp

@@ -1,12 +1,20 @@
-#define DEBUG_TYPE "joao"
+#define DEBUG_TYPE "uSSA"
 
-#include "joao.h"
+#include "uSSA.h"
+#include "llvm/Support/CommandLine.h"
 
 using namespace llvm;
 
 static const std::string newdefstr = "_newdef";
 
-void joao::getAnalysisUsage(AnalysisUsage &AU) const {
+static cl::opt<bool, false>
+TruncInstrumentation("insert-trunc-checks", cl::desc("Insert overflow checks in the trunc instructions."), cl::NotHidden);
+
+bool getTruncInstrumentation(){
+	return TruncInstrumentation;
+}
+
+void uSSA::getAnalysisUsage(AnalysisUsage &AU) const {
 	AU.addRequiredTransitive<DominanceFrontier>();
 	AU.addRequiredTransitive<DominatorTree>();
 	
@@ -14,7 +22,7 @@ void joao::getAnalysisUsage(AnalysisUsage &AU) const {
 	AU.setPreservesCFG();
 }
 
-bool joao::runOnFunction(Function &F) {
+bool uSSA::runOnFunction(Function &F) {
 	DT_ = &getAnalysis<DominatorTree>();
 	DF_ = &getAnalysis<DominanceFrontier>();
 	
@@ -26,7 +34,7 @@ bool joao::runOnFunction(Function &F) {
 	return false;
 }
 
-void joao::createNewDefs(BasicBlock *BB)
+void uSSA::createNewDefs(BasicBlock *BB)
 {
 	for (BasicBlock::iterator BBit = BB->begin(), BBend = BB->end(); BBit != BBend; ++BBit) {
 		Instruction *I = BBit;
@@ -58,27 +66,29 @@ void joao::createNewDefs(BasicBlock *BB)
 				break;
 			
 			case Instruction::Trunc:
-				// Here we don't have a constant, only an interval due to the truncation number of bits
-				// Check if first operand is non-constant
-				op = I->getOperand(0);
 				
-				if (!isa<ConstantInt>(op)) {
-					std::string newname = op->getName().str() + newdefstr;
+				if (TruncInstrumentation){
+					// Here we don't have a constant, only an interval due to the truncation number of bits
+					// Check if first operand is non-constant
+					op = I->getOperand(0);
 					
-					BinaryOperator *newdef = BinaryOperator::Create(Instruction::Add, op, ConstantInt::get(op->getType(), 0), Twine(newname), next);
-					
[…]

### 2012-08-14 src/uSSA/uSSA.h

@@ -21,10 +21,10 @@
 
 namespace llvm {
 
-class joao : public FunctionPass {
+class uSSA : public FunctionPass {
 public:
 	static char ID; // Pass identification, replacement for typeid.
-	joao() : FunctionPass(ID) {}
+	uSSA() : FunctionPass(ID) {}
 	void getAnalysisUsage(AnalysisUsage &AU) const;
 	bool runOnFunction(Function&);
 	
@@ -36,5 +36,7 @@ private:
 	DominatorTree *DT_;
 	DominanceFrontier *DF_;
 };
+
+bool getTruncInstrumentation();
 
 }

### 2012-08-14 doc/Manuscript/CGO/paper.tex
Added more text to the paper.

@@
 less than $\mathit{MaxInt} - c$.
-If that were not the case, an overflow would have happened and the
+If that were not the case, then an overflow would have happened and the
 program would have terminated.
@@
 We then associate $u$ with the constraint $I[U'] \sqsubseteq I[U] \sqcap [-\infty, \mathit{MaxInt} - c]$.
+Figure~\ref{fig:ex_ir} compares the u-SSA form with the SSA and e-SSA
+intermediate program representations.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\columnwidth]{images/ex_ir}
+\end{center}
+\caption{\label{fig:ex_ir}
+(a) Example program.
+(b) SSA form~\cite{Cytron91}.
+(c) e-SSA form~\cite{Bodik00}.
+(d) u-SSA form.}
+\end{figure}
 

### 2012-08-14 src/OverflowDetect/OverflowDetect.cpp
Correcting a little bug

@@ -36,7 +36,7 @@ static cl::opt<bool, false>
 InsertAborts("insert-ovf-aborts", cl::desc("Abort the instrumented program when an overflow occurs."), cl::NotHidden);
 
 //Declaration of this argument is in uSSA.cpp. Public function is in uSSA.h
-static bool TruncInstrumentation = getTruncInstrumentation();
+bool TruncInstrumentation = getTruncInstrumentation();
 
 using namespace llvm;
 

### 2012-08-15 doc/Manuscript/CGO/paper.tex
Added new references to the paper.

@@
 lead a program to produce unexpected results.
-As an example, in 19XX, XX happened because of an integer overflow~\cite{TODO}.
-% TODO: we need to do some research to improve this paragraph:
-% - We need to explain better the wrapping-arithmetics, and we need to cite it
+As an example, in 1996, the Ariane 5 rocket was lost due to an arithmetic
+overflow -- a bug that resulted in a loss of more than U\$370 million
+dollars~\cite{Dowson97}.
+% TODO We need to explain better the wrapping-arithmetics, and we need to cite it
 % properly. What could be a good (official) citation?
@@
 % Why there is still the need to devise efficient methods to eliminate them.
-There exist recent research about the dynamic detection of
-integer overflows~\cite{Brumley07,Dietz12}.
-This approach consists in instrumenting the program in such a way that the
-transformed binary is overflow aware.
+Programming languages such as Ada or Lisp can be customized to throw exceptions
+whenever integer overflows are detected.
+Furthermore, there exist recent work proposing to instrument binaries derived
+from C, C++ and Java programs to detect the occurrence of overflows
+dynamically~\cite{Brumley07,Dietz12}.
 Thus, if an overflow happens, then the instrumented program can take some
-action, such as logging that event, or terminating the program.
-There has been also some work aiming at reducing the instrumentation overhead
-via static analysis~\cite{Molnar09,Zhang09}.
-However, many authors, such as XX {\em et al.}~\cite[p.xx]{TODO} and
-XX {\em et al.}~\cite[p.x]{TODO}, leave instrumentation pruning as future
-work.
-The goal of this paper is to bring this future back to the present.
+action, such as to log the event, or to terminate the program.
+However, this safety comes with a price: arithmetic operations need to be
+surveilled, and the runtime checks cost time.
+Zhang {\em et al.}~\cite{Zhang09} have proposed to eliminate some of this
+overhead via a tainted flow analysis.
+We have a similar goal, yet, our approach is substantially different.
 
@@
 
-\section{Background}
-\label{sec:bck}
+\subsection{The Danger that Integer Overflows Represent}
+\label{sub:danger}
 
 % A brief overview about integer overflows.
+The $n$-bits two's complement arithmetics wraps around at multiples of
+$2^n$.
+As an example the representation of $16 \times 16$ in 8-bits two's complement
+is $0$, because $0 \equiv (16\times16) \mod 2^8$.
+Programmers can use this well-defined behavior in benign ways.
+For instance, the wrapping arithmetics provides a cheap $\mod 2^n$ operator,
+which is used in the implementation of hash-functions or random number
+generators.
+However, integer overflows can also be the sources of bugs that are hard to
+identify.
+As pointed by Dietz {\em et al.}~\cite{Dietz12}, integer overflows are the
+sources of vulnerabilities in widely used software, such as OpenSSH and
+Firefox.
+These vulnerabilities could give an adversary the opportunity to execute
+arbitrary code in the compromised application.
 
-% Harmless and malefic overflows.
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\columnwidth]{images/ex_buffer_overflow}
+\end{center}
+\caption{\label{fig:ex_buffer_overflow}
+An example of an exploitable integer overflow vulnerability.}
+\end{figure}
+
+% An example of vulnerability.
+Figure~\ref{fig:ex_buffer_overflow} illustrates an example of integer
+overflow vulnerability.
+This vulnerability allows an adversary to perform a buffer overrun attack
+on a program that is apparently guarded against this type of exploit.
+For simplicity, we will use the C datatype \texttt{char}, which represents
+8-bit long numbers.
+Examples of actual vulnerabilities, with larger primitive types, are given
+by Brumley {\em et al.}~\cite{Brumley07}, Dietz {\em et al.}~\cite{Dietz12}, and
+Zhang {\em et al.}~\cite{Zhang09}.
+The function \texttt{read\_matrix} reads a matrix of bytes from a file.
+This matrix is stored as a linear sequence of bytes in a vector \texttt{out}.
+The first and second bytes in the file, e.g., \texttt{w} and \texttt{h},
+determine the number of rows and columns in the matrix.
+If the \texttt{char} datatype could represent arbitrarily long numbers, than this
+function would be safe against buffer overruns, because it would only write data
+on allocated memory.
+However, if, for instance, we have that \texttt{w}$= 16$ and \texttt{h}$= 16$,
+then \texttt{s}$= 0$, as pointed before.
+In this case, a buffer of length zero would be allocated, and all the data
+stored in the file would overwrite memory in the stack of activation records.
+By carefully crafting an input file, and adversary can, in this way, overwrite
+the return address of \texttt{read\_matrix}, diverting program's execution to
+one of \texttt{libc}'s function, such as \texttt{telnet}, for instance.
 
@@
 We then associate $u$ with the constraint $I[U'] \sqsubseteq I[U] \sqcap [-\infty, \mathit{MaxInt} - c]$.
-Figure~\ref{fig:ex_ir} compares the u-SSA form with the SSA and e-SSA
-intermediate program representations.
 
@@
 
+Figure~\ref{fig:ex_ir} compares the u-SSA form with the SSA and e-SSA
+intermediate program representations.
+We use the notation $v = \bullet$ to denote a definition of variable $v$, and
+$\bullet = v$ to denote a use of that variable.
+Figure~\ref{fig:ex_ir}(b) shows the example program converted to the SSA format.
+The different definitions of variable $u$ have been renamed, and a
+$\phi$-function re-combines these definitions into a single name.
+The SSA form sparsifies a dataflow analysis that only extracts information from
+the definition sites of variables, such as constant propagation.
+Figure~\ref{fig:ex_ir}(c) shows the same program in e-SSA form.
+This time we have renamed variable $v$ right after the conditional test where
+this variable is used.
+The e-SSA form serves dataflow analyses that acquire information from definition
+sites and conditional tests.
+Examples of these analyses include array bounds checking
+elimination~\cite{Bodik00} and traditional implementations of range
+analyses~\cite{Gough94,Patterson95}.
+Finally, Figure~\ref{fig:ex_ir}(d) shows our example in u-SSA form.
+The live range of variable $v_1$ has been divided right after its use.
+This representation is useful to analyses that learn information from the way
+that variables are used, and propagate this information forwardly.
+
 \section{The Dynamic Instrumentation Library}
@@
 
+\noindent
+\paragraph{Dynamic Detection of Integer Overflows: }
+Brumley \em{et al.}~\cite{Brumley07} have developed a tool, RICH, to detect
+integer integer overflows.
+Their approach consists in instrumenting every integer operation that might cause
+to an overflow, underflow, or data loss.
+Different from our method, RICH uses specific features of the x86 architecture
+to reduce the instrumentation overhead.
+Moreover, no static analysis is used to avoid unnecessary assertions, although
+this improvement is cited as future work.
+
+\paragraph{Static Detection of Integer Overflows: }
+\noindent
 Zhang {\em et al.}~\cite{Zhang09} have designed a static analysis to sanitize
@@
 checks inserted in the transformed code.
+Wang {\em et a.}~\cite{Wang09} have implemented a tool, IntScope, that combines symbolic execution and taint analysis to detect integer overflow vulnerabilities.
+The authors have been able to use this tool to successfully identify many vulnerabilities in industrial quality software.
+Our work, and Wang {\em et a.}Õs work are essentially different: they use symbolic execution, whereas we rely on
+range analysis.
+Furthermore, contrary to them, we do not try to detected statically the
+possibility of an integer overflow happening.
+Contrary to us, they do not transform the program to prevent or detect such event dynamically.
+Still in the field of symbolic execution, Molnar
+{\em et al.}~\cite{Molnar09} have implemented a tool, SmartFuzz, that analyzes
+Linux x86 binaries to find integer overflow bugs.
+They prove the existence of bugs by generating test cases for them.
+
+\noindent
+\paragraph{Intermediate Program Representations: }
 

### 2012-08-15 src/OverflowDetect/OverflowDetect.cpp

@@ -24,8 +24,9 @@
 #include <map>
 #include <stdio.h>
 
-
 #include "../RangeAnalysis/RangeAnalysis.h"
+
+using namespace llvm;
 
 typedef enum { OvUnknown, OvCanHappen, OvWillHappen, OvWillNotHappen } OvfPrediction;
 
@@ -38,7 +39,7 @@ InsertAborts("insert-ovf-aborts", cl::desc("Abort the instrumented program when 
 //Declaration of this argument is in uSSA.cpp. Public function is in uSSA.h
 bool TruncInstrumentation = getTruncInstrumentation();
 
-using namespace llvm;
+
 
 namespace {
 

### 2012-08-15 src/OverflowDetect/OverflowDetect.cpp

@@ -418,7 +418,9 @@ bool OverflowDetect::runOnModule(Module &M) {
 
 	NrSignedInsts = 0;
 	NrUnsignedInsts = 0;
-
+	NrInsts = 0;
+	NrOvfStaticallyDetected = 0;
+	NrPossibleOvfStaticallyDetected = 0;
 
 	//Insert the global declarations (fPrintf, stderr, etc...)
 	InsertGlobalDeclarations();

### 2012-08-15 src/uSSA/uSSA.h

@@ -37,6 +37,6 @@ private:
 	DominanceFrontier *DF_;
 };
 
-bool getTruncInstrumentation();
-
 }
+
+bool getTruncInstrumentation();

### 2012-08-29 src/OverflowDetect/OverflowDetect.cpp
Improving static analysis of unsigned instructions

@@ -175,7 +175,7 @@ OvfPrediction OverflowDetect::ovfStaticAnalysis(Instruction* I, InterProceduralR
 			
 			APInt MinValue = APInt::getSignedMinValue(numBits);
 			APInt MaxValue = APInt::getSignedMaxValue(numBits);
-			
+
 			//If needed, do the signed extend of everyone to the same bit width
 			if (MinValue.getBitWidth() < r.getUpper().getBitWidth()){
 				MinValue = MinValue.sext(r.getUpper().getBitWidth());
@@ -201,20 +201,29 @@ OvfPrediction OverflowDetect::ovfStaticAnalysis(Instruction* I, InterProceduralR
 			//FIXME: Our RA doesn't handle unsigned instructions very well.
 			// So, I'm using the signed max value to do the analysis. It needs to be fixed.
 			APInt MaxValue = APInt::getSignedMaxValue(numBits);
-			
+
 			//If needed, do the unsigned extend of everyone to the same bit width
 			if (MaxValue.getBitWidth() < r.getUpper().getBitWidth()){
 				Zero = Zero.zext(r.getUpper().getBitWidth());
 				MaxValue = MaxValue.zext(r.getUpper().getBitWidth());					
 			}
 
-			if (r.getLower().uge(Zero) && r.getUpper().ult(MaxValue)) {
+			/* We don't have the precision needed to say that an overflow will happen.
+			if (MaxValue.ult(r.getLower())) {
+				NrOvfStaticallyDetected++;
+				return OvWillHappen;
+			}
+			else*/ if (MaxValue.ult(r.getUpper())) {
+				NrPossibleOvfStaticallyDetected++;
+				return OvCanHappen;
+			}
+			else if (r.getLower().uge(Zero) && r.getUpper().ult(MaxValue)) {
 				NrPrunnedInsts++;
 				return OvWillNotHappen;
 			}
-			else {
+			else
 				return OvUnknown;
-			}
+
 		}
 
 	}
@@ -326,7 +335,7 @@ void OverflowDetect::InsertGlobalDeclarations(){
 	stringConstant = llvm::ConstantArray::get(*context, "(Suspected) Overflow occurred in %s, line %d. [%d]\n", true);
 	messageStr = new GlobalVariable(*module, stringConstant->getType(), true,
 	                                                llvm::GlobalValue::InternalLinkage,
-	                                                stringConstant, "TruncErrorMessage");
[…]

### 2012-08-29 doc/Manuscript/CLSC/paper.tex
CLSC: added hyperrefs
Changed paper into a technical report.

@@
 \usepackage{mathpartir} % To use \inferrule
-
-\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
-T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+\usepackage{hyperref}	% FIXME: Don't know if it's necessary
+
+%\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
+%T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
 
@@
 
-\title{Speed and Precision in Range Analysis}
+\title{Speed and Precision in Range Analysis \\
+Technical Report LLP002-2012}
 

### 2012-08-29 doc/Manuscript/CGO/paper.tex
Removed two directories.

@@
 \paragraph{Dynamic Detection of Integer Overflows: }
-Brumley \em{et al.}~\cite{Brumley07} have developed a tool, RICH, to detect
+Brumley {\em et al.}~\cite{Brumley07} have developed a tool, RICH, to detect
 integer integer overflows.

### 2012-08-29 doc/Manuscript/CGO/paper.tex
Added first experiments to the paper.

@@
 $G = (\varset{C} \cup \varset{V}, E)$.
+We call the vertices in \varset{C} the {\em constraint nodes}, and we call the
+nodes in \varset{V} the {\em variable nodes}.
 If variable $V \in \varset{V}$ is used in constraint $C \in \varset{C}$, then
@@
 
+We have implemented our range analysis algorithm in LLVM 3.0, and have run
+experiments on a Intel quad core CPU with a 2.40GHz clock, and 3.6GB of RAM.
+Each core has a 4,096KB L1 cache.
+We use Linux Ubuntu 10.04.4.
+Our implementation of range analysis has 3,958 lines of
+commented C++ code, our e/u-SSA conversion module has 672 lines, and our
+instrumentation pass has XXX lines.
+We have analyzed 428 C programs that constitute the LLVM test suite plus the
+integer benchmarks in SPEC CPU 2006.
+Together, these programs contain 4.76 million assembly instructions.
+This section has two main goals.
+First, we want to show that our range analysis is fast and precise.
+Second, we want to demonstrate the effectiveness of our framework to detect
+integer overflows.
+
+\noindent
+\textbf{Time and Memory Complexity.}
+Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
+run our range analysis and the size of the input programs.
+We show data for the 100 largest benchmarks in our test suite, in number
+of variable nodes in the constraint graph.
+We perform function inlining before running our analysis.
+Each point in the X line corresponds to a benchmark.
+We analyze the smallest benchmark in this set, \texttt{Prolangs-C/deriv2}, which
+has 1,131 variable nodes in the constraint graph, in 20ms.
+We take 15.91 sec to analyze our largest benchmark, \texttt{403.gcc}, which,
+after function inlining, has 1,266,273 assembly instructions, and a
+constraint graph with 679,652 variable nodes.
+For this data set, the coefficient of determination $(R^2)$ is 0.967, which
+provides very strong evidence about the linear asymptotic complexity of our
+implementation.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\columnwidth]{images/TimeCorr}
+\end{center}
+\caption{\label{fig:TimeCorr}
+Correlation between program size (number of var nodes in constraint
+graphs after inlining) and analysis runtime (ms).
+Coefficient of determination = 0.967.
+}
+\end{figure}
+
+The experiments also reveal that the memory consumption of our implementation
+is linear with the program size.
+Figure~\ref{fig:MemCorr} plots these two quantities together.
+The linear correlation, in this case, is even stronger than that found in
+Figure~\ref{fig:TimeCorr}, which compares runtime and program size: the
+coefficient of determination is 0.9947.
+The figure only shows our 100 largest benchmarks.
+Again, SPEC \texttt{403.gcc} is the heaviest benchmark, requiring
+265,588KB to run.
+Memory includes stack, heap and the executable program code.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\columnwidth]{images/MemCorr}
+\end{center}
+\caption{\label{fig:MemCorr}
+Comparison between program size (number of var nodes in constraint
+graphs) and memory consumption (KB).
+Coefficient of determination = 0.9947.
+}
+\end{figure}
+
+\noindent
+\textbf{Precision.}
+Our implementation of range analysis is remarkably precise, considering its
+runtime.
+Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
+about 25 minutes to go over a program with almost 900 basic blocks.
+We analyze programs of similar size in less than one second.
+We do not claim our approach is as precise as such algorithms, even though we
+are able to find exact bounds to 4/5 of the examples presented
+in~\cite{Lakhdar11}.
+On the contrary, we present a compromise between precision and speed
+that scales to very large programs.
+Nevertheless, our results are far from being trivial.
+We have implemented a dynamic profiler that measures, for each variable,
+its upper and lower limits, given an execution of the target program.
+Figure~\ref{fig:precision} compares our results with those measured
+dynamically for the Stanford benchmark suite, which is publicly
+available~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
+We chose Stanford because these benchmarks do not read data from external
+files; hence, imprecisions are due to library functions that we cannot
+analyze.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\columnwidth]{images/precUpperBound}
+\includegraphics[width=\columnwidth]{images/precLowerBound}
+\end{center}
+\caption{\label{fig:precision}
+(Upper) Comparison between static range analysis and dynamic profiler for
+upper bounds.
+(Lower) Comparison between static range analysis and dynamic profiler for
+lower bounds. The numbers above the benchmark names give the number of
+variables in each program.}
+\end{figure}
+
+We have classified the bounds estimated by the static analysis into four
+categories.
+The first category, which we call $1$, contains those bounds that are tight:
+during the execution of the program, the variable has been assigned an upper,
+or lower limit, that equals the limit inferred statically.
+The second category, which we call $n$, contains the bounds that are
+within twice the value inferred statically.
+For instance, if the range analysis estimates that a variable $v$ is in the
+range $[0, 100]$, and during the execution the dynamic profiler finds that
+its maximum value is $51$, then $v$ falls into this category.
+The third category, $n^2$, contains variables whose actual value is within
+a quadratic factor from the estimated value.
+In our example, $v$'s upper bound would have to be at most $10$ for it to
+be in this category.
+Finally, the fourth category contains variables whose estimated value lays
+outside a quadratic factor of the actual value.
+We call this category {\em imprecise}, and it contains mostly the limits that
+our static analysis has marked as either $+\infty$ or $-\infty$.
+As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
+we have estimated statically are exact.
+Similarly, 51.99\% of our upper bounds are also tight.
+The figure also shows that, on average, 37.39\% of our lower limits are
+imprecise, and 35.40\% of our upper limits are imprecise.
+This result is on pair with those obtained by more costly analysis, such as
+Stephenson {\em et al.}'s~\cite{Stephenson00}.
+
 \section{Related Work}

### 2012-08-30 doc/Manuscript/CGO/paper.tex
Adding the section about the Instrumentation Framework

@@
 semantics~\cite{TODO} to perform integer operations.
-If a number $n$ is too large to fit into a primitive data type $T$, then $N$'s
-value wraps around, and $N \mbox{module} T_{max}$ wounds up represented
+If a number $n$ is too large to fit into a primitive data type $T$, then $n$'s
+value wraps around, and $n \mbox{module} T_{max}$ wounds up represented
 instead.
@@
 % A brief overview about integer overflows.
+In addition to being a source of several bugs that are hard to identify, integer 
+overflows can represent a serious security risk.
+As pointed by Dietz {\em et al.}~\cite{Dietz12}, integer overflows are the
+sources of vulnerabilities in widely used software, such as OpenSSH and
+Firefox.
+These vulnerabilities could give an adversary the opportunity to execute
+arbitrary code in the compromised application. 
+We will demonstrate this with an example.
 The $n$-bits two's complement arithmetics wraps around at multiples of
@@
 generators.
-However, integer overflows can also be the sources of bugs that are hard to
-identify.
-As pointed by Dietz {\em et al.}~\cite{Dietz12}, integer overflows are the
-sources of vulnerabilities in widely used software, such as OpenSSH and
-Firefox.
-These vulnerabilities could give an adversary the opportunity to execute
-arbitrary code in the compromised application.
 
@@
 
+%What is our instrumentation library
+We have implemented our instrumentation library as a LLVM transformation pass.
+This means our library works on the LLVM IR\footnote{LLVM Intermediate Representation, available at \url{http://llvm.org/docs/LangRef.html}}
+instead of working directly on the original source code. 
+This fact brings the advantage of allowing us to rely on
+other analysis, such as the range analysis, to check if it is safe to state that
+the possible values of an instruction will always be within the range of
+the declared type of the instruction. 
+If we can state this, it means that it is impossible for an overflow to 
+occur in the analyzed instruction, because the range analysis is a conservative analysis; 
+in this case, we don't need to insert dynamic checks, and the instrumented program
+will be smaller and run faster than it would run without this prunning.
+%How our instrumentation works
+Our instrumentation works by identifying the instructions that may 
+lead to an overflow, and inserting assertions after that instructions. 
+The LLVM IR has five instructions that may lead to an overflow: 
+ADD, SUB, MUL, SHL, and TRUNC.
+
+%ADD instrumentation
+The ADD instruction performs the sum of two operands $op_1$ and $op_2$ and
+stores the result in a variable $x$.
+The semantics of an ADD instruction is $x = op_1 + op_2$. 
+The goal of the instrumentation of the instruction $x$ is to ensure that after the
+execution of the ADD instruction, $x$ stores the correct result of the sum,
+without any data loss.
+For this, we insert different checks for signed and unsigned instructions.
+If we are dealing with an unsigned ADD instruction $x$, we insert the following check:
+\code{if $x < op_1$ or $x < op_2$ then} handle overflow \code{else} continue to 
+$x$'s next instruction. If, instead, we are instrumenting a signed instruction $x$, 
+we will have the following assertion:
+\code{if ($op_1 > 0$ and $op_2 > 0$ and $x < 0$) or ($op_1 < 0$ and $op_2 < 0$ and $x > 0$) then} 
+handle overflow \code{else} continue to $x$'s next instruction.
+
+%SUB instrumentation
+The SUB instruction performs the subtraction of two operands $op_1$ and $op_2$ and
+stores the result in a variable $x$.
+The semantics of an SUB instruction is $x = op_1 - op_2$. 
+Note that the SUB semantics is close to the ADD semantics. For this reason, the
+instrumentation of the SUB instructions looks like the ADD instrumentation.
+In this case, we insert different checks for signed and unsigned instructions, too.
+If we are dealing with an unsigned SUB instruction $x$, we insert the following check:
+\code{if $op_1 < op_2$ then} handle overflow \code{else} continue to 
+$x$'s next instruction. Note that the assertion makes a single comparison. This is sound
+because if $op_2$ is greater than $op_1$, the result would be negative. However, unsigned
+integers can't represent negative numbers. In this case, we have a overflow.
+The instrumentation of a signed SUB instruction $x$ will have the following assertion:
+\code{if ($op_1 < 0$ and $op_2 > 0$ and $x > 0$) or ($op_1 > 0$ and $op_2 < 0$ and $x < 0$) then} 
+handle overflow \code{else} continue to $x$'s next instruction.
+
+%MUL instrumentation
+The MUL instruction performs the multiplication of two operands $op_1$ and $op_2$ and
+stores the result in a variable $x$.
+The semantics of an SUB instruction is $x = op_1 \times{} op_2$. 
+The dynamic check of the MUL instruction is the same for signed or unsigned
+instructions, because it relies on DIV, the inverse operation. 
+To verify if the overflow has occurred, we divide $x$ by $op_1$. If the result
+is different of $op_2$, then we have an overflow to handle. 
+In this case, we will have the following assertion: 
+\code{if $op_1 \neq{} 0$ and $(x \div{} op_1) \neq{} op_2$ then} 
+handle overflow \code{else} continue to $x$'s next instruction.
+Note that if $op_1 = 0$, the result of the multiplication will never overflow,
+because the result of the MUL will be $0$, too. In this case, if $op_1 \eq{} 0$,
+we don't perform the DIV, to prevent a division by zero exception.
+This technique of using the inverse operation to check for overflows doesn't work
+for ADD and SUB instructions because the $n$-bits two's complement wrapping-arithmetics 
+would give us false negatives in the assertions.
+
+%SHL instrumentation
+The SHL instruction performs the shift left operation. It receives a value $op_1$
+and the number of bits $n$ the SHL will shift $op_1$ by; the operation's result
+is stored in a variable $x$.
+The semantics os a SHL instruction is $x = op_1 \times{} 2^n$. 
+The SHL instructions result in overflows when the $op_1$ is positive and 
+$x$ is negative or when $op_1$ is negative and $n \neq{} 0$, because 
+the SHL in negative numbers would produce undefined results that are
+as harmful as integer overflows. This assertion has the same semantics for
+signed and unsigned instructions.
+In this case, we will have the following check: 
+\code{if ($op_1 > 0$ and $x < op_1$) or ($op_1 < 0$ and $n \neq{} 0$) then} 
+handle overflow \code{else} continue to $x$'s next instruction.
+
+%TRUNC instrumentation
+The TRUNC instruction performs the truncation of the operand $op_1$
+to a number $n$ of bits and stores the result in a variable $x$.
+The semantics of a TRUNC instruction is $x = n$LessSignificantBits$(op_1)$.
+The dynamic checking of a TRUNC instruction is done by expanding $x$ to
+the datatype of $op_1$ and comparing the expanded $x$ with $op_1$. If they 
+don't have the same value, some data has been lost by the TRUNC instruction.
+In this case, we have the following check:
+\code{if $cast(x, getDataType(op_1)) \neq{} op_1$ then} 
+handle overflow \code{else} continue to $x$'s next instruction.
+Note that the instrumentation of the TRUNC instructions is a very fragile point, 
+because it would catch any truncation with loss of precision, even if it is a
+benign one.
+To avoid trouble and make it easier to our users, the instrumentation library 
+can receive an argument to let the user decide whether to insert or not
+instrumentation in the TRUNC instructions.
+
+%Overview about the instrumentation technique
+Our instrumentation library works by adding new instructions to
+the target programs. The instrumentation of all the instructions works
+in the same way: first, the pass inserts instructions to gather the
+information needed to do the assertion.
+Then, the pass inserts compare instructions to generate a boolean value
+that indicates if there is an overflow or not. 
+After this, the pass inserts a conditional branch: 
+the false clause of the branch, where the control goes when there is no 
+overflow and the program, leads to the first instruction after the instruction
+which has received the instrumentation.
+the true clause of the branch leads to a block where we do the
+overflow handling.
+
+%What we can do in the overflow handling
+Our library has three options to handle the overflows. 
+First option: Do nothing and ignore the overflows. This option
+was used to do our experiments and verify the slowdown produced 
+by the new instructions.
+Second option: Give a message in stderr whenever an overflow occurs.
+This is the standard option. The instrumented program remains with 
+the same behavior, but whenever an overflow occurs, the program
+prints \texttt{Overflow occurred in FileName.cpp, line X.} in the
+standard error stream.
+Third option: Abort the execution when an overflow occurs.
+This is the more aggressive option. With this option enabled,
+whenever an overflow occurs, the program is terminated, not allowing
+the execution of the program with undefined behavior due to 
+integer overflows.
+
 \section{Experimental Results}
@@
 commented C++ code, our e/u-SSA conversion module has 672 lines, and our
-instrumentation pass has XXX lines.
+instrumentation pass has 762 lines.
 We have analyzed 428 C programs that constitute the LLVM test suite plus the
@@
 dynamically for the Stanford benchmark suite, which is publicly
-available~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
+available~\footnote{\url{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
 We chose Stanford because these benchmarks do not read data from external

### 2012-08-30 doc/Manuscript/CGO/paper.tex
Little corrections

@@
 stores the result in a variable $x$.
-The semantics of an SUB instruction is $x = op_1 - op_2$. 
+The semantics of a SUB instruction is $x = op_1 - op_2$. 
 Note that the SUB semantics is close to the ADD semantics. For this reason, the
@@
 stores the result in a variable $x$.
-The semantics of an SUB instruction is $x = op_1 \times{} op_2$. 
+The semantics of a MUL instruction is $x = op_1 \times{} op_2$. 
 The dynamic check of the MUL instruction is the same for signed or unsigned
@@
 Note that if $op_1 = 0$, the result of the multiplication will never overflow,
-because the result of the MUL will be $0$, too. In this case, if $op_1 \eq{} 0$,
+because the result of the MUL will be $0$, too. In this case, if $op_1 = 0$,
 we don't perform the DIV, to prevent a division by zero exception.
@@
 is stored in a variable $x$.
-The semantics os a SHL instruction is $x = op_1 \times{} 2^n$. 
+The semantics of a SHL instruction is $x = op_1 \times{} 2^n$. 
 The SHL instructions result in overflows when the $op_1$ is positive and 
@@
 the false clause of the branch, where the control goes when there is no 
-overflow and the program, leads to the first instruction after the instruction
+overflow, leads to the first instruction after the instruction
 which has received the instrumentation.

### 2012-08-30 doc/Manuscript/CGO/paper.tex
Added Raphael's text to the paper.

@@
 We have implemented our instrumentation library as a LLVM transformation pass.
-This means our library works on the LLVM IR\footnote{LLVM Intermediate Representation, available at \url{http://llvm.org/docs/LangRef.html}}
+This means our library works on the LLVM IR\footnote{LLVM Intermediate Representation, available at \texttt{http://llvm.org/docs/LangRef.html}}
 instead of working directly on the original source code. 
@@
 Figure~\ref{fig:precision} compares our results with those measured
-dynamically for the Stanford benchmark suite, which is publicly
-available~\footnote{\url{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
+dynamically for the Stanford benchmark, which is publicly
+available in the LLVM test suite.
 We chose Stanford because these benchmarks do not read data from external
@@
 \paragraph{Dynamic Detection of Integer Overflows: }
-Brumley {\em et al.}~\cite{Brumley07} have developed a tool, RICH, to detect
-integer integer overflows.
-Their approach consists in instrumenting every integer operation that might cause
-to an overflow, underflow, or data loss.
-Different from our method, RICH uses specific features of the x86 architecture
+Brumley {\em et al.}~\cite{Brumley07} have developed a tool, RICH, to
+dynamically detect integer overflows.
+Their goal is to secure C programs against integer overflow based
+vulnerabilities.
+The author's approach consists in instrumenting every integer operation that
+might cause an overflow, underflow, or data loss.
+The main result of Brumley {\em et al.} is the verification that instrumenting
+programs to guard them against integer overflows does not compromise their
+performance significantly: the average slowdown across four large applications
+is 5\%.
+RICH, Brumley {\em et al}'s tool, uses specific features of the x86 architecture
 to reduce the instrumentation overhead.
-Moreover, no static analysis is used to avoid unnecessary assertions, although
-this improvement is cited as future work.
+Chinchani {\em et al.}~\cite{Chinchani04} follow a similar approach.
+In this work, the authors describe each arithmetic operation formally, and then
+use characteristics of the computer architecture to detect overflows at
+runtime.
+% TODO bibtex: cite Archerr: Runtime environment driven program safety.
+Contrary to these previous work, we instrument programs at LLVM's intermediate
+representation level, which is machine independent.
+Nevertheless, the performance of the programs that we instrument is on par with
+Brumley's.
+
+Similar to Brumley {\em et al.}, Dietz {\em et al.}~\cite{Dietz12} have 
+implemented a tool, IOC, that instruments C/C++ programs to detect integer
+overflows.
+They approach the problem of detecting integer overflows from a software
+engineering point-of-view; hence, performance is not a concern.
+The authors have used IOC to perform a lengthy study about the occurrences
+of overflows in real-world programs, and have found that these events are very
+common.
+It is possible to implement a dynamic analysis without instrumenting the
+target program.
+In this case, developers must use some form of code emulation.
+Chen {\em et al.}~\cite{Chen09}, for instance, uses a modified
+Valgrind~\cite{Nethercote07} virtual machine to detect integer overflows.
+The main drawback of emulation is performance: Chen {\em et al.} report a
+50X slowdown.
+% TODO: cite "Brick: A  binary tool for run-time detecting and locating
+% integer-based vulnerability.
+The main difference between all these previous works and ours is the fact that
+we focus on generating less instrumentation, an endeavor that we accomplish via
+static analysis.
 
@@
 The authors have been able to use this tool to successfully identify many vulnerabilities in industrial quality software.
-Our work, and Wang {\em et a.}Õs work are essentially different: they use symbolic execution, whereas we rely on
+Our work, and Wang {\em et a.}'s work are essentially different: they use symbolic execution, whereas we rely on
 range analysis.

### 2012-08-30 doc/Manuscript/CGO/paper.tex
Revised Raphael's section.

@@
 \usepackage{amssymb}    % To use \mathbb
+\usepackage{skak}
 
@@
 %What is our instrumentation library
-We have implemented our instrumentation library as a LLVM transformation pass.
-This means our library works on the LLVM IR\footnote{LLVM Intermediate Representation, available at \texttt{http://llvm.org/docs/LangRef.html}}
-instead of working directly on the original source code. 
-This fact brings the advantage of allowing us to rely on
-other analysis, such as the range analysis, to check if it is safe to state that
-the possible values of an instruction will always be within the range of
-the declared type of the instruction. 
-If we can state this, it means that it is impossible for an overflow to 
-occur in the analyzed instruction, because the range analysis is a conservative analysis; 
-in this case, we don't need to insert dynamic checks, and the instrumented program
-will be smaller and run faster than it would run without this prunning.
-%How our instrumentation works
+We have implemented our instrumentation library as a LLVM transformation pass;
+thus, we work at the level of the compiler's intermediate
+representation~\footnote{\texttt{http://llvm.org/docs/LangRef.html}}.
+This is in contrast to previous work, which either transforms the
+source code~\cite{Dietz12}, or the machine dependent code~\cite{Brumley07}.
+We work at this level because we want to couple our library with static
+analyses, such as the algorithm that we describe in Section~\ref{sec:range}.
+% TODO: add text about the interaction between static an dynamic analysis at
+% the end of this section.
 Our instrumentation works by identifying the instructions that may 
-lead to an overflow, and inserting assertions after that instructions. 
+lead to an overflow, and inserting assertions after those instructions. 
 The LLVM IR has five instructions that may lead to an overflow:
-ADD, SUB, MUL, SHL, and TRUNC.
+\textsc{Add}, \textsc{Sub}, \textsc{Mul},  \textsc{Trunc} and \textsc{Shl}
+(left shift).
+Figure~\ref{fig:instrumentation} shows the conditions that we use to
+detect overflows.
 
-%ADD instrumentation
-The ADD instruction performs the sum of two operands $op_1$ and $op_2$ and
-stores the result in a variable $x$.
-The semantics of an ADD instruction is $x = op_1 + op_2$. 
-The goal of the instrumentation of the instruction $x$ is to ensure that after the
-execution of the ADD instruction, $x$ stores the correct result of the sum,
-without any data loss.
-For this, we insert different checks for signed and unsigned instructions.
-If we are dealing with an unsigned ADD instruction $x$, we insert the following check:
-\code{if $x < op_1$ or $x < op_2$ then} handle overflow \code{else} continue to 
-$x$'s next instruction. If, instead, we are instrumenting a signed instruction $x$, 
-we will have the following assertion:
-\code{if ($op_1 > 0$ and $op_2 > 0$ and $x < 0$) or ($op_1 < 0$ and $op_2 < 0$ and $x > 0$) then} 
-handle overflow \code{else} continue to $x$'s next instruction.
+\begin{figure}[t!]
+\begin{center}
+\begin{tabular}{ll}
+$x = o_1 \ +_s \ o_2$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
+                  & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
+$x = o_1 \ +_u \ o_2$ & $x < o_1 \vee x < o_2$ \\ \\
+$x = o_1 \ -_s \ o_2$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \vee$ \\
+                  & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
+$x = o_1 \ -_u \ o_2$ & $o_1 < o_2$ \\ \\
+$x = o_1 \ \times_{u/s} \ o_2$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\ \\
+$x = o_1 \ \qside \ n$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
+$x = \ \downarrow_n \ o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
+$x = \ (\mbox{type}_n) o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
+\end{tabular}
+\end{center}
+\caption{\label{fig:instrumentation}Overflow checks. We use $\downarrow_n$ for
+the operation that truncates to $n$ bits, and we use $(\mbox{type}_n)$ for a
+bit-cast to $n$ bits. Both operations have the same semantics in our setting.
+The subscript $s$ indicates a signed instruction; the subscript $u$ indicate
+an unsigned operation.}
+\end{figure}
 
 %SUB instrumentation
-The SUB instruction performs the subtraction of two operands $op_1$ and $op_2$ and
-stores the result in a variable $x$.
-The semantics of a SUB instruction is $x = op_1 - op_2$. 
-Note that the SUB semantics is close to the ADD semantics. For this reason, the
-instrumentation of the SUB instructions looks like the ADD instrumentation.
-In this case, we insert different checks for signed and unsigned instructions, too.
-If we are dealing with an unsigned SUB instruction $x$, we insert the following check:
-\code{if $op_1 < op_2$ then} handle overflow \code{else} continue to 
-$x$'s next instruction. Note that the assertion makes a single comparison. This is sound
-because if $op_2$ is greater than $op_1$, the result would be negative. However, unsigned
-integers can't represent negative numbers. In this case, we have a overflow.
-The instrumentation of a signed SUB instruction $x$ will have the following assertion:
-\code{if ($op_1 < 0$ and $op_2 > 0$ and $x > 0$) or ($op_1 > 0$ and $op_2 < 0$ and $x < 0$) then} 
-handle overflow \code{else} continue to $x$'s next instruction.
+The instrumentation that we insert is mostly straightforward.
+We discuss in the rest of this section a few cases.
+If we are dealing with an unsigned \textsc{Sub} instruction, e.g,
+$x = o_1 \ -_u \ o_2$, then a single check is enough to detect the bad
+behavior: $o_1 < o_2$.
+This is sound because if $o_2$ is greater than $o_1$, then we assume that it is
+a bug to try to represent a negative number in unsigned arithmetics.
+Regarding multiplication, e.g., $x = o_1 \times o_2$, if $o_1 = 0$, then this
+operation can never cause an overflow.
+This test is necessary to us, because we check integer overflows in
+multiplication via the inverse operation of integer division.
+Thus, the test prevents a division by zero from happening.
+The semantics of a \textsc{Trunc} instruction $x = \ \downarrow_n \ o_1$ is
+to assign to $x$ the $n$ least significant bits of $o_1$.
+The dynamic checking in this case consists in expanding $x$ to
+the datatype of $o_1$ and comparing the expanded value with $o_1$.
+The LLVM IR provides instructions to perform these type expansions.
+Note that our instrumentation catches any truncation with loss of precision, even
+if it is a benign one.
+However, we give users the possibility of disabling checks on truncation.
 
-%MUL instrumentation
-The MUL instruction performs the multiplication of two operands $op_1$ and $op_2$ and
-stores the result in a variable $x$.
-The semantics of a MUL instruction is $x = op_1 \times{} op_2$. 
-The dynamic check of the MUL instruction is the same for signed or unsigned
-instructions, because it relies on DIV, the inverse operation. 
-To verify if the overflow has occurred, we divide $x$ by $op_1$. If the result
-is different of $op_2$, then we have an overflow to handle. 
-In this case, we will have the following assertion: 
-\code{if $op_1 \neq{} 0$ and $(x \div{} op_1) \neq{} op_2$ then} 
-handle overflow \code{else} continue to $x$'s next instruction.
-Note that if $op_1 = 0$, the result of the multiplication will never overflow,
-because the result of the MUL will be $0$, too. In this case, if $op_1 = 0$,
-we don't perform the DIV, to prevent a division by zero exception.
-This technique of using the inverse operation to check for overflows doesn't work
-for ADD and SUB instructions because the $n$-bits two's complement wrapping-arithmetics 
-would give us false negatives in the assertions.
-
-%SHL instrumentation
-The SHL instruction performs the shift left operation. It receives a value $op_1$
-and the number of bits $n$ the SHL will shift $op_1$ by; the operation's result
-is stored in a variable $x$.
-The semantics of a SHL instruction is $x = op_1 \times{} 2^n$. 
-The SHL instructions result in overflows when the $op_1$ is positive and 
-$x$ is negative or when $op_1$ is negative and $n \neq{} 0$, because 
-the SHL in negative numbers would produce undefined results that are
-as harmful as integer overflows. This assertion has the same semantics for
-signed and unsigned instructions.
-In this case, we will have the following check: 
-\code{if ($op_1 > 0$ and $x < op_1$) or ($op_1 < 0$ and $n \neq{} 0$) then} 
-handle overflow \code{else} continue to $x$'s next instruction.
-
-%TRUNC instrumentation
-The TRUNC instruction performs the truncation of the operand $op_1$
-to a number $n$ of bits and stores the result in a variable $x$.
-The semantics of a TRUNC instruction is $x = n$LessSignificantBits$(op_1)$.
-The dynamic checking of a TRUNC instruction is done by expanding $x$ to
-the datatype of $op_1$ and comparing the expanded $x$ with $op_1$. If they 
-don't have the same value, some data has been lost by the TRUNC instruction.
-In this case, we have the following check:
-\code{if $cast(x, getDataType(op_1)) \neq{} op_1$ then} 
-handle overflow \code{else} continue to $x$'s next instruction.
-Note that the instrumentation of the TRUNC instructions is a very fragile point, 
-because it would catch any truncation with loss of precision, even if it is a
-benign one.
-To avoid trouble and make it easier to our users, the instrumentation library 
-can receive an argument to let the user decide whether to insert or not
-instrumentation in the TRUNC instructions.
-
+\paragraph{Practical Considerations: }
 %Overview about the instrumentation technique
 Our instrumentation library works by adding new instructions to
-the target programs. The instrumentation of all the instructions works
+the target programs.
+The instrumentation of all the instructions works
 in the same way: first, the pass inserts instructions to gather the

### 2012-08-31 src/OverflowDetect/OverflowDetect.cpp

@@ -131,6 +131,7 @@ static bool isSignedInst(Value* V){
 			return dyn_cast<OverflowingBinaryOperator>(I)->hasNoSignedWrap();
 			break;
 
+		case Instruction::BitCast:
 		case Instruction::Trunc:
 			return isSignedInst(I->getOperand(0));
 			break;
@@ -462,6 +463,9 @@ bool OverflowDetect::isValidInst(Instruction *I)
 		return I->getType()->isIntegerTy() && I->getOperand(0)->getType()->isIntegerTy() && I->getOperand(1)->getType()->isIntegerTy();
 	else if (dyn_cast<TruncInst>(I))
 		return TruncInstrumentation && I->getType()->isIntegerTy() && I->getOperand(0)->getType()->isIntegerTy();
+	else if (dyn_cast<BitCastInst>(I))
+		//Only do the instrumentation of the bitcast if the cast is to a lower size datatype (semantically equivalent to a trunc)
+		return TruncInstrumentation && I->getType()->isIntegerTy() && I->getOperand(0)->getType()->isIntegerTy() && I->getType()->getPrimitiveSizeInBits() < I->getOperand(0)->getType()->getPrimitiveSizeInBits();
 	else
 		return false;
 }
@@ -555,6 +559,7 @@ void OverflowDetect::insertInstrumentation(Instruction* I, BasicBlock* AbortBB, 
     ICmpInst *positiveOp2;
 
     ICmpInst *negativeOp1;
+    ICmpInst *negativeOp2;
 
     ICmpInst *lessThanOp1;
     ICmpInst *lessThanOp2;
@@ -577,6 +582,9 @@ void OverflowDetect::insertInstrumentation(Instruction* I, BasicBlock* AbortBB, 
 
 	Value* op1;
 	Value* op2;
+
+	Value* hasIntegerBug1 = NULL;
+	Value* hasIntegerBug2 = NULL;
 	Value* hasIntegerBug = NULL;
     Value* canCauseOverflow;
     Value* tmpValue;
@@ -599,14 +607,21 @@ void OverflowDetect::insertInstrumentation(Instruction* I, BasicBlock* AbortBB, 
 			if (isSigned){
 				//How to do an integer bug with a signed Add instruction:
 				//		add two big positive numbers and get a negative number as result
-				positiveOp1 = new ICmpInst(nextInstruction, CmpInst::ICMP_SGT, op1, constZero);
[…]

### 2012-08-31 doc/Manuscript/CGO/paper.tex
Revised Raphael's section, adding rules for static analysis.

@@
 
-\title{An Algorithm to Prune Integer Overflow Tests}
+\title{Combining Static and Dynamic Analyses to Secure Programs against
+Integer Overflows}
 %\subtitle{A case study of the Dinamica EGO toolkit}
@@
 source code~\cite{Dietz12}, or the machine dependent code~\cite{Brumley07}.
-We work at this level because we want to couple our library with static
-analyses, such as the algorithm that we describe in Section~\ref{sec:range}.
-% TODO: add text about the interaction between static an dynamic analysis at
-% the end of this section.
+We work at the intermediate representation level to be able to couple our
+library with static analyses, such as the algorithm that we described in
+Section~\ref{sec:range}.
 Our instrumentation works by identifying the instructions that may 
@@
 (left shift).
-Figure~\ref{fig:instrumentation} shows the conditions that we use to
+Figure~\ref{fig:instrumentation} shows the dynamic tests that we perform to
 detect overflows.
+This figure also shows how the static analysis helps us to avoid inserting
+some of those tests.
+We only generate the dynamic tests on the right if the conditions in the middle
+are false.
 
-\begin{figure}[t!]
+\begin{figure*}[t!]
 \begin{center}
-\begin{tabular}{ll}
-$x = o_1 \ +_s \ o_2$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
-                  & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
-$x = o_1 \ +_u \ o_2$ & $x < o_1 \vee x < o_2$ \\ \\
-$x = o_1 \ -_s \ o_2$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \vee$ \\
-                  & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
-$x = o_1 \ -_u \ o_2$ & $o_1 < o_2$ \\ \\
-$x = o_1 \ \times_{u/s} \ o_2$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\ \\
-$x = o_1 \ \qside \ n$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
-$x = \ \downarrow_n \ o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
-$x = \ (\mbox{type}_n) o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
+\begin{tabular}{lll}
+Instruction & Static Check & Dynamic Check \\ \\
+$x = o_1 \ +_s \ o_2$ & $\lb{I[o_1]} + \lb{I[o_2]} > \mbox{MIN\_INT} \wedge \ub{I[o_1]} + \ub{I[o_2]} < \mbox{MAX\_INT}$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
+                  & & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
+$x = o_1 \ +_u \ o_2$ & $\ub{I[o_1]} + \ub{I[o_2]} < \mbox{MAX\_INT}$ & $x < o_1 \vee x < o_2$ \\ \\
+$x = o_1 \ -_s \ o_2$ & $\lb{I[o_1]} - \lb{I[o_2]} > \mbox{MIN\_INT} \wedge \ub{I[o_1]} - \ub{I[o_2]} < \mbox{MAX\_INT}$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \ \ \vee$ \\
+                      & & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
+$x = o_1 \ -_u \ o_2$ & $\lb{I[o_1]} > \ub{I[o_2]}$ & $o_1 < o_2$ \\ \\
+$x = o_1 \ \times_{u/s} \ o_2$ & $\mbox{MIN\_INT} < \{ \lb{I[o_1]} \times \lb{I[o_2]}, \lb{I[o_1]} \times \ub{I[o_2]} \}< \mbox{MAX\_INT}$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\
+                       & $\mbox{MIN\_INT} < \{ \ub{I[o_1]}\times \lb{I[o_2]}, \ub{I[o_1]}\times \ub{I[o_2]} \}< \mbox{MAX\_INT}$ & \\ \\
+$x = o_1 \ \qside \ n$ & $2^n \times \ub{I[o_1]} < \mbox{MAX\_INT}$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
+$x = \ \downarrow_n \ o_1$  & $\lb{I[o_1]} > -2^n \wedge \ub{I[o_1]} < 2^n_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
+$x = \ (\mbox{type}_n) o_1$ & $\lb{I[o_1]} > -2^n \wedge \ub{I[o_1]} < 2^n_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
 \end{tabular}
@@
 The subscript $s$ indicates a signed instruction; the subscript $u$ indicate
-an unsigned operation.}
-\end{figure}
+an unsigned operation. If the static check is true, we do not insert the dynamic
+check. The values \{MAX/MIN\}\_INT depend on the type of the instruction.}
+\end{figure*}
 
@@
 The instrumentation that we insert is mostly straightforward.
-We discuss in the rest of this section a few cases.
+We discuss in the rest of this section a few interesting cases.
 If we are dealing with an unsigned \textsc{Sub} instruction, e.g,
@@
 Thus, the test prevents a division by zero from happening.
-The semantics of a \textsc{Trunc} instruction $x = \ \downarrow_n \ o_1$ is
-to assign to $x$ the $n$ least significant bits of $o_1$.
+The \textsc{Trunc} instruction, e.g., $x = \ \downarrow_n \ o_1$ assigns to $x$
+the $n$ least significant bits of $o_1$.
 The dynamic checking in this case consists in expanding $x$ to
@@
 The LLVM IR provides instructions to perform these type expansions.
-Note that our instrumentation catches any truncation with loss of precision, even
-if it is a benign one.
-However, we give users the possibility of disabling checks on truncation.
+Note that our instrumentation catches any truncation that might result in
+data loss, even if this loss is benign.
+To make the dynamic checks more liberal, we give users the possibility of
+disabling tests over truncations.
 
@@
 Our instrumentation library works by adding new instructions to
-the target programs.
-The instrumentation of all the instructions works
-in the same way: first, the pass inserts instructions to gather the
-information needed to do the assertion.
-Then, the pass inserts compare instructions to generate a boolean value
-that indicates if there is an overflow or not. 
-After this, the pass inserts a conditional branch: 
-the false clause of the branch, where the control goes when there is no 
-overflow, leads to the first instruction after the instruction
-which has received the instrumentation.
-the true clause of the branch leads to a block where we do the
-overflow handling.
+the target program.
+Although the dynamic check depends on the instruction that is instrumented,
+the general modus operandi is the same.
+Our dynamic tests check for overflows after they happen.
+The code that we insert to detect the overflow diverts the program flow in case
+such an event takes place.
 
-%What we can do in the overflow handling
-Our library has three options to handle the overflows. 
-First option: Do nothing and ignore the overflows. This option
-was used to do our experiments and verify the slowdown produced 
-by the new instructions.
-Second option: Give a message in stderr whenever an overflow occurs.
-This is the standard option. The instrumented program remains with 
-the same behavior, but whenever an overflow occurs, the program
-prints \texttt{Overflow occurred in FileName.cpp, line X.} in the
-standard error stream.
-Third option: Abort the execution when an overflow occurs.
-This is the more aggressive option. With this option enabled,
-whenever an overflow occurs, the program is terminated, not allowing
-the execution of the program with undefined behavior due to 
-integer overflows.
+% TODO: add a snapshot of a function with, and without the integer overflow
+% instrumentation. Let this function be very simple, e.g.,
+% int foo(int x, int y) {return x + y; }
+
+Which actions are performed once the overflow is detected depends on the
+user.
+Our library gives the user three options to handle them. 
+First option: no-op.
+This option allows us to verify the slowdown produced by the new instructions.
+Second option: logging.
+This is the standard option, and it preserves the behavior of the instrumented
+program.
+Whenever an overflow is detected, we print \texttt{Overflow detected in
+FileName.cpp, line X.} in the standard error stream.
+Third option: abort.
+This option terminates the program once an overflow is detected.
+Thus, it disallows undefined behavior due to integer overflows.
 
@@
 graphs after inlining) and analysis runtime (ms).
+Each point represents a benchmark.
 Coefficient of determination = 0.967.

### 2012-08-31 doc/Manuscript/CGO/paper.tex
Added new figs.
Added more text to Raphael's section.

@@
 detect overflows.
-This figure also shows how the static analysis helps us to avoid inserting
-some of those tests.
-We only generate the dynamic tests on the right if the conditions in the middle
-are false.
 
-\begin{figure*}[t!]
+\begin{figure}[t!]
 \begin{center}
-\begin{tabular}{lll}
-Instruction & Static Check & Dynamic Check \\ \\
-$x = o_1 \ +_s \ o_2$ & $\lb{I[o_1]} + \lb{I[o_2]} > \mbox{MIN\_INT} \wedge \ub{I[o_1]} + \ub{I[o_2]} < \mbox{MAX\_INT}$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
-                  & & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
-$x = o_1 \ +_u \ o_2$ & $\ub{I[o_1]} + \ub{I[o_2]} < \mbox{MAX\_INT}$ & $x < o_1 \vee x < o_2$ \\ \\
-$x = o_1 \ -_s \ o_2$ & $\lb{I[o_1]} - \lb{I[o_2]} > \mbox{MIN\_INT} \wedge \ub{I[o_1]} - \ub{I[o_2]} < \mbox{MAX\_INT}$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \ \ \vee$ \\
-                      & & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
-$x = o_1 \ -_u \ o_2$ & $\lb{I[o_1]} > \ub{I[o_2]}$ & $o_1 < o_2$ \\ \\
-$x = o_1 \ \times_{u/s} \ o_2$ & $\mbox{MIN\_INT} < \{ \lb{I[o_1]} \times \lb{I[o_2]}, \lb{I[o_1]} \times \ub{I[o_2]} \}< \mbox{MAX\_INT}$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\
-                       & $\mbox{MIN\_INT} < \{ \ub{I[o_1]}\times \lb{I[o_2]}, \ub{I[o_1]}\times \ub{I[o_2]} \}< \mbox{MAX\_INT}$ & \\ \\
-$x = o_1 \ \qside \ n$ & $2^n \times \ub{I[o_1]} < \mbox{MAX\_INT}$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
-$x = \ \downarrow_n \ o_1$  & $\lb{I[o_1]} > -2^n \wedge \ub{I[o_1]} < 2^n_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
-$x = \ (\mbox{type}_n) o_1$ & $\lb{I[o_1]} > -2^n \wedge \ub{I[o_1]} < 2^n_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
+\begin{tabular}{ll}
+Instruction & Dynamic Check \\ \\
+$x = o_1 \ +_s \ o_2$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
+                      & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
+$x = o_1 \ +_u \ o_2$ & $x < o_1 \vee x < o_2$ \\ \\
+$x = o_1 \ -_s \ o_2$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \ \ \vee$ \\
+                      & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
+$x = o_1 \ -_u \ o_2$ & $o_1 < o_2$ \\ \\
+$x = o_1 \ \times_{u/s} \ o_2$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\ \\
+$x = o_1 \ \qside \ n$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
+$x = \ \downarrow_n \ o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
+$x = \ (\mbox{type}_n) o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
 \end{tabular}
@@
 The subscript $s$ indicates a signed instruction; the subscript $u$ indicate
-an unsigned operation. If the static check is true, we do not insert the dynamic
-check. The values \{MAX/MIN\}\_INT depend on the type of the instruction.}
-\end{figure*}
+an unsigned operation.}
+\end{figure}
 
@@
 operation can never cause an overflow.
-This test is necessary to us, because we check integer overflows in
-multiplication via the inverse operation of integer division.
+This test is necessary, because we check integer overflows in
+multiplication via the inverse operation, e.g., integer division.
 Thus, the test prevents a division by zero from happening.
@@
 the $n$ least significant bits of $o_1$.
-The dynamic checking in this case consists in expanding $x$ to
+The dynamic check, in this case, consists in expanding $x$ to
 the datatype of $o_1$ and comparing the expanded value with $o_1$.
@@
 the general modus operandi is the same.
-Our dynamic tests check for overflows after they happen.
+The dynamic tests check for overflows after they happen.
 The code that we insert to detect the overflow diverts the program flow in case
 such an event takes place.
+Figure~\ref{fig:instrumented_cfg} shows an actual control flow graph, before
+and after the instrumentation.
+From this figure, it is clear that the instrumented program will be larger
+than the original code.
+However, as we show empirically, this growth is not too large when compared
+to the total size of our benchmarks, because most of the instructions in these
+programs do not demand instrumentation.
+Furthermore, none of the instructions used to instrument integer arithmetics
+access memory.
+Therefore, the overall slowdown that the instrumentation causes is usually
+small.
 
-% TODO: add a snapshot of a function with, and without the integer overflow
-% instrumentation. Let this function be very simple, e.g.,
-% int foo(int x, int y) {return x + y; }
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\columnwidth]{images/instrumented_cfg}
+\end{center}
+\caption{\label{fig:instrumented_cfg}
+(a) A simple C function.
+(b) The same function converted to the LLVM intermediate representation.
+(c) The instrumented code. The boldface lines were part of the original
+program.}
+\end{figure}
 
 Which actions are performed once the overflow is detected depends on the
-user.
+user, who has the option to overwrite the \texttt{handle\_overflow}
+function in Figure~\ref{fig:instrumented_cfg}.
 Our library gives the user three options to handle them. 
@@
 Thus, it disallows undefined behavior due to integer overflows.
+
+\paragraph{Using the static analysis to avoid some overflow checks.}
+Our library can, optionally, use the range analysis to avoid having to
+insert some overflow checks into the instrumented program.
+We give the user the possibility of calling the range analysis with either the
+e-SSA or the u-SSA live range splitting strategies.
+Our static analysis classifies variables into four categories, depending on
+their bounds:
+\begin{itemize}
+\item \textbf{Safe}: a variable is safe if its bounds are fully contained
+inside its declared type.
+For instance, if $x$ is declared as an unsigned 8-bits integer, then $x$ is
+safe if its bounds are within the interval $[0, 255]$.
+\item \textbf{Suspicious}: we say that a variable is suspicious if its bounds
+go beyond the interval of its declared type, but the intersection between
+these two ranges is non-empty.
+For instance, the same variable $x$ would be suspicious if
+$I[x] = [0, 257]$, as $\ub{I[x]} > \ub{\mathtt{int8}}$.
+\item \textbf{Uncertain}: we classify a variable as uncertain if at least one
+of its limits is unbounded.
+Our variable $x$ would be uncertain if $I[x] = [0, \infty]$.
+We distinguish suspicious from uncertain variables because we speculate that
+actual overflows are more common among elements in the former category.
+\item \textbf{buggy}: a variable is buggy if the intersection between its
+inferred range and the range of its declared type is empty.
+This is a definitive case of an overflow.
+Continuing with our example, $x$ would be buggy if, for instance,
+$I[x] = [257, \infty]$, given that $[257, \infty] \sqcap [0, 255] = \emptyset$.
+\end{itemize}
+Independent on the arithmetic instruction that is being analyzed, the
+instrumentation library performs the same test:
+if the result $x$ of an arithmetic instruction such as $x = o_1 +_s \ o_2$ is
+safe, then the overflow check is not necessary, otherwise it must be
+created.
 

### 2012-09-02 doc/Manuscript/CGO/paper.tex
Changed a few parts of the text of the tech report

@@
 Nevertheless, the performance of the programs that we instrument is on par with
-Brumley's.
-
-Similar to Brumley {\em et al.}, Dietz {\em et al.}~\cite{Dietz12} have 
-implemented a tool, IOC, that instruments C/C++ programs to detect integer
-overflows.
+Brumley's, even without the support of the static range analysis.
+
+Dietz {\em et al.}~\cite{Dietz12} have 
+implemented a tool, IOC, that instruments the source code of C/C++ programs to
+detect integer overflows.
 They approach the problem of detecting integer overflows from a software
 engineering point-of-view; hence, performance is not a concern.
-The authors have used IOC to perform a lengthy study about the occurrences
+The authors have used IOC to carry out a lengthy study about the occurrences
 of overflows in real-world programs, and have found that these events are very
@@
 % integer-based vulnerability.
-The main difference between all these previous works and ours is the fact that
-we focus on generating less instrumentation, an endeavor that we accomplish via
-static analysis.
+We differ from all this previous work because we focus on generating less
+instrumentation, an endeavor that we accomplish via static analysis.
 
 \paragraph{Static Detection of Integer Overflows: }
-\noindent
-Zhang {\em et al.}~\cite{Zhang09} have designed a static analysis to sanitize
+Zhang {\em et al.}~\cite{Zhang09} have used static analysis to sanitize
 programs against integer overflow based vulnerabilities.
@@
 
+\paragraph{Range Analysis:}
+One of the goals of this paper is to describe a state-of-the-art range analysis
+algorithm.
+Many parts of this algorithm are well-known, others are novel.
+The widening and narrowing operators from Section~\ref{sub:micro} were
+introduced by Cousot and Cousot's seminal paper~\cite{Cousot77}.
+The widening operator that we use in our implementation is well-known among
+compiler writers~\cite[p.228]{Nielson99}.
+The interval lattice from Section~\ref{sub:lattice} has been used in a
+plethora of old and recent range analysis
+solvers~\cite{Gough94,Mahlke01,Oh12,Patterson95,Stephenson00,Su05}.
+The use of futures to handle comparisons between variables is a contribution
+of this paper.
+It is well-known that the use of strongly connected components speeds up
+constraint resolution systems~\cite[Sec 6.3]{Nielson99}.
+However, publicly available implementations of range analysis such as those in
+FLEX~\footnote{The MIT's FLEX/Harpoon compiler
+provides an implementation of Stephenson's algorithm~\cite{Stephenson00}, and
+is available at \texttt{http://flex.cscott. net/Harpoon/}.},
+gcc~\footnote{Gcc's VRP pass (at \texttt{http://gcc.gnu.org/svn/gcc/trunk/gcc/
+tree-vrp.c}) implements a variant of Patterson's
+algorithm~\cite{Patterson95}.} or Mozilla's IonMonkey~\footnote{\texttt{hg.mozilla.org/projects/ionmonkey/file/629d1e6251b9/js/
+src/ion/RangeAnalysis.cpp}. This code is implemented after Gough and
+Klaeren~\cite{Gough94}} do not resort to this technique.
+Furthermore, they only deal with comparisons between variables and constants.
+
 \noindent
-\paragraph{Intermediate Program Representations: }
+\paragraph{Sparse Dataflow Analysis: }
+In this paper we provide a {\em sparse} implementation of range analysis.
+Sparsity, in our context, means that we associate points in the lattice of
+interest -- intervals in our case -- directly to variables.
+Dense analyses map such information to pairs formed by variables and program
+points.
+The compiler related literature contains many descriptions of sparse
+data-flow analyses.
+Some analyses obtain sparsity by using specific program
+representations, like we did.
+Others rely on data-structures.
+In terms of data-structures, the first, and best known method proposed to
+support sparse data-flow analyses is Choi {\em et al.}'s {\em Sparse Evaluation
+Graph} (SEG)~\cite{Choi91}.
+
+One of the program representation that we discuss in
+Section~\ref{sub:splitting} -- the {\em
+Extended Static Single Assignment} (e-SSA) form -- has been used by Bodik
+{\em et al.}~\cite{Bodik00} and Gough {\em et al.}~\cite{Gough94}.
+The other representation, which we call u-SSA, is very similar to
+the \emph{Static Single Use} form (SSU).
+There exists many variants of SSU~\cite{Plevyak96,George03,Lo98}.
+The main difference is that our dataflow analysis propagates information
+forwardly, whereas the SSU form has been conceive for dataflow analyses
+that propagate information backwardly.
+Thus, it requires special nodes at branches to merge information from
+different program paths.
+The SSU form programs are a subset of the programs in the so called
+{\em Static Single Information} form, a program representation proposed by
+Scott Ananian~\cite{Ananian99}.
+SSI makes sparse the dataflow analyses that propagate information both
+forwardly and backwardly.
 

### 2012-09-02 doc/Manuscript/CGO/paper.tex
Added number of instructions necessary for instrumentation.

@@
 In general, these programming languages resort to a {\em wrapping-arithmetics}
-semantics~\cite{TODO} to perform integer operations.
+semantics~\cite{Warren02} to perform integer operations.
 If a number $n$ is too large to fit into a primitive data type $T$, then $n$'s
@@
 instead.
-There are situations in which this semantics is acceptable~\cite{TODO}.
+There are situations in which this semantics is acceptable~\cite{Dietz12}.
 For instance, programers might rely on this behavior to implement hash functions
@@
 paper we focus on SPEC CPU 2006.
-Our implementation has been able to reproduce the results recently obtained by XX
-{\em et al.}~\cite{TODO}.
+Our implementation has been able to reproduce the results obtained by
+previous work.
+We have re-discovered the integer overflows recently observed by
+Dietz {\em et al.}~\cite{Dietz12}, and the performance of our instrumentation
+library, without the support of range analysis, matches the state-of-the-art
+implementation of Brumley {\em et al.}
 Our range analysis algorithm has been able to remove XX\% of the overflow
@@
 We inline small functions to make our implementation more context sensitive.
-Our experiments show that the speed and memory consumption of our range analysis
-grows linearly with the program size.
+Our experiments show that the speed and memory consumption of our range
+analysis grows linearly with the program size.
 By comparing the ranges that we determine statically with the results found by
@@
 
-We will use the program in Figure~\ref{fig:ex1}(a) to illustrate our range
-analysis.
-Figure~\ref{fig:ex1}(b) shows the same program in e-SSA form~\cite{Bodik00},
-and Figure~\ref{fig:ex1}(c) outlines the constraints that we extract from this
-program.
+We will use the program in Figure~\ref{fig:ex_eSSA_cgo}(a) to illustrate our
+range analysis.
+Figure~\ref{fig:ex_eSSA_cgo}(b) shows the same program in e-SSA
+form~\cite{Bodik00},
+and Figure~\ref{fig:ex_eSSA_cgo}(c) outlines the constraints that we extract
+from this program.
 There is a clear correspondence between instructions and constraints.
@@
 A possible solution to the range analysis problem, as obtained via the
-techniques that we will introduce in Section~\ref{sub:algo}, is given in
+techniques that we will introduce in Section~\ref{sub:prop}, is given in
 Figure~\ref{fig:ex_eSSA_cgo}(d).
@@
 The LLVM IR has five instructions that may lead to an overflow:
-\textsc{Add}, \textsc{Sub}, \textsc{Mul},  \textsc{Trunc} and \textsc{Shl}
-(left shift).
+\textsc{Add}, \textsc{Sub}, \textsc{Mul},  \textsc{Trunc} (also bit-casts) and
+\textsc{Shl} (left shift).
 Figure~\ref{fig:instrumentation} shows the dynamic tests that we perform to
 detect overflows.
+Instrumenting \textsc{Add} or \textsc{Sub} requires 12 LLVM instructions.
+Left shifts need 8 instructions, \textsc{Mul} requires 7 and
+\textsc{Trunc} requires only 3.
 

### 2012-09-02 doc/Manuscript/CGO/paper.tex
Changed a few lines in the paper.

@@
 detect overflows.
-Instrumenting \textsc{Add} or \textsc{Sub} requires 12 LLVM instructions.
-Left shifts need 8 instructions, \textsc{Mul} requires 7 and
-\textsc{Trunc} requires only 3.
 
@@
 than the original code.
-However, as we show empirically, this growth is not too large when compared
-to the total size of our benchmarks, because most of the instructions in these
-programs do not demand instrumentation.
+Instrumenting \textsc{Add} or \textsc{Sub} requires 12 LLVM instructions.
+Left shifts need 8 instructions, \textsc{Mul} requires 7 and
+\textsc{Trunc} requires only 3.
+These numbers do not include the instructions necessary to handle the overflow,
+e.g., block \texttt{\%11} in Figure~\ref{fig:instrumented_cfg}, as this code is
+not in the program's main path.
+Nevertheless, as we show empirically, this growth is not too large when
+compared to the total size of our benchmarks, because most of the instructions
+in these programs do not demand instrumentation.
 Furthermore, none of the instructions used to instrument integer arithmetics

### 2012-09-02 doc/Manuscript/CGO/paper.tex
Added table with Raphael's numbers.

@@
 \titlebanner{Code Generation and Optimization}        % These are ignored unless
-\preprintfooter{Dinamica EGO}   % 'preprint' option specified.
+\preprintfooter{Overflow Elimination}   % 'preprint' option specified.
 
@@
 than the original code.
-Instrumenting \textsc{Add} or \textsc{Sub} requires 12 LLVM instructions.
-Left shifts need 8 instructions, \textsc{Mul} requires 7 and
-\textsc{Trunc} requires only 3.
-These numbers do not include the instructions necessary to handle the overflow,
-e.g., block \texttt{\%11} in Figure~\ref{fig:instrumented_cfg}, as this code is
-not in the program's main path.
+Figure~\ref{fig:num_instructions} shows how many LLVM instructions are
+necessary to instrument each arithmetic operation.
+These numbers do not include the instructions necessary to handle the overflow
+itself, e.g., block \texttt{\%11} in Figure~\ref{fig:instrumented_cfg}, as this
+code is not in the program's main path.
 Nevertheless, as we show empirically, this growth is not too large when
@@
 small.
+
+\begin{figure}[t!]
+\begin{center}
+\begin{small}
+\renewcommand{\arraystretch}{1.5}
+\begin{tabular}{lccccc}
+& \textsc{Add} & \textsc{Sub} & \textsc{Mul} & \textsc{Shl} & \textsc{Trunc} \\
+signed   & 12 & 12 & 6 & 8 & 3 \\
+unsigned & 4  & 2  & 6 & 2 & 3
+\end{tabular}
+\end{small}
+\caption{\label{fig:num_instructions}
+Number of instructions used in each dynamic check.}
+\end{center}
+\end{figure}
 
@@
 Which actions are performed once the overflow is detected depends on the
-user, who has the option to overwrite the \texttt{handle\_overflow}
+user, who has the option to overwrite the \texttt{handle\_ overflow}
 function in Figure~\ref{fig:instrumented_cfg}.

### 2012-09-04 doc/Manuscript/CGO/paper.tex
Changing the title and inserting a number in the abstract

@@
 
-\title{Combining Static and Dynamic Analyses to Secure Programs against
+\title{A fast and Low-Overhead Technique to Secure Programs Against
 Integer Overflows}
+%\title{Combining Static and Dynamic Analyses to Secure Programs against
+%Integer Overflows}
 %\subtitle{A case study of the Dinamica EGO toolkit}
@@
 This framework has been used to detect overflows in hundreds of C/C++ programs.
-As a testimony of its effectiveness, our range analysis has been able to speedup
-the SPEC CPU 2006 instrumented programs by XX\%.
+As a testimony of its effectiveness, our range analysis has been able to eliminate
+19\% of all the overflow checks necessary to secure the C programs in the
+SPEC CPU 2006 benchmark suite.
 \end{abstract}

### 2012-09-05 doc/Manuscript/CGO/paper.tex

@@
 \usepackage{skak}
+\usepackage{pbox}
 
@@
 
+%How we have done the dynamic tests  
+We have executed the instrumented programs of the integer benchmarks of SPEC
+2006 CPU to measure the overhead produced by our instrumentation in 
+real world programs.
+To do this, we have developed a little test framework that compiles, runs and
+collects the statistics about the execution of all the programs of SPEC 2006.
+The execution of the programs was made using the arguments and inputs of the "test"
+mode, which is faster to execute than the "ref" inputs.
+
+%Analysis about the Table 1 of the spreadsheet (Unpruned Instrumentation)
+Figure~\ref{fig:unpruned_analysis} shows the number of instructions that has
+received instrumentation, without avoiding inserting the instrumentation by
+using any static analysis. We can see that the number of these instructions is low 
+(3,96 in the average). 
+Figure~\ref{fig:prunning_analysis} shows that it is possible to use static 
+analysis to avoid inserting the instrumentation in some instructions.
+
 \section{Related Work}

### 2012-09-05 doc/Manuscript/CGO/paper.tex

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%Static Analysis
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%Analysis about the amount of instructions
+\begin{figure*}[t!]
+\begin{center}
+\begin{small}
+\begin{tabular}{|l|r|r|r|r|}
+\hline
+Program Name & \multicolumn{1}{l|}{Num.
+of Instructions} & \multicolumn{1}{l|}{Num.
+Instrumented Instructions} & \multicolumn{1}{l|}{Perc.
+Instrumented Instructions} & \multicolumn{1}{l|}{Num.
+Instructions Where Overflows Occurred} \\ \hline
+470.lbm & 13.724 & 1.142 & 8,32\% & 0 \\ \hline
+433.milc & 44.236 & 1.602 & 3,62\% & 11 \\ \hline
+444.namd & 100.276 & 3.234 & 3,23\% & 12 \\ \hline
+447.dealII & 1.381.408 & 36.157 & 2,62\% & 50 \\ \hline
+450.soplex & 136.367 & 3.158 & 2,32\% & 13 \\ \hline
+464.h264ref & 271.627 & 13.846 & 5,10\% & 167 \\ \hline
+473.astar & 19.243 & 857 & 4,45\% & 0 \\ \hline
+458.sjeng & 54.051 & 2.504 & 4,63\% & 68 \\ \hline
+429.mcf & 4.725 & 165 & 3,49\% & 8 \\ \hline
+471.omnetpp & 203.201 & 1.972 & 0,97\% & 2 \\ \hline
+403.gcc & 1.419.456 & 18.669 & 1,32\% & N/A \\ \hline
+445.gobmk & 308.475 & 14.129 & 4,58\% & 4 \\ \hline
+462.libquantum & 16.297 & 928 & 5,69\% & 7 \\ \hline
+401.bzip2 & 38.831 & 2.158 & 5,56\% & N/A \\ \hline
+456.hmmer & 114.136 & 4.001 & 3,51\% & 0 \\ \hline
+\multicolumn{1}{|r|}{Total (Average)} & 275.070 & 6.968 & 3,96\% &  \\ \hline
+\end{tabular}
+\end{small}
+\end{center}
+\caption{\label{fig:unprunedTable}
+Unpruned analysis.}
+\end{figure*}
+
+%Prunning analysis
+\begin{figure*}[t!]
+\begin{center}
+\begin{small}
+\begin{tabular}{|l|r|r|r|r|r|}
+\hline
+Name & \multicolumn{1}{l|}{Num.
+instrumented instructions} & \multicolumn{1}{l|}{Num.
+instrumented instructions with e-SSA form} & \multicolumn{1}{l|}{Perc.
+reduction of e-SSA compared with original} & \multicolumn{1}{l|}{Num.
+instrumented instructions with u-SSA form} & \multicolumn{1}{l|}{Perc.
+reduction of u-SSA compared with original} \\ \hline
+470.lbm & 1.142 & 4 & 99,65\% & 4 & 99,65\% \\ \hline
+433.milc & 1.573 & 1.070 & 31,98\% & 1.065 & 32,29\% \\ \hline
+444.namd & 2.965 & 2.900 & 2,19\% & 2.900 & 2,19\% \\ \hline
+447.dealII & 37.234 & 29.870 & 19,78\% & 28.779 & 22,71\% \\ \hline
+450.soplex & 3.268 & 2.927 & 10,43\% & 2.897 & 11,35\% \\ \hline
+464.h264ref & 13.548 & 11.301 & 16,59\% & 11.342 & 16,28\% \\ \hline
+473.astar & 903 & 808 & 10,52\% & 806 & 10,74\% \\ \hline
+458.sjeng & 2.483 & 2.354 & 5,20\% & 2.190 & 11,80\% \\ \hline
+429.mcf & 165 & 164 & 0,61\% & 164 & 0,61\% \\ \hline
+471.omnetpp & 1.533 & 1.313 & 14,35\% & 1.313 & 14,35\% \\ \hline
+403.gcc & 16.330 & 15.282 & 6,42\% & 15.110 & 7,47\% \\ \hline
+445.gobmk & 13.947 & 12.563 & 9,92\% & 12.478 & 10,53\% \\ \hline
+462.libquantum & 943 & 820 & 13,04\% & 817 & 13,36\% \\ \hline
+401.bzip2 & 2.350 & 1.966 & 16,34\% & 1.966 & 16,34\% \\ \hline
+456.hmmer & 3.963 & 3.346 & 15,57\% & 3.304 & 16,63\% \\ \hline
+Total (Average) & 6.823 & 5.779 & 18,17\% & 5.676 & 19,09\% \\ \hline
+\end{tabular}
+\end{small}
+\end{center}
+\caption{\label{fig:prunningTable}
+Prunning analysis.}
+\end{figure*}
+
+%Classification analysis
+\begin{figure*}[t!]
+\begin{center}
+\begin{small}
+\begin{tabular}{|l|r|r|r|r|r|r|r|r|}
+\hline
+Name & \multicolumn{1}{l|}{Number of safe instructions} & \multicolumn{1}{l|}{Number of suspicious instructions} & \multicolumn{1}{l|}{Number of uncertain instructions} & \multicolumn{1}{l|}{Number of buggy instructions} & \multicolumn{1}{l|}{Number of suspicious instructions that produced overflows} & \multicolumn{1}{l|}{Percentage of suspicious instructions that produced overflows} & \multicolumn{1}{l|}{Number of Uncertain instructions that produced overflows} & \multicolumn{1}{l|}{Percentage of Uncertain instructions that produced overflows} \\ \hline
+470.lbm & 1.138 & 0 & 4 & 0 & 0 & 0,00\% & 0 & 0,00\% \\ \hline
+433.milc & 536 & 17 & 1.048 & 0 & 0 & 0,00\% & 11 & 1,05\% \\ \hline
+444.namd & 334 & 480 & 2.420 & 0 & 0 & 0,00\% & 12 & 0,50\% \\ \hline
+447.dealII & 6.188 & 39 & 28.740 & 0 & 0 & 0,00\% & 50 & 0,17\% \\ \hline
+450.soplex & 229 & 16 & 2.881 & 0 & 0 & 0,00\% & 13 & 0,45\% \\ \hline
+464.h264ref & 2.539 & 1.195 & 10.147 & 0 & 7 & 0,59\% & 167 & 1,65\% \\ \hline
+473.astar & 48 & 11 & 795 & 0 & 0 & 0,00\% & 0 & 0,00\% \\ \hline
+458.sjeng & 150 & 213 & 1.977 & 0 & 0 & 0,00\% & 68 & 3,44\% \\ \hline
+429.mcf & 1 & 0 & 164 & 0 & 0 & 0,00\% & 8 & 4,88\% \\ \hline
+471.omnetpp & 659 & 25 & 1.288 & 0 & 1 & 4,00\% & 2 & 0,16\% \\ \hline
+403.gcc & 3.365 & 1.045 & 14.065 & 0 & N/A & N/A & N/A & N/A \\ \hline
+445.gobmk & 1.509 & 742 & 11.736 & 0 & 0 & 0,00\% & 4 & 0,03\% \\ \hline
+462.libquantum & 104 & 12 & 805 & 0 & 0 & 0,00\% & 7 & 0,87\% \\ \hline
+401.bzip2 & 192 & 40 & 1.926 & 0 & N/A & N/A & N/A & N/A \\ \hline
+456.hmmer & 663 & 222 & 3.082 & 0 & 0 & 0,00\% & 0 & 0,00\% \\ \hline
+\end{tabular}
+\end{small}
+\end{center}
+\caption{\label{fig:classificationTable}
+Classification analysis.}
+\end{figure*}
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\columnwidth]{images/prunningChart}
+\end{center}
+\caption{\label{fig:prunningChart}
+Chart showing that our analysis was able to avoid inserting 24.93\%
+of the instrumentations, in average.}
+\end{figure}
+
+%Analysis about the tables
+
+%Analysis about the area chart (Amount of instrumentations pruned)
+
 %How we have done the dynamic tests  
@@
 
-%Analysis about the Table 1 of the spreadsheet (Unpruned Instrumentation)
-Figure~\ref{fig:unpruned_analysis} shows the number of instructions that has
-received instrumentation, without avoiding inserting the instrumentation by
-using any static analysis. We can see that the number of these instructions is low 
-(3,96 in the average). 
-Figure~\ref{fig:prunning_analysis} shows that it is possible to use static 
-analysis to avoid inserting the instrumentation in some instructions.
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\columnwidth]{images/performanceChart}
+\end{center}
+\caption{\label{fig:performanceChart}
+Comparison between the execution times with and without prunning, 
+normalized by the original program's execution time.}
+\end{figure}
+
+%Performance
+Figure~\ref{fig:performanceChart} compares the execution times of the original program (baseline),
+the instrumented program without avoiding to insert any instrumentation (black), and 
+the instrumented program avoiding to insert the instrumentation whenever possible (gray).
+We can see that the maximum overhead we have got was near 10\% in 464.h264ref, a program
+that relies heavily on arithmetics instructions. 
+As pointed by Dietz {\em et al.}~\cite{Dietz12}, we had some cases where surprisingly we had
+a little speedup, too. Nevertheless, the fully instrumented programs had an average slowdown of 3,2\%
+and the programs instrumented using the range analysis to do the prunning had an average slowdown
+of 1,7\%. That results indicate that even avoiding to insert just 25\% of the overflow checks,
+we still can recover almost 50\% of the overhead, because many of the 
+analyzed instructions are within loops.
 

### 2012-09-06 doc/Manuscript/CGO/paper.tex

@@
 size of primitive numeric types.
-For instance, the \texttt{int} type, in Java, ranges from $-2^{31}-1$ to
-$2^{31}$.
+For instance, the \texttt{int} type, in C++, ranges from $-2^{31}$ to
+$2^{31}-1$.
 Consequently, there exist numbers that cannot be represented by these types.
@@
 As an example, in 1996, the Ariane 5 rocket was lost due to an arithmetic
-overflow -- a bug that resulted in a loss of more than U\$370 million
-dollars~\cite{Dowson97}.
+overflow -- a bug that resulted in a loss of more than US\$370 million~\cite{Dowson97}.
+
 % TODO We need to explain better the wrapping-arithmetics, and we need to cite it
 % properly. What could be a good (official) citation?
-% And we need to cite
 
@@
 implementation of Brumley {\em et al.}
-Our range analysis algorithm has been able to remove XX\% of the overflow
+Our range analysis algorithm has been able to remove 24.93\% of the overflow
 checks created by the dynamic instrumentation framework.
@@
 \begin{tabular}{lcl}
-$x + \infty = \infty, x \neq -\infty$ & \mbox{\hspace{0.1cm}} & $x - \infty = - \infty, x \neq +\infty$ \\
+$x + \infty = \infty$ & \mbox{\hspace{0.1cm}} & $x - \infty = - \infty, x \neq +\infty$ \\
 $x \times \infty = \infty$ if $x > 0$ & & $x \times \infty = -\infty$ if $x < 0$ \\
@@
 Given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
-$\ub{\iota} = u$.
+$\ub{\iota} = u$, where $\lb{\iota}$ is the lower bound and $\ub{\iota}$ is
+the upper bound of a variable.
 We let \varset{V} be a set of constraint variables, and

### 2012-09-06 doc/Manuscript/CGO/paper.tex
Experimental Section

--- 2012-09-05 doc/Manuscript/CGO/paper.tex
@@
 
+\subsection{Static Range Analysis}
+\label{sub:range}
+
 \noindent
-\textbf{Time and Memory Complexity.}
+\textbf{Time and Memory Complexity: }
 Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
@@
 \noindent
-\textbf{Precision.}
-Our implementation of range analysis is remarkably precise, considering its
-runtime.
-Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
-about 25 minutes to go over a program with almost 900 basic blocks.
+\textbf{Precision: }
+Our implementation of range analysis offers a good tradeoff between
+precision and runtime.
+Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance,
+takes about 25 minutes to go over a program with almost 900 basic blocks.
 We analyze programs of similar size in less than one second.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%Static Analysis
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %How we have done the dynamic tests
+\subsection{The Instrumentation Library}
+\label{sub:inst}
+
 We have executed the instrumented programs of the integer benchmarks of SPEC
-2006 CPU to measure the overhead produced by our instrumentation in 
-real world programs.
-To do this, we have developed a little test framework that compiles, runs and
-collects the statistics about the execution of all the programs of SPEC 2006.
-The execution of the programs was made using the arguments and inputs of the "test"
-mode, which is faster to execute than the "ref" inputs.
-
-%Analysis about the Table 1 of the spreadsheet (Unpruned Instrumentation)
-Figure~\ref{fig:unpruned_analysis} shows the number of instructions that has
-received instrumentation, without avoiding inserting the instrumentation by
-using any static analysis. We can see that the number of these instructions is low 
-(3,96 in the average). 
-Figure~\ref{fig:prunning_analysis} shows that it is possible to use static 
-analysis to avoid inserting the instrumentation in some instructions.
+2006 CPU to probe the overhead imposed by our instrumentation.
+All the programs have been executed with its ``test" input set.
+We have not been able to run the binary that LLVM produces for SPEC's
+\texttt{gcc} in our environment, even without any of our passes.
+Figure~\ref{fig:unprunedTable} shows the percentage of instructions that
+we instrument, without the intervention of the range analysis.
+The number of instrumented instructions is relatively low, compared to the
+total number of instructions, because we only instrument six different
+LLVM bitcodes, in a set of 57 opcodes, not counting intrinsics.
+Figure~\ref{fig:unprunedTable} also shows how many instructions have caused
+overflows.
+On the average, 4.90\% of the instrumented sites have caused integer overflows.
+
+%Analysis about the amount of instructions
+\begin{figure}[htb]
+\begin{center}
+\begin{small}
+\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|}
+\hline
+Benchmark & \#I & \#II & \#II/\#I & \#O \\ \hline
+470.lbm & 13,724 & 1,142 & 8.32\% & 0 \\ \hline
+433.milc & 44,236 & 1,602 & 3.62\% & 11 \\ \hline
+444.namd & 100,276 & 3,234 & 3.23\% & 12 \\ \hline
+447.dealII & 1,381,408 & 36,157 & 2.62\% & 50 \\ \hline
+450.soplex & 136,367 & 3,158 & 2.32\% & 13 \\ \hline
+464.h264ref & 271,627 & 13,846 & 5.10\% & 167 \\ \hline
+473.astar & 19,243 & 857 & 4.45\% & 0 \\ \hline
+458.sjeng & 54,051 & 2,504 & 4.63\% & 68 \\ \hline
+429.mcf & 4,725 & 165 & 3,49\% & 8 \\ \hline
+471.omnetpp & 203,201 & 1,972 & 0.97\% & 2 \\ \hline
+403.gcc & 1,419,456 & 18,669 & 1.32\% & N/A \\ \hline
+445.gobmk & 308,475 & 14,129 & 4.58\% & 4 \\ \hline
+462.libquantum & 16,297 & 928 & 5.69\% & 7 \\ \hline
+401.bzip2 & 38,831 & 2,158 & 5.56\% & N/A \\ \hline
+456.hmmer & 114,136 & 4,001 & 3.51\% & 0 \\ \hline
+\multicolumn{1}{|r|}{Total (Average)} & 275.070 & 6.968 & 3,96\% &  \\ \hline
+\end{tabular*}
+\end{small}
+\end{center}
+\caption{\label{fig:unprunedTable}
+Instrumentation library without support of static range analysis.
+\#I: number of LLVM bitcode instructions in the original program.
+\#II: number of instructions that have been instrumented.
+\#O: number of actual overflow tests detected dynamically.}
+\end{figure}
+
+Figure~\ref{fig:prunningTable} shows how many checks our range analysis
+can avoid.
+In some benchmarks we obtain impressive results:
+For instance, the range analysis avoids 1,138 out of 1,142 checks in
+\texttt{470.lbm}.
+In other benchmarks, such as in \texttt{429.mcf}, we have been able to
+avoid only 1, out of 165 tests.
+In general we fare better in programs that bound input sizes via conditional
+tests, as \texttt{lbm} does.
+Using u-SSA, instead of e-SSA adds a negligible improvement onto our results.
+We speculate that this improvement is small because variables tend to be used
+a small number of times.
+Benoit {\em et al.}~\cite{Benoit08} have demonstrated that the vast majority
+of all the program variables are used less than five times in the program
+code.
+The u-SSA form only helps to avoid checks upon variables that are used more
+than once.
+
+%Prunning analysis
+\begin{figure}[t!]
+\begin{center}
+\begin{small}
+\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|r|}
+\hline
+Benchmark & \#II & \#E & \#I/\#E & \#U &
+\#I/\#U \\ \hline
+470.lbm & 1.142 & 4 & 99,65\% & 4 & 99,65\% \\ \hline
+433.milc & 1.573 & 1.070 & 31,98\% & 1.065 & 32,29\% \\ \hline
+444.namd & 2.965 & 2.900 & 2,19\% & 2.900 & 2,19\% \\ \hline
+447.dealII & 37.234 & 29.870 & 19,78\% & 28.779 & 22,71\% \\ \hline
+450.soplex & 3.268 & 2.927 & 10,43\% & 2.897 & 11,35\% \\ \hline
+464.h264ref & 13.548 & 11.342 & 16,59\% & 11.301 & 16,28\% \\ \hline
+473.astar & 903 & 808 & 10,52\% & 806 & 10,74\% \\ \hline
+458.sjeng & 2.483 & 2.354 & 5,20\% & 2.190 & 11,80\% \\ \hline
+429.mcf & 165 & 164 & 0,61\% & 164 & 0,61\% \\ \hline
+471.omnetpp & 1.533 & 1.313 & 14,35\% & 1.313 & 14,35\% \\ \hline
+403.gcc & 16.330 & 15.282 & 6,42\% & 15.110 & 7,47\% \\ \hline
+445.gobmk & 13.947 & 12.563 & 9,92\% & 12.478 & 10,53\% \\ \hline
+462.libquantum & 943 & 820 & 13,04\% & 817 & 13,36\% \\ \hline
+401.bzip2 & 2.350 & 1.966 & 16,34\% & 1.966 & 16,34\% \\ \hline
+456.hmmer & 3.963 & 3.346 & 15,57\% & 3.304 & 16,63\% \\ \hline
+Total (Average) & 6.823 & 5.779 & 18,17\% & 5.676 & 19,09\% \\ \hline
+\end{tabular*}
+\end{small}
+\end{center}
+\caption{\label{fig:prunningTable}
+Instrumentation library with support of static range analysis.
+\#II: number of instructions that have been instrumented without range
+analysis.
+\#E: number of instructions instrumented in the e-SSA form program.
+\#U: number of instructions instrumented in the u-SSA form program.}
+\end{figure}
+
+Figure~\ref{fig:classificationTable} shows how our range analysis
+classifies instructions.
+Out of all the 102,790 instructions that we have instrumented throughout the
+entire SPEC suite, 3.92\% are suspicious, 17.19\% are safe, and
+78.89\% are uncertain.
+We had, at first, speculated that overflows would be more common among
+suspicious instructions, as their bounds, inferred statically, go beyond the
+limits of their declared types.
+However, our experiments did not let us confirm this hypothesis.
+
+%Classification analysis
+\begin{figure}[t!]
+\begin{center}
+\begin{scriptsize}
+\begin{tabular}{|l|r|r|r|r|r|r|r|r|}
+\hline
+Bench & \#Sf & \#S & \#U & \#SO & \#SO/\#O & \#UO & \#UO/\#O
+\\ \hline
+lbm     & 1138 & 0    & 4 & 0 & 0,00\% & 0 & 0,00\% \\ \hline
+milc    & 536  & 17   & 1048  & 0 & 0,00\% & 11 & 1,05\% \\ \hline
+namd    & 334  & 480  & 2420  & 0 & 0,00\% & 12 & 0,50\% \\ \hline
+dealII  & 6188 & 39   & 28740  & 0 & 0,00\% & 50 & 0,17\% \\ \hline
+soplex  & 229  & 16   & 2881 & 0 & 0,00\% & 13 & 0,45\% \\ \hline
+h264ref & 2539 & 1195 & 10147 & 7 & 0,59\% & 167 & 1,65\% \\ \hline
+astar   & 48   & 11   & 795 & 0 & 0,00\% & 0 & 0,00\% \\ \hline
+sjeng   & 150  & 213  & 1977 &0 & 0,00\% & 68 & 3,44\% \\ \hline
+mcf     & 1    & 0    & 164 & 0 & 0,00\% & 8 & 4,88\% \\ \hline
+omnetpp & 659  & 25   & 1288 & 1 & 4,00\% & 2 & 0,16\% \\ \hline
+gcc     & 3365 & 1045 & 14065 & N/A & N/A & N/A & N/A \\ \hline
+gobmk   & 1509 & 742  & 11736 & 0 & 0,00\% & 4 & 0,03\% \\ \hline
+libqtum & 104  & 12   & 805 & 0 & 0,00\% & 7 & 0,87\% \\ \hline
+bzip2   & 192  & 40   & 1926 & N/A & N/A & N/A & N/A \\ \hline
+hmmer   & 663  & 222  & 3082 & 0 & 0,00\% & 0 & 0,00\% \\ \hline
+\end{tabular}
+\end{scriptsize}
+\end{center}
+\caption{\label{fig:classificationTable}
+How the range analysis classified arithmetic instructions.
+\#Sf: safe.
+\#S: suspicious.
+\#U: uncertain.
+\#SO: number of suspicious instructions that overflowed.
+\#UO: number of uncertain instructions that overflowed.
+}
+\end{figure}
+
+Figure~\ref{fig:percentage_elimination} shows, for the entire LLVM test
+suite, the percentage of overflow checks that our range analysis, with the
+e-SSA intermediate representation, could avoid.
+Each bar refers to a specific benchmark in the test suite.
+We only consider applications that had at least one instrumented
+instruction; the total number of benchmarks that meet this requirement is
+333.
+On the average, our range analysis avoid 24.93\% of the overflow checks.
+Considering the benchmarks in SPEC2006 only, this number is 20.57\%.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\columnwidth]{images/percentage_elimination}
+\end{center}
+\caption{\label{fig:percentage_elimination}
+Percentage of overflow checks that our range analysis has been able to remove.
+Each bar is a benchmark in the LLVM test suite.
+Benchmarks have been ordered by the effectiveness of the range analysis.
+On average, we have eliminated 24.93\% of the checks (geomean).}
+\end{figure}
+
+Figure~\ref{fig:spec_runtime} shows the impact of our instrumentation in the
+runtime of the SPEC benchmarks.
+We run each benchmark 20 times.
+The largest slowdown that we have observed, 11.83\%, happened in
+\texttt{h264ref}, the benchmark that presented the largest number of distinct
+sites where overflows happened dynamically. 
+On the average, the instrumented programs are 3.24\% slower than the
+original benchmarks.
+If we use the range analysis to eliminate overflow checks, this slowdown falls
+to 1.73\%.
+The range analysis, in this case, reduces the instrumentation overhead by
+53.40\%.
+This improvement is larger than the percentage of overflow checks that we
+avoid, e.g,. 20.57\%.
+We believe that this difference is due to the fact that we are able to
+eliminate checks on induction variables, as our range analysis can rely on
+the loop boundaries to achieve this end.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\columnwidth]{images/spec_runtime}
+\end{center}
+\caption{\label{fig:spec_runtime}
+Comparison between the execution times with and without prunning, 
+normalized by the original program's execution time.}
+\end{figure}
 
@@
 \noindent
-\paragraph{Dynamic Detection of Integer Overflows: }
+\textbf{Dynamic Detection of Integer Overflows: }
 Brumley {\em et al.}~\cite{Brumley07} have developed a tool, RICH, to
@@
 
-\paragraph{Static Detection of Integer Overflows: }
+\noindent
+\textbf{Static Detection of Integer Overflows: }
 Zhang {\em et al.}~\cite{Zhang09} have used static analysis to sanitize
@@
 
-\paragraph{Range Analysis:}
+\noindent
+\textbf{Range Analysis and sparse representations:}
 One of the goals of this paper is to describe a state-of-the-art range analysis
@@
 Furthermore, they only deal with comparisons between variables and constants.
-
-\noindent
-\paragraph{Sparse Dataflow Analysis: }
+The only work that we are aware of that has used range analysis to avoid
+overflow checks is Sol {\em et al}'s~\cite{Sol11}; however, their algorithm
+is much more limited than ours.
+Contrary to us, Sol {\em et al.} eliminate overflow checks in
+straight-line code traces produced by a just-in-time compiler, and they use the
+runtime values of variables to improve their precision.
+
 In this paper we provide a {\em sparse} implementation of range analysis.
@@
 points.
-The compiler related literature contains many descriptions of sparse
-data-flow analyses.
-Some analyses obtain sparsity by using specific program
+Some previous analyses have achieved sparsity by using specific program
 representations, like we did.
-Others rely on data-structures.
-In terms of data-structures, the first, and best known method proposed to
-support sparse data-flow analyses is Choi {\em et al.}'s {\em Sparse Evaluation
-Graph} (SEG)~\cite{Choi91}.
-
 One of the program representation that we discuss in
@@
 There exists many variants of SSU~\cite{Plevyak96,George03,Lo98}.
-The main difference is that our dataflow analysis propagates information
-forwardly, whereas the SSU form has been conceive for dataflow analyses
-that propagate information backwardly.
+The main difference to our representation is that our dataflow analysis
+propagates information forwardly, whereas the SSU form has been conceive for
+dataflow analyses that propagate information backwardly.
 Thus, it requires special nodes at branches to merge information from
@@
 
+This paper has presented a static range analysis algorithm that reduces
+the overhead necessary to secure programs against integer overflows.
+This algorithm is able to analyze inter-procedurally programs with
+half-a-million lines of code, which produce almost one million constraints, in
+ten seconds.
+We proposed the notion of ``future bounds" as a way to handle comparisons
+between variables, and relied on strong connected components to scale up to
+large programs.
+Although the overhead of guarding programs against integer overflows is small,
+as previous work has demonstrated, we believe that our technique is still
+important, as its gains are cumulative with improvements in hardware.
+Our entire code is publicly available, as a series of LLVM patches.
+The data that we show in this paper is also anonymously available, to ensure
+the reproducibility of our experiments.
+
 \bibliographystyle{plain}

### 2012-09-07 doc/Manuscript/CGO/paper.tex

@@
 \#II: number of instructions that have been instrumented.
-\#O: number of actual overflow tests detected dynamically.}
+\#O: number of instructions that actually overflowed in the dynamic tests.}
 \end{figure}
@@
 \hline
-Bench & \#Sf & \#S & \#U & \#SO & \#SO/\#O & \#UO & \#UO/\#O
+Bench & \#Sf & \#S & \#U & \#SO & \#SO/\#S & \#UO & \#UO/\#U
 \\ \hline
@@
 soplex  & 229  & 16   & 2881 & 0 & 0,00\% & 13 & 0,45\% \\ \hline
-h264ref & 2539 & 1195 & 10147 & 7 & 0,59\% & 167 & 1,65\% \\ \hline
+h264ref & 2539 & 1195 & 10147 & 7 & 0,59\% & 160 & 1,58\% \\ \hline
 astar   & 48   & 11   & 795 & 0 & 0,00\% & 0 & 0,00\% \\ \hline
@@
 mcf     & 1    & 0    & 164 & 0 & 0,00\% & 8 & 4,88\% \\ \hline
-omnetpp & 659  & 25   & 1288 & 1 & 4,00\% & 2 & 0,16\% \\ \hline
+omnetpp & 659  & 25   & 1288 & 1 & 4,00\% & 1 & 0,07\% \\ \hline
 gcc     & 3365 & 1045 & 14065 & N/A & N/A & N/A & N/A \\ \hline
@@
 \caption{\label{fig:classificationTable}
-How the range analysis classified arithmetic instructions.
+How the range analysis classified arithmetic instructions in the 
+u-SSA form program.
 \#Sf: safe.
@@
 The range analysis, in this case, reduces the instrumentation overhead by
-53.40\%.
+46.60\%.
 This improvement is larger than the percentage of overflow checks that we
@@
 Brumley {\em et al.}~\cite{Brumley07} have developed a tool, RICH, to
-dynamically detect integer overflows.
-Their goal is to secure C programs against integer overflow based
-vulnerabilities.
+secure C programs against integer overflow based vulnerabilities.
 The author's approach consists in instrumenting every integer operation that
@@
 range analysis.
-Furthermore, contrary to them, we do not try to detected statically the
-possibility of an integer overflow happening.
 Contrary to us, they do not transform the program to prevent or detect such event dynamically.

### 2012-09-09 doc/Manuscript/CGO/paper.tex
changed the paper to make it fit into 10 pages.

@@
 
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
+% TODO: explain that we have not observed overflows in the safe instructions.
 
@@
 \keywords
-Integer arithmetics, Overflow, Compiler, Range analysis
+Integer Overflow, Compiler, Range analysis
 
@@
 % Our contribution: range analysis.
-In this paper we discuss a simple and efficient range analysis algorithm that
+This paper describes the range analysis algorithm that
 we have developed to eliminate overflow checks in instrumented programs.
-Our algorithm has three core insights.
-Firstly, we rely on a three-phases approach to extract information from
-comparisons between variables, e.g., $x < y$.
-Previous algorithms either deal with these comparisons by resorting to
+As we show in Section~\ref{sec:range}, our algorithm has three core insights.
+Firstly, in Section~\ref{sub:micro} we propose a three-phases approach to extract
+information from comparisons between variables, e.g., $x < y$.
+Previous algorithms either deal with these comparisons via
 expensive relational analyses~\cite{Cousot78,Lakhdar11,Mine06}, or only
@@
 constants~\cite{Mahlke01,Patterson95,Stephenson00}.
-Secondly, we rely on strongly connected components to achieve better speed and
-precision.
-It is well-known that this technique is effective in speeding up static analysis
-algorithms~\cite[Sec 6.3]{Nielson99}.
-However, given our three-phases approach, we also achieve better precision
-by solving strong components in a topological ordering.
-Finally, we propose a program representation that allows us to perform a sparse
-range analysis while taking benefit from the knowledge that the execution of
-an instrumented program is overflow-free.
-This extensive live range splitting strategy is only valid if the instrumented
+Secondly, in Section~\ref{sub:prop} we show how we rely on strongly connected
+components to achieve speed and precision.
+It is well-known that this technique is effective in speeding up constraint
+resolution~\cite[Sec 6.3]{Nielson99}.
+However, given our three-phases approach, we improve precision
+by solving strong components in topological order.
+Finally, in Section~\ref{sub:splitting} we propose a new program representation that
+is more precise than other intermediate forms used to solve range
+analysis sparsely.
+This new live range splitting strategy is only valid if the instrumented
 program terminates whenever an integer overflow is detected.
-If we cannot rely on this guarantee, then our live range splitting strategy
-is more conservative, and produces the program representation that Bodik
-{\em et al.}~\cite{Bodik00} have called the Extended Static Single Assignment
+If we cannot rely on this guarantee, then our more conservative live range splitting
+strategy produces the program representation that Bodik
+{\em et al.}~\cite{Bodik00} call Extended Static Single Assignment
 form.
+In Section~\ref{sec:exp} we show that an interprocedural implementation of our
+algorithm analyzes programs with half-a-million lines of code in less than ten
+seconds.
+Furthermore, the speed and memory consumption of our range analysis grows linearly
+with the program size.
 
@@
 instrumentation library.
-This instrumentation framework, also a contribution of this paper, has been
-implemented in the LLVM compiler~\cite{Lattner04}.
-We have used it to log overflows in a vast number of programs, and in this
+This instrumentation framework, a contribution of this paper that we describe in
+Section~\ref{sec:dyn}, has been implemented in the LLVM compiler~\cite{Lattner04}.
+We have logged overflows in a vast number of programs, and in this
 paper we focus on SPEC CPU 2006.
-Our implementation has been able to reproduce the results obtained by
-previous work.
 We have re-discovered the integer overflows recently observed by
-Dietz {\em et al.}~\cite{Dietz12}, and the performance of our instrumentation
-library, without the support of range analysis, matches the state-of-the-art
-implementation of Brumley {\em et al.}
-Our range analysis algorithm has been able to remove 24.93\% of the overflow
+Dietz {\em et al.}~\cite{Dietz12}.
+The performance of our instrumentation library, even without the support of range
+analysis, is within the 5\% runtime overhead of Brumley {\em et al.}'s~\cite{Brumley07}
+state-of-the-art algorithm.
+The range analysis halves down this overhead.
+Our static analysis algorithm avoids 24.93\% of the overflow
 checks created by the dynamic instrumentation framework.
-All our experiments use an interprocedural version of range analysis.
-We inline small functions to make our implementation more context sensitive.
-Our experiments show that the speed and memory consumption of our range
-analysis grows linearly with the program size.
-By comparing the ranges that we determine statically with the results found by
-a profiler, we see that more than 50\% of the bounds that we found are within a
-factor of 50\% of the values that variables actually assume.
+With this support, the instrumented SPEC programs are only 1.73\% slower.
+Therefore, we show in this paper that securing programs against integer overflows
+is very cheap.
 
@@
 % The macro algorithm
-The range analysis algorithm that we propose in this paper works in a number
-of steps.
+Our range analysis algorithm works in a number of steps.
 Firstly, we convert the program to an intermediate representation that gives
@@
 $G = (\varset{C} \cup \varset{V}, E)$.
-We call the vertices in \varset{C} the {\em constraint nodes}, and we call the
-nodes in \varset{V} the {\em variable nodes}.
-If variable $V \in \varset{V}$ is used in constraint $C \in \varset{C}$, then
-we create an edge from $V$ to $C$.
+The vertices in \varset{C} are the {\em constraint nodes}, and the
+nodes in \varset{V} are the {\em variable nodes}.
+If $V \in \varset{V}$ is used in constraint $C \in \varset{C}$, then
+we create an edge $\overrightarrow{VC}$.
 If constraint $C \in \varset{C}$ defines variable $\varset{V} \in V$, then we
-create an edge from $C$ to $V$.
+create an edge from $\overrightarrow{CV}$.
 Figure~\ref{fig:ex_graph} shows the constraint graph that we build for the
@@
 
-We determine the ranges for each strongly connected component in a three phases
-approach.
-For each SCC, first we determine the growth pattern of each variable via a
-widening phase.
+We determine the ranges for each strongly connected component in three phases.
+First we determine the growth pattern of each variable in the component via
+widening.
 In the second step, once we know in which directions the values stored in each
 variable might grow, we replace futures by actual bounds.
-Finally, we have a narrowing phase that uses conditional tests to improve the
+Finally, a narrowing phase starting from conditional tests improves the
 precision of our results.
@@
 For instance, if a variable is only updated by sums with positive numbers, then
-we say that it only grows up.
+it only grows up.
 If, instead, a variable is only updated by sums with negative numbers, then it
@@
 \rightarrow [-\infty, \infty]$.
-Therefore, we can determine the growth behavior of each constraint variable in
-a strong component in linear time on the number of constraints in that
-component.
+Therefore, we determine the growth behavior of each constraint variable in
+a strong component in linear time on the number of constraints.
 Figure~\ref{fig:ex_partition_grow_crop}(b) shows the abstract state of the
@@
 and $v_{2f} = v_2 \sqcap [\fun{ft}(v_1) + 1, +\infty]$.
-Notice that a variable $v$ can never be associated with a future bound to
+A variable $v$ can never be associated with a future bound to
 itself, e.g., $\fun{ft}(v)$.
@@
 $u$ with $\fun{ft}(v)$, then $u$ is a fresh name created to split the live range
-of $v$, given that $v$ was used in a conditional.
+of $v$.
 
@@
 \textbf{Time and Memory Complexity: }
-Figure~\ref{fig:TimeCorr} provides a visual comparison between the time to
-run our range analysis and the size of the input programs.
+Figure~\ref{fig:TimeCorr} compares the time to run our range analysis with the size of
+the input programs.
 We show data for the 100 largest benchmarks in our test suite, in number
@@
 categories.
-The first category, which we call $1$, contains those bounds that are tight:
-during the execution of the program, the variable has been assigned an upper,
+The first category, called $1$, contains tight bounds:
+during program execution, the variable has been assigned an upper,
 or lower limit, that equals the limit inferred statically.
-The second category, which we call $n$, contains the bounds that are
+The second category, called $n$, contains the bounds that are
 within twice the value inferred statically.
@@
 2006 CPU to probe the overhead imposed by our instrumentation.
-All the programs have been executed with its ``test" input set.
+These programs have been executed with their ``test" input sets.
 We have not been able to run the binary that LLVM produces for SPEC's
-\texttt{gcc} in our environment, even without any of our passes.
+\texttt{gcc} in our environment, even without any of our passes, due to a missing
+\texttt{ctype.h} header file.
 Figure~\ref{fig:unprunedTable} shows the percentage of instructions that
@@
 \caption{\label{fig:unprunedTable}
-Instrumentation library without support of static range analysis.
+Instrumentation without support of range analysis.
 \#I: number of LLVM bitcode instructions in the original program.
@@
 Figure~\ref{fig:prunningTable} shows how many checks our range analysis
-can avoid.
-In some benchmarks we obtain impressive results:
-For instance, the range analysis avoids 1,138 out of 1,142 checks in
+avoids.
+Some results are expressive: the range analysis avoids 1,138 out of 1,142 checks in
 \texttt{470.lbm}.
@@
 classifies instructions.
-Out of all the 102,790 instructions that we have instrumented throughout the
-entire SPEC suite, 3.92\% are suspicious, 17.19\% are safe, and
+Out of all the 102,790 instructions that we have instrumented in SPEC, 3.92\% are
+suspicious, 17.19\% are safe, and
 78.89\% are uncertain.
@@
 \caption{\label{fig:percentage_elimination}
-Percentage of overflow checks that our range analysis has been able to remove.
+Percentage of overflow checks that our range analysis removes.
 Each bar is a benchmark in the LLVM test suite.
@@
 \caption{\label{fig:spec_runtime}
-Comparison between the execution times with and without prunning, 
+Comparison between execution times with and without prunning, 
 normalized by the original program's execution time.}
@@
 Brumley {\em et al.}~\cite{Brumley07} have developed a tool, RICH, to
-secure C programs against integer overflow based vulnerabilities.
+secure C programs against integer overflows.
 The author's approach consists in instrumenting every integer operation that
 might cause an overflow, underflow, or data loss.
-The main result of Brumley {\em et al.} is the verification that instrumenting
-programs to guard them against integer overflows does not compromise their
+The main result of Brumley {\em et al.} is the verification that guarding programs
+against integer overflows does not compromise their
 performance significantly: the average slowdown across four large applications
@@
 engineering point-of-view; hence, performance is not a concern.
-The authors have used IOC to carry out a lengthy study about the occurrences
+The authors have used IOC to carry out a study about the occurrences
 of overflows in real-world programs, and have found that these events are very
@@
 the \emph{Static Single Use} form (SSU).
-There exists many variants of SSU~\cite{Plevyak96,George03,Lo98}.
+There exists other variants of SSU~\cite{Plevyak96,George03}.
 The main difference to our representation is that our dataflow analysis
@@
 the overhead necessary to secure programs against integer overflows.
-This algorithm is able to analyze inter-procedurally programs with
-half-a-million lines of code, which produce almost one million constraints, in
+This algorithm analyzes inter-procedurally programs with
+half-a-million lines of code, i.e., almost one million constraints, in
 ten seconds.
-We proposed the notion of ``future bounds" as a way to handle comparisons
-between variables, and relied on strong connected components to scale up to
-large programs.
+We proposed the notion of ``future bounds" to handle comparisons
+between variables, and tested different program representations to improve our
+precision.
 Although the overhead of guarding programs against integer overflows is small,
 as previous work has demonstrated, we believe that our technique is still
-important, as its gains are cumulative with improvements in hardware.
+important, as some of these programs will be executed millions of times.
 Our entire code is publicly available, as a series of LLVM patches.

### 2012-09-09 doc/Manuscript/CGO/paper.tex
Polished the paper a little bit.

@@
 % TODO: explain that we have not observed overflows in the safe instructions.
+% TODO: explain that u-SSA does not yield noticeable performance improvements.
 
@@
 The integer primitive type has upper and lower bounds in many programming
-languages, including C, C++ and Java.
-This limitation might lead programs that manipulate large integer numbers to
+languages, including C, and Java.
+These limits might lead programs that manipulate large integer numbers to
 produce unexpected results due to overflows.
@@
 of these overflows.
-In this paper we present an algorithm that relies on static range analysis to
-eliminate this instrumentation whenever possible.
-We present a novel live range splitting strategy that refine the precision of
-our range analysis.
-We have used this algorithm to improve the quality of a dynamic instrumentation
+In this paper we present the first algorithm that uses static range analysis to
+avoid this instrumentation whenever possible.
+Our range analysis contains novel techniques, such as a live range splitting
+strategy that refines the precision of our results.
+We have used this algorithm to avoid some checks created by a dynamic instrumentation
 library that we have implemented in LLVM.
 This framework has been used to detect overflows in hundreds of C/C++ programs.
-As a testimony of its effectiveness, our range analysis has been able to eliminate
-19\% of all the overflow checks necessary to secure the C programs in the
-SPEC CPU 2006 benchmark suite.
+As a testimony of its effectiveness, our range analysis has been able to avoid
+25\% of all the overflow checks necessary to secure the C programs in the
+LLVM test suite.
+This reduction has reduced 50\% of the runtime overhead of instrumentation.
 \end{abstract}
@@
 If a number $n$ is too large to fit into a primitive data type $T$, then $n$'s
-value wraps around, and $n \mbox{module} T_{max}$ wounds up represented
-instead.
+value wraps around, and we obtain $n \ \mbox{module} \ T_{max}$.
 There are situations in which this semantics is acceptable~\cite{Dietz12}.
@@
 form.
-In Section~\ref{sec:exp} we show that an interprocedural implementation of our
+In Section~\ref{sub:range} we show that an interprocedural implementation of our
 algorithm analyzes programs with half-a-million lines of code in less than ten
@@
 From the lattice $\varset{Z}$ we define the product lattice
-$\varset{Z}^2$, which is defined as follows:
+$\varset{Z}^2$ as follows:
 %
@@
 \begin{cases}
-[\mbox{max}(a_1, b_1), \mbox{min}(a_2, b_2)], \ \mbox{if} \ a_1 \leq b_1 \leq a_2  \\ \mbox{or} \ b_1 \leq a_1 \leq b_2 \\
+[\mbox{max}(a_1, b_1), \mbox{min}(a_2, b_2)], \ \mbox{if} \ a_1 \leq b_1 \leq a_2  \\ \ \ \ \ \mbox{or} \ b_1 \leq a_1 \leq b_2 \\
 [a_1, a_2] \sqcap [b_1, b_2] = \emptyset, \ \mbox{otherwise}
@@
 If constraint $C \in \varset{C}$ defines variable $\varset{V} \in V$, then we
-create an edge from $\overrightarrow{CV}$.
+create an edge $\overrightarrow{CV}$.
 Figure~\ref{fig:ex_graph} shows the constraint graph that we build for the
 program in Figure~\ref{fig:ex_eSSA_cgo}(b).
-If $V$ is used by $C$ as the input of a future, then the edge from
+Our algorithm introduces the notion of {\em future ranges}, which we use to extract
+range information from comparisons between variables.
+In Section~\ref{sub:splitting} we explain how futures are created.
+If $V$ is used by $C$ as the input of a future range, then the edge from
 $V$ to $C$ represents what Ferrante {\em et al.} call a {\em control
@@
 After building the constraint graph, we find its strongly connected components.
-We collapse these components in super nodes, and then proceed to propagate
+We collapse these components in super nodes, and then propagate
 ranges along the resulting digraph.
@@
 singletons.
-For instance, 99.11\% of the SCCs in SPEC CPU 2006 gcc (\texttt{403.gcc}) have
+For instance, 99.11\% of the SCCs in SPEC CPU 2006 dealII (\texttt{447.dealII}) have
 only one node.
 Moreover, the composite components usually contain a small number of nodes.
-As an example, the largest component in \texttt{403.gcc}, has 2,131 nodes,
-even though gcc's constraint graph has over one million nodes.
+As an example, the largest component in dealII has 2,131 nodes,
+even though dealII's constraint graph has over one million nodes.
 This large SCC exists due to a long chain of mutually recursive function calls.
@@
 %
-In order to correctly replace a future $\fun{ft}(V)$ that limits a constraint
+To correctly replace a future $\fun{ft}(V)$ that limits a constraint
 variable $V'$, we need to have already applied the growth analysis onto $V$.
 Had we considered only data dependence edges, then it would be possible
-that $V'$ be analyzed before $V$.
+that $V'$'s strong component would be analyzed before $V$'s.
 However, because of control dependence edges, this case cannot happen.
@@
 operator in our running example.
-Ranges initially improve due to the two conditional tests in the program.
+Ranges improve due to the two conditional tests in the program.
 Firstly, we have that $I[I_t] = I[I_1] \sqcap [-\infty, 98]$, which gives us
@@
 We use the notation $v = \bullet$ to denote a definition of variable $v$, and
-$\bullet = v$ to denote a use of that variable.
+$\bullet = v$ to denote a use of it.
 Figure~\ref{fig:ex_ir}(b) shows the example program converted to the SSA format.
-The different definitions of variable $u$ have been renamed, and a
-$\phi$-function re-combines these definitions into a single name.
+Different definitions of variable $u$ have been renamed, and a
+$\phi$-function joins these definitions into a single name.
 The SSA form sparsifies a dataflow analysis that only extracts information from
@@
 The live range of variable $v_1$ has been divided right after its use.
-This representation is useful to analyses that learn information from the way
+This representation assists analyses that learn information from the way
 that variables are used, and propagate this information forwardly.
@@
 We discuss in the rest of this section a few interesting cases.
-If we are dealing with an unsigned \textsc{Sub} instruction, e.g,
+When dealing with an unsigned \textsc{Sub} instruction, e.g,
 $x = o_1 \ -_u \ o_2$, then a single check is enough to detect the bad
 behavior: $o_1 < o_2$.
-This is sound because if $o_2$ is greater than $o_1$, then we assume that it is
+If $o_2$ is greater than $o_1$, then we assume that it is
 a bug to try to represent a negative number in unsigned arithmetics.
@@
 %Overview about the instrumentation technique
-Our instrumentation library works by adding new instructions to
+Our instrumentation library inserts new instructions into
 the target program.
@@
 the general modus operandi is the same.
-The dynamic tests check for overflows after they happen.
+Dynamic tests check for overflows after they happen.
 The code that we insert to detect the overflow diverts the program flow in case
@@
 and after the instrumentation.
-From this figure, it is clear that the instrumented program will be larger
-than the original code.
+Clearly the instrumented program will be larger than the original code.
 Figure~\ref{fig:num_instructions} shows how many LLVM instructions are
@@
 code is not in the program's main path.
-Nevertheless, as we show empirically, this growth is not too large when
+Nevertheless, as we show empirically, this growth is small when
 compared to the total size of our benchmarks, because most of the instructions
@@
 Therefore, the overall slowdown that the instrumentation causes is usually
-small.
+small, and the experiments in Section~\ref{sub:inst} confirm this
+observation.
 
@@
 function in Figure~\ref{fig:instrumented_cfg}.
-Our library gives the user three options to handle them. 
+Our library gives the user three options to handle overflows. 
 First option: no-op.
@@
 This option terminates the program once an overflow is detected.
-Thus, it disallows undefined behavior due to integer overflows.
+Thus, it disallows undefined behavior due to integer overflows, and gives us the
+opportunity to use the u-SSA form to get extra precision.
 
@@
 the input programs.
-We show data for the 100 largest benchmarks in our test suite, in number
+We show data for the 100 largest benchmarks in our test suite, considering the number
 of variable nodes in the constraint graph.
@@
 has 1,131 variable nodes in the constraint graph, in 20ms.
-We take 15.91 sec to analyze our largest benchmark, \texttt{403.gcc}, which,
-after function inlining, has 1,266,273 assembly instructions, and a
+We take 9.91 sec to analyze our largest benchmark, \texttt{403.gcc}, which,
+after function inlining, has 1,419,456 assembly instructions, and a
 constraint graph with 679,652 variable nodes.
@@
 Correlation between program size (number of var nodes in constraint
-graphs after inlining) and analysis runtime (ms).
+graphs) and analysis runtime (ms).
 Each point represents a benchmark.
@@
 The experiments also reveal that the memory consumption of our implementation
-is linear with the program size.
-Figure~\ref{fig:MemCorr} plots these two quantities together.
+grows linearly with the program size.
+Figure~\ref{fig:MemCorr} plots these two quantities.
 The linear correlation, in this case, is even stronger than that found in
-Figure~\ref{fig:TimeCorr}, which compares runtime and program size: the
-coefficient of determination is 0.9947.
+Figure~\ref{fig:TimeCorr}: the coefficient of determination is 0.9947.
 The figure only shows our 100 largest benchmarks.
-Again, SPEC \texttt{403.gcc} is the heaviest benchmark, requiring
+Again, SPEC \texttt{403.gcc} is the largest benchmark, requiring
 265,588KB to run.
@@
 that scales to very large programs.
-Nevertheless, our results are far from being trivial.
+Nevertheless, our results are not trivial.
 We have implemented a dynamic profiler that measures, for each variable,
@@
 We chose Stanford because these benchmarks do not read data from external
-files; hence, imprecisions are due to library functions that we cannot
+files; hence, imprecisions are due exclusively to library functions that we cannot
 analyze.
@@
 The third category, $n^2$, contains variables whose actual value is within
-a quadratic factor from the estimated value.
+a quadratic factor of the estimated value.
 In our example, $v$'s upper bound would have to be at most $10$ for it to
@@
 We have not been able to run the binary that LLVM produces for SPEC's
-\texttt{gcc} in our environment, even without any of our passes, due to a missing
-\texttt{ctype.h} header file.
+\texttt{gcc} in our environment, even without any of our passes, due to an
+incompatible \texttt{ctype.h} header.
 Figure~\ref{fig:unprunedTable} shows the percentage of instructions that
@@
 462.libquantum & 16,297 & 928 & 5.69\% & 7 \\ \hline
-401.bzip2 & 38,831 & 2,158 & 5.56\% & N/A \\ \hline
+401.bzip2 & 38,831 & 2,158 & 5.56\% & 0 \\ \hline
 456.hmmer & 114,136 & 4,001 & 3.51\% & 0 \\ \hline
@@
 However, our experiments did not let us confirm this hypothesis.
+To check the correctness of our approach, we have instrumented the safe instructions,
+but have not observed any overflow caused by them.
 
@@
 the loop boundaries to achieve this end.
+We have not noticed any runtime difference between programs converted to
+e-SSA form or u-SSA form.
 
@@
 important, as some of these programs will be executed millions of times.
-Our entire code is publicly available, as a series of LLVM patches.
-The data that we show in this paper is also anonymously available, to ensure
-the reproducibility of our experiments.
+Our entire code plus the data that we show in this paper is publicly available.
+A more extensive list of experiments is also available in our technical report.
 

### 2012-09-10 doc/Manuscript/CGO/paper.tex

@@
 \newcommand{\code}[1]{\textbf{#1}}
+
+\setlength{\textfloatsep}{12.0pt plus 1.0pt minus 1.0pt}
 
@@
 stored in the file would overwrite memory in the stack of activation records.
-By carefully crafting an input file, and adversary can, in this way, overwrite
+By carefully crafting an input file, an adversary can, in this way, overwrite
 the return address of \texttt{read\_matrix}, diverting program's execution to
@@
 incompatible \texttt{ctype.h} header.
+In addition, we have not been able to collect the statistics regarding
+the overflows occurred in SPEC's \texttt{bzip2}, because the log file
+has became too large, due to the occurrence of more than 3.000.000.000 
+overflows. Thus, the machine ran out of memory.
 Figure~\ref{fig:unprunedTable} shows the percentage of instructions that
@@
 462.libquantum & 16,297 & 928 & 5.69\% & 7 \\ \hline
-401.bzip2 & 38,831 & 2,158 & 5.56\% & 0 \\ \hline
+401.bzip2 & 38,831 & 2,158 & 5.56\% & N/A \\ \hline
 456.hmmer & 114,136 & 4,001 & 3.51\% & 0 \\ \hline

### 2012-11-06 doc/Manuscript/CGO/paper.tex
Added comments from the reviewers to the paper.

@@
 Integer Overflows}
-%\title{Combining Static and Dynamic Analyses to Secure Programs against
-%Integer Overflows}
 %\subtitle{A case study of the Dinamica EGO toolkit}
 
-\authorinfo{Removed due to blind review}
-           {No affiliation given}
-           {No e-mail given}
+\authorinfo{Raphael Ernani Rodrigues, Victor Hugo Sperle Campos,
+Fernando Magno Quint\~{a}o Pereira}
+           {Department of Computer Science -- The Federal University of Minas gerais (UFMG) -- Brazil}
+           {\{raphael,victorsc,fernando\}@dcc.ufmg.br}
 
@@
 surveilled, and the runtime checks cost time.
-Zhang {\em et al.}~\cite{Zhang09} have proposed to eliminate some of this
+Zhang {\em et al.}~\cite{Zhang10} have proposed to eliminate some of this
 overhead via a tainted flow analysis.
@@
 As we show in Section~\ref{sec:range}, our algorithm has three core insights.
-Firstly, in Section~\ref{sub:micro} we propose a three-phases approach to extract
+Firstly, in Section~\ref{sub:micro} we propose a three-phase approach to extract
 information from comparisons between variables, e.g., $x < y$.
@@
 resolution~\cite[Sec 6.3]{Nielson99}.
-However, given our three-phases approach, we improve precision
+However, given our three-phase approach, we improve precision
 by solving strong components in topological order.
@@
 by Brumley {\em et al.}~\cite{Brumley07}, Dietz {\em et al.}~\cite{Dietz12}, and
-Zhang {\em et al.}~\cite{Zhang09}.
+Zhang {\em et al.}~\cite{Zhang10}.
 The function \texttt{read\_matrix} reads a matrix of bytes from a file.
@@
 
+Notice that $(\infty -\infty)$ is not well-defined.
 From the lattice $\varset{Z}$ we define the product lattice
@@
 mapping from these variables to intervals in $\varset{Z}^2$.
-Our objective is to solve a constraint system $C$, formed by constraints such
-as those seen in Figure~\ref{fig:eval_function}(left).
+Our objective is to solve a constraint system \varset{C}, formed by constraints
+such as those seen in Figure~\ref{fig:eval_function}(left).
 We let the $\phi$-functions be as defined by Cytron
@@
 \textsc{Range Analysis Problem} \\
-\textbf{Input:} a set $C$ of constraints ranging over a set \varset{V} of
-variables. \\
+\textbf{Input:} a set \varset{C} of constraints ranging over a set
+\varset{V} of variables. \\
 \textbf{Output:} a mapping I such that, for any variable
@@
 resulting digraph topologically.
-Finally, for each strong component, we apply a three-phases approach to determine
+Finally, for each strong component, we apply a three-phase approach to determine
 the ranges of the variables.
@@
 In Section~\ref{sub:splitting} we explain how futures are created.
-If $V$ is used by $C$ as the input of a future range, then the edge from
-$V$ to $C$ represents what Ferrante {\em et al.} call a {\em control
+If $V$ is used by constraint $C$ as the input of a future range, then the edge
+from $V$ to $C$ represents what Ferrante {\em et al.} call a {\em control
 dependence}~\cite[p.323]{Ferrante87}.
@@
 
-\subsection{A Three-Phases Approach to Solve Strong Components}
+\subsection{A Three-Phase Approach to Solve Strong Components}
 \label{sub:micro}
 
-We determine the ranges for each strongly connected component in three phases.
+We find the ranges of the variables in each strongly connected component in
+three phases.
 First we determine the growth pattern of each variable in the component via
@@
 \noindent
-\textbf{Widening: } we start solving constraints by determining how each program
-variable might grow.
+\textbf{Widening: } we start solving constraints by determining how each
+program variable might grow.
 For instance, if a variable is only updated by sums with positive numbers, then
@@
 Some variables can also grow in both directions.
-We discover these growth patterns by abstractly interpreting the constraints that
-constitute the strongly connected component.
+We discover these growth patterns by abstractly interpreting the constraints
+that constitute the strongly connected component.
 We ensure termination via a widening operator.
@@
 We let $\lb{[l, u]} = l$ and $\ub{[l, u]} = u$.
+We let $\bot$ denote non-itialized intervals, so that $[\bot, \bot] \sqcup
+[l, u] = [l, u]$.
+This operation only happens at $\phi$-nodes, because we evaluate constraints in
+topological order.
 The map $I$ and the abstract evaluation function $e$ are determined as in
 Figure~\ref{fig:eval_function}.
-We have a different implementation of $e$ for each operation that the
+We have an implementation of $e$ for each operation that the
 target programming language provides.
@@
 imprecise, and 35.40\% of our upper limits are imprecise.
-This result is on pair with those obtained by more costly analysis, such as
+This result is on par with those obtained by more costly analysis, such as
 Stephenson {\em et al.}'s~\cite{Stephenson00}.
@@
 incompatible \texttt{ctype.h} header.
-In addition, we have not been able to collect the statistics regarding
-the overflows occurred in SPEC's \texttt{bzip2}, because the log file
-has became too large, due to the occurrence of more than 3.000.000.000 
-overflows. Thus, the machine ran out of memory.
+In addition, we have not been able to collect the statistics about
+the overflows that occurred in SPEC's \texttt{bzip2}, because the log file
+was too large.
+We verified more than 3.000.000.000  overflows in this program.
 Figure~\ref{fig:unprunedTable} shows the percentage of instructions that
@@
 \hline
-Benchmark & \#II & \#E & \#I/\#E & \#U &
-\#I/\#U \\ \hline
-470.lbm & 1.142 & 4 & 99,65\% & 4 & 99,65\% \\ \hline
-433.milc & 1.573 & 1.070 & 31,98\% & 1.065 & 32,29\% \\ \hline
-444.namd & 2.965 & 2.900 & 2,19\% & 2.900 & 2,19\% \\ \hline
-447.dealII & 37.234 & 29.870 & 19,78\% & 28.779 & 22,71\% \\ \hline
-450.soplex & 3.268 & 2.927 & 10,43\% & 2.897 & 11,35\% \\ \hline
-464.h264ref & 13.548 & 11.342 & 16,59\% & 11.301 & 16,28\% \\ \hline
-473.astar & 903 & 808 & 10,52\% & 806 & 10,74\% \\ \hline
-458.sjeng & 2.483 & 2.354 & 5,20\% & 2.190 & 11,80\% \\ \hline
-429.mcf & 165 & 164 & 0,61\% & 164 & 0,61\% \\ \hline
-471.omnetpp & 1.533 & 1.313 & 14,35\% & 1.313 & 14,35\% \\ \hline
-403.gcc & 16.330 & 15.282 & 6,42\% & 15.110 & 7,47\% \\ \hline
-445.gobmk & 13.947 & 12.563 & 9,92\% & 12.478 & 10,53\% \\ \hline
-462.libquantum & 943 & 820 & 13,04\% & 817 & 13,36\% \\ \hline
-401.bzip2 & 2.350 & 1.966 & 16,34\% & 1.966 & 16,34\% \\ \hline
-456.hmmer & 3.963 & 3.346 & 15,57\% & 3.304 & 16,63\% \\ \hline
-Total (Average) & 6.823 & 5.779 & 18,17\% & 5.676 & 19,09\% \\ \hline
+Benchmark & \#II & \#E & \%(II, E) & \#U &
+\%(II, U) \\ \hline
+lbm & 1.142 & 4 & 99,65\% & 4 & 99,65\% \\ \hline
+milc & 1.573 & 1.070 & 31,98\% & 1.065 & 32,29\% \\ \hline
+namd & 2.965 & 2.900 & 2,19\% & 2.900 & 2,19\% \\ \hline
+dealII & 37.234 & 29.870 & 19,78\% & 28.779 & 22,71\% \\ \hline
+soplex & 3.268 & 2.927 & 10,43\% & 2.897 & 11,35\% \\ \hline
+h264ref & 13.548 & 11.342 & 16,59\% & 11.301 & 16,28\% \\ \hline
+astar & 903 & 808 & 10,52\% & 806 & 10,74\% \\ \hline
+sjeng & 2.483 & 2.354 & 5,20\% & 2.190 & 11,80\% \\ \hline
+mcf & 165 & 164 & 0,61\% & 164 & 0,61\% \\ \hline
+omnetpp & 1.533 & 1.313 & 14,35\% & 1.313 & 14,35\% \\ \hline
+gcc & 16.330 & 15.282 & 6,42\% & 15.110 & 7,47\% \\ \hline
+gobmk & 13.947 & 12.563 & 9,92\% & 12.478 & 10,53\% \\ \hline
+libquantum & 943 & 820 & 13,04\% & 817 & 13,36\% \\ \hline
+bzip2 & 2.350 & 1.966 & 16,34\% & 1.966 & 16,34\% \\ \hline
+hmmer & 3.963 & 3.346 & 15,57\% & 3.304 & 16,63\% \\ \hline
+Average & 6.823 & 5.779 & 18,17\% & 5.676 & 19,09\% \\ \hline
 \end{tabular*}
@@
 classifies instructions.
-Out of all the 102,790 instructions that we have instrumented in SPEC, 3.92\% are
-suspicious, 17.19\% are safe, and
-78.89\% are uncertain.
+Out of all the 102,790 instructions that we have instrumented in SPEC,
+3.92\% are suspicious, 17.19\% are safe, and 78.89\% are uncertain.
+This means that we found precise bounds to $3.92 + 17.19 = 21.11\%$ of all
+the program variables, and that 78.98\% of them are bound to intervals with
+at least one unknown limit.
 We had, at first, speculated that overflows would be more common among
@@
 \caption{\label{fig:spec_runtime}
-Comparison between execution times with and without prunning, 
+Comparison between execution times with and without pruning, 
 normalized by the original program's execution time.}
@@
 Brumley's, even without the support of the static range analysis.
+None of the above works use range analysis to eliminate overflow checks.
+This analysis, in our case, could eliminate approximately 45\% of the
+tests.
 
@@
 \textbf{Static Detection of Integer Overflows: }
-Zhang {\em et al.}~\cite{Zhang09} have used static analysis to sanitize
+Zhang {\em et al.}~\cite{Zhang10} have used static analysis to sanitize
 programs against integer overflow based vulnerabilities.
@@
 plethora of old and recent range analysis
-solvers~\cite{Gough94,Mahlke01,Oh12,Patterson95,Stephenson00,Su05}.
+solvers~\cite{Gough94,Gawlitza09,Logozzo10,Mahlke01,Oh12,Patterson95,Stephenson00,Su05}.
 The use of futures to handle comparisons between variables is a contribution
@@
 the \emph{Static Single Use} form (SSU).
-There exists other variants of SSU~\cite{Plevyak96,George03}.
+There exists other variants of SSU~\cite{Lo98,Plevyak96,George03}.
 The main difference to our representation is that our dataflow analysis
-propagates information forwardly, whereas the SSU form has been conceive for
+propagates information forwardly, whereas the SSU form has been conceived for
 dataflow analyses that propagate information backwardly.

### 2012-11-09 doc/Manuscript/CGO/paper.tex
Put the paper in the CGO format.

@@
-\documentclass[preprint]{sigplanconf}[10pt]
+\documentclass{sigplanconf}[10pt]
 
@@
 of these overflows.
-In this paper we present the first algorithm that uses static range analysis to
+In this paper we present an algorithm that uses static range analysis to
 avoid this instrumentation whenever possible.
-Our range analysis contains novel techniques, such as a live range splitting
-strategy that refines the precision of our results.
+Our range analysis contains novel techniques, such as a notion of ``future"
+bounds to handle comparisons between variables.
 We have used this algorithm to avoid some checks created by a dynamic instrumentation
@@
 
-\subsection{The Danger that Integer Overflows Represent}
-\label{sub:danger}
-
-% A brief overview about integer overflows.
-In addition to being a source of several bugs that are hard to identify, integer 
-overflows can represent a serious security risk.
-As pointed by Dietz {\em et al.}~\cite{Dietz12}, integer overflows are the
-sources of vulnerabilities in widely used software, such as OpenSSH and
-Firefox.
-These vulnerabilities could give an adversary the opportunity to execute
-arbitrary code in the compromised application. 
-We will demonstrate this with an example.
-The $n$-bits two's complement arithmetics wraps around at multiples of
-$2^n$.
-As an example the representation of $16 \times 16$ in 8-bits two's complement
-is $0$, because $0 \equiv (16\times16) \mod 2^8$.
-Programmers can use this well-defined behavior in benign ways.
-For instance, the wrapping arithmetics provides a cheap $\mod 2^n$ operator,
-which is used in the implementation of hash-functions or random number
-generators.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\columnwidth]{images/ex_buffer_overflow}
-\end{center}
-\caption{\label{fig:ex_buffer_overflow}
-An example of an exploitable integer overflow vulnerability.}
-\end{figure}
-
-% An example of vulnerability.
+\noindent
+\textbf{The Danger that Integer Overflows Represent: }
 Figure~\ref{fig:ex_buffer_overflow} illustrates an example of integer
@@
 on a program that is apparently guarded against this type of exploit.
-For simplicity, we will use the C datatype \texttt{char}, which represents
-8-bit long numbers.
+We will use the C datatype \texttt{char}, which represents
+signed 8-bit numbers, so that the interested reader can easily
+reconstruct this exploit.
 Examples of actual vulnerabilities, with larger primitive types, are given
-by Brumley {\em et al.}~\cite{Brumley07}, Dietz {\em et al.}~\cite{Dietz12}, and
-Zhang {\em et al.}~\cite{Zhang10}.
-The function \texttt{read\_matrix} reads a matrix of bytes from a file.
-This matrix is stored as a linear sequence of bytes in a vector \texttt{out}.
-The first and second bytes in the file, e.g., \texttt{w} and \texttt{h},
-determine the number of rows and columns in the matrix.
-If the \texttt{char} datatype could represent arbitrarily long numbers, than this
-function would be safe against buffer overruns, because it would only write data
-on allocated memory.
-However, if, for instance, we have that \texttt{w}$= 16$ and \texttt{h}$= 16$,
-then \texttt{s}$= 0$, as pointed before.
-In this case, a buffer of length zero would be allocated, and all the data
-stored in the file would overwrite memory in the stack of activation records.
-By carefully crafting an input file, an adversary can, in this way, overwrite
-the return address of \texttt{read\_matrix}, diverting program's execution to
-one of \texttt{libc}'s function, such as \texttt{telnet}, for instance.
+by Brumley {\em et al.}~\cite{Brumley07}, Dietz {\em et al.}~\cite{Dietz12},
+and Zhang {\em et al.}~\cite{Zhang10}.
+The function \texttt{read\_matrix} copies the linearized matrix \texttt{data}
+into \texttt{buf}, to later process it, in line 12.
+The arguments \texttt{w} and \texttt{h} determine the number of rows and
+columns in the matrix.
+If the \texttt{char} datatype could represent arbitrarily long numbers, than
+this function would be safe against buffer overruns, due to the test at
+line 3.
+However, if, for instance, we have that \texttt{w}$= 6$ and \texttt{h}$= 22$,
+then \texttt{buf\_size}$= -124$, due to the wrapping semantics of
+integer arithmetics in C.
+In this case, twelve words in the activation record of \texttt{read\_matrix}
+would be overwritten.
+By carefully crafting the contents of \texttt{data}, an adversary can, in this
+way, change the return address of \texttt{read\_matrix}, diverting program's
+execution to a system's function, such as \texttt{telnet}, for instance.
+
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=\columnwidth]{images/ex_buffer_overflow}
+\end{center}
+\caption{\label{fig:ex_buffer_overflow}
+An example of an exploitable integer overflow vulnerability.}
+\end{figure}
 
@@
 \varset{V} of variables. \\
-\textbf{Output:} a mapping I such that, for any variable
+\textbf{Output:} a mapping I such that, for any
 $V \in \varset{V}$, e(V) = I[V].
@@
 widening.
-In the second step, once we know in which directions the values stored in each
-variable might grow, we replace futures by actual bounds.
+In the second step, we replace future bounds by actual limits.
 Finally, a narrowing phase starting from conditional tests improves the
@@
 
+\acks
+We thank Douglas do Couto for helping us with the LLVM code base.
+We also thank the anonymous reviewers for their useful comments and
+suggestions.
+This project is supported by FAPEMIG, grant 16899*01.
+
 \bibliographystyle{plain}

### 2012-11-11 doc/Manuscript/CGO/paper.tex

@@
 LLVM test suite.
-This reduction has reduced 50\% of the runtime overhead of instrumentation.
+This reduction has decreased 50\% of the runtime overhead of instrumentation.
 \end{abstract}
@@
 by $p$ to $u'$.
-We then associate $u$ with the constraint $I[U'] \sqsubseteq I[U] \sqcap [-\infty, \mathit{MaxInt} - c]$.
+We then associate $u'$ with the constraint $I[U'] \sqsubseteq I[U] \sqcap [-\infty, \mathit{MaxInt} - c]$.
 
@@
 checks inserted in the transformed code.
-Wang {\em et a.}~\cite{Wang09} have implemented a tool, IntScope, that combines symbolic execution and taint analysis to detect integer overflow vulnerabilities.
+Wang {\em et al.}~\cite{Wang09} have implemented a tool, IntScope, that combines symbolic execution and taint analysis to detect integer overflow vulnerabilities.
 The authors have been able to use this tool to successfully identify many vulnerabilities in industrial quality software.

### 2012-11-13 doc/Manuscript/CGO/paper.tex
Added Vic/Raph's suggestions.

@@
 Fernando Magno Quint\~{a}o Pereira}
-           {Department of Computer Science -- The Federal University of Minas gerais (UFMG) -- Brazil}
+           {Department of Computer Science -- The Federal University of Minas Gerais (UFMG) -- Brazil}
            {\{raphael,victorsc,fernando\}@dcc.ufmg.br}
@@
 LLVM test suite.
-This reduction has decreased 50\% of the runtime overhead of instrumentation.
+This optimization has reduced the runtime overhead of instrumentation by 50\%.
 \end{abstract}
@@
 If a number $n$ is too large to fit into a primitive data type $T$, then $n$'s
-value wraps around, and we obtain $n \ \mbox{module} \ T_{max}$.
+value wraps around, and we obtain $n \ \mbox{modulo} \ T_{max}$.
 There are situations in which this semantics is acceptable~\cite{Dietz12}.
-For instance, programers might rely on this behavior to implement hash functions
-and random number generators.
+For instance, programmers might rely on this behavior to implement hash
+functions and random number generators.
 On the other hand, there exist also situations in which this behavior might
@@
 whenever integer overflows are detected.
-Furthermore, there exist recent work proposing to instrument binaries derived
-from C, C++ and Java programs to detect the occurrence of overflows
+Furthermore, there exist recent works proposing to instrument binaries
+derived from C, C++ and Java programs to detect the occurrence of overflows
 dynamically~\cite{Brumley07,Dietz12}.
@@
 columns in the matrix.
-If the \texttt{char} datatype could represent arbitrarily long numbers, than
+If the \texttt{char} datatype could represent arbitrarily long numbers, then
 this function would be safe against buffer overruns, due to the test at
@@
 interval domain.
-Armed with these concepts, we define the range analysis problem as follows:
-
-\begin{definition}
-\label{def:rcp}
-\textsc{Range Analysis Problem} \\
+Armed with these concepts, we define the range analysis problem as follows: \\
+
+\noindent
+\textsc{Definition: Range Analysis Problem} \\
 \textbf{Input:} a set \varset{C} of constraints ranging over a set
@@
 $V \in \varset{V}$, e(V) = I[V].
-\end{definition}
 
@@
 (a) Example program.
-(b) Control Flow Graph in SSA form.
+(b) Control Flow Graph in e-SSA form.
 (c) Constraints that we extract from the program.
@@
 \caption{\label{fig:ex_graph}
-The constraint graph that we build to the program in
+The constraint graph that we build for the program in
 Figure~\ref{fig:ex_eSSA_cgo}(b).}
@@
 Therefore, we determine the growth behavior of each constraint variable in
-a strong component in linear time on the number of constraints.
+a strong component in linear time on the number of constraints in that
+component.
 Figure~\ref{fig:ex_partition_grow_crop}(b) shows the abstract state of the
@@
 \textbf{Future resolution: }
-The next phase of the algorithm to determine intervals inside a strong
+the next phase of the algorithm to determine intervals inside a strong
 component consists in replacing futures by actual bounds, a task that we
@@
 phase.
-For instance, the growth analysis has found out that the value stored in variable
-\texttt{i}$_i$ can only increase.
+For instance, the growth analysis has found out that the value stored in
+variable \texttt{i}$_1$ can only increase.
 Given that this variable is assigned the initial value zero, we can replace
@@
 %
-Figure~\ref{fig:ex_partition_grow_crop}(d) show the result of our narrowing
+Figure~\ref{fig:ex_partition_grow_crop}(d) shows the result of our narrowing
 operator in our running example.
@@
 (i) $l_f$ or $l_t$ dominate any use of $v$;
-(ii) there exist a use of $v$ at the dominance frontier of $l_f$ or $l_t$.
+(ii) there exists a use of $v$ at the dominance frontier of $l_f$ or $l_t$.
 For the notions of dominance and dominance-frontier, see Aho
@@
 $v_{1v} = v_1 \sqcap [-\infty, \fun{ft}(v_2) - 1]$
-and $v_{2f} = v_2 \sqcap [\fun{ft}(v_1) + 1, +\infty]$.
+and $v_{2v} = v_2 \sqcap [\fun{ft}(v_1) + 1, +\infty]$.
 A variable $v$ can never be associated with a future bound to
@@
 
-\paragraph{Practical Considerations: }
+\paragraph{Practical Considerations. }
 %Overview about the instrumentation technique
@@
 For instance, the same variable $x$ would be suspicious if
-$I[x] = [0, 257]$, as $\ub{I[x]} > \ub{\mathtt{int8}}$.
+$I[x] = [0, 257]$, as $\ub{I[x]} > \ub{\mathtt{uint8}}$.
 \item \textbf{Uncertain}: we classify a variable as uncertain if at least one
@@
 actual overflows are more common among elements in the former category.
-\item \textbf{buggy}: a variable is buggy if the intersection between its
+\item \textbf{Buggy}: a variable is buggy if the intersection between its
 inferred range and the range of its declared type is empty.
@@
 Each core has a 4,096KB L1 cache.
-We use Linux Ubuntu 10.04.4.
+We have used Linux Ubuntu 10.04.4.
 Our implementation of range analysis has 3,958 lines of
@@
 (Lower) Comparison between static range analysis and dynamic profiler for
-lower bounds. The numbers above the benchmark names give the number of
-variables in each program.}
+lower bounds.}
 \end{figure}
@@
 We have not been able to run the binary that LLVM produces for SPEC's
-\texttt{gcc} in our environment, even without any of our passes, due to an
-incompatible \texttt{ctype.h} header.
+\texttt{gcc} in our environment, even without any of our transformations, due
+to an incompatible \texttt{ctype.h} header.
 In addition, we have not been able to collect the statistics about
@@
 was too large.
-We verified more than 3.000.000.000  overflows in this program.
+We verified more than 3,000,000,000  overflows in this program.
 Figure~\ref{fig:unprunedTable} shows the percentage of instructions that
@@
 456.hmmer      & 114,136 & 4,001 & 3.51\% & 0 \\ \hline
-\multicolumn{1}{|r|}{Total (Average)} & 275.070 & 6.968 & 3,96\% &  \\ \hline
+\multicolumn{1}{|r|}{Total (Average)} & 275,070 & 6,968 & 3.96\% &  \\ \hline
 \end{tabular*}
@@
 In other benchmarks, such as in \texttt{429.mcf}, we have been able to
-avoid only 1, out of 165 tests.
+avoid only 1 out of 165 tests.
 In general we fare better in programs that bound input sizes via conditional
 tests, as \texttt{lbm} does.
-Using u-SSA, instead of e-SSA adds a negligible improvement onto our results.
+Using u-SSA, instead of e-SSA, adds a negligible improvement onto our results.
 We speculate that this improvement is small because variables tend to be used
@@
 \hline
-Benchmark & \#II & \#E & \%(II, E) & \#U &
-\%(II, U) \\ \hline
-lbm & 1.142 & 4 & 99,65\% & 4 & 99,65\% \\ \hline
-milc & 1.573 & 1.070 & 31,98\% & 1.065 & 32,29\% \\ \hline
-namd & 2.965 & 2.900 & 2,19\% & 2.900 & 2,19\% \\ \hline
-dealII & 37.234 & 29.870 & 19,78\% & 28.779 & 22,71\% \\ \hline
-soplex & 3.268 & 2.927 & 10,43\% & 2.897 & 11,35\% \\ \hline
-h264ref & 13.548 & 11.342 & 16,59\% & 11.301 & 16,28\% \\ \hline
-astar & 903 & 808 & 10,52\% & 806 & 10,74\% \\ \hline
-sjeng & 2.483 & 2.354 & 5,20\% & 2.190 & 11,80\% \\ \hline
-mcf & 165 & 164 & 0,61\% & 164 & 0,61\% \\ \hline
-omnetpp & 1.533 & 1.313 & 14,35\% & 1.313 & 14,35\% \\ \hline
-gcc & 16.330 & 15.282 & 6,42\% & 15.110 & 7,47\% \\ \hline
-gobmk & 13.947 & 12.563 & 9,92\% & 12.478 & 10,53\% \\ \hline
-libquantum & 943 & 820 & 13,04\% & 817 & 13,36\% \\ \hline
-bzip2 & 2.350 & 1.966 & 16,34\% & 1.966 & 16,34\% \\ \hline
-hmmer & 3.963 & 3.346 & 15,57\% & 3.304 & 16,63\% \\ \hline
-Average & 6.823 & 5.779 & 18,17\% & 5.676 & 19,09\% \\ \hline
+Benchmark  &   \#II &  \#E & \%(II, E) &  \#U & \%(II, U) \\ \hline
+lbm        &  1,142 &      4 & 99.65\% &      4 & 99.65\% \\ \hline
+milc       &  1,573 &  1,070 & 31.98\% &  1,065 & 32.29\% \\ \hline
+namd       &  2,965 &  2,900 &  2.19\% &  2,900 &  2.19\% \\ \hline
+dealII     & 37,234 & 29,870 & 19.78\% & 28,779 & 22.71\% \\ \hline
+soplex     &  3,268 &  2,927 & 10.43\% &  2,897 & 11.35\% \\ \hline
+h264ref    & 13,548 & 11,342 & 16.59\% & 11,301 & 16.28\% \\ \hline
+astar      &    903 &    808 & 10.52\% &    806 & 10.74\% \\ \hline
+sjeng      &  2,483 &  2,354 &  5.20\% &  2,190 & 11.80\% \\ \hline
+mcf        &    165 &    164 &  0.61\% &    164 &  0.61\% \\ \hline
+omnetpp    &  1,533 &  1,313 & 14.35\% &  1,313 & 14.35\% \\ \hline
+gcc        & 16,330 & 15,282 &  6.42\% & 15,110 &  7.47\% \\ \hline
+gobmk      & 13,947 & 12,563 &  9.92\% & 12,478 & 10.53\% \\ \hline
+libquantum &    943 &    820 & 13.04\% &    817 & 13.36\% \\ \hline
+bzip2      &  2,350 &  1,966 & 16.34\% &  1,966 & 16.34\% \\ \hline
+hmmer      &  3,963 &  3,346 & 15.57\% &  3,304 & 16.63\% \\ \hline
+Total      & 102,347 & 86,688 &        & 85,135 &         \\ \hline
 \end{tabular*}
@@
 How the range analysis classified arithmetic instructions in the 
-u-SSA form program.
+u-SSA form programs.
 \#Sf: safe.
@@
 333.
-On the average, our range analysis avoid 24.93\% of the overflow checks.
-Considering the benchmarks in SPEC2006 only, this number is 20.57\%.
+On the average, our range analysis avoids 24.93\% of the overflow checks.
+Considering the benchmarks in SPEC 2006 only, this number is 20.57\%.
 
@@
 runtime of the SPEC benchmarks.
-We run each benchmark 20 times.
+We ran each benchmark 20 times.
 The largest slowdown that we have observed, 11.83\%, happened in
@@
 e-SSA form or u-SSA form.
+Surprisingly, some of the instrumented programs run faster than the original
+code.
+This behavior has also been observed by Dietz {\em et al.}~\cite{Dietz12}.
 
@@
 % TODO bibtex: cite Archerr: Runtime environment driven program safety.
-Contrary to these previous work, we instrument programs at LLVM's intermediate
+Contrary to these previous works, we instrument programs at LLVM's intermediate
 representation level, which is machine independent.
@@
 Brumley's, even without the support of the static range analysis.
-None of the above works use range analysis to eliminate overflow checks.
-This analysis, in our case, could eliminate approximately 45\% of the
-tests.
+Furthermore, our range analysis could eliminate approximately
+45\% of the tests that a naive implementation of Brumley's technique
+would insert; hence, halving down the runtime overhead of instrumentation.
 
@@
 The main drawback of emulation is performance: Chen {\em et al.} report a
-50X slowdown.
+50x slowdown.
 % TODO: cite "Brick: A  binary tool for run-time detecting and locating
@@
 representations, like we did.
-One of the program representation that we discuss in
+One of the program representations that we discuss in
 Section~\ref{sub:splitting} -- the {\em
@@
 the \emph{Static Single Use} form (SSU).
-There exists other variants of SSU~\cite{Lo98,Plevyak96,George03}.
+There exist other variants of SSU~\cite{Lo98,Plevyak96,George03}.
 The main difference to our representation is that our dataflow analysis
@@
 \acks
-We thank Douglas do Couto for helping us with the LLVM code base.
+We thank Douglas do Couto and Igor Costa for helping us with the LLVM code
+base.
 We also thank the anonymous reviewers for their useful comments and

### 2012-11-15 doc/Manuscript/CGO/paper.tex

@@
 lbm        &  1,142 &      4 & 99.65\% &      4 & 99.65\% \\ \hline
-milc       &  1,573 &  1,070 & 31.98\% &  1,065 & 32.29\% \\ \hline
-namd       &  2,965 &  2,900 &  2.19\% &  2,900 &  2.19\% \\ \hline
-dealII     & 37,234 & 29,870 & 19.78\% & 28,779 & 22.71\% \\ \hline
-soplex     &  3,268 &  2,927 & 10.43\% &  2,897 & 11.35\% \\ \hline
-h264ref    & 13,548 & 11,342 & 16.59\% & 11,301 & 16.28\% \\ \hline
-astar      &    903 &    808 & 10.52\% &    806 & 10.74\% \\ \hline
-sjeng      &  2,483 &  2,354 &  5.20\% &  2,190 & 11.80\% \\ \hline
+milc       &  1,602 &  1,070 & 33.21\% &  1,065 & 33.52\% \\ \hline
+namd       &  3,234 &  2,900 & 10.33\% &  2,900 & 10.33\% \\ \hline
+dealII     & 36,157 & 29,870 & 17.39\% & 28,779 & 20.41\% \\ \hline
+soplex     &  3,158 &  2,927 &  7.31\% &  2,897 &  8.26\% \\ \hline
+h264ref    & 13,846 & 11,342 & 18.38\% & 11,301 & 18.08\% \\ \hline
+astar      &    857 &    808 &  5.72\% &    806 &  5.95\% \\ \hline
+sjeng      &  2,504 &  2,354 &  5.99\% &  2,190 & 12.54\% \\ \hline
 mcf        &    165 &    164 &  0.61\% &    164 &  0.61\% \\ \hline
-omnetpp    &  1,533 &  1,313 & 14.35\% &  1,313 & 14.35\% \\ \hline
-gcc        & 16,330 & 15,282 &  6.42\% & 15,110 &  7.47\% \\ \hline
-gobmk      & 13,947 & 12,563 &  9.92\% & 12,478 & 10.53\% \\ \hline
-libquantum &    943 &    820 & 13.04\% &    817 & 13.36\% \\ \hline
-bzip2      &  2,350 &  1,966 & 16.34\% &  1,966 & 16.34\% \\ \hline
-hmmer      &  3,963 &  3,346 & 15.57\% &  3,304 & 16.63\% \\ \hline
-Total      & 102,347 & 86,688 &        & 85,135 &         \\ \hline
+omnetpp    &  1,972 &  1,313 & 33.42\% &  1,313 & 33.42\% \\ \hline
+gcc        & 18,669 & 15,282 & 18.14\% & 15,110 & 19.06\% \\ \hline
+gobmk      & 14,129 & 12,563 & 11.08\% & 12,478 & 11.69\% \\ \hline
+libquantum &    928 &    820 & 11.64\% &    817 & 11.96\% \\ \hline
+bzip2      &  2,158 &  1,966 &  8.90\% &  1,966 &  8.90\% \\ \hline
+hmmer      &  4,001 &  3,346 & 16.37\% &  3,304 & 17.42\% \\ \hline
+Total      & 104,522 & 86,688 &        & 85,135 &         \\ \hline
 \end{tabular*}

### 2012-12-14 doc/Manuscript/CGO/paper.tex
Added final version of the paper to SVN

@@
-\documentclass{sigplanconf}[10pt]
-
-% TODO: explain that we have not observed overflows in the safe instructions.
-% TODO: explain that u-SSA does not yield noticeable performance improvements.
+\documentclass[10pt]{cgo13}
 
@@
 
-\conferenceinfo{CGO '13}{February 23 - February 27. Shenzhen, China} 
+\conferenceinfo{CGO '13}{23-27 February 2013, Shenzhen China.}
 \copyrightyear{2013}
-\copyrightdata{[to be supplied]}
+\copyrightdata{978-1-4673-5525-4/13/\$31.00 \copyright2013 IEEE}
 
@@
 
-\title{A fast and Low-Overhead Technique to Secure Programs Against
+\title{A Fast and Low-Overhead Technique to Secure Programs Against
 Integer Overflows}
@@
 
-% TODO We need to explain better the wrapping-arithmetics, and we need to cite it
-% properly. What could be a good (official) citation?
-
 % Why there is still the need to devise efficient methods to eliminate them.
@@
 dynamically~\cite{Brumley07,Dietz12}.
-Thus, if an overflow happens, then the instrumented program can take some
-action, such as to log the event, or to terminate the program.
-However, this safety comes with a price: arithmetic operations need to be
+Thus, the instrumented program can take some action when an overflow happens,
+such as to log the event, or to terminate the program.
+However, this safety has a price: arithmetic operations need to be
 surveilled, and the runtime checks cost time.
-Zhang {\em et al.}~\cite{Zhang10} have proposed to eliminate some of this
+Zhang {\em et al.}~\cite{Zhang10} have eliminated some of this
 overhead via a tainted flow analysis.
@@
 As we show in Section~\ref{sec:range}, our algorithm has three core insights.
-Firstly, in Section~\ref{sub:micro} we propose a three-phase approach to extract
-information from comparisons between variables, e.g., $x < y$.
+Firstly, in Section~\ref{sub:prop} we show how we rely on strongly connected
+components to achieve speed and precision.
+It is well-known that this technique is effective in speeding up constraint
+resolution~\cite[Sec 6.3]{Nielson99}.
+Yet, we go beyond: given our three-phase approach, we improve precision
+by solving strong components in topological order.
+Secondly, in Section~\ref{sub:micro} we describe this three-phase approach to
+extract information from comparisons between variables, e.g., $x < y$.
 Previous algorithms either deal with these comparisons via
@@
 constants~\cite{Mahlke01,Patterson95,Stephenson00}.
-Secondly, in Section~\ref{sub:prop} we show how we rely on strongly connected
-components to achieve speed and precision.
-It is well-known that this technique is effective in speeding up constraint
-resolution~\cite[Sec 6.3]{Nielson99}.
-However, given our three-phase approach, we improve precision
-by solving strong components in topological order.
 Finally, in Section~\ref{sub:splitting} we propose a new program representation that
@@
 instrumentation library.
-This instrumentation framework, a contribution of this paper that we describe in
+This instrumentation framework, which we describe in
 Section~\ref{sec:dyn}, has been implemented in the LLVM compiler~\cite{Lattner04}.
@@
 
-\noindent
-\textbf{The Danger that Integer Overflows Represent: }
-Figure~\ref{fig:ex_buffer_overflow} illustrates an example of integer
-overflow vulnerability.
-This vulnerability allows an adversary to perform a buffer overrun attack
-on a program that is apparently guarded against this type of exploit.
-We will use the C datatype \texttt{char}, which represents
-signed 8-bit numbers, so that the interested reader can easily
-reconstruct this exploit.
-Examples of actual vulnerabilities, with larger primitive types, are given
-by Brumley {\em et al.}~\cite{Brumley07}, Dietz {\em et al.}~\cite{Dietz12},
-and Zhang {\em et al.}~\cite{Zhang10}.
-The function \texttt{read\_matrix} copies the linearized matrix \texttt{data}
-into \texttt{buf}, to later process it, in line 12.
-The arguments \texttt{w} and \texttt{h} determine the number of rows and
-columns in the matrix.
-If the \texttt{char} datatype could represent arbitrarily long numbers, then
-this function would be safe against buffer overruns, due to the test at
-line 3.
-However, if, for instance, we have that \texttt{w}$= 6$ and \texttt{h}$= 22$,
-then \texttt{buf\_size}$= -124$, due to the wrapping semantics of
-integer arithmetics in C.
-In this case, twelve words in the activation record of \texttt{read\_matrix}
-would be overwritten.
-By carefully crafting the contents of \texttt{data}, an adversary can, in this
-way, change the return address of \texttt{read\_matrix}, diverting program's
-execution to a system's function, such as \texttt{telnet}, for instance.
-
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\columnwidth]{images/ex_buffer_overflow}
-\end{center}
-\caption{\label{fig:ex_buffer_overflow}
-An example of an exploitable integer overflow vulnerability.}
-\end{figure}
-
 \section{Range Analysis}
 \label{sec:range}
-
-\subsection{The Interval Lattice}
-\label{sub:lattice}
 
@@
 from this program.
-There is a clear correspondence between instructions and constraints.
-Our analysis is sparse~\cite{Choi91}; thus, we associate one, and only one,
-constraint with each integer variable defined in the program.
+There is a correspondence between instructions and constraints.
+Our analysis is sparse~\cite{Choi91}; thus, it associates one, and only one,
+constraint with each integer variable.
 A possible solution to the range analysis problem, as obtained via the
@@
 Our range analysis algorithm works in a number of steps.
-Firstly, we convert the program to an intermediate representation that gives
+Firstly, we convert the program to a representation that gives
 us subsidies to perform a sparse analysis.
@@
 We let $\lb{[l, u]} = l$ and $\ub{[l, u]} = u$.
-We let $\bot$ denote non-itialized intervals, so that $[\bot, \bot] \sqcup
+We let $\bot$ denote non-initialized intervals, so that $[\bot, \bot] \sqcup
 [l, u] = [l, u]$.
@@
 $x = o_1 \ \qside \ n$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
-$x = \ \downarrow_n \ o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
-$x = \ (\mbox{type}_n) o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
+$x = \ \downarrow_n \ o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
 \end{tabular}
@@
 \caption{\label{fig:instrumentation}Overflow checks. We use $\downarrow_n$ for
-the operation that truncates to $n$ bits, and we use $(\mbox{type}_n)$ for a
-bit-cast to $n$ bits. Both operations have the same semantics in our setting.
+the operation that truncates to $n$ bits.
 The subscript $s$ indicates a signed instruction; the subscript $u$ indicate
@@
 \caption{\label{fig:num_instructions}
-Number of instructions used in each dynamic check.}
+Number of instructions used in each check.}
 \end{center}
@@
 Which actions are performed once the overflow is detected depends on the
-user, who has the option to overwrite the \texttt{handle\_ overflow}
+user, who has the option to overwrite the \texttt{handle\_overflow}
 function in Figure~\ref{fig:instrumented_cfg}.
@@
 The linear correlation, in this case, is even stronger than that found in
-Figure~\ref{fig:TimeCorr}: the coefficient of determination is 0.9947.
+Figure~\ref{fig:TimeCorr}: the coefficient of determination is 0.994.
 The figure only shows our 100 largest benchmarks.
@@
 graphs) and memory consumption (KB).
-Coefficient of determination = 0.9947.
+Coefficient of determination = 0.994.
 }
@@
 \begin{center}
-\begin{small}
+\begin{scriptsize}
 \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|r|}
@@
 \end{tabular*}
-\end{small}
+\end{scriptsize}
 \end{center}
@@
 The authors have been able to use this tool to successfully identify many vulnerabilities in industrial quality software.
-Our work, and Wang {\em et a.}'s work are essentially different: they use symbolic execution, whereas we rely on
+Our work, and Wang {\em et al.}'s work are essentially different: they use symbolic execution, whereas we rely on
 range analysis.
@@
 They prove the existence of bugs by generating test cases for them.
-
-\noindent
-\textbf{Range Analysis and sparse representations:}
-One of the goals of this paper is to describe a state-of-the-art range analysis
-algorithm.
-Many parts of this algorithm are well-known, others are novel.
-The widening and narrowing operators from Section~\ref{sub:micro} were
-introduced by Cousot and Cousot's seminal paper~\cite{Cousot77}.
-The widening operator that we use in our implementation is well-known among
-compiler writers~\cite[p.228]{Nielson99}.
-The interval lattice from Section~\ref{sub:lattice} has been used in a
-plethora of old and recent range analysis
-solvers~\cite{Gough94,Gawlitza09,Logozzo10,Mahlke01,Oh12,Patterson95,Stephenson00,Su05}.
-The use of futures to handle comparisons between variables is a contribution
-of this paper.
-It is well-known that the use of strongly connected components speeds up
-constraint resolution systems~\cite[Sec 6.3]{Nielson99}.
-However, publicly available implementations of range analysis such as those in
-FLEX~\footnote{The MIT's FLEX/Harpoon compiler
-provides an implementation of Stephenson's algorithm~\cite{Stephenson00}, and
-is available at \texttt{http://flex.cscott. net/Harpoon/}.},
-gcc~\footnote{Gcc's VRP pass (at \texttt{http://gcc.gnu.org/svn/gcc/trunk/gcc/
-tree-vrp.c}) implements a variant of Patterson's
-algorithm~\cite{Patterson95}.} or Mozilla's IonMonkey~\footnote{\texttt{hg.mozilla.org/projects/ionmonkey/file/629d1e6251b9/js/
-src/ion/RangeAnalysis.cpp}. This code is implemented after Gough and
-Klaeren~\cite{Gough94}} do not resort to this technique.
-Furthermore, they only deal with comparisons between variables and constants.
-The only work that we are aware of that has used range analysis to avoid
-overflow checks is Sol {\em et al}'s~\cite{Sol11}; however, their algorithm
-is much more limited than ours.
-Contrary to us, Sol {\em et al.} eliminate overflow checks in
-straight-line code traces produced by a just-in-time compiler, and they use the
-runtime values of variables to improve their precision.
-
-In this paper we provide a {\em sparse} implementation of range analysis.
-Sparsity, in our context, means that we associate points in the lattice of
-interest -- intervals in our case -- directly to variables.
-Dense analyses map such information to pairs formed by variables and program
-points.
-Some previous analyses have achieved sparsity by using specific program
-representations, like we did.
-One of the program representations that we discuss in
-Section~\ref{sub:splitting} -- the {\em
-Extended Static Single Assignment} (e-SSA) form -- has been used by Bodik
-{\em et al.}~\cite{Bodik00} and Gough {\em et al.}~\cite{Gough94}.
-The other representation, which we call u-SSA, is very similar to
-the \emph{Static Single Use} form (SSU).
-There exist other variants of SSU~\cite{Lo98,Plevyak96,George03}.
-The main difference to our representation is that our dataflow analysis
-propagates information forwardly, whereas the SSU form has been conceived for
-dataflow analyses that propagate information backwardly.
-Thus, it requires special nodes at branches to merge information from
-different program paths.
-The SSU form programs are a subset of the programs in the so called
-{\em Static Single Information} form, a program representation proposed by
-Scott Ananian~\cite{Ananian99}.
-SSI makes sparse the dataflow analyses that propagate information both
-forwardly and backwardly.
 
@@
 as previous work has demonstrated, we believe that our technique is still
-important, as some of these programs will be executed millions of times.
-Our entire code plus the data that we show in this paper is publicly available.
-A more extensive list of experiments is also available in our technical report.
+important, as some of these programs will be executed millions of times. \\
+\noindent
+\textbf{Software: } Our implementation is
+publicly available at \\
+\texttt{http://code.google.com/p/range-analysis/}.
 
@@
 base.
-We also thank the anonymous reviewers for their useful comments and
+We also thank the anonymous reviewers for comments and
 suggestions.
-This project is supported by FAPEMIG, grant 16899*01.
+This project is supported by FAPEMIG and Intel.
 

### 2013-04-23 doc/Manuscript/SBLP13/paper.tex
Adding SBLP13 paper

\documentclass{llncs}

\usepackage[applemac]{inputenc}
\usepackage[brazil]{babel}

\usepackage{proof}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{skak}

\sloppy

\begin{document}

\title{Prevenção Automática de Ataques de Não-Terminação}

\author{Raphael Ernani Rodrigues and Fernando Magno Quintão Pereira}

\institute{Departamento de Ciência da Computação -- UFMG -- Brasil
\email{\{raphael,fernando\}@dcc.ufmg.br}
}

\maketitle

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\begin{abstract}
Dizemos que um programa é vulnerável a um ataque de não terminação quando
um adversário pode lhe fornecer valores de entrada que façam algum de seus laços iterar para sempre.
A prevenção de ataques desse tipo é difícil, pois eles não se originam de bugs que infringem a semântica da linguagem em que o programa foi feito.
Ao contrário, essas vulnerabilidades têm origem na aritmética modular inteira de linguagens como C, C++ e Java, a qual possui semântica bem definida.
Neste artigo nós apresentamos uma ferramenta que detecta tais problemas, e que saneia código vulnerável.
A detecção da vulnerabilidade dá-se via uma análise de fluxo de informação;
a sua cura decorre de guardas que nosso compilador insere no código vulnerável.
Nós implementamos esse arcabouço em LLVM, um compilador de qualidade industrial, e testa-mo-no em um conjunto de programas que compraz mais de
2.5 milhões de linhas de código escrito em C.
Descobrimos que, em média, caminhos em que informação perigosa trafega são pequenos, sendo compostos por não mais que 10 instruções assembly.
A instrumentação que inserimos para prevenir ataques de não terminação aumenta o tamanho do programa saneado em cerca de 5\% em média, e torna-os menos que 2\% mais lentos.
\end{abstract}

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\begin{abstract}
We say that a program is vulnerable to a non-termination attack if (i) it
contains a loop that is bounded by values coming from public inputs, and (ii)
an adversary can manipulate these values to force this loop to iterate forever.
Preventing this kind of attack is difficult because they do not originate
from bugs that break the semantics of the programming
language, such as buffer overflows.
Instead, they usually are made possible by the wrapping integer arithmetics
used by C, C++ and Java-like languages, which have well-defined semantics.
In this paper we present the diagnosis and the cure for this type of attack.
Firstly, we describe a tainted-flow analysis that detects non-termination
vulnerabilities.
Secondly, we provide a compiler transformation that inserts arithmetic guards on
loop conditions that may not terminate due to integer overflows.
We have implemented our framework in the LLVM compiler, and have tested it on
a benchmark suite containing over 2.5 million lines of C code.
We have found out that the typical path from inputs to loop conditions is,
on the average, XX instructions long.
Our instrumentation that prevents this kind of attacks adds less than XX\%
extra code on the secured program.
The final, protected code, is, on the average, less than XX\% slower than the
original, unprotected program.
\end{abstract}

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\section{Introdução}
\label{sec:int}

Um ataque de Negação de Serviços ({\em Denial-of-Service -- DoS}) consiste
em sobrecarregar um servidor com uma quantidade de falsas requisições grande o
suficiente para lhe comprometer a capacidade de atender contatos
legítimos.
Existem hoje diversas maneiras diferentes de detectar e reduzir a efetividade
desse tipo de ataque~\cite{Moore06}.
Neste artigo, contudo, descreveremos uma forma de negação de serviço que é de
difícil detecção: {\em os ataques de não-terminação}.
Um adversário realiza um ataque desse tipo fornecendo ao programa alvo entradas
cuidadosamente produzidas para forçar iterações eternas sobre um laço
vulnerável.
Um ataque de não terminação demanda conhecimento do código fonte do sistema
a ser abordado.
Não obstante tal limitação, esse tipo de ataque pode ser muito efetivo,
pois bastam algumas requisições para comprometer o sistema alvo.
Uma vez que essa quantidade de incursões é pequena, os métodos tradicionais
de detecção de negação de serviço não podem ser usados para reconhecer ataques
de não-terminação.
Além disso, dada a vasta quantidade de código aberto usado nos mais diversos
ramos da indústria de {\em software}, usuários maliciosos têm a sua
disposição um vasto campo de ação.

A detecção de código vulnerável a ataques de não-terminação é difícil.
Tal dificuldade existe, sobretudo, porque esse tipo de ataque não decorre de
deficiências de tipagem fraca, normalmente presentes em programas escritos em
C ou C++.
Programas escritos em linguagens fortemente tipadas, como Java, por exemplo,
também apresentam a principal fonte de vulnerabilidades a ataques de
não terminação: a {\em aritmética modular inteira}.
Em outras palavras, uma operação como \texttt{a + 1}, em Java, C ou C++, pode
resultar em um valor menor que aquele inicialmente encontrado na variável
\texttt{a}.
Esse fenômeno ocorrerá quando a variável \texttt{a} guardar o maior inteiro
presente em cada uma dessas linguagens.
Nesse caso, ao fim da operação, \texttt{a + 1} representará o menor inteiro
possível em complemento de dois.
Em outras palavras, um laço como \texttt{for (i = 0; i <= N; i++)} nunca
terminará se \texttt{N} for \texttt{MAX\_INT}, o maior inteiro da linguagem.

Este artigo traz duas contribuições relacionadas a ataques de não-terminação.
Em primeiro lugar, ele descreve uma técnica que descobre vulnerabilidades
relacionadas a esse tipo de ataque.
Em segundo lugar, o artigo mostra como código pode ser protegido contra tais
ataques.
A nossa técnica de detecção de vulnerabilidades é baseada em análise de
fluxo contaminado.
Tal análise é parte do arcabouço teórico de rastreamento de informação
inicialmente proposto por Denning e Denning nos anos setenta~\cite{Denning77}.
Um ataque de fluxo contaminado pode ser efetivo somente em programas cujas
operações críticas dependam de dados de entrada.
Em nosso contexto, uma operação crítica é o teste de controle de laço.
Em conjunto com o algoritmo de detecção de vulnerabilidades, nós propomos
também uma técnica para sanear programas contra ataques de não-terminação.
Nossa estratégia consiste na inserção de verificações sobre as
operações aritméticas realizadas pelo programa alvo.
Essas verificações ocorrem durante a execução do programa, e invocam código de
segurança sempre que estouros de precisão em variáveis inteira são percebidos.
Nós instrumentamos somente código que controla o número de iterações em laços.
Consequentemente, o arcabouço que propomos incorre em uma perda de
desempenho muito pequena, e, em nossa opinião, completamente justificável em
decurso do benefício que assegura.

Nós implementamos todas as idéias que discutimos neste artigo em LLVM,
um compilador de qualidade industrial~\cite{Lattner04}.
Na seção~\ref{sec:exp} descreveremos uma série de experimentos que
validam nossa análise.
Ao analisar os programas presentes na coleção SPEC CPU 2006, fomos capazes
de descobrir XX laços que são influenciados por dados provenientes de
entradas públicas, isto é, que podem ser manipuladas por um adversário.
Dentre esses laços, pelo menos XX estão sujeitos à não terminação.
Para obter esse número, utilizamos um padrão muito simples: procuramos por
laços cujo teste de parada é do tipo \texttt{i <= N}, sendo \texttt{N}
dependente da entrada.
Uma vez que nos atemos a esse tipo de condição de parada, especulamos que
a quantidade de laços vulneráveis presente nos programas de SPEC CPU 2006
seja bem maior que o valor que apuramos.
A nossa instrumentação -- usada para impedir os ataques de não-terminação --
mostra-se extremamente eficiente.
Os testes que inserimos antes de cada operação aritmética que pode levar a
um ataque de não terminação custa-nos uma perda de desempenho de menos que
XX\%.

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\section{Ataques de Não-Terminação}
\label{sec:bkg}

De acordo com Appel e Palsberg~\cite[p.376]{Appel02}, um laço natural é um
conjunto de nodos $S$ do grafo de fluxo de controle de um programa (CFG),
includindo um nodo cabeçalho $H$, com as seguintes três propriedades:
%
\begin{itemize}
\item a partir de qualquer nodo em $S$ existe um caminho que chega a $H$;
\item existe um caminho de $H$ até qualquer nodo que faz parte de $S$;
\item não existe caminho que não contenha $H$ de qualquer nodo fora de $S$ para
qualquer nodo em $S$.
\end{itemize}
%
A {\em condição de parada} de um laço é uma expressão booleana
$E = f(e_1, e_2, \ldots, e_n)$, sendo cada $e_j, 1 \leq j \leq n$ um valor que
contribui para a computação de $E$.
Seja $P$ um programa que possui um laço $L$ limitado por uma condição
de parada $E = f(e_1, e_2, \ldots, e_n)$.
Dizemos que $P$ é vulnerável a um ataque de não
terminação quando as duas condições abaixo são verdadeiras sobre ele:
\begin{enumerate}
\item existe um subconjunto $E' \subseteq \{e_1, e_2, \ldots, e_n\}$ de
valores que dependem de um conjunto $I = \{i_1, i_2, \ldots, i_m\}$ de
dados lidos a partir da entrada do programa.
\item existe uma atribuição de valores $i_1 \mapsto v_1,
i_2 \mapsto v_2, \ldots, i_m \mapsto v_m$ que, ao influenciar $E'$, faz com
que o laço $L$ não termine.
\end{enumerate}
Note que a nossa definição de vulnerabilidade de não-terminação requer a
noção de dependência de dados.
Se um programa $P$ possui uma instrução que usa a variável $u$ e define a
variável $v$, então $v$ {\em depende} de $u$.
Dependências de dados são transitivas, e embora possam ser cíclicas, não
são necessariamente comutativas.

Ilustraremos ataques de não terminação via o programa mostrado na
Figura~\ref{fig:fact}(a).
Esse programa calcula o fatorial de um número inteiro na linguagem C.
O padrão que rege essa linguagem de programação não determina o tamanho do
tipo \texttt{int}.
Essa informação depende da implementação do compilador usado.
Entretanto, é usual que inteiros sejam representados como números de 32
bits em complemento de dois. 
Nesse caso, o maior inteiro representável é
$\mathit{MAX\_INT} = 2^{31} - 1 = 2,147,483,647$.
Se o parâmetro \texttt{n} for igual a $\mathit{MAX\_INT}$, então a condição
da linha 4 sempre será verdadeira, e o laço nunca terminará.
A não-terminação ocorre porque quando \texttt{i} finalmente chega a
$\mathit{MAX\_INT}$, a soma $i + 1$ nos retorna o menor inteiro possível, isto
é, $- 2^{31}$.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{images/fact}
\caption{(a) Uma função em C, que calcula o fatorial de um número inteiro.
(b) O CFG da função \texttt{fact}.
(c) Exemplo de laço cuja condição de parada depende de valores de entrada
mas que sempre termina.}
\label{fig:fact}
\end{center}
\end{figure}

O programa da figura~\ref{fig:fact}(a) é vulnerável a ataques de
não-terminação.
Para explicitar tal fato, a figura~\ref{fig:fact}(b) mostra o grafo de
fluxo de controle do programa.
Os blocos básicos que começam nos rótulos 3 e 7 formam um laço natural,
segundo a definição de Appel e Palsberg.
Esse laço é controlado pela condição de parada $i_1 \leq n$.
A variável $n$, o limite do laço, é dependente da entrada.
Existe um valor de $n$, a saber $\mathit{MAX\_INT}$, que força o laço a
não-terminar.

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\section{Detecção e Prevenção de Não-Terminações}
\label{sec:sol}

Nesta seção descreveremos nossa técnica para detectar vulnerabilidades de
não-terminação.
Esse algoritmo de detecção fornece os subsídios necessários a uma segunda
técnica que introduzimos neste artigo: o {\em saneamento de laços}.

\subsection{Detecção Automática de Não-Terminação}
\label{sub:det}

Dizemos que um laço é {\em potencialmente vulnerável} a ataque de não
terminação quando as condições que o controlam usam valores que dependem de
dados de entrada do programa.
Note que uma vulnerabilidade potencial não implica em uma vulnerabilidade
real.
A título de exemplo, o programa da Figura~\ref{fig:fact}(c), uma sutil
alteração da função \texttt{fact} inicialmente vista na
figura~\ref{fig:fact}(a), termina para qualquer entrada.
A fim de detectar vulnerabilidades potenciais, nós utilizamos o
{\em grafo de dependências de dados}.
Esse grafo é definido da seguinte forma:
para cada variável $v$ no programa, nós criamos um nodo $n_v$, e para cada
instrução $i$ no programa, nós criamos um nodo $n_i$.
Para cada instrução $i: v = f(\ldots, u, \ldots)$ que define uma variável $v$ e
usa uma variável $u$ n\'{o}s criamos duas areastas: $n_u \leftarrow n_i$ e
$n_i \leftarrow n_v$.
O grafo de dependências que extraímos a partir d CFG visto na
figura~\ref{fig:fact}(b) é mostrado na figura~\ref{fig:depGraph}.

\begin{figure}[t!]
\begin{center}
%\includegraphics[width=1\textwidth]{images/depGraph}
\includegraphics{images/depGraph}
\caption{(a) Grafo de dependências da função \texttt{fact}, construído a
partir do CFG visto na figura~\ref{fig:fact}(b).}
\label{fig:depGraph}
\end{center}
\end{figure}

Um caminho entre uma entrada do programa, e o predicado que controla o laço
indica uma vulnerabilidade potencial.
O grafo de dependências de nosso exemplo possui uma tal vulnerabilidade:
existe um caminho que une o nodo correspondente ao parâmetro \texttt{n}, uma
entrada, ao nodo que corresponde a \texttt{p}, o predicado de controle do
laço.
Esse tipo de caminho, uma vez contruído o grafo, pode ser encontrado em
tempo linear no número de arestas do grafo - normalmente proporcional ao
número de nodos - via a simples busca em profundidade ou largura.

\subsection{Saneamento de Laços}
\label{sub:san}

Uma vez encontrado um caminho vulnerável, passamos à fase de saneamento de
laços.
Um laço pode ser saneado via a inserção de testes que detectam e tratam a
ocorrência de estouros de precisão inteira.
Nós inserimos tais testes sobre as operações aritméticas {\em internas} que
controlam a condição de parada do laço.
Se $S$ é o conjunto de blocos básicos que fazem parte do laço, uma operação é
dita interna quando ela ocorre nalgum bloco em $S$.
Continuando com o nosso exemplo, o laço alvo possui dois blocos básicos:
o primeiro começa no rótulo três, e o segundo começa no rótulo sete.
O laço possui duas operações aritméticas internas, todas ocorrendo no segundo
bloco básico.
Dentre essas operações, aquela no rótulo sete é inofensiva: ela define a
variável $r_2$, que não participa da condição de parada do laço.
Por outro lado, a operação no rótulo oito, que define a variável $i_2$, é
usada no cálculo daquela condição, e precisa ser instrumentada.

Novamente, o grafo de dependências ajuda-nos a encontrar quais operações
precisam ser instrumentadas para sanear um laço controlado por um
predicado $p$.
Nesse caso, usamos o seguinte critério para determinar se uma operação
$i : v = f(v_1, \ldots, v_n)$ precisa ser instrumentada:
%
\begin{itemize}
\item existe um caminho no nodo $n_i$ até o nodo $n_p$.
\item O nodo $n_i$ encontra-se em um ciclo.
\end{itemize}
%
A título de exemplo, a operação de incremento $++$ no grafo de dependências
visto na figura~\ref{fig:depGraph} precisa ser instrumentada.
Em primeiro lugar, porque essa operação encontra-se em um ciclo.
Em segundo lugar, porque existe um caminho do nodo $n_{++}$ até o nodo
$n_p$.

\paragraph{Instrumentação de Saneamento. }
A fim de garantir que estouros de precisão inteiros não venham a causar
a não-terminação de laços, nós inserimos testes no código binário do
programa alvo.
O código que constitui cada um desses testes é formado por uma guarda, mas
um tratador de eventos.
Nossas guardas usam as condições mostradas na figura~\ref{fig:ovfCheck} para
verificar a ocorrência de estouros de precisão.
Atualmente instrumentamos quatro tipos diferentes de instrução:
adição, subtração, multiplicação e arredamentos para a esquerda.
As operações de adição, subtração e multiplicação podem ser com ou sem sinal
aritmético.

\begin{figure}[t!]
\begin{center}
\begin{tabular}{ll}
Instrução & Verificação \\ \\
$x = o_1 \ +_s \ o_2$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
                      & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
$x = o_1 \ +_u \ o_2$ & $x < o_1 \vee x < o_2$ \\ \\
$x = o_1 \ -_s \ o_2$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \ \ \vee$ \\
                      & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
$x = o_1 \ -_u \ o_2$ & $o_1 < o_2$ \\ \\
$x = o_1 \ \times_{u/s} \ o_2$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\ \\
$x = o_1 \ \qside \ n$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
$x = \ \downarrow_n \ o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
\end{tabular}
\end{center}
\caption{\label{fig:ovfCheck}Overflow checks. Usamos $\downarrow_n$ para
descrever a operação que trunca em $n$ bits.
O subscrito $s$ indica uma operação aritmética com sinal, e o subscrito $u$
indica uma operação sem sinal.}
\end{figure}

Os testes são implementados como sequências de operações binárias, executados
logo após a instrução guardada.
Para ilustrar esse ponto, mostramos, na figura~\ref{fig:instrumented_cfg},
o código necessário para instrumentar uma soma com sinal de duas variáveis.
Essa figura mostra código no formato intermediário usado por LLVM, o compilador
que utilizamos para implementar as idéias descritas neste artigo.
Omitimos, nesse exemplo, o código do tratador de evento de estouro, pois ele
simplesmente chamada uma rotina implementada em uma biblioteca
dinamicamente compartilhada.
Conforme podemos observar pela figura, uma guarda aumenta o código instrumentado
substancialmente.
Nesse exemplo em particular a verificação requer a inserção de 14 novas
instruções no programa guardado.
Embora tal crescimento a princípio possa parecer proibitivamente grande,
os experimentos que mostraremos na seção~\ref{sec:exp} indicam que somente
uma parcela muito pequena das instruções do programa alvo precisam ser
guardadas.
Consequentemente, o custo, em termos de crescimento de código e perda de
desempenho, é negligível.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{images/instrumented_cfg}
\caption{(a) }
\label{fig:instrumented_cfg}
\end{center}
\end{figure}

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\section{Resultados Experimentais}
\label{sec:exp}

\begin{figure}[t!]
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|} \hline
Benchmark & Num. Laços & Alcançáveis & Vulneráveis &
Caminhos \\ \hline
470.lbm        & & & &  \\ \hline
433.milc       & & & &  \\ \hline
444.namd       & & & &  \\ \hline
447.dealII     & & & &  \\ \hline
450.soplex     & & & &  \\ \hline
464.h264ref    & & & &  \\ \hline
473.astar      & & & &  \\ \hline
458.sjeng      & & & &  \\ \hline
429.mcf        & & & &  \\ \hline
471.omnetpp    & & & &  \\ \hline
403.gcc        & & & &  \\ \hline
445.gobmk      & & & &  \\ \hline
462.libquantum & & & &  \\ \hline
401.bzip2      & & & &  \\ \hline
456.hmmer      & & & &  \\ \hline
\end{tabular*}
\end{center}
\caption{Informações estáticas inferidas pela análise de não-terminação.
Num. laços: número de laços no programa.
Alcançáveis: quantidade de laços que são dependentes de dados produzidos a
partir de canais de entrada.
Vulneráveis: número de laços que preenchem nossos requisitos de
vulnerabilidade.
Caminhos: tamanho médio do menor caminho de dependência de dados da entrada
até a operação de controle do laço.}
\label{fig:staticLoopData}
\end{figure}

\begin{figure}[t!]
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|} \hline
Benchmark & Num. Instruções & Num. Arit. & Instrumentação &
Crescimento \\ \hline
470.lbm        & & & &  \\ \hline
433.milc       & & & &  \\ \hline
444.namd       & & & &  \\ \hline
447.dealII     & & & &  \\ \hline
450.soplex     & & & &  \\ \hline
464.h264ref    & & & &  \\ \hline
473.astar      & & & &  \\ \hline
458.sjeng      & & & &  \\ \hline
429.mcf        & & & &  \\ \hline
471.omnetpp    & & & &  \\ \hline
403.gcc        & & & &  \\ \hline
445.gobmk      & & & &  \\ \hline
462.libquantum & & & &  \\ \hline
401.bzip2      & & & &  \\ \hline
456.hmmer      & & & &  \\ \hline
\end{tabular*}
\end{center}
\caption{Impacto da instrumentação no código saneado.
Num. Instruções: número de instruções no benchmark.
Num. Arit: número de instruções que podem causar estouro de precisão inteira.
Instrumentação: quantidade de testes inseridos para sanear o programa.
Crescimento: razão entre o tamanho do programa instrumentado e o tamanho do
programa original.}
\label{fig:instImpact}
\end{figure}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/Runtime}
\caption{Variação no tempo de execução dos benchmarks devido à instrumentação
de saneamento.}
\label{fig:Runtime}
\end{center}
\end{figure}

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\section{Trabalhos Relacionados}
\label{sec:rel}

Este trabalho aborda temas relacionados a diferentes áreas da análise estática
e dinâmica de programas, a saber: teoria de fluxo de informação, detecção de
estouros de precisão inteira e análise de não-terminação.
Além disso, este trabalho utiliza o conceito de {\em grafos de dependências},
inicialmente proposto por Ferrante {\em et al.}~\cite{Ferrante87}.
Em nosso caso, o grafo de dependência dá-nos a estrutura de dados básica sobre
a qual caminhos que levam à não-terminação podem ser encontrados.
Esses grafos, contudo, historicamente vêm se prestando a muitos outros
propósitos, como escalonamento de instruções, detecção de condições de corrida e
propagação de constantes, por exemplo.

Neste artigo usamos o grafo de dependências para rastrear o fluxo de
informação contaminada.
O rastreamento de fluxo de informação é uma grande sub-área dentro do campo de
análise estática de programas~\cite{Denning77}.
Existem duas formas principais de rastrear a informação.
Pode-se traçar o fluxo de dados a partir de operações sigilosas até entradas
que um adversário pode ler.
Esse modo de rastreamento é popularmente conhecido como detecção de vazamento
de segredos~\cite{Hammer06}.
E, no sentido inverso, pode-se traçar o fluxo de informação de entradas que
um adversário pode manipular até operações críticas dentro do
programa~\cite{Tripp09}.
Essa categoria inclui nosso trabalho, além de diversos outros tipos de
vulnerabilidades, tais como {\em Injeção de Código SQL}~\cite{Wassermann07},
{\em Injeção de Scripts}~\cite{Rimsa10} e {\em Ataques de Estouro de
Buffer}~\cite{Levy96}.

Nós instrumentamos código considerado vulnerável para detectar estouros de
precisão que podem levar à não-terminação.
Esses mesmos testes já foram usados com vários outros objetivos em trabalhos
anteriores.
O mais importante trabalho nessa área deve-se, provavelmente, a Brumley
{\em et al.}~\cite{Brumley07}.
O grupo de David Brumley desenvolveu uma ferramenta, RICH, que instrumenta
cada operação aritmética passível de estouro de precisão inteira em um programa
C.
A principal conclusão daquele trabalho foi que esse tipo de instrumentação
não compromete sobremaneira o desempenho do programa modificado.
RICH, por exemplo, aumenta o tempo de execução dos programas instrumentados
em menos que 6\% em média.
Outro trabalho importante nesse campo foi publicado por Dietz
{\em et al.}~\cite{Dietz12}.
Esse grupo implementou IOC, uma ferramenta que, assim como RICH, detecta a
ocorrência de estouros de precisão em operações aritméticas inteiras.
Porém, ao contrário de Brumley {\em et al.}, Dietz {\em et al.} usaram sua
ferramenta para desenvolver um amplo estudo sobre a ocorrência de estouros
em programas reais.
Nosso trabalho difere desses outros em propósito: estamos interessados em
prevenir ataques de não terminação; e em método: nós instrumentamos somente uma
pequena parte dos programas alvo.

Finalmente, nosso trabalho relaciona-se com outros que também procuram
detectar, estaticamente, a não-terminação de programas.
A maior parte desses trabalhos utilizam análise simbólica de código para criar
expressões que levem um laço à não terminação.
Exemplos desse tipo de pesquisa incluem os trabalhos de Burnim
{\em et al.}~\cite{Burnim09}, Brockschmidt {\em et al}~\cite{Brockschmidt11} e
Veroyen {\em et al.}~\cite{Velroyen08}.
Esses trabalhos não levam em consideração possibilidade de não-terminação
devido à estouros de precisão, tampouco procuram detectar possíveis
vulnerabilidades baseadas em negação de serviço.
Existem, contudo, trabalhos na linha de detecção de não-terminação que são
bastante próximos do nosso.

Um trabalho que prova não-terminação, mesmo em face de estouros de precisão
deve-se à Gupta {\em et al.}~\cite{Gupta08}.
Gupta, assim como os trabalhos anteriormente relacionados, utiliza análise
simbólica para provar a não-terminação de programas.
A ferramenta implementada por Gupta {\em et al.}, denominada TNT, é capaz de
encontrar uma expressão algébrica que leva um laço de programa a iterar
para sempre.
Porém, TNT não aponta quais laços podem ser controlados a partir da entrada
do programa.
Por outro lado, a ferramenta  SAFERPHP, proposta por Son
{\em et al.}~\cite{Son11} possui exatamente esse objetivo.
SAFERPHP analisa o código de programas escritos em PHP, procurando por laços
que um adversário pode controlar, com o propósito, justamente, de evitar
ataques de não-terminação.
A principal diferença entre nosso trabalho, e aquele de Son {\em et al.}, é
que, enquanto nossa ferramenta busca detectar a não-terminação devido à
estouros de precisão inteira, SAFERPHP considera a aritmética de
precisão infinita.
Além disso, tanto SAFERPHP quanto TNT utilizam execução simbólica sobre caminhos
possíveis no programa alvo.
Essa abordagem, em nossa opinião, não é prática.
Testemunho disso é o fato de tais ferramentas terem sido usadas, até a presente
data, somente para analisar programas muito pequenos.

%mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
\section{Conclusão}
\label{sec:con}

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2013-04-23 doc/Manuscript/SBLP13/resources/fact.c

@@ -1,0 +1,18 @@
+#include <stdio.h>
+#include <stdlib.h>
+
+int fact(int n) {
+  int r = 1;
+  int i = 2;
+  while (i <= n) {
+    r *= i;
+    i++;
+  }
+  return r;
+}
+
+int main(int argc, char** argv) {
+  int n = atoi(argv[1]);
+  printf("%d\n", n);
+  printf("The factorial of %d is %d\n", n, fact(n));
+}

### 2013-04-24 doc/Manuscript/SBLP13/paper.tex
Including numbers in table 1 and 2

@@
 
-\usepackage[applemac]{inputenc}
+\usepackage[utf8]{inputenc}
 \usepackage[brazil]{babel}
@@
 Dizemos que um programa é vulnerável a um ataque de não terminação quando
-um adversário pode lhe fornecer valores de entrada que façam algum de seus laços iterar para sempre.
-A prevenção de ataques desse tipo é difícil, pois eles não se originam de bugs que infringem a semântica da linguagem em que o programa foi feito.
-Ao contrário, essas vulnerabilidades têm origem na aritmética modular inteira de linguagens como C, C++ e Java, a qual possui semântica bem definida.
-Neste artigo nós apresentamos uma ferramenta que detecta tais problemas, e que saneia código vulnerável.
+um adversário pode lhe fornecer valores de entrada que façam algum de seus laços 
+iterar para sempre.
+A prevenção de ataques desse tipo é difícil, pois eles não se originam de bugs 
+que infringem a semântica da linguagem em que o programa foi feito.
+Ao contrário, essas vulnerabilidades têm origem na aritmética modular inteira de 
+linguagens como C, C++ e Java, a qual possui semântica bem definida.
+Neste artigo nós apresentamos uma ferramenta que detecta tais problemas, 
+e que saneia código vulnerável.
 A detecção da vulnerabilidade dá-se via uma análise de fluxo de informação;
 a sua cura decorre de guardas que nosso compilador insere no código vulnerável.
-Nós implementamos esse arcabouço em LLVM, um compilador de qualidade industrial, e testa-mo-no em um conjunto de programas que compraz mais de
+Nós implementamos esse arcabouço em LLVM, um compilador de qualidade industrial, 
+e testa-mo-no em um conjunto de programas que compraz mais de
 2.5 milhões de linhas de código escrito em C.
-Descobrimos que, em média, caminhos em que informação perigosa trafega são pequenos, sendo compostos por não mais que 10 instruções assembly.
-A instrumentação que inserimos para prevenir ataques de não terminação aumenta o tamanho do programa saneado em cerca de 5\% em média, e torna-os menos que 2\% mais lentos.
+Descobrimos que, em média, caminhos em que informação perigosa trafega são 
+pequenos, sendo compostos por não mais que 10 instruções assembly.
+A instrumentação que inserimos para prevenir ataques de não terminação aumenta o 
+tamanho do programa saneado em cerca de 5\% em média, e torna-os menos que 2\% 
+mais lentos.
 \end{abstract}
@@
 We have found out that the typical path from inputs to loop conditions is,
-on the average, XX instructions long.
-Our instrumentation that prevents this kind of attacks adds less than XX\%
+on the average, less than 10 instructions long.
+Our instrumentation that prevents this kind of attacks adds less than 5\%
 extra code on the secured program.
-The final, protected code, is, on the average, less than XX\% slower than the
+The final, protected code, is, on the average, less than 2\% slower than the
 original, unprotected program.
@@
 Caminhos \\ \hline
-470.lbm        & & & &  \\ \hline
-433.milc       & & & &  \\ \hline
-444.namd       & & & &  \\ \hline
-447.dealII     & & & &  \\ \hline
-450.soplex     & & & &  \\ \hline
-464.h264ref    & & & &  \\ \hline
-473.astar      & & & &  \\ \hline
-458.sjeng      & & & &  \\ \hline
-429.mcf        & & & &  \\ \hline
-471.omnetpp    & & & &  \\ \hline
-403.gcc        & & & &  \\ \hline
-445.gobmk      & & & &  \\ \hline
-462.libquantum & & & &  \\ \hline
-401.bzip2      & & & &  \\ \hline
-456.hmmer      & & & &  \\ \hline
+433.milc	&	326	&	8	&	4	&	3   \\ \hline
+444.namd	&	508	&	7	&	1	&	2  \\ \hline
+447.dealII	&	4657	&	540	&	225	&	3  \\ \hline
+450.soplex	&	547	&	21	&	12	&	3  \\ \hline
+470.lbm		&	17	&	0	&	0	&	- \\ \hline
+401.bzip2	&	205	&	1	&	1	&	3  \\ \hline
+403.gcc		&	3853	&	109	&	59	&	4  \\ \hline
+429.mcf		&	38	&	2	&	0	&	2  \\ \hline
+445.gobmk	&	1073	&	37	&	19	&	4  \\ \hline
+456.hmmer	&	687	&	133	&	48	&	2  \\ \hline
+458.sjeng	&	235	&	9	&	3	&	4  \\ \hline
+462.libquantum	&	67	&	2	&	1	&	4  \\ \hline
+464.h264ref	&	1586	&	27	&	22	&	15  \\ \hline
+473.astar	&	86	&	7	&	7	&	18  \\ \hline
 \end{tabular*}
@@
 Crescimento \\ \hline
-470.lbm        & & & &  \\ \hline
-433.milc       & & & &  \\ \hline
-444.namd       & & & &  \\ \hline
-447.dealII     & & & &  \\ \hline
-450.soplex     & & & &  \\ \hline
-464.h264ref    & & & &  \\ \hline
-473.astar      & & & &  \\ \hline
-458.sjeng      & & & &  \\ \hline
-429.mcf        & & & &  \\ \hline
-471.omnetpp    & & & &  \\ \hline
-403.gcc        & & & &  \\ \hline
-445.gobmk      & & & &  \\ \hline
-462.libquantum & & & &  \\ \hline
-401.bzip2      & & & &  \\ \hline
-456.hmmer      & & & &  \\ \hline
+433.milc	&	22615	&	1165	&	7	&	0,50\%   \\ \hline
+444.namd	&	65045	&	3136	&	1	&	0,03\%   \\ \hline
+447.dealII	&	415756	&	14948	&	279	&	0,58\%   \\ \hline
+450.soplex	&	60772	&	1815	&	19	&	0,46\%   \\ \hline
+470.lbm	&	3614	&	1130	&	0	&	0,00\%   \\ \hline
+401.bzip2	&	16359	&	1688	&	1	&	0,10\%   \\ \hline
+403.gcc	&	792039	&	14853	&	92	&	0,14\%   \\ \hline
+429.mcf	&	2541	&	158	&	0	&	0,00\%   \\ \hline
+445.gobmk	&	130183	&	12158	&	54	&	0,63\%   \\ \hline
+456.hmmer	&	61908	&	3585	&	86	&	2,00\%   \\ \hline
+458.sjeng	&	27708	&	2438	&	20	&	1,10\%   \\ \hline
+462.libquantum	&	5848	&	593	&	6	&	1,04\%   \\ \hline
+464.h264ref	&	131905	&	13413	&	248	&	2,28\%   \\ \hline
+473.astar	&	8261	&	639	&	9	&	1,73\%   \\ \hline
 \end{tabular*}

### 2013-04-24 doc/Manuscript/SBLP13/paper.tex
Added more text to the paper.

@@
 fluxo de controle do programa.
+Esse CFG está convertido para o formato de atribuição estática única
+(SSA)~\cite{Cytron91}.
+Usaremos essa representação de programas porque ela facilita a nossa análise
+de dependência de dados.
 Os blocos básicos que começam nos rótulos 3 e 7 formam um laço natural,
@@
 
-Dizemos que um laço é {\em potencialmente vulnerável} a ataque de não
-terminação quando as condições que o controlam usam valores que dependem de
-dados de entrada do programa.
-Note que uma vulnerabilidade potencial não implica em uma vulnerabilidade
-real.
+Dizemos que um laço é {\em alcançável} quando as condições que o controlam
+usam valores que dependem de dados de entrada do programa.
+Note que um laço alcançável não é necessariamente vulnerável.
 A título de exemplo, o programa da Figura~\ref{fig:fact}(c), uma sutil
@@
 figura~\ref{fig:fact}(a), termina para qualquer entrada.
-A fim de detectar vulnerabilidades potenciais, nós utilizamos o
+A fim de determinar quais laços são alcançáveis, nós utilizamos o
 {\em grafo de dependências de dados}.
@@
 
-Um caminho entre uma entrada do programa, e o predicado que controla o laço
-indica uma vulnerabilidade potencial.
-O grafo de dependências de nosso exemplo possui uma tal vulnerabilidade:
+Um caminho entre uma entrada do programa, e o predicado que controla o laço é
+uma condição necessária para um ataque de não-terminação.
+O grafo de dependências de nosso exemplo apresenta tal condição:
 existe um caminho que une o nodo correspondente ao parâmetro \texttt{n}, uma
@@
 número de nodos - via a simples busca em profundidade ou largura.
+
+Diremos que um laço é {\em vulnerável} quando ele é alcançável, e, além disso,
+sua condição de parada é dependente de alguma operação {\em cíclica}
+passível de estouro de precisão.
+Seguindo a definição de laços de Appel e Palsberg, uma operação cíclica é
+qualquer instrução que ocorre no corpo $S$ do laço.
+Por exemplo, no CFG da figura~\ref{fig:fact}(b), as instruções
+$i_2 = i_1 + 1$ e $r_2 = r_1 \times i_1$ são cíclicas.
+O laço daquele exemplo encaixa-se em nossa definição de vulnerabilidade,
+pois sua condição de parada é alcançável a partir da entrada, e depende de
+uma instrução cíclica passível de estouro de precisão: $i_2 = i_1 + 1$.
+
+A nossa definição de vulnerabilidade inclui muitos laços que não são, de fato,
+vulneráveis, tais como aquele visto na figura~\ref{fig:fact}(c).
+Seria possível utilizar técnicas computacionalmente intensivas, tais como
+algoritmos de satisfabilidade, para refinar a nossa definição, eliminando
+alguns desses falsos positivos.
+Tal abordagem já foi utilizada em trabalhos anteriores ao
+nosso~\cite{Brockschmidt11,Burnim09,Gupta08,Son11,Velroyen08}.
+Por outro lado, os próprios autores desses trabalhos reportam que dificilmente
+suas técnicas poderiam lidar com programas muito grandes.
+Nós optamos por usar uma definição mais conservadora de laço vulnerável para
+termos uma ferramenta prática.
+Nós sanearemos todo laço considerado perigoso, inclusive aqueles que, devido
+à nossa definição liberal de vulnerabilidade, de fato não o são.
+Ainda assim, conforme mostraremos na seção~\ref{sec:exp}, o impacto dessa
+instrumentação é negligível.
 
@@
 
+Nós implementamos as técnicas descritas neste artigo em LLVM versão X.X.
+Nossa implementação foi testada em uma máquina Intel XX, com XX Gigabtyes de
+RAM, e XX GHz de Clock.
+Executamos nossa análise com sucesso sobre o arcabouço de testes de LLVM,
+um conjunto de programas com mais de 4.3 milhões de linhas de código C.
+No restante desta seção mostraremos somente resultados obtidos sobre os
+programas de SPEC CPU 2006.
+
+\noindent
+\textbf{Definição de Entrada de Dados: } nos experimentos apresentados nesta
+seção, consideraremos como entrada de dados os seguintes valores:
+\begin{itemize}
+\item os argumentos do método \texttt{main};
+\item o resultado retornado por funções {\em externas};
+\item ponteiros passados como argumento de funções {\em externas}.
+\end{itemize}
+As funções externas são a união dos seguintes três conjuntos:
+\begin{itemize}
+\item funções que não foram declaradas em nenhum dos arquivos que compõem
+o programa compilado;
+\item funções sem corpo;
+\item funções que podem ser chamadas via um ponteiro de funções.
+\end{itemize}
+
+\paragraph{Laços Alcançáveis e Vulneráveis. }
+A figura~\ref{fig:staticLoopData} mostra a quantidade de laços alcançáveis e
+vulneráveis que encontramos por programa.
+Estamos considerando, nesse caso, somente dependências de dados que não
+envolvam memória.
+Em outras palavras, todos os nodos do grafo de dependência são valores que
+podem residir em registradores.
+Essa versão de nosso detector não é segura: podemos deixar de marcar alguns
+laços vulneráveis como tal.
+Por outro lado, ela nos dá um número mínimo de laços que precisaríamos marcar,
+de acordo com a definição de laços alcançáveis da seção~\ref{sub:det}.
+Vemos, pela figura, que entre 1.3\% e 19.3\% de todos os laços do programa
+são alcançáveis.
+Aproximadamente a metade dos laços alcançáveis é vulnerável.
+Os laços que não são considerados vulneráveis em geral usam condições de
+paradas que comparam valores via igualdade, por exemplo,
+\verb^while(a[i] != '\0') {i++;}^.
+
 \begin{figure}[t!]
@@
 \caption{Informações estáticas inferidas pela análise de não-terminação.
-Num. laços: número de laços no programa.
-Alcançáveis: quantidade de laços que são dependentes de dados produzidos a
+{\bf Num. laços}: número de laços no programa.
+{\bf Alcançáveis}: quantidade de laços que são dependentes de dados produzidos a
 partir de canais de entrada.
-Vulneráveis: número de laços que preenchem nossos requisitos de
+{\bf Vulneráveis}: número de laços que preenchem nossos requisitos de
 vulnerabilidade.
-Caminhos: tamanho médio do menor caminho de dependência de dados da entrada
+{\bf Caminhos}: tamanho médio do menor caminho de dependência de dados da entrada
 até a operação de controle do laço.}
@@
 \caption{Impacto da instrumentação no código saneado.
-Num. Instruções: número de instruções no benchmark.
-Num. Arit: número de instruções que podem causar estouro de precisão inteira.
-Instrumentação: quantidade de testes inseridos para sanear o programa.
-Crescimento: razão entre o tamanho do programa instrumentado e o tamanho do
+{\bf Num. Instruções}: número de instruções no benchmark.
+{\bf Num. Arit}: número de instruções que podem causar estouro de precisão inteira.
+{\bf Instrumentação}: quantidade de testes inseridos para sanear o programa.
+{\bf Crescimento}: razão entre o tamanho do programa instrumentado e o tamanho do
 programa original.}

### 2013-04-25 doc/Manuscript/SBLP13/paper.tex
Added more text into the experiments.

@@
 
-\title{Prevenção Automática de Ataques de Não-Terminação}
+\title{Prevenção de Ataques de Não-Terminação baseados em
+Estouros de Precisão}
 
@@
 De acordo com Appel e Palsberg~\cite[p.376]{Appel02}, um laço natural é um
-conjunto de nodos $S$ do grafo de fluxo de controle de um programa (CFG),
+conjunto de nodos $S$ do grafo de fluxo de controle (CFG) de um programa,
 includindo um nodo cabeçalho $H$, com as seguintes três propriedades:
@@
 \item existe um caminho de $H$ até qualquer nodo que faz parte de $S$;
-\item não existe caminho que não contenha $H$ de qualquer nodo fora de $S$ para
-qualquer nodo em $S$.
+\item qualquer caminho de um nodo fora de $S$ para um nodo em $S$
+contém $H$.
 \end{itemize}
@@
 alteração da função \texttt{fact} inicialmente vista na
-figura~\ref{fig:fact}(a), termina para qualquer entrada.
-A fim de determinar quais laços são alcançáveis, nós utilizamos o
-{\em grafo de dependências de dados}.
+figura~\ref{fig:fact}(a), termina para qualquer entrada, embora ele contenha
+um laço alcançável.
+Utilizamos o {\em grafo de dependências de dados} para determinar laços
+alcançáveis.
 Esse grafo é definido da seguinte forma:
@@
 
-A nossa definição de vulnerabilidade inclui muitos laços que não são, de fato,
-vulneráveis, tais como aquele visto na figura~\ref{fig:fact}(c).
+A nossa definição de vulnerabilidade inclui muitos laços que não são
+concretamente vulneráveis, tais como aquele visto na figura~\ref{fig:fact}(c).
 Seria possível utilizar técnicas computacionalmente intensivas, tais como
@@
 ocorrência de estouros de precisão inteira.
-Nós inserimos tais testes sobre as operações aritméticas {\em internas} que
+Nós inserimos tais testes sobre as operações aritméticas cíclicas que
 controlam a condição de parada do laço.
-Se $S$ é o conjunto de blocos básicos que fazem parte do laço, uma operação é
-dita interna quando ela ocorre nalgum bloco em $S$.
 Continuando com o nosso exemplo, o laço alvo possui dois blocos básicos:
 o primeiro começa no rótulo três, e o segundo começa no rótulo sete.
-O laço possui duas operações aritméticas internas, todas ocorrendo no segundo
+O laço possui duas operações aritméticas cíclicas, todas ocorrendo no segundo
 bloco básico.
@@
 \paragraph{Instrumentação de Saneamento. }
-A fim de garantir que estouros de precisão inteiros não venham a causar
-a não-terminação de laços, nós inserimos testes no código binário do
-programa alvo.
+Para evitar que estouros de precisão venham a causar a não-terminação
+de laços, nós inserimos testes no código binário do programa alvo.
 O código que constitui cada um desses testes é formado por uma guarda, mas
@@
 Omitimos, nesse exemplo, o código do tratador de evento de estouro, pois ele
-simplesmente chamada uma rotina implementada em uma biblioteca
+simplesmente invoca uma rotina implementada em uma biblioteca
 dinamicamente compartilhada.
@@
 \includegraphics[width=1\textwidth]{images/instrumented_cfg}
-\caption{(a) }
+\caption{(a) Programa que será instrumentado.
+(b) Representação do programa em bytecodes LLVM.
+(c) Operação que está sendo instrumentada.
+(d) Teste de detecção de estouro de precisão.
+(e) Programa instrumentado, em bytecodes LLVM.}
 \label{fig:instrumented_cfg}
@@
 \begin{itemize}
-\item os argumentos do método \texttt{main};
+\item os argumentos do método \texttt{main}, isto é, as variáveis \texttt{argc}
+e \texttt{argv};
 \item o resultado retornado por funções {\em externas};
@@
 Essa versão de nosso detector não é segura: podemos deixar de marcar alguns
-laços vulneráveis como tal.
+laços vulneráveis que são controlados por dado contaminado que tenha sido
+armazenado em memória.
 Por outro lado, ela nos dá um número mínimo de laços que precisaríamos marcar,
@@
 \verb^while(a[i] != '\0') {i++;}^.
+A figura também reporta o tamanho médio dos caminhos vulneráveis.
+Um caminho vulnerável é a sequência de instruções que precisa ser executada
+para que a expressão de controle do laço seja influenciada por algum valor de
+entrada.
+Conforme vemos pela figura, esses caminhos são geralmente pequenos.
+Os dois caminhos vulneráveis em \texttt{429.mcf}, por exemplo, possuem somente
+duas instruções cada um.
+Concluí-se que o desenvolvedor, ao procurar por possíveis vulnerabilidades
+a ataques de não terminação, teria de ater-se a uma quantidade muito pequena
+de nodos do grafo de dependência do programa.
 
@@
 445.gobmk	&	1073	&	37	&	19	&	4  \\ \hline
-456.hmmer	&	687	&	133	&	48	&	2  \\ \hline
+456.hmmer	&	687	&	133	&	48	&	3  \\ \hline
 458.sjeng	&	235	&	9	&	3	&	4  \\ \hline
@@
 \end{figure}
+
+A figura~\ref{fig:instImpact} mostra o impacto da instrumentação que inserimos
+no tamanho do programa.
+Em média, cada laço vulnerável custou-nos a criação de 1.44 guardas, como
+aquela vista na figura~\ref{fig:instrumented_cfg}(e).
+O aumento de tamanho do programa guardado é negligível, conforme podemos
+observar na última coluna da figura~\ref{fig:instImpact}.
+O maior crescimento, observado em \texttt{464.h264ref}, foi de 2.28\%.
+Uma vez que o número de guardas inseridas nos programas é tão pequeno, o
+crescimento de seu tempo de execução também é pequeno.
+Executamos todos os programas instrumentados, passando-lhes suas entradas de
+referência, conforme especificado no manual de uso de SPEC CPU 2006, e as
+diferenças de tempo de execução ficaram dentro da margem de erro dos benchmarks.
+Em outras palavras, não pudemos perceber quaisquer diferenças, em termos de
+tempo de execução, entre os programas originais e instrumentados.
 
@@
 
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{images/Runtime}
-\caption{Variação no tempo de execução dos benchmarks devido à instrumentação
-de saneamento.}
-\label{fig:Runtime}
-\end{center}
-\end{figure}
+% Describe which programs were aborted once you tried to execute them.
 
@@
 
+Neste artigo nós descrevemos uma forma de ataque de negação de serviço que
+busca levar o programa alvo à não-terminação.
+Ao contrário da literatura relacionada, ate-mo-nos a ataques baseados em
+estouro de precisão de aritmética de inteiros.
+Esse tipo de fenômeno caracteriza linguagens de programação como Java, C e
+C++.
+Nós definimos algumas propriedades necessárias para a efetiva realização de
+um ataque de não-terminação, a saber, condição de controle controlada por
+adversário, e por operações cíclicas passíveis de estouro de precisão.
+Em seguida, mostramos como eliminar a última dessas condições via guardas
+inseridas pelo compilador.
+Finalmente, mostramos experimentalmente que nossas guardas, ainda que inseridas
+conservadoramente, não comprometem o tempo de execução do programa
+instrumentado.
+Demonstramos assim que a prevenção de ataques de não-terminação baseados em
+estouros de precisão é barata e efetiva.
+
+Neste trabalho nós adotamos uma definição muito conservadora de laços
+vulneráveis.
+De fato, muitos dos laços que indicamos como vulneráveis, em nossos
+experimentos, de fato não o são.
+Nossa decisão foi fruto de um compromisso entre a precisão e a eficiência:
+instrumentamos todos os laços possivelmente vulneráveis, mesmo aqueles que
+não são perigosos, e ainda assim mantivemos estável o tempo de execução dos
+programas.
+Por outro lado, é nossa intenção, em trabalho futuro, estreitar essa definição
+de laço vulnerável, a fim de fornecer a desenvolvedores uma ferramenta que lhes
+auxilie na descoberta de exemplos de vulnerabilidades.
+
+\noindent
+\textbf{Software: } nossas técnicas foram todas implementadas em LLVM, e estão
+disponíveis publicamente na URL \texttt{http://code.google.com/p/range-analysis/}.
+
 \bibliographystyle{plain}

### 2013-05-04 doc/Manuscript/SBLP13/paper.tex
partial commit

@@
 Caminhos \\ \hline
-433.milc	&	326	&	8	&	4	&	3   \\ \hline
-444.namd	&	508	&	7	&	1	&	2  \\ \hline
-447.dealII	&	4657	&	540	&	225	&	3  \\ \hline
-450.soplex	&	547	&	21	&	12	&	3  \\ \hline
-470.lbm		&	17	&	0	&	0	&	- \\ \hline
-401.bzip2	&	205	&	1	&	1	&	3  \\ \hline
-403.gcc		&	3853	&	109	&	59	&	4  \\ \hline
-429.mcf		&	38	&	2	&	0	&	2  \\ \hline
-445.gobmk	&	1073	&	37	&	19	&	4  \\ \hline
-456.hmmer	&	687	&	133	&	48	&	3  \\ \hline
-458.sjeng	&	235	&	9	&	3	&	4  \\ \hline
-462.libquantum	&	67	&	2	&	1	&	4  \\ \hline
-464.h264ref	&	1586	&	27	&	22	&	15  \\ \hline
-473.astar	&	86	&	7	&	7	&	18  \\ \hline
+433.milc 		&	339		&	150		&	139		&	25	\\ \hline
+444.namd	 		&	490		&	444		&	409		&	9	\\ \hline
+447.dealII		&	4747		&	3482		&	2653		&	12	\\ \hline
+450.soplex		&	535		&	506		&	453		&	14	\\ \hline
+470.lbm			&	19		&	0		&	0		&	0	\\ \hline
+445.gobmk		&	1068		&	574		&	497		&	17	\\ \hline
+458.sjeng		&	232		&	140		&	110		&	15	\\ \hline
+462.libquantum	&	75		&	44		&	41		&	10	\\ \hline
+464.h264ref		&	1624 	&	194		&	164		&	17	\\ \hline
 \end{tabular*}
@@
 Crescimento \\ \hline
-433.milc	&	22615	&	1165	&	7	&	0,50\%   \\ \hline
-444.namd	&	65045	&	3136	&	1	&	0,03\%   \\ \hline
-447.dealII	&	415756	&	14948	&	279	&	0,58\%   \\ \hline
-450.soplex	&	60772	&	1815	&	19	&	0,46\%   \\ \hline
-470.lbm	&	3614	&	1130	&	0	&	0,00\%   \\ \hline
-401.bzip2	&	16359	&	1688	&	1	&	0,10\%   \\ \hline
-403.gcc	&	792039	&	14853	&	92	&	0,14\%   \\ \hline
-429.mcf	&	2541	&	158	&	0	&	0,00\%   \\ \hline
-445.gobmk	&	130183	&	12158	&	54	&	0,63\%   \\ \hline
-456.hmmer	&	61908	&	3585	&	86	&	2,00\%   \\ \hline
-458.sjeng	&	27708	&	2438	&	20	&	1,10\%   \\ \hline
-462.libquantum	&	5848	&	593	&	6	&	1,04\%   \\ \hline
-464.h264ref	&	131905	&	13413	&	248	&	2,28\%   \\ \hline
-473.astar	&	8261	&	639	&	9	&	1,73\%   \\ \hline
+433.milc 		&	24971	&	1101		&	150		&	9.49\%	\\ \hline
+444.namd	 		&	77922	&	3136		&	937		&	15.73\%	\\ \hline
+447.dealII		&	483614	&	14910	&	3772		&	6.42\%	\\ \hline
+450.soplex		&	67808	&	1779		&	762		&	17.63\%	\\ \hline
+470.lbm			&	3788		&	1130		&	0		&	0.00\%	\\ \hline
+445.gobmk		&	146298	&	11856	&	939		&	9.80\%	\\ \hline
+458.sjeng		&	25473	&	2138		&	167		&	9.97\%	\\ \hline
+462.libquantum	&	6562		&	593		&	66		&	14.04\%	\\ \hline
+464.h264ref		&	141772	&	13398	&	469		&	4.26\%	\\ \hline
 \end{tabular*}
@@
 \label{fig:instImpact}
+\end{figure}
+
+\begin{figure}[t!]
+\begin{center}
+\renewcommand{\arraystretch}{1.2}
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|} \hline
+Benchmark & Num. Op. & Num. Var. & Num. Mem. & Num. Arestas \\ \hline
+433.milc			&	24799	&	20435	&	6901		&	74599	\\ \hline
+444.namd			&	78043	&	72866	&	10468	&	232792	\\ \hline
+447.dealII		&	512373	&	441576	&	131986	&	1516203	\\ \hline
+450.soplex		&	69861	&	56172	&	21764	&	204126	\\ \hline
+470.lbm			&	3787		&	3490		&	626		&	10859	\\ \hline
+445.gobmk		&	152342	&	167197	&	21146	&	458405	\\ \hline
+458.sjeng		&	25169	&	26313	&	2522		&	73355	\\ \hline
+462.libquantum	&	6552		&	5845		&	940		&	19142	\\ \hline
+464.h264ref		&	141606	&	106292	&	45223	&	409813	\\ \hline
+\end{tabular*}
+\end{center}
+\caption{Dados do grafo de dependência dos programas analisados}.
+{\bf Num. Op.}: número de nós de operação.
+{\bf Num. Var.}: número de nós de variáveis e constantes.
+{\bf Num. Var.}: número de nós de memória.
+{\bf Num. Arestas}: número de arestas do grafo.}
+\label{fig:tblDepGraph}
 \end{figure}

### 2013-05-04 src/OverflowSanitizer/OverflowSanitizer.cpp

@@ -10,8 +10,12 @@
 
 using namespace std;
 
-#define InsertAborts true
+#define InsertAborts false
 #define TruncInstrumentation true
+
+
+static cl::opt<bool, false>
+OnlySLE("only-SLE", cl::desc("Only consider SLE, ULE, SGE and UGE as loop conditions."), cl::NotHidden);
 
 //Table 1
 STATISTIC(NrLoops					, "Number of Loops");
@@ -218,7 +222,7 @@ bool ::OverflowSanitizer::runOnModule(Module& M) {
 
 				Value* loopCondition = BI->getCondition();
 
-				std::pair<GraphNode*,int> nearestDependency = depGraph.depGraph->getNearestDependency(loopCondition, inputValues, true);
+				std::pair<GraphNode*,int> nearestDependency = depGraph.depGraph->getNearestDependency(loopCondition, inputValues, false);
 
 				if(nearestDependency.first){
 
@@ -244,35 +248,40 @@ bool ::OverflowSanitizer::runOnModule(Module& M) {
 	for (std::list<Value*>::iterator Cit = inpDepLoopConditions.begin(), Cend = inpDepLoopConditions.end(); Cit != Cend; Cit++){
 
 
-//		//The ICmpInst has 2 operands: X comes from input. Y is updated inside the loop. We must make sure that Y will never overflow.
-//		if (ICmpInst* CI = dyn_cast<ICmpInst>(*Cit)) {
-//
-//			BasicBlock* parentBB = CI->getParent();
-//			BasicBlock* loopHeader = loopHeaders[loopBlocks[parentBB]];
-//
-//			switch (CI->getPredicate()){
-//
-//                case CmpInst::ICMP_SLE:
-//                case CmpInst::ICMP_ULE:
-//                case CmpInst::ICMP_SGE:
-//                case CmpInst::ICMP_UGE:
-//
-//        			lookForValuesToSafe(CI, loopHeader);
-//
-//                	break;
-//                default:
-//                	break;
-//			}
-//
-//		}
+		if (OnlySLE) {
 
+			//The ICmpInst has 2 operands: X comes from input. Y is updated inside the loop. We must make sure that Y will never overflow.
+			if (ICmpInst* CI = dyn_cast<ICmpInst>(*Cit)) {
 
-		if (Instruction* Inst = dyn_cast<Instruction>(*Cit)) {
[…]

### 2013-05-04 src/OverflowSanitizer/OverflowSanitizer.h

@@ -40,7 +40,7 @@
 #include <list>
 #include <stdio.h>
 
-#include "../RangeAnalysis/RangeAnalysis.h"
+//#include "../RangeAnalysis/RangeAnalysis.h"
 #include "../../Analysis/DepGraph/InputValues.h"
 #include "../../Analysis/DepGraph/DepGraph.h"
 #include "../uSSA/uSSA.h"

### 2013-05-07 doc/Manuscript/SBLP13/paper.tex
partial commit

@@
 A instrumentação que inserimos para prevenir ataques de não terminação aumenta o 
-tamanho do programa saneado em cerca de 5\% em média, e torna-os menos que 2\% 
+tamanho do programa saneado em cerca de 10\% em média, e torna-os menos que 2\% 
 mais lentos.
@@
 on the average, less than 10 instructions long.
-Our instrumentation that prevents this kind of attacks adds less than 5\%
-extra code on the secured program.
+Our instrumentation that prevents this kind of attacks adds in average less than 
+10\% extra code on the secured program.
 The final, protected code, is, on the average, less than 2\% slower than the
@@
 
-Nós implementamos todas as idéias que discutimos neste artigo em LLVM,
+Nós implementamos todas as ideias que discutimos neste artigo em LLVM,
 um compilador de qualidade industrial~\cite{Lattner04}.
@@
 de laços, nós inserimos testes no código binário do programa alvo.
-O código que constitui cada um desses testes é formado por uma guarda, mas
+O código que constitui cada um desses testes é formado por uma guarda, mais
 um tratador de eventos.
@@
 Atualmente instrumentamos quatro tipos diferentes de instrução:
-adição, subtração, multiplicação e arredamentos para a esquerda.
+adição, subtração, multiplicação e arredamentos para a esquerda, conhecidos como {\em shift left}.
 As operações de adição, subtração e multiplicação podem ser com ou sem sinal
@@
 \end{center}
-\caption{Dados do grafo de dependência dos programas analisados}.
+\caption{Dados do grafo de dependência dos programas analisados.
 {\bf Num. Op.}: número de nós de operação.

### 2013-05-07 doc/Manuscript/SBLP13/paper.tex
partial update

@@
 \renewcommand{\arraystretch}{1.2}
-\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|} \hline
-Benchmark & Num. Laços & Alcançáveis & Vulneráveis &
-Caminhos \\ \hline
-433.milc 		&	339		&	150		&	139		&	25	\\ \hline
-444.namd	 		&	490		&	444		&	409		&	9	\\ \hline
-447.dealII		&	4747		&	3482		&	2653		&	12	\\ \hline
-450.soplex		&	535		&	506		&	453		&	14	\\ \hline
-470.lbm			&	19		&	0		&	0		&	0	\\ \hline
-445.gobmk		&	1068		&	574		&	497		&	17	\\ \hline
-458.sjeng		&	232		&	140		&	110		&	15	\\ \hline
-462.libquantum	&	75		&	44		&	41		&	10	\\ \hline
-464.h264ref		&	1624 	&	194		&	164		&	17	\\ \hline
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|r|r|} \hline
+Benchmark 		& Inst. 			& Entr. 		& Nós Op. 		& Nós Var. 		& Nós Mem. 		& Arestas 		\\ \hline
+433.milc			&	24,971		&	3,995	&	24,799		&	20,435		&	6,901		&	74,599		\\ \hline
+444.namd			&	77,922		&	15,964	&	78,043		&	72,866		&	10,468		&	232,792		\\ \hline
+447.dealII		&	483,614		&	98,180	&	512,374		&	441,576		&	131,986		&	1,516,202	\\ \hline
+450.soplex		&	67,808		&	11,182	&	69,870		&	56,172		&	21,764		&	204,159		\\ \hline
+470.lbm			&	3,788		&	239		&	3,787		&	3,490		&	626			&	10,859		\\ \hline
+400.perlbench	&	288,429		&	25,224	&	287,050		&	230,982		&	98,886		&	819,872		\\ \hline
+401.bzip2		&	17,999		&	1,862	&	18,007		&	15,019		&	4,423		&	53,414		\\ \hline
+403.gcc			&	830,861		&	65,118	&	830,054		&	660,772		&	321,279		&	2,440,279	\\ \hline
+429.mcf			&	2,851		&	373		&	2,897		&	2,354		&	905			&	8,608		\\ \hline
+445.gobmk		&	146,298		&	21,420	&	152,342		&	167,197		&	21,146		&	458,405		\\ \hline
+456.hmmer		&	62,704		&	8,487	&	63,004		&	51,549		&	20,727		&	182,736		\\ \hline
+458.sjeng		&	25,473		&	2,610	&	25,169		&	26,313		&	2,522		&	73,355		\\ \hline
+462.libquantum	&	6,562		&	921		&	6,552		&	5,845		&	940			&	19,142		\\ \hline
+464.h264ref		&	141,772		&	11,995	&	141,606		&	106,292		&	45,219		&	409,813		\\ \hline
+471.omnetpp		&	96,929		&	15,989	&	101,197		&	88,819		&	31,686		&	305,170		\\ \hline
+473.astar		&	9,386		&	1,506	&	9,476		&	8,199		&	1,982		&	28,067		\\ \hline
+483.xalancbmk	&	648,941		&	132,976	&	689,176		&	569,780		&	244,874		&	1,971,945	\\ \hline
+Total			&	2,936,308	&	418,041	&	3,015,403	&	2,527,660	&	966,334		&	8,809,417	\\ \hline
 \end{tabular*}

### 2013-05-08 doc/Manuscript/SBLP13/paper.tex
Updating paper

--- 2013-05-07 doc/Manuscript/SBLP13/paper.tex
@@
 \usepackage{skak}
+
+\usepackage{booktabs}
 
@@
 Descobrimos que, em média, caminhos em que informação perigosa trafega são 
-pequenos, sendo compostos por não mais que 10 instruções assembly.
+pequenos, sendo compostos por não mais que 13 instruções assembly.
 A instrumentação que inserimos para prevenir ataques de não terminação aumenta o 
-tamanho do programa saneado em cerca de 10\% em média, e torna-os menos que 2\% 
+tamanho do programa saneado em cerca de 5\% em média, e torna-os menos que 2\% 
 mais lentos.
@@
 We have found out that the typical path from inputs to loop conditions is,
-on the average, less than 10 instructions long.
+on the average, less than 13 instructions long.
 Our instrumentation that prevents this kind of attacks adds in average less than 
-10\% extra code on the secured program.
+5\% extra code on the secured program.
 The final, protected code, is, on the average, less than 2\% slower than the
@@
 Ao analisar os programas presentes na coleção SPEC CPU 2006, fomos capazes
-de descobrir XX laços que são influenciados por dados provenientes de
+de descobrir 12.304 laços que são influenciados por dados provenientes de
 entradas públicas, isto é, que podem ser manipuladas por um adversário.
-Dentre esses laços, pelo menos XX estão sujeitos à não terminação.
+Dentre esses laços, pelo menos 920 estão sujeitos à não terminação.
 Para obter esse número, utilizamos um padrão muito simples: procuramos por
@@
 
-Nós implementamos as técnicas descritas neste artigo em LLVM versão X.X.
-Nossa implementação foi testada em uma máquina Intel XX, com XX Gigabtyes de
-RAM, e XX GHz de Clock.
-Executamos nossa análise com sucesso sobre o arcabouço de testes de LLVM,
+Nós implementamos as técnicas descritas neste artigo em LLVM versão 3.3.
+Nossa implementação foi testada em uma máquina Intel\textsuperscript{®} Core™ i7-3770, com 16 Gigabtyes de
+RAM, e 3.40 GHz de Clock.
+Executamos nossa análise com sucesso sobre o arcabouço de testes do LLVM,
 um conjunto de programas com mais de 4.3 milhões de linhas de código C.
@@
 
+\paragraph{Grafo de dependências de dados. }
+A figura~\ref{fig:tblDepGraph} mostra informações estáticas a respeito da
+estrutura dos programas analisados.
+Essa estrutura compreende o tamanho absoluto do programa, em número de 
+instruções e as informações do grafo de dependências de dados extraído de cada 
+programa.
+É possível observar que em média 14\% dos valores podem conter informações 
+vindas da entrada de dados. Esse numero é influenciado por todas as chamadas a 
+funções externas ao código fonte avaliado, para as quais não há como saber se 
+há ou não acesso a dados externos.
+Outra informação importante é que há mais nós de operação no grafo do que instruções no programa.
+Esse fato acontece pois nossa análise é interprocedural e é preciso inserir nós 
+de operação para ligar parâmetros formais a parâmetros reais e também para ligar
+valores de retorno aos valores que recebem o resultado de funções.
+Além disso, a tabela apresenta a quantidade de nós de memória, que são vértices do grafo que
+agrupam variáveis que contém ponteiros e que podem apontar
+para a mesma região de memória. Para identificar quais variáveis pertencem a cada vértice de memória, 
+é necessária a realização de uma análise de ponteiros seguida da criação de 
+conjuntos disjuntos de áreas de memória.
+Finalmente, o número de arestas dos grafos gerados indica que no caso médio, o grafo é esparso.
+Nos programas do SPEC, a relação é de 1,38 arestas por vértice, em média.
+Essa característica é evidenciada pela média de relação de 1,38 arestas por vértice, nos programas do SPEC; 
+e pela correlação de XX entre o número de vértices e o número de arestas de cada grafo, considerando 
+os 100 maiores casos de teste padrão do LLVM.
+
+\begin{figure}[t!]
+\begin{center}
+\renewcommand{\arraystretch}{1.2}
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lrrrrrr} \toprule
+Benchmark 		& Num. Inst. 	& Entr.		& Num. Op. 		& Num. Var. 		& Num. Mem. 		& Num. Arestas 		\\ \midrule
+433.milc			&	24,971		&	3,995	&	24,799		&	20,435		&	6,901		&	74,599		\\ 
+444.namd			&	77,922		&	15,964	&	78,043		&	72,866		&	10,468		&	232,792		\\ 
+447.dealII		&	483,614		&	98,180	&	512,374		&	441,576		&	131,986		&	1,516,202	\\ 
+450.soplex		&	67,808		&	11,182	&	69,870		&	56,172		&	21,764		&	204,159		\\ 
+470.lbm			&	3,788		&	239		&	3,787		&	3,490		&	626			&	10,859		\\ 
+400.perlbench	&	288,429		&	25,224	&	287,050		&	230,982		&	98,886		&	819,872		\\ 
+401.bzip2		&	17,999		&	1,862	&	18,007		&	15,019		&	4,423		&	53,414		\\ 
+403.gcc			&	830,861		&	65,118	&	830,054		&	660,772		&	321,279		&	2,440,279	\\ 
+429.mcf			&	2,851		&	373		&	2,897		&	2,354		&	905			&	8,608		\\ 
+445.gobmk		&	146,298		&	21,420	&	152,342		&	167,197		&	21,146		&	458,405		\\ 
+456.hmmer		&	62,704		&	8,487	&	63,004		&	51,549		&	20,727		&	182,736		\\ 
+458.sjeng		&	25,473		&	2,610	&	25,169		&	26,313		&	2,522		&	73,355		\\ 
+462.libquantum	&	6,562		&	921		&	6,552		&	5,845		&	940			&	19,142		\\ 
+464.h264ref		&	141,772		&	11,995	&	141,606		&	106,292		&	45,219		&	409,813		\\ 
+471.omnetpp		&	96,929		&	15,989	&	101,197		&	88,819		&	31,686		&	305,170		\\ 
+473.astar		&	9,386		&	1,506	&	9,476		&	8,199		&	1,982		&	28,067		\\ 
+483.xalancbmk	&	648,941		&	132,976	&	689,176		&	569,780		&	244,874		&	1,971,945	\\ \midrule
+Total			&	2,936,308	&	418,041	&	3,015,403	&	2,527,660	&	966,334		&	8,809,417	\\ \bottomrule
+\end{tabular*}
+\end{center}
+\caption{Dados do grafo de dependência dos programas analisados.
+{\bf Num. Inst.}: número de instruções do programa.
+{\bf Entr.}: número de valores que podem representar entradas de dados.
+{\bf Num. Op.}: número de nós de operação.
+{\bf Num. Var.}: número de nós de variáveis e constantes.
+{\bf Num. Var.}: número de nós de memória.
+{\bf Num. Arestas}: número de arestas do grafo.}
+\label{fig:tblDepGraph}
+\end{figure}
+
 \paragraph{Laços Alcançáveis e Vulneráveis. }
@@
 vulneráveis que encontramos por programa.
-Estamos considerando, nesse caso, somente dependências de dados que não
-envolvam memória.
-Em outras palavras, todos os nodos do grafo de dependência são valores que
-podem residir em registradores.
-Essa versão de nosso detector não é segura: podemos deixar de marcar alguns
-laços vulneráveis que são controlados por dado contaminado que tenha sido
-armazenado em memória.
 Por outro lado, ela nos dá um número mínimo de laços que precisaríamos marcar,
 de acordo com a definição de laços alcançáveis da seção~\ref{sub:det}.
-Vemos, pela figura, que entre 1.3\% e 19.3\% de todos os laços do programa
+Vemos, pela figura, que Aproximadamente 70\% de todos os laços do programa
 são alcançáveis.
 Aproximadamente a metade dos laços alcançáveis é vulnerável.
-Os laços que não são considerados vulneráveis em geral usam condições de
-paradas que comparam valores via igualdade, por exemplo,
-\verb^while(a[i] != '\0') {i++;}^.
-A figura também reporta o tamanho médio dos caminhos vulneráveis.
-Um caminho vulnerável é a sequência de instruções que precisa ser executada
-para que a expressão de controle do laço seja influenciada por algum valor de
-entrada.
-Conforme vemos pela figura, esses caminhos são geralmente pequenos.
-Os dois caminhos vulneráveis em \texttt{429.mcf}, por exemplo, possuem somente
-duas instruções cada um.
-Concluí-se que o desenvolvedor, ao procurar por possíveis vulnerabilidades
+Dos laços vulneráveis, aproximadamente 12\% são controlados por comparações do 
+tipo \texttt{i <= N};
+Conforme vemos pela figura, os caminhos entre as entradas de dados e as condições 
+de parada dos loops são geralmente pequenos.
+Os dez caminhos vulneráveis em \texttt{429.mcf}, por exemplo, possuem somente
+sete instruções, em média.
+Conclui-se que o desenvolvedor, ao procurar por possíveis vulnerabilidades
 a ataques de não terminação, teria de ater-se a uma quantidade muito pequena
@@
 \renewcommand{\arraystretch}{1.2}
-\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|} \hline
-Benchmark & Num. Laços & Alcançáveis & Vulneráveis &
-Caminhos \\ \hline
-433.milc 		&	339		&	150		&	139		&	25	\\ \hline
-444.namd	 		&	490		&	444		&	409		&	9	\\ \hline
-447.dealII		&	4747		&	3482		&	2653		&	12	\\ \hline
-450.soplex		&	535		&	506		&	453		&	14	\\ \hline
-470.lbm			&	19		&	0		&	0		&	0	\\ \hline
-445.gobmk		&	1068		&	574		&	497		&	17	\\ \hline
-458.sjeng		&	232		&	140		&	110		&	15	\\ \hline
-462.libquantum	&	75		&	44		&	41		&	10	\\ \hline
-464.h264ref		&	1624 	&	194		&	164		&	17	\\ \hline
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lrrrrrr} \toprule
+Benchmark			&	Num. Laços (L)	&	Alcançáveis	&	Caminhos	&	\multicolumn{2}{c}{Vulneráveis} &	\% (II/L) \\ 
+                                                                     \cmidrule{5-6} 
+{}				&	{}		&	{}			&	{}				&	I		&	II	&	{}	\\ \midrule
+433.milc			&	329		&	146			&	25				&	138		&	0	&	0.00\%	\\ 
+444.namd			&	484		&	438			&	9				&	408		&	2	&	0.41\%	\\ 
+447.dealII		&	4,759	&	3,493		&	12				&	2,657	&	73	&	1.53\%	\\ 
+450.soplex		&	542		&	513			&	14				&	453		&	175	&	32.29\%	\\ 
+470.lbm			&	18		&	0			&	0				&	0		&	0	&	0.00\%	\\ 
+400.perlbench	&	1,160	&	1,034		&	14				&	315		&	72	&	6.21\%	\\ 
+401.bzip2		&	211		&	151			&	19				&	95		&	24	&	11.37\%	\\ 
+403.gcc			&	3,824	&	2,966		&	24				&	1,297	&	310	&	8.11\%	\\ 
+429.mcf			&	39		&	10			&	7				&	2		&	0	&	0.00\%	\\
+445.gobmk		&	1,082	&	588			&	17				&	499		&	25	&	2.31\%	\\ 
+456.hmmer		&	681		&	376			&	8				&	255		&	84	&	12.33\%	\\ 
+458.sjeng		&	235		&	139			&	15				&	109		&	2	&	0.85\%	\\ 
+462.libquantum	&	79		&	44			&	10				&	41		&	3	&	3.80\%	\\ 
+464.h264ref		&	1,614	&	193			&	17				&	161		&	9	&	0.56\%	\\ 
+471.omnetpp		&	363		&	280			&	8				&	201		&	41	&	11.29\%	\\ 
+473.astar		&	88		&	53			&	8				&	50		&	18	&	20.45\%	\\ 
+483.xalancbmk	&	2,212	&	1,880		&	10				&	1,061	&	82	&	3.71\%	\\ \midrule
+Total			&	17,720	&	12,304		&	13				&	7,742	&	920	&	5.19\%	\\ \bottomrule
+
 \end{tabular*}
@@
 \caption{Informações estáticas inferidas pela análise de não-terminação.
-{\bf Num. laços}: número de laços no programa.
+{\bf Num. Laços}: número de laços no programa.
 {\bf Alcançáveis}: quantidade de laços que são dependentes de dados produzidos a
 partir de canais de entrada.
+{\bf Caminhos}: tamanho médio do menor caminho de dependência de dados da entrada
+até a operação de controle do laço.
 {\bf Vulneráveis}: número de laços que preenchem nossos requisitos de
 vulnerabilidade.
-{\bf Caminhos}: tamanho médio do menor caminho de dependência de dados da entrada
-até a operação de controle do laço.}
+{\bf I}: análise considerando todas as condições de parada de loop potencialmente vulneráveis.
+{\bf II}: análise considerando somente as condições de parada de loop controlados por comparações do tipo \texttt{i <= N}.}
 \label{fig:staticLoopData}
@@
 
+\paragraph{Instrumentação. }
 A figura~\ref{fig:instImpact} mostra o impacto da instrumentação que inserimos
 no tamanho do programa.
-Em média, cada laço vulnerável custou-nos a criação de 1.44 guardas, como
+Em média, cada laço vulnerável custou-nos a criação de 1.7 guardas, como
 aquela vista na figura~\ref{fig:instrumented_cfg}(e).
-O aumento de tamanho do programa guardado é negligível, conforme podemos
-observar na última coluna da figura~\ref{fig:instImpact}.
-O maior crescimento, observado em \texttt{464.h264ref}, foi de 2.28\%.
-Uma vez que o número de guardas inseridas nos programas é tão pequeno, o
-crescimento de seu tempo de execução também é pequeno.
+O aumento de tamanho do programa guardado é pequeno, conforme podemos
+observar nas duas últimas colunas da figura~\ref{fig:instImpact}.
+O maior crescimento, observado em \texttt{450.soplex}, foi de 17.83\%.
+Na média, entretanto, os programas cresceram apenas 5\% em número de instruções.
+Uma vez que o número de instruções inseridas nos programas é tão pequeno, o
+crescimento de seu tempo de execução também é pequeno, pois nenhuma das 
+instruções realiza operações demoradas, como acesso a memórias lentas.
 Executamos todos os programas instrumentados, passando-lhes suas entradas de
@@
 diferenças de tempo de execução ficaram dentro da margem de erro dos benchmarks.
+% XX - Melhorar esse final de parágrafo, colocando os resultados do teste dinâmico
 Em outras palavras, não pudemos perceber quaisquer diferenças, em termos de
@@
 \renewcommand{\arraystretch}{1.2}
-\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|} \hline
-Benchmark & Num. Instruções & Num. Arit. & Instrumentação &
-Crescimento \\ \hline
-433.milc 		&	24971	&	1101		&	150		&	9.49\%	\\ \hline
-444.namd	 		&	77922	&	3136		&	937		&	15.73\%	\\ \hline
-447.dealII		&	483614	&	14910	&	3772		&	6.42\%	\\ \hline
-450.soplex		&	67808	&	1779		&	762		&	17.63\%	\\ \hline
-470.lbm			&	3788		&	1130		&	0		&	0.00\%	\\ \hline
-445.gobmk		&	146298	&	11856	&	939		&	9.80\%	\\ \hline
-458.sjeng		&	25473	&	2138		&	167		&	9.97\%	\\ \hline
-462.libquantum	&	6562		&	593		&	66		&	14.04\%	\\ \hline
-464.h264ref		&	141772	&	13398	&	469		&	4.26\%	\\ \hline
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lrrrrrr} \toprule
+Benchmark		&	Num. Inst.	&	Num. Arit.	&	\multicolumn{2}{c}{Instrumentação}		&	\multicolumn{2}{c}{Crescimento}			\\ 
+													\cmidrule{4-5} 			\cmidrule{6-7}
+{}				&	{}			&	{}			&	I		&	II		&	I		&	II		\\ \midrule
+433.milc			&	24,971		&	1,101		&	150		&	0		&	9.49\%	&	0.00\%	\\
+444.namd			&	77,922		&	3,136		&	932		&	2		&	15.75\%	&	0.04\%	\\
+447.dealII		&	483,614		&	14,910		&	3,762	&	77		&	6.40\%	&	0.21\%	\\
+450.soplex		&	67,808		&	1,779		&	771		&	261		&	17.83\%	&	6.09\%	\\
+470.lbm			&	3,788		&	1,130		&	0		&	0		&	0.00\%	&	0.00\%	\\
+400.perlbench	&	288,429		&	7,983		&	655		&	226		&	2.92\%	&	1.09\%	\\
+401.bzip2		&	17,999		&	1,684		&	240		&	56		&	17.20\%	&	4.61\%	\\
+403.gcc			&	830,861		&	14,338		&	2,072	&	473		&	3.26\%	&	0.83\%	\\
+429.mcf			&	2,851		&	158			&	4		&	0		&	2.00\%	&	0.00\%	\\
+445.gobmk		&	146,298		&	11,856		&	932		&	76		&	9.73\%	&	0.78\%	\\
+456.hmmer		&	62,704		&	3,210		&	419		&	143		&	10.08\%	&	3.37\%	\\
+458.sjeng		&	25,473		&	2,138		&	167		&	22		&	9.97\%	&	1.31\%	\\
+462.libquantum	&	6,562		&	593			&	68		&	4		&	14.49\%	&	0.98\%	\\
+464.h264ref		&	141,772		&	13,398		&	411		&	102		&	3.81\%	&	0.88\%	\\
+471.omnetpp		&	96,929		&	2,029		&	249		&	46		&	3.91\%	&	0.73\%	\\
+473.astar		&	9,386		&	639			&	99		&	37		&	16.46\%	&	6.13\%	\\
+483.xalancbmk	&	648,941		&	12,528		&	1,562	&	116		&	2.52\%	&	0.24\%	\\ \midrule
+Total			&	2,936,308	&	92,610		&	12,493	&	1,641	&	5.02\%	&	0.81\%	\\ \bottomrule
 \end{tabular*}
@@
 \caption{Impacto da instrumentação no código saneado.
-{\bf Num. Instruções}: número de instruções no benchmark.
-{\bf Num. Arit}: número de instruções que podem causar estouro de precisão inteira.
+{\bf Num. Inst.}: número de instruções no benchmark.
+{\bf Num. Arit.}: número de instruções que podem causar estouro de precisão inteira.
 {\bf Instrumentação}: quantidade de testes inseridos para sanear o programa.
 {\bf Crescimento}: razão entre o tamanho do programa instrumentado e o tamanho do
-programa original.}
+programa original.
+{\bf I}: análise considerando todas as condições de parada de loop potencialmente vulneráveis.
+{\bf II}: análise considerando somente as condições de parada de loop controlados por comparações do tipo \texttt{i <= N}.}
 \label{fig:instImpact}
-\end{figure}
-
-\begin{figure}[t!]
-\begin{center}
-\renewcommand{\arraystretch}{1.2}
-\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|} \hline
-Benchmark & Num. Op. & Num. Var. & Num. Mem. & Num. Arestas \\ \hline
-433.milc			&	24799	&	20435	&	6901		&	74599	\\ \hline
-444.namd			&	78043	&	72866	&	10468	&	232792	\\ \hline
-447.dealII		&	512373	&	441576	&	131986	&	1516203	\\ \hline
-450.soplex		&	69861	&	56172	&	21764	&	204126	\\ \hline
-470.lbm			&	3787		&	3490		&	626		&	10859	\\ \hline
-445.gobmk		&	152342	&	167197	&	21146	&	458405	\\ \hline
-458.sjeng		&	25169	&	26313	&	2522		&	73355	\\ \hline
-462.libquantum	&	6552		&	5845		&	940		&	19142	\\ \hline
-464.h264ref		&	141606	&	106292	&	45223	&	409813	\\ \hline
-\end{tabular*}
-\end{center}
-\caption{Dados do grafo de dependência dos programas analisados.
-{\bf Num. Op.}: número de nós de operação.
-{\bf Num. Var.}: número de nós de variáveis e constantes.
-{\bf Num. Var.}: número de nós de memória.
-{\bf Num. Arestas}: número de arestas do grafo.}
-\label{fig:tblDepGraph}
 \end{figure}
@@
 estouro de precisão de aritmética de inteiros.
-Esse tipo de fenômeno caracteriza linguagens de programação como Java, C e
+Esse tipo de fenômeno é característico de linguagens de programação como Java, C e
 C++.

### 2013-05-09 src/OverflowSanitizer/OverflowSanitizer.cpp
Turning off fprintf

@@ -12,6 +12,7 @@ using namespace std;
 
 #define InsertAborts false
 #define TruncInstrumentation true
+#define InsertFprintfs false
 
 
 static cl::opt<bool, false>
@@ -145,7 +146,7 @@ BasicBlock* OverflowSanitizer::NewOverflowOccurrenceBlock(Instruction* I, BasicB
 	BasicBlock* result = BasicBlock::Create(*context, "", I->getParent()->getParent(), NextBlock);
 	BranchInst* branch = BranchInst::Create(NextBlock, result);
 
-	//if (InsertFprintfs){
+	if (InsertFprintfs){
 		Value* vStderr = new LoadInst(GVstderr, "loadstderr", branch);
 
 		std::vector<Value*> args;
@@ -156,7 +157,7 @@ BasicBlock* OverflowSanitizer::NewOverflowOccurrenceBlock(Instruction* I, BasicB
 		args.push_back(InstructionIdentifier);
 		ArrayRef<Value*> ARargs(args);
 		CallInst::Create(FPrintF, ARargs, "", branch);
-	//}
+	}
 
 	return result;
 }

### 2013-05-09 doc/Manuscript/SBLP13/paper.tex
Changed a bit sec:exp

@@
 No restante desta seção mostraremos somente resultados obtidos sobre os
-programas de SPEC CPU 2006.
+programas disponíveis em SPEC CPU 2006.
 
 \noindent
-\textbf{Definição de Entrada de Dados: } nos experimentos apresentados nesta
-seção, consideraremos como entrada de dados os seguintes valores:
+\textbf{Definição de Entrada de Dados.}
+As entradas de dados são as funções que um adversário pode usar para forçar
+a não-terminação de um programa.
+Nos experimentos apresentados nesta seção, consideraremos como entrada de
+dados os seguintes valores:
 \begin{itemize}
@@
 estrutura dos programas analisados.
-Essa estrutura compreende o tamanho absoluto do programa, em número de 
-instruções e as informações do grafo de dependências de dados extraído de cada 
+Mostramos o tamanho absoluto do programa, em número de 
+instruções e as informações sobre o seu grafo de dependências.
+Em média 14\% dos valores que o programa manipula podem
+conter informações vindas da entrada de dados.
+Esses valores são produzidos por funções externas, para as quais não é
+possível saber se há ou não acesso a dados externos.
+Observa-se que existem mais nós de operação no grafo do que instruções no
 programa.
-É possível observar que em média 14\% dos valores podem conter informações 
-vindas da entrada de dados. Esse numero é influenciado por todas as chamadas a 
-funções externas ao código fonte avaliado, para as quais não há como saber se 
-há ou não acesso a dados externos.
-Outra informação importante é que há mais nós de operação no grafo do que instruções no programa.
-Esse fato acontece pois nossa análise é interprocedural e é preciso inserir nós 
-de operação para ligar parâmetros formais a parâmetros reais e também para ligar
-valores de retorno aos valores que recebem o resultado de funções.
-Além disso, a tabela apresenta a quantidade de nós de memória, que são vértices do grafo que
-agrupam variáveis que contém ponteiros e que podem apontar
-para a mesma região de memória. Para identificar quais variáveis pertencem a cada vértice de memória, 
-é necessária a realização de uma análise de ponteiros seguida da criação de 
-conjuntos disjuntos de áreas de memória.
-Finalmente, o número de arestas dos grafos gerados indica que no caso médio, o grafo é esparso.
-Nos programas do SPEC, a relação é de 1,38 arestas por vértice, em média.
-Essa característica é evidenciada pela média de relação de 1,38 arestas por vértice, nos programas do SPEC; 
-e pela correlação de XX entre o número de vértices e o número de arestas de cada grafo, considerando 
-os 100 maiores casos de teste padrão do LLVM.
+Esse fato verifica-se porque nossa análise é interprocedural.
+Assim, inserimos nós de operação para criar dependências entre parâmetros
+formais a parâmetros reais e também para ligar valores de retorno aos valores
+que recebem o resultado de funções.
+A tabela também apresenta a quantidade de nós de memória.
+Esses nós são vértices do grafo de dependência que representam os ponteiros
+e os blocos de memória alocados no código do programa.
+A fim de determinar se dois ponteiros são sinônimos ou não, implementamos
+a análise de apontadores de Andersen~\cite{Andersen94}.
+Sinônimos são agrupados em um mesmo nó de memória.
+Analisando a figura~\ref{fig:tblDepGraph}, constatamos que os grafos de
+dependências são esparsos.
+Nos programas de SPEC CPU 2006, a razão entre o número de vértices e o número
+de arestas é 1.38.
+Essa relação fica ainda mais evidente quando extraímos o coeficiente de
+determinação entre essas duas grandezas -- número de vértices e arestas.
+Obtemos o valor de 0.99, o que indica uma forte correlação linear entre os
+dois valores.
 
@@
 \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lrrrrrr} \toprule
-Benchmark 		& Num. Inst. 	& Entr.		& Num. Op. 		& Num. Var. 		& Num. Mem. 		& Num. Arestas 		\\ \midrule
+Benchmark 		&      Insts. 	& Entr.		&    Ops. 		&   Vars. 		&  Mems. 		& Arestas 		\\ \midrule
 433.milc			&	24,971		&	3,995	&	24,799		&	20,435		&	6,901		&	74,599		\\ 
@@
 \end{center}
-\caption{Dados do grafo de dependência dos programas analisados.
-{\bf Num. Inst.}: número de instruções do programa.
-{\bf Entr.}: número de valores que podem representar entradas de dados.
-{\bf Num. Op.}: número de nós de operação.
-{\bf Num. Var.}: número de nós de variáveis e constantes.
-{\bf Num. Var.}: número de nós de memória.
-{\bf Num. Arestas}: número de arestas do grafo.}
-\label{fig:tblDepGraph}
+\caption{\label{fig:tblDepGraph}Dados do grafo de dependência dos programas analisados.
+{\bf Insts.}: número de instruções do programa.
+{\bf Entr.}: número de instruções que podem representar entradas de dados.
+{\bf Ops.}: número de nós de operação no grafo de dependência.
+{\bf Vars.}: número de nós que representam variáveis.
+{\bf Mems.}: número de nós que representam blocos de memória.
+{\bf Arestas}: número de arestas do grafo.}
 \end{figure}

### 2013-05-09 doc/Manuscript/SBLP13/paper.tex
revised discussion in sec:exp

@@
 
-\paragraph{Grafo de dependências de dados. }
-A figura~\ref{fig:tblDepGraph} mostra informações estáticas a respeito da
-estrutura dos programas analisados.
-Mostramos o tamanho absoluto do programa, em número de 
-instruções e as informações sobre o seu grafo de dependências.
-Em média 14\% dos valores que o programa manipula podem
+\noindent
+\textbf{Grafo de dependências de dados. }
+A figura~\ref{fig:tblDepGraph} mostra informações estáticas a respeito dos
+grafos de dependências dos programas analisados.
+Como ponto de referência, mostramos o tamanho absoluto de cada programa, em
+número de  instruções.
+Em média 14\% dos valores que cada programa manipula podem
 conter informações vindas da entrada de dados.
 Esses valores são produzidos por funções externas, para as quais não é
-possível saber se há ou não acesso a dados externos.
+possível saber se há ou não contaminação por dados externos.
 Observa-se que existem mais nós de operação no grafo do que instruções no
@@
 que recebem o resultado de funções.
-A tabela também apresenta a quantidade de nós de memória.
-Esses nós são vértices do grafo de dependência que representam os ponteiros
-e os blocos de memória alocados no código do programa.
-A fim de determinar se dois ponteiros são sinônimos ou não, implementamos
-a análise de apontadores de Andersen~\cite{Andersen94}.
-Sinônimos são agrupados em um mesmo nó de memória.
-Analisando a figura~\ref{fig:tblDepGraph}, constatamos que os grafos de
-dependências são esparsos.
-Nos programas de SPEC CPU 2006, a razão entre o número de vértices e o número
-de arestas é 1.38.
-Essa relação fica ainda mais evidente quando extraímos o coeficiente de
-determinação entre essas duas grandezas -- número de vértices e arestas.
-Obtemos o valor de 0.99, o que indica uma forte correlação linear entre os
-dois valores.
 
@@
 
-\paragraph{Laços Alcançáveis e Vulneráveis. }
-A figura~\ref{fig:staticLoopData} mostra a quantidade de laços alcançáveis e
-vulneráveis que encontramos por programa.
-Por outro lado, ela nos dá um número mínimo de laços que precisaríamos marcar,
-de acordo com a definição de laços alcançáveis da seção~\ref{sub:det}.
-Vemos, pela figura, que Aproximadamente 70\% de todos os laços do programa
-são alcançáveis.
-Aproximadamente a metade dos laços alcançáveis é vulnerável.
-Dos laços vulneráveis, aproximadamente 12\% são controlados por comparações do 
-tipo \texttt{i <= N};
-Conforme vemos pela figura, os caminhos entre as entradas de dados e as condições 
-de parada dos loops são geralmente pequenos.
-Os dez caminhos vulneráveis em \texttt{429.mcf}, por exemplo, possuem somente
-sete instruções, em média.
-Conclui-se que o desenvolvedor, ao procurar por possíveis vulnerabilidades
-a ataques de não terminação, teria de ater-se a uma quantidade muito pequena
-de nodos do grafo de dependência do programa.
+A figura~\ref{fig:tblDepGraph} também apresenta a quantidade de nós de memória.
+Esses nós são vértices do grafo de dependência que representam os ponteiros
+e os blocos de memória alocados no código do programa.
+A fim de determinar se dois ponteiros são sinônimos ou não, implementamos
+a análise de apontadores de Andersen~\cite{Andersen94}.
+Sinônimos, isto é, apontadores que indicam regiões de memória sobrepostas, são
+agrupados em um mesmo nó.
+Quanto mais precisa a análise de ponteiros, menores os conjuntos contidos em
+cada uma dessas unidades.
+A análise de Andersen é um compromisso entre eficiência e precisão.
+
+Analisando a figura~\ref{fig:tblDepGraph}, constatamos que os grafos de
+dependências são esparsos.
+Nos programas de SPEC CPU 2006, a razão entre o número de vértices e o número
+de arestas é 1.38.
+Essa relação fica ainda mais evidente quando extraímos o coeficiente de
+determinação entre essas duas grandezas -- número de vértices e arestas.
+Obtemos o valor de 0.99, o que indica uma forte correlação linear entre os
+dois valores.
+Essa baixa densidade ocorre porque programas reais variáveis tendem a ser
+usadas um número baixo de vezes.
+Boissinot {\em et al.}~\cite{Benoit08} demonstraram, empiricamente, que
+a maior parte das variáveis é usada somente uma vez no programa, e mais de
+99\% das variáveis são usadas menos que cinco vezes.
+Assim, a maior parte dos vértices que representam variáveis em nossos grafos de
+dependências possuem grau de saída inferior a cinco.
 
@@
 \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lrrrrrr} \toprule
-Benchmark			&	Num. Laços (L)	&	Alcançáveis	&	Caminhos	&	\multicolumn{2}{c}{Vulneráveis} &	\% (II/L) \\ 
+Benchmark			&		Laços (L)	&	Alcançáveis	&	Caminhos	&	\multicolumn{2}{c}{Vulneráveis} &	\% (II/L) \\ 
                                                                      \cmidrule{5-6} 
@@
 \caption{Informações estáticas inferidas pela análise de não-terminação.
-{\bf Num. Laços}: número de laços no programa.
+{\bf Laços}: número de laços no programa.
 {\bf Alcançáveis}: quantidade de laços que são dependentes de dados produzidos a
@@
 vulnerabilidade.
-{\bf I}: análise considerando todas as condições de parada de loop potencialmente vulneráveis.
-{\bf II}: análise considerando somente as condições de parada de loop controlados por comparações do tipo \texttt{i <= N}.}
+{\bf I}: laços vulneráveis segundo a definição da seção~\ref{sub:det}.
+{\bf II}: laços vulneráveis controlados por comparações do tipo
+\texttt{i <= N}.}
 \label{fig:staticLoopData}
@@
 
-\paragraph{Instrumentação. }
-A figura~\ref{fig:instImpact} mostra o impacto da instrumentação que inserimos
-no tamanho do programa.
-Em média, cada laço vulnerável custou-nos a criação de 1.7 guardas, como
-aquela vista na figura~\ref{fig:instrumented_cfg}(e).
-O aumento de tamanho do programa guardado é pequeno, conforme podemos
-observar nas duas últimas colunas da figura~\ref{fig:instImpact}.
-O maior crescimento, observado em \texttt{450.soplex}, foi de 17.83\%.
-Na média, entretanto, os programas cresceram apenas 5\% em número de instruções.
-Uma vez que o número de instruções inseridas nos programas é tão pequeno, o
-crescimento de seu tempo de execução também é pequeno, pois nenhuma das 
-instruções realiza operações demoradas, como acesso a memórias lentas.
-Executamos todos os programas instrumentados, passando-lhes suas entradas de
-referência, conforme especificado no manual de uso de SPEC CPU 2006, e as
-diferenças de tempo de execução ficaram dentro da margem de erro dos benchmarks.
-% XX - Melhorar esse final de parágrafo, colocando os resultados do teste dinâmico
-Em outras palavras, não pudemos perceber quaisquer diferenças, em termos de
-tempo de execução, entre os programas originais e instrumentados.
+\noindent
+\textbf{Laços Alcançáveis e Vulneráveis. }
+A figura~\ref{fig:staticLoopData} mostra a quantidade de laços alcançáveis e
+vulneráveis que encontramos por programa.
+A noção de ``laço vulnerável" é definida na seção~\ref{sub:det}.
+A quantidade de laços vulneráveis que reportamos representa um limite inferior
+no número de estruturas de iteração que precisamos instrumentar para prevenir
+ataques de não terminação baseados em estouro de precisão.
+A figura indica que aproximadamente 70\% de todos os laços do programa
+são alcançáveis.
+Dessa quantidade, a metade é vulnerável.
+Cerca de 12\% dos laços vulneráveis são controlados por comparações do
+tipo \texttt{i <= N}.
+Esse tipo de condição é particularmente perigosa, pois, conforme visto no
+exemplo da figura~\ref{fig:fact}, se o limite \texttt{N} for o
+maior inteiro possível, então a condição será sempre verdade.
+A figura~\ref{fig:staticLoopData} mostra que os caminhos entre as entradas de
+dados e as condições de parada de laços são geralmente pequenos.
+Por exemplo, os dez caminhos vulneráveis em \texttt{429.mcf} possuem em média
+sete instruções.
+Conclui-se que um auditor que procure por vulnerabilidades a ataques de
+não-terminação tem, em geral, de analisar sequências relativamente
+pequenas de código.
 
@@
 \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lrrrrrr} \toprule
-Benchmark		&	Num. Inst.	&	Num. Arit.	&	\multicolumn{2}{c}{Instrumentação}		&	\multicolumn{2}{c}{Crescimento}			\\ 
+Benchmark		&		Insts.	&		Arits.	&	\multicolumn{2}{c}{Instrumentação}		&	\multicolumn{2}{c}{Crescimento}			\\ 
 													\cmidrule{4-5} 			\cmidrule{6-7}
@@
 \caption{Impacto da instrumentação no código saneado.
-{\bf Num. Inst.}: número de instruções no benchmark.
-{\bf Num. Arit.}: número de instruções que podem causar estouro de precisão inteira.
+{\bf Insts.}: número de instruções no benchmark.
+{\bf Arits.}: número de instruções que podem causar estouro de precisão inteira.
 {\bf Instrumentação}: quantidade de testes inseridos para sanear o programa.
@@
 programa original.
-{\bf I}: análise considerando todas as condições de parada de loop potencialmente vulneráveis.
-{\bf II}: análise considerando somente as condições de parada de loop controlados por comparações do tipo \texttt{i <= N}.}
+{\bf I}: análise considerando todos os laços vulneráveis.
+{\bf II}: análise considerando somente os laços controlados por condições do
+tipo \texttt{i <= N}.}
 \label{fig:instImpact}
 \end{figure}
+
+\noindent
+\textbf{Instrumentação. }
+A figura~\ref{fig:instImpact} mostra que nossa instrumentação tem um
+impacto ínfimo sobre o tamanho do programa guardado.
+Nossos programas possuem poucas operações de tipo inteiro passíveis de estouro
+de precisão -- em média somente 3.15\% das instruções são desse tipo.
+Dessas instruções, um número ainda menor é usado dentro de laços vulneráveis.
+No caso médio, cada laço vulnerável custou-nos a criação de 1.64 guardas, como
+aquela vista na figura~\ref{fig:instrumented_cfg}(e).
+O aumento de tamanho do programa guardado é pequeno, conforme podemos
+observar nas duas últimas colunas da figura~\ref{fig:instImpact}.
+O maior crescimento, observado em \texttt{450.soplex}, foi de 17.83\%.
+Na média, entretanto, os programas cresceram apenas 5\% em número de
+instruções.
+Um dos benchmarks, \texttt{470.lbm} não recebeu qualquer instrumentação,
+uma vez que ele não possui laços controlados por dados de entrada.
+Todos os laços desse benchmark dependem de constantes criadas dentro do próprio
+programa.
+
+\noindent
+\textbf{Tempo de execução. }
+Uma vez que o número de instruções inseridas nos programas é tão pequeno, o
+crescimento de seu tempo de execução é irrisório.
+Para justificar tal fato, alegamos que nenhuma das instruções presentes em
+uma guarda realiza operações demoradas, como acesso a memórias lentas.
+Executamos todos os programas instrumentados, passando-lhes suas entradas de
+referência, conforme especificado no manual de uso de SPEC CPU 2006, e as
+diferenças de tempo de execução ficaram dentro da margem de erro
+prevista pelo manual SPEC CPU 2006.
+% XX - Melhorar esse final de parágrafo, colocando os resultados do teste dinâmico
+Em outras palavras, não pudemos perceber quaisquer diferenças, em termos de
+tempo de execução, entre os programas originais e instrumentados.
 

### 2013-05-10 doc/Manuscript/SBLP13/paper.tex
Added discussion about runtime.

@@
 referência, conforme especificado no manual de uso de SPEC CPU 2006, e as
-diferenças de tempo de execução ficaram dentro da margem de erro
-prevista pelo manual SPEC CPU 2006.
-% XX - Melhorar esse final de parágrafo, colocando os resultados do teste dinâmico
-Em outras palavras, não pudemos perceber quaisquer diferenças, em termos de
-tempo de execução, entre os programas originais e instrumentados.
+diferenças de tempo de execução são mostradas na figura~\ref{fig:runtime}.
+Cada programa foi amostrado 30 vezes, e o resultado que apresentamos é a
+média aritmética dessas trinta amostras.
+A margem de erro é negligível.
+Testamos dois modos de instrumentação.
+No primeiro deles guardamos todos os laços considerados vulneráveis contra
+estouros de precisão.
+Nesse caso, observamos que os programas instrumentados ficaram 0.61\% mais
+lentos.
+No segundo modo de instrumentação guardamos somente os laços cujas condições
+de controle usam comparadores do tipo \texttt{<=}.
+Registramos nesse segundo experimento que os programas modificados ficaram
+0.24\% mais lentos.
+Caso houvéssemos instrumentado todas as operações aritméticas nos programas
+disponíveis em SPEC CPU 2006, a taxa de lentidão seria de 3.24\%.
+Não mostramos esse resultado na figura, para não comprometermos a sua
+leitura.
+Em alguns programas, como \texttt{401.bzip2}, por exemplo, pudemos registrar
+diminuição do tempo de execução.
+Não encontramos razão aparente para tal comportamento, pois não efetuamos
+qualquer otimização sobre o código instrumentado.
 
-% Describe which programs were aborted once you tried to execute them.
+\begin{figure}[t!]
+\begin{center}
+\includegraphics[width=1\textwidth]{images/runtime}
+\caption{Variação no tempo de execução dos programas instrumentados.}
+\label{fig:runtime}
+\end{center}
+\end{figure}
 

### 2013-05-10 doc/Manuscript/SBLP13/paper.tex
Fixed typos in the text.

@@
 A instrumentação que inserimos para prevenir ataques de não terminação aumenta o 
-tamanho do programa saneado em cerca de 5\% em média, e torna-os menos que 2\% 
+tamanho do programa saneado em cerca de 5\% em média, e torna-os menos que 1\% 
 mais lentos.
@@
 an adversary can manipulate these values to force this loop to iterate forever.
-Preventing this kind of attack is difficult because they do not originate
+Preventing this kind of attack is difficult because it does not originate
 from bugs that break the semantics of the programming
@@
 on the average, less than 13 instructions long.
-Our instrumentation that prevents this kind of attacks adds in average less than 
+Our instrumentation that prevents this kind of attack adds on average less than
 5\% extra code on the secured program.
-The final, protected code, is, on the average, less than 2\% slower than the
+The final, protected code, is, on the average, less than 1\% slower than the
 original, unprotected program.
@@
 
-Nós implementamos todas as ideias que discutimos neste artigo em LLVM,
+Implementamos todas as ideias que discutimos neste artigo em LLVM,
 um compilador de qualidade industrial~\cite{Lattner04}.
@@
 validam nossa análise.
-Ao analisar os programas presentes na coleção SPEC CPU 2006, fomos capazes
+Examinando os programas presentes na coleção SPEC CPU 2006, fomos capazes
 de descobrir 12.304 laços que são influenciados por dados provenientes de
 entradas públicas, isto é, que podem ser manipuladas por um adversário.
-Dentre esses laços, pelo menos 920 estão sujeitos à não terminação.
-Para obter esse número, utilizamos um padrão muito simples: procuramos por
+Pelo menos 920 desses laços estão sujeitos à não terminação.
+Esse número advém de um padrão simples: procuramos por
 laços cujo teste de parada é do tipo \texttt{i <= N}, sendo \texttt{N}
@@
 seja bem maior que o valor que apuramos.
+Para testar essa hipótese, alargamos a nossa definição de laços vulneráveis
+para englobar qualquer iterador cuja condição de controle esteja sujeita a
+estouro de precisão.
+Dentre os laços encontrados em SPEC CPU 2006, 7,742 deles atendem esse
+requisito.
 A nossa instrumentação -- usada para impedir os ataques de não-terminação --
@@
 Os testes que inserimos antes de cada operação aritmética que pode levar a
-um ataque de não terminação custa-nos uma perda de desempenho de menos que
-XX\%.
+um ataque de não terminação custa-nos uma perda de desempenho que vai de
+0.61\% no pior caso, a 0.24\% no melhor.
+Protegemos laços conservadoramente.
+Assim, ainda que não estejamos fornecendo uma prova formal de que um laço é
+vulnerável, nós garantimos que qualquer programa instrumentado é invulnerável
+a ataques de não-terminação baseados em estouro de precisão.
 

### 2013-05-10 doc/Manuscript/SBLP13/paper.tex
15 pages.

@@
 crescimento de seu tempo de execução é irrisório.
-Para justificar tal fato, alegamos que nenhuma das instruções presentes em
-uma guarda realiza operações demoradas, como acesso a memórias lentas.
+Além disso, nenhuma das instruções presentes em uma guarda realiza operações
+demoradas, como acesso a memórias lentas.
 Executamos todos os programas instrumentados, passando-lhes suas entradas de
@@
 estouro de precisão de aritmética de inteiros.
-Esse tipo de fenômeno é característico de linguagens de programação como Java, C e
-C++.
+Esse fenômeno caracteriza linguagens como Java, C e C++.
 Nós definimos algumas propriedades necessárias para a efetiva realização de

### 2013-05-10 doc/Manuscript/SBLP13/paper.tex
Correcting typos.

@@
 Além disso, dada a vasta quantidade de código aberto usado nos mais diversos
-ramos da indústria de {\em software}, usuários maliciosos têm a sua
+ramos da indústria de {\em software}, usuários maliciosos têm à sua
 disposição um vasto campo de ação.
@@
 Para cada instrução $i: v = f(\ldots, u, \ldots)$ que define uma variável $v$ e
-usa uma variável $u$ n\'{o}s criamos duas areastas: $n_u \leftarrow n_i$ e
-$n_i \leftarrow n_v$.
+usa uma variável $u$ n\'{o}s criamos duas areastas: $n_u \rightarrow n_i$ e
+$n_i \rightarrow n_v$.
 O grafo de dependências que extraímos a partir d CFG visto na
@@
 laço.
-Esse tipo de caminho, uma vez contruído o grafo, pode ser encontrado em
+Esse tipo de caminho, uma vez construído o grafo, pode ser encontrado em
 tempo linear no número de arestas do grafo - normalmente proporcional ao
@@
 \begin{itemize}
-\item existe um caminho no nodo $n_i$ até o nodo $n_p$.
+\item Existe um caminho no nodo $n_i$ até o nodo $n_p$.
 \item O nodo $n_i$ encontra-se em um ciclo.
+\item A operação que define a variável $i$ pode provocar um estouro aritmético.
 \end{itemize}
@@
 Assim, inserimos nós de operação para criar dependências entre parâmetros
-formais a parâmetros reais e também para ligar valores de retorno aos valores
+formais e parâmetros reais e também para ligar valores de retorno aos valores
 que recebem o resultado de funções.

### 2013-07-10 doc/Manuscript/SBLP13/paper.tex
Last ref.
Little adjusts for SBLP'13

@@
 \usepackage[utf8]{inputenc}
-\usepackage[brazil]{babel}
+\usepackage[english,brazil]{babel}
 
@@
 
+\addtolength{\parskip}{-3mm}
 %mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
@@
 Ao contrário, essas vulnerabilidades têm origem na aritmética modular inteira de 
-linguagens como C, C++ e Java, a qual possui semântica bem definida.
+linguagens como C, C++ e Java, que possui semântica bem definida.
 Neste artigo nós apresentamos uma ferramenta que detecta tais problemas, 
@@
 %mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
+{
+\selectlanguage{english}
 \begin{abstract}
-We say that a program is vulnerable to a non-termination attack if (i) it
-contains a loop that is bounded by values coming from public inputs, and (ii)
+We say that a program is vulnerable to a non-termination attack if it
+contains a loop that is bounded by values dependent on public inputs, and
 an adversary can manipulate these values to force this loop to iterate forever.
@@
 used by C, C++ and Java-like languages, which have well-defined semantics.
-In this paper we present the diagnosis and the cure for this type of attack.
+In this paper we present a diagnosis and a solution for this type of attack.
 Firstly, we describe a tainted-flow analysis that detects non-termination
@@
 Our instrumentation that prevents this kind of attack adds on average less than
-5\% extra code on the secured program.
-The final, protected code, is, on the average, less than 1\% slower than the
+5\% extra code on the secured programs and make them less than 1\% slower than the
 original, unprotected program.
 \end{abstract}
+}
+
+\addtolength{\parskip}{3mm}
 
@@
 De acordo com Appel e Palsberg~\cite[p.376]{Appel02}, um laço natural é um
-conjunto de nodos $S$ do grafo de fluxo de controle (CFG) de um programa,
-includindo um nodo cabeçalho $H$, com as seguintes três propriedades:
+conjunto de nós $S$ do grafo de fluxo de controle (CFG) de um programa,
+includindo um nó cabeçalho $H$, com as seguintes três propriedades:
 %
 \begin{itemize}
-\item a partir de qualquer nodo em $S$ existe um caminho que chega a $H$;
-\item existe um caminho de $H$ até qualquer nodo que faz parte de $S$;
-\item qualquer caminho de um nodo fora de $S$ para um nodo em $S$
+\item a partir de qualquer nó em $S$ existe um caminho que chega a $H$;
+\item existe um caminho de $H$ até qualquer nó que faz parte de $S$;
+\item qualquer caminho de um nó fora de $S$ para um nó em $S$
 contém $H$.
@@
 i_2 \mapsto v_2, \ldots, i_m \mapsto v_m$ que, ao influenciar $E'$, faz com
-que o laço $L$ não termine.
+que $E$ nunca tenha um valor que faça o laço $L$ parar.
 \end{enumerate}
@@
 Esse grafo é definido da seguinte forma:
-para cada variável $v$ no programa, nós criamos um nodo $n_v$, e para cada
-instrução $i$ no programa, nós criamos um nodo $n_i$.
+para cada variável $v$ no programa, nós criamos um nó $n_v$, e para cada
+instrução $i$ no programa, nós criamos um nó $n_i$.
 Para cada instrução $i: v = f(\ldots, u, \ldots)$ que define uma variável $v$ e
@@
 $n_i \rightarrow n_v$.
-O grafo de dependências que extraímos a partir d CFG visto na
+O grafo de dependências que extraímos a partir do CFG visto na
 figura~\ref{fig:fact}(b) é mostrado na figura~\ref{fig:depGraph}.
@@
 
-Um caminho entre uma entrada do programa, e o predicado que controla o laço é
+Um caminho entre uma entrada do programa e o predicado que controla o laço é
 uma condição necessária para um ataque de não-terminação.
 O grafo de dependências de nosso exemplo apresenta tal condição:
-existe um caminho que une o nodo correspondente ao parâmetro \texttt{n}, uma
-entrada, ao nodo que corresponde a \texttt{p}, o predicado de controle do
+existe um caminho que une o nó correspondente ao parâmetro \texttt{n}, uma
+entrada, ao nó que corresponde a \texttt{p}, o predicado de controle do
 laço.
@@
 tempo linear no número de arestas do grafo - normalmente proporcional ao
-número de nodos - via a simples busca em profundidade ou largura.
+número de nós - via uma simples busca em profundidade ou largura.
 
@@
 \begin{itemize}
-\item Existe um caminho no nodo $n_i$ até o nodo $n_p$.
-\item O nodo $n_i$ encontra-se em um ciclo.
+\item Existe um caminho no nó $n_i$ até o nó $n_p$.
+\item O nó $n_i$ encontra-se em um ciclo.
 \item A operação que define a variável $i$ pode provocar um estouro aritmético.
@@
 Em primeiro lugar, porque essa operação encontra-se em um ciclo.
-Em segundo lugar, porque existe um caminho do nodo $n_{++}$ até o nodo
+Em segundo lugar, porque existe um caminho do nó $n_{++}$ até o nó
 $n_p$.
@@
 No restante desta seção mostraremos somente resultados obtidos sobre os
-programas disponíveis em SPEC CPU 2006.
+programas escritos na linguagem C disponíveis em SPEC CPU 2006.
 
@@
 número de  instruções.
-Em média 14\% dos valores que cada programa manipula podem
+Em média, 14\% dos valores que cada programa manipula podem
 conter informações vindas da entrada de dados.
@@
 dois valores.
-Essa baixa densidade ocorre porque programas reais variáveis tendem a ser
+Essa baixa densidade ocorre porque em programas reais variáveis tendem a ser
 usadas um número baixo de vezes.
@@
 são alcançáveis.
-Dessa quantidade, a metade é vulnerável.
+Dessa quantidade, metade é vulnerável.
 Cerca de 12\% dos laços vulneráveis são controlados por comparações do
@@
 Esse modo de rastreamento é popularmente conhecido como detecção de vazamento
-de segredos~\cite{Hammer06}.
+de segredos~\cite{Hammer06,Quadros11}.
 E, no sentido inverso, pode-se traçar o fluxo de informação de entradas que
@@
 Esses trabalhos não levam em consideração possibilidade de não-terminação
-devido à estouros de precisão, tampouco procuram detectar possíveis
+devido a estouros de precisão, tampouco procuram detectar possíveis
 vulnerabilidades baseadas em negação de serviço.
